---
title: "Running univariate models to calculate Heritability_ASReml"
author: "Dinesh Ghimire"
date: "2025-06-18"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr","asreml", "readxl", "tidyr", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "GGally")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```

#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NMSU"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NMSU 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa_NMSU"
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is misssing please let me know. 
key<-read.csv("2022/meta_data/cultivar_list_key.csv") # contains both CU and NMSU cultivars here. 
# its same for 2022 and 2023 (cultivar list)
```




1. *First treatment NMSU_Normal 2022 *
with spatial in the model, random = ~ vm(Material_Id, ct_geno_inverse) + ar1(Row):ar1(Range),



*Normal irrigation treatment*
Loading and preparing data files
```{r loading files 06182025}
#import merged phenotype file 
nmsu22_nor_pheno <- read.csv("2022/MSI_data/Output_files/nmsu22_nor_combined_phenotypes_MSI.csv", sep = ",") %>% 
  select(-X) # remove index column

# Before running models, make sure the variables are in correct form. Need to convert into factors and numeric if needed.
  # converting all factorial variables into factor
nmsu22_nor_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu22_nor_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)



#import genotype file
grm_aa <- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1)
grm_aa <- as.matrix(grm_aa)
mean(diag(grm_aa)) # 1
hist(diag(grm_aa))

grm_agh <- grm_aa+diag(.0005,length(grm_aa[,1]),length(grm_aa[,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


Normally if the range of numbers vary a lot between vi's and yield, then we can scale yield to variance level of vi's. but here since yield is in lb and upto 2 digit, should have that issue. If got any, we can always use scaled yield.
```{r scaling yield }
# we only have to scale 6 harvest values based on NDVI's value. Of course the variance between VI's differ but it should be okay to use Yield scaled onto NDVI for analysis with NDRE and GNDVI as well. 

# Define the mapping of yield traits to NDVI columns
yield_traits <- paste0("yld", 1:6)
ndvi_times <- c("NDVI_110", "NDVI_144", "NDVI_180", "NDVI_206", "NDVI_236", "NDVI_270") # selecting one of the time point in each harvest window. selecting based on higher variation time point from heatmap, not actually calculating all. 

# Loop over each yield trait
for (i in 1:length(yield_traits)) {
  
  trait <- yield_traits[i]
  ndvi_col <- ndvi_times[i]
  
  # Calculate variance of NDVI and yield
  ndvi_var <- var(nmsu22_nor_pheno[[ndvi_col]], na.rm = TRUE)
  yld_var <- var(nmsu22_nor_pheno[[trait]], na.rm = TRUE)
  
  # Calculate scaling factor
  sf <- sqrt(ndvi_var) / sqrt(yld_var)
  
  # Create new scaled yield column
  new_col <- paste0("sca_", trait)
  nmsu22_nor_pheno[[new_col]] <- nmsu22_nor_pheno[[trait]] * sf
  
  #  Print scaling factor
  cat("Scaling factor for", trait, "using", ndvi_col, ":", sf, "\n")
}


# thus all scaled yields are added as a new column at last. 
```

one but not confirming now
```{r for checking outliers GZA_F}
# Fit a linear model
model <- lm(yld1 ~ Rep, data = nmsu22_nor_pheno, na.action = na.exclude)

# Calculate residuals
nmsu22_nor_pheno$residuals <- residuals(model, na.action = na.exclude)
# Standardize the residuals
nmsu22_nor_pheno$residuals_standard <- scale(nmsu22_nor_pheno$residuals, center = T, scale = T) 
# Identify outliers
outliers <- nmsu22_nor_pheno %>% filter(abs(residuals_standard) > 3)
# Print outliers
View(outliers)
# one, not removing for now.

```


Run the models

```{r run univariate models with spatial and GRM 08122025}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
# first for NDVI
# we have 32 imaging events so 32 heritabilities here. 
# 62, 76, 90, 96, 104, 110, 114, 124, 130, 138 (10th), 144, 150, 160, 166, 174, 180, 188, 194, 200, 206 (20th), 216, 222, 228, 236, 240, 250, 256, 264, 270, 278, 284, 292

#  For NDVI_104
st_nmsu22_nor_104 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_104 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_104)$varcomp

asreml::vpredict(st_nmsu22_nor_104, heritA ~ V1 / (V1 + V5)) # 0.38
# without grm 0.34

# run for all time points and yield below

```



For yield 
now with new GRM... slightly changed numbers (close).. haven't changed it yet
```{r Heritability of yield with GRM 081225}

asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
# for yield 1
st_nmsu22_nor_yld1 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld1)$varcomp
asreml::vpredict(st_nmsu22_nor_yld1, heritA ~ V1 / (V1 + V5)) # 0.58 with 0.12 SE

# for yield 2
 # asreml.options(extra=1000,maxit =1000) # to setup more iterations for successful burn-in or convergence

st_nmsu22_nor_yld2 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld2)$varcomp
asreml::vpredict(st_nmsu22_nor_yld2, heritA ~ V1 / (V1 + V5)) # 0.63 with 0.11 SE

# for yield 3
st_nmsu22_nor_yld3 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld3)$varcomp
asreml::vpredict(st_nmsu22_nor_yld3, heritA ~ V1 / (V1 + V5)) # 0.63 with 0.10 SE


# for yield 4
st_nmsu22_nor_yld4 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld4)$varcomp
asreml::vpredict(st_nmsu22_nor_yld4, heritA ~ V1 / (V1 + V5)) # 0.65 with 0.09 SE


# for yield 5
st_nmsu22_nor_yld5 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld5)$varcomp
asreml::vpredict(st_nmsu22_nor_yld5, heritA ~ V1 / (V1 + V5)) # 0.68 with 0.09 SE


# for yield 6
st_nmsu22_nor_yld6 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld6 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld6)$varcomp
asreml::vpredict(st_nmsu22_nor_yld6, heritA ~ V1 / (V1 + V5)) # 0.62 with 0.099 SE

```





# Run for all VI's
### STORE BOTH HERITABILITY AND VARIANCE COMPONENTS

 To check variance components as well: store VC in another data frame
```{r For all VI's 08122025}
asreml.options(extra=300,maxit =300) # to setup more iterations for successful burn-in or convergence

# Define time points
times <- c(62, 76, 90, 96, 104, 110, 114, 124, 130, 138, 
           144, 150, 160, 166, 174, 180, 188, 194, 200, 206,
           216, 222, 228, 236, 240, 250, 256, 264, 270, 278, 
           284, 292)

# Initialize heritability results
herit_results <- data.frame(
  Time = times,
  Herit_ndvi = NA, Herit_ndvi_spa = NA,
  Herit_gndvi = NA, Herit_gndvi_spa = NA,
  Herit_ndre = NA, Herit_ndre_spa = NA
)

# Initialize variance component storage
varcomp_results <- data.frame(
  Time = numeric(0),
  Index = character(0),
  Model = character(0),
  Component = character(0),
  Estimate = numeric(0),
  stringsAsFactors = FALSE
)

# Function to fit model and return heritability and variance components
fit_heritability <- function(index_prefix, time_point, data) {
  col_name <- paste0(index_prefix, "_", time_point)
  
  if (!col_name %in% colnames(data)) {
    warning(paste("Column", col_name, "not found. Skipping."))
    return(list(h2 = c(NA, NA), vc = NULL))
  }
  
  # --- Simple Model ---
  mod_simple <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Simple model failed for", col_name))
    return(NULL)
  })
  
  h2_simple <- NA
  vc_simple <- NULL
  if (!is.null(mod_simple)) {
    h2_simple <- tryCatch({
      vpredict(mod_simple, herit ~ V1 / (V1 + V2))$Estimate
    }, error = function(e) NA)
    
    vc_simple <- tryCatch({
      vc <- summary(mod_simple)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Simple")
    }, error = function(e) NULL)
  }
  
  # --- Spatial Model ---
  mod_spatial <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Spatial model failed for", col_name))
    return(NULL)
  })
  
  h2_spatial <- NA
  vc_spatial <- NULL
  if (!is.null(mod_spatial)) {
    h2_spatial <- tryCatch({
      vpredict(mod_spatial, herit_spatial ~ V1 / (V1 + V5))$Estimate
    }, error = function(e) NA)
    
    vc_spatial <- tryCatch({
      vc <- summary(mod_spatial)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Spatial")
    }, error = function(e) NULL)
  }

  return(list(h2 = c(h2_simple, h2_spatial),
              vc = rbind(vc_simple, vc_spatial)))
}

# Loop through all time points and indices
for (i in seq_along(times)) {
  time <- times[i]
  cat("\nProcessing time point:", time, "\n")
  
  for (index in c("NDVI", "GNDVI", "NDRE")) {
    result <- fit_heritability(index, time, nmsu22_nor_pheno)
    
    # Store heritabilities
    herit_results[i, paste0("Herit_", tolower(index))] <- result$h2[1]
    herit_results[i, paste0("Herit_", tolower(index), "_spa")] <- result$h2[2]
    
    # Store variance components
    if (!is.null(result$vc)) {
      result$vc$Index <- index
      result$vc$Time <- time
      varcomp_results <- bind_rows(varcomp_results, result$vc)
    }
  }
}

# Round heritability results to 2 decimals
herit_results_nor <- herit_results %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

# Done!
print(head(herit_results_nor))
print(head(varcomp_results))

```

check for NA's
- NDVI with spatial 278
- GNDVi with spatial: 160 and 250
- NDRE with spatial: 160
*conclusion*: remove 160 and 250 ( poor heritability, all indices < 0.1)
```{r}
# NDVI: 278, 0.42 with row
st_nmsu22_nor_ndvi_278 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_278 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_ndvi_278)$varcomp
asreml::vpredict(st_nmsu22_nor_ndvi_278, heritA ~ V3 / (V3 + V4)) # 0.42 with row only.. restrained with range


# GNDVI: bound with row , can't trust REMOVE
st_nmsu22_nor_gndvi_160 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_160 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_160)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_160, heritA ~ V3 / (V3 + V4)) # 0.28 with row only, but bound F.. restrained with row

# bound with row , can't trust REMOVE (also bad correlation with yield 6 ; found out in later analysis)
st_nmsu22_nor_gndvi_250 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_250 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_250)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_250, heritA ~ V3 / (V3 + V4)) # 0.28 with row only, but bound F.. restrained with row


# lets remove this or set as 0
herit_results_nor$Herit_ndvi_spa[herit_results_nor$Time == 160] <- 0
herit_results_nor$Herit_gndvi_spa[herit_results_nor$Time == 160] <- 0
herit_results_nor$Herit_ndre_spa[herit_results_nor$Time == 160] <- 0

# also 250 gndvi, and 278 ndre, 278 ndvi as 0.42
herit_results_nor$Herit_gndvi_spa[herit_results_nor$Time == 250] <- 0
herit_results_nor$Herit_ndre_spa[herit_results_nor$Time == 278] <- 0

# remove time points: 160, 250 ( <0.2 so won't have any impact on mt models)
```


Checking some very high heritabilities (eg. 0.8 in GNDVI but 0 in NDVI) to make sure the realibility of the estimates.
- NDVI 250
- NDRE 216
- GNDVI 62, 76, 90, 174, 166, 180, 188, 216, 222, 228
conclusion: no issues here
```{r}
asreml.options(maxit = 500, extra = 500)
# 250 in NDVI: just to make sure it's catching something real as other two are not explaining anything..
st_nmsu22_nor_ndvi_250 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_250 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_ndvi_250)$varcomp
asreml::vpredict(st_nmsu22_nor_ndvi_250, heritA ~ V1 / (V1 + V5)) # no issues here, 0.15


# NDRE:  216
st_nmsu22_nor_ndre_216 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_216 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_ndre_216)$varcomp
asreml::vpredict(st_nmsu22_nor_ndre_216, heritA ~ V1 / (V1 + V5)) # no issues here, 0.95



# GNDVI: 62, 76, 90, 174, 166, 180, 188, 216, 222, 228

st_nmsu22_nor_gndvi_62 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_62 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_62)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_62, heritA ~ V1 / (V1 + V5)) # no issues here. 

st_nmsu22_nor_gndvi_76 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_76 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_76)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_76, heritA ~ V1 / (V1 + V5)) # no issues here. 

st_nmsu22_nor_gndvi_90 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_90 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_90)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_90, heritA ~ V1 / (V1 + V5)) # no issues here. 

st_nmsu22_nor_gndvi_174 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_174 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_174)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_174, heritA ~ V1 / (V1 + V5)) # no issues here. 

st_nmsu22_nor_gndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_166 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_166)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_166, heritA ~ V1 / (V1 + V5)) # no issues here. 


st_nmsu22_nor_gndvi_180 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_180 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_180)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_180, heritA ~ V1 / (V1 + V5)) # no issues here. 


st_nmsu22_nor_gndvi_188 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_188 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_188)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_188, heritA ~ V1 / (V1 + V5)) # no issues here. 

st_nmsu22_nor_gndvi_216 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_216 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_216)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_216, heritA ~ V1 / (V1 + V5)) # no issues here. 


st_nmsu22_nor_gndvi_222 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_222 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_222)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_222, heritA ~ V1 / (V1 + V5)) # no issues here. 


st_nmsu22_nor_gndvi_228 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_228 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_228)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_228, heritA ~ V1 / (V1 + V5)) # no issues here, 0.80


# no issues with high values

write.csv(herit_results_nor, file = "2022/MSI_data/output_files/Heritabilities_withGRM_final_MSI_nor_22_aa.csv", row.names = F)


```





make box plot to visulaize and compare 

```{r}
herit_results_nor <- read.csv("2022/MSI_data/output_files/Heritabilities_withGRM_final_MSI_nor_22_aa.csv", sep = ",") 


# Reshape all six heritability columns to long format
herit_nor_long_all <- herit_results_nor %>%
  select(Time, Herit_ndvi, Herit_ndvi_spa,
               Herit_gndvi, Herit_gndvi_spa,
               Herit_ndre, Herit_ndre_spa) %>%
  pivot_longer(cols = -Time, names_to = "Type", values_to = "Heritability") %>%
  mutate(
    Index = case_when(
  grepl("gndvi", Type) ~ "GNDVI",
  grepl("ndre", Type) ~ "NDRE",
  grepl("ndvi", Type) ~ "NDVI"
),
    Model = ifelse(grepl("_spa", Type), "Spatial", "Simple")
  )


gg_nor <- ggplot(herit_nor_long_all, aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") + # Randomly "jitter" (spread out) the data points horizontally by up to ±0.2 units on the x-axis (i.e., within each box)
  # alternate geom_point() where all will be stacked in a vertical line so overalp issue.
  facet_wrap(~ Model) +
  labs(title = "h² by VI and Model Type in Normal NMSU 2022",
       x = "Vegetation Index",
       y = "Heritability (h²)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(c(0,1)) +
  scale_fill_brewer(palette = "Set2")

gg_nor

ggsave(filename = "Boxplot_Heritability_h2_VI_nmsu_normal_2022.png", plot = gg_nor, path = "2022/output_images", width = 8, height = 7,dpi = 400)

```







08112025 also testing if by rescaling gndvi early phase where the numbers are too low or negative (multiply by 10)
### variance component values increase by constant^2 but final heritability stays the same

```{r}
asreml.options(maxit = 200, extra= 200)
st_nmsu22_nor_gndvi_62 <- { # trying without GRM. no specific reasons, tried with GRM too, still same
  asreml(fixed = GNDVI_62 ~ 1 + Rep, #rep effect as fixed
         random = ~ id(Germplasm) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_62)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_62, heritA ~ V1 / (V1 + V5))  # 0.77

rescale # same result, the variance component increased by 100 (10^2) but the final heritability was same...

st_nmsu22_nor_gndvi_62_scaled <- { # trying without GRM
  asreml(fixed = GNDVI_62*10 ~ 1 + Rep, #rep effect as fixed
         random = ~ id(Germplasm) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_62_scaled)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_62_scaled, heritA ~ V1 / (V1 + V5))  # 0.77



# lets test one more
st_nmsu22_nor_gndvi_188 <- { # trying without GRM
  asreml(fixed = GNDVI_188 ~ 1 + Rep, #rep effect as fixed
         random = ~ id(Germplasm) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_188)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_188, heritA ~ V1 / (V1 + V5))  # 0.71

rescale # same result, the variance component increased by 100 (10^2) but the final heritability was same...

st_nmsu22_nor_gndvi_188_scaled <- { # trying without GRM
  asreml(fixed = GNDVI_188*10 ~ 1 + Rep, #rep effect as fixed
         random = ~ id(Germplasm) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi_188_scaled)$varcomp
asreml::vpredict(st_nmsu22_nor_gndvi_188_scaled, heritA ~ V1 / (V1 + V5))  # 0.71
```




For showing spatial variation in manuscript, made a table out of simple vs spatial models
- did for Nor 2022 (NMSU trial with higher spatial)
- and for Snyder 2023 (less spatial plot)

GNDVI_114  <->  yld1
GNDVI_150  <->  yld2
GNDVI_180  <->  yld3
GNDVI_206  <->  yld4
GNDVI_240  <->  yld5
GNDVI_292  <->  yld6
```{r NOR 22 Variance components of yield and last flight }
### =====================================================
### Harvest 1
### =====================================================

# GNDVI_114
st_nmsu22_nor_gndvi114 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_114 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi114)$varcomp

st_nmsu22_nor_gndvi114_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_114 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi114_spa)$varcomp

# yld1
st_nmsu22_nor_yld1 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld1 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld1)$varcomp

st_nmsu22_nor_yld1_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld1 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld1_spa)$varcomp


### =====================================================
### Harvest 2: GNDVI_150
### =====================================================

# GNDVI_150
st_nmsu22_nor_gndvi150 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_150 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi150)$varcomp

st_nmsu22_nor_gndvi150_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_150 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi150_spa)$varcomp

# yld2
st_nmsu22_nor_yld2 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld2 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld2)$varcomp

st_nmsu22_nor_yld2_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld2 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld2_spa)$varcomp

### =====================================================
### Harvest 3: GNDVI_180  <->  yld3
### =====================================================

# GNDVI_180
st_nmsu22_nor_gndvi180 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_180 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi180)$varcomp

st_nmsu22_nor_gndvi180_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_180 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi180_spa)$varcomp

# yld3
st_nmsu22_nor_yld3 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld3 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld3)$varcomp

st_nmsu22_nor_yld3_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld3 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld3_spa)$varcomp

### =====================================================
### Harvest 4: GNDVI_206  <->  yld4
### =====================================================

# GNDVI_206
st_nmsu22_nor_gndvi206 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_206 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi206)$varcomp

st_nmsu22_nor_gndvi206_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_206 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi206_spa)$varcomp

# yld4
st_nmsu22_nor_yld4 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld4 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld4)$varcomp

st_nmsu22_nor_yld4_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld4 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld4_spa)$varcomp

### =====================================================
### Harvest 5: GNDVI_240  <->  yld5
### =====================================================

# GNDVI_240
st_nmsu22_nor_gndvi240 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_240 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi240)$varcomp

st_nmsu22_nor_gndvi240_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_240 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi240_spa)$varcomp

# yld5
st_nmsu22_nor_yld5 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld5 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld5)$varcomp

st_nmsu22_nor_yld5_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld5 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld5_spa)$varcomp

### =====================================================
### Harvest 6: GNDVI_292  <->  yld6
### =====================================================

# GNDVI_292
st_nmsu22_nor_gndvi292 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_292 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi292)$varcomp

st_nmsu22_nor_gndvi292_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_292 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_gndvi292_spa)$varcomp

# yld6
st_nmsu22_nor_yld6 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld6 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld6)$varcomp

st_nmsu22_nor_yld6_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld6 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld6_spa)$varcomp

```


















################################# Early Termination ################################






2. *First treatment NMSU_Early_Termination 2022 *
with spatial in the model, random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),



*Early_Termination irrigation treatment*
Loading and preparing data files
```{r loading files 06272025}
#import merged phenotype file 
nmsu22_et_pheno <- read.csv("2022/MSI_data/Output_files/nmsu22_et_combined_phenotypes_MSI.csv", sep = ",") %>% 
  select(-X) # remove index column

# Before running models, make sure the variables are in correct form. Need to convert into factors and numeric if needed.
  # converting all factorial variables into factor
nmsu22_et_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu22_et_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
  ## make sure all variables are in correct form (factors or numeric)

#import genotype file
grm_aa <- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1)
grm_aa <- as.matrix(grm_aa)
mean(diag(grm_aa)) # 1
hist(diag(grm_aa))

grm_agh <- grm_aa+diag(.0005,length(grm_aa[,1]),length(grm_aa[,1]))
grm_inverse <- solve(grm_agh)

#now run the model for now without GRM but need to find GRM.

```


Normally if the range of numbers vary a lot between vi's and yield, then we can scale yield to variance level of vi's. but here since yield is in lb and upto 2 digit, should have that issue. If got any, we can always use scaled yield.
```{r scaling yield }
# we only have to scale 5 harvest values based on NDVI's value. Of course the variance between VI's differ but it should be okay to use Yield scaled onto NDVI for analysis with NDRE and GNDVI as well. 

# Define the mapping of yield traits to NDVI columns
yield_traits <- paste0("yld", 1:5)
ndvi_times <- c("NDVI_110", "NDVI_144", "NDVI_180", "NDVI_200", "NDVI_270") # selecting one of the time point in each harvest window. selecting based on higher variation time point from heatmap, not actually calculating all. 

# Loop over each yield trait
for (i in 1:length(yield_traits)) {
  
  trait <- yield_traits[i]
  ndvi_col <- ndvi_times[i]
  
  # Calculate variance of NDVI and yield
  ndvi_var <- var(nmsu22_et_pheno[[ndvi_col]], na.rm = TRUE)
  yld_var <- var(nmsu22_et_pheno[[trait]], na.rm = TRUE)
  
  # Calculate scaling factor
  sf <- sqrt(ndvi_var) / sqrt(yld_var)
  
  # Create new scaled yield column
  new_col <- paste0("sca_", trait)
  nmsu22_et_pheno[[new_col]] <- nmsu22_et_pheno[[trait]] * sf
  
  #  Print scaling factor
  cat("Scaling factor for", trait, "using", ndvi_col, ":", sf, "\n")
}

# thus all scaled yields are added as a new column at last. 
```


one on yield 5 but not removing as its close to 3. 
```{r for checking outliers}
# for all yields
library(dplyr)
library(purrr)


# Traits
yield_traits_et <- paste0("yld", 1:5)

# Function to fit model, calculate residuals, standardize, and extract outliers
process_outliers_et <- function(trait, data) {
  # Build formula dynamically
  formula <- as.formula(paste(trait, "~ Rep + Row + Range + Germplasm"))
  
  # Fit model
  model <- lm(formula, data = data, na.action = na.exclude)
  
  # Get residuals and standardized residuals
  data[[paste0(trait, "_resid")]] <- residuals(model, na.action = na.exclude)
  data[[paste0(trait, "_resid_std")]] <- scale(data[[paste0(trait, "_resid")]], center = TRUE, scale = TRUE)
  
  # Extract outliers
  outliers <- data %>% filter(abs(.data[[paste0(trait, "_resid_std")]]) > 3)
  
  return(outliers)
}

# Apply the function to all yield traits
outlier_list_et <- map(yield_traits_et, ~process_outliers(.x, nmsu22_et_combined))

# Name the list elements for clarity
names(outlier_list_et) <- yield_traits_et

# View outliers for each trait

# plot and standard residual value

View(outlier_list_et$yld1) # none
View(outlier_list_et$yld2) # none
View(outlier_list_et$yld3) # none
View(outlier_list_et$yld4) # none
View(outlier_list_et$yld5) # NM1703PAR germplasm, plot 316 with -3.18 (no need to remove)

```


```{r run univariate models with spatial 08132025}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
# first for NDVI
# we have 32 imaging events so 32 heritabilities here. 
# 62, 76, 90, 96, 104, 110, 114, 124, 130, 138 (10th), 144, 150, 160, 166, 174, 180, 188, 194, 200, 206 (20th), 216, 222, 228, 236, 240, 250, 256, 264, 270, 278, 284, 292
 # plus heritability of yields too


# For NDVI_114
st_nmsu22_et_114 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_114 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_114)$varcomp

asreml::vpredict(st_nmsu22_et_114, heritA ~ V1 / (V1 + V5)) # 0.56

# run for all time points and all VI's
```



# for yield first and then VI's
```{r Heritability of yield (Broad Sense as we don't have GRM) 06282025}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
# for yield 1
st_nmsu22_et_yld1 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld1)$varcomp
asreml::vpredict(st_nmsu22_et_yld1, heritA ~ V1 / (V1 + V5)) # 0.49 with 0.14 SE with both row and range


# for yield 2
 # asreml.options(extra=1000,maxit =1000) # to setup more iterations for successful burn-in or convergence

st_nmsu22_et_yld2 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld2)$varcomp
asreml::vpredict(st_nmsu22_et_yld2, heritA ~ V1 / (V1 + V5)) # 0.61 with 0.14 SE

# for yield 3
st_nmsu22_et_yld3 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld3)$varcomp
asreml::vpredict(st_nmsu22_et_yld3, heritA ~ V1 / (V1 + V5)) # 0.73 with 0.096 SE


# for yield 4
st_nmsu22_et_yld4 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld4)$varcomp
asreml::vpredict(st_nmsu22_et_yld4, heritA ~ V1 / (V1 + V5)) # 0.54 with 0.12 SE


# for yield 5
st_nmsu22_et_yld5 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld5)$varcomp
asreml::vpredict(st_nmsu22_et_yld5, heritA ~ V1 / (V1 + V5)) # 0.83 with 0.07 SE

```



# Run for all VI's
### STORE BOTH HERITABILITY AND VARIANCE COMPONENTS

#### Ran all and then filtered based on model ran without GRM. Codes in without GRM file. Here will remove those not useful time points and the time points during ET.

----------------------------------------------- FILTER TIME POINTS ------------------------------------------------- 

#### DOY for June 22nd through September 9th, 2022

```{r now removing such untrusty time points as well as termination time points and exporting final clean copy of dataset}
# 1. Remove termination time points
 # filtering time points based on heritability values and the fact that the flights during termination period  have no information about the plants. so removing all dates from July 1- 182 doy and 252 doy.( the actual time of ET is 173 to 252). But the impact started observing in 182 flight.

colnames(nmsu22_et_pheno)
#removing all dates from July 1- 182 doy and early sept - 252 doy.
# Identify DOYs to exclude from colnames above
exclude_doys <- 182:252
 # these include 188, 194, 200, 206, 216, 222, 228, 236, 240, 250

# Get column names to exclude (VI columns with these DOYs)
exclude_cols <- grep(paste0("_(", paste(exclude_doys, collapse = "|"), ")$"), colnames(nmsu22_et_pheno), value = TRUE)

# Make a filtered dataset
nmsu22_et_pheno_filtered <- nmsu22_et_pheno[, !(colnames(nmsu22_et_pheno) %in% exclude_cols)]
colnames(nmsu22_et_pheno_filtered) # now the time points jump from 180 to 256 straight. 


# 2. remove rest of time points with either very low heritability (<0.1) or unusually high of 1 and then later found those were not converging which were tested in above chunks
# use above filtered dataset and do further filtering
 ## for ndvi, remove 
    # 110(0.03 heritability, so out), 
    # 160 (bound, not converging with either row or range or both), 
       # 194 also has issue but anyways removing bcz of in termination zone. 
    # 250 ( no heritability)

 ## for gndvi, remove:
    #160


 ## for ndre, remove:
  # 110 ( o heritability), 160 (0 heritability), 174(didn't converge), 250 (0 H2), 256(0 H2)

# now lets remove
nmsu22_et_pheno_filtered <- nmsu22_et_pheno_filtered %>%
  select(-c(NDVI_110, NDVI_160, GNDVI_160, NDRE_110, NDRE_160, NDRE_174, NDRE_256))


# manually checked and all values are in right place, now lets export this file for future use. 

write.csv(nmsu22_et_pheno_filtered, file = "2022/MSI_data/output_files/nmsu_phenotypes_et_filtered_22", row.names = F)

```



# now that we have filtered data set. lets run H2 on this file and change some NAs as above to get final H2 estimates
 # for yield, nothing changed, its same. 
 # for VI: lets run the loop
 

# Run for all VI's without termination time point
### STORE BOTH HERITABILITY AND VARIANCE COMPONENTS

 To check variance components as well: store VC in another data frame
```{r For all VI's 08132025}
nmsu22_et_pheno_filtered <- read.csv("2022/MSI_data/output_files/nmsu_phenotypes_et_filtered_22")
# Before running models, make sure the variables are in correct form. Need to convert into factors and numeric if needed.
  # converting all factorial variables into factor
nmsu22_et_pheno_filtered[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu22_et_pheno_filtered[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(nmsu22_et_pheno_filtered)
# Define time points and only select those that are in the filtered file. even if only present in 1 index.
times <- c(62, 76, 90, 96, 104, 110, 114, 124, 130, 138, 144, 150, 166, 174, 180, 256, 264, 270, 278, 284, 292) 

# Initialize heritability results
herit_results <- data.frame(
  Time = times,
  Herit_ndvi = NA, Herit_ndvi_spa = NA,
  Herit_gndvi = NA, Herit_gndvi_spa = NA,
  Herit_ndre = NA, Herit_ndre_spa = NA
)

# Initialize variance component storage
varcomp_results <- data.frame(
  Time = numeric(0),
  Index = character(0),
  Model = character(0),
  Component = character(0),
  Estimate = numeric(0),
  stringsAsFactors = FALSE
)

# Function to fit model and return heritability and variance components
fit_heritability <- function(index_prefix, time_point, data) {
  col_name <- paste0(index_prefix, "_", time_point)
  
  if (!col_name %in% colnames(data)) {
    warning(paste("Column", col_name, "not found. Skipping."))
    return(list(h2 = c(NA, NA), vc = NULL))
  }
  
  # --- Simple Model ---
  mod_simple <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Simple model failed for", col_name))
    return(NULL)
  })
  
  h2_simple <- NA
  vc_simple <- NULL
  if (!is.null(mod_simple)) {
    h2_simple <- tryCatch({
      vpredict(mod_simple, herit ~ V1 / (V1 + V2))$Estimate
    }, error = function(e) NA)
    
    vc_simple <- tryCatch({
      vc <- summary(mod_simple)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Simple")
    }, error = function(e) NULL)
  }
  
  # --- Spatial Model ---
  mod_spatial <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Spatial model failed for", col_name))
    return(NULL)
  })
  
  h2_spatial <- NA
  vc_spatial <- NULL
  if (!is.null(mod_spatial)) {
    h2_spatial <- tryCatch({
      vpredict(mod_spatial, herit_spatial ~ V1 / (V1 + V5))$Estimate
    }, error = function(e) NA)
    
    vc_spatial <- tryCatch({
      vc <- summary(mod_spatial)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Spatial")
    }, error = function(e) NULL)
  }

  return(list(h2 = c(h2_simple, h2_spatial),
              vc = rbind(vc_simple, vc_spatial)))
}

# Loop through all time points and indices
for (i in seq_along(times)) {
  time <- times[i]
  cat("\nProcessing time point:", time, "\n")
  
  for (index in c("NDVI", "GNDVI", "NDRE")) {
    result <- fit_heritability(index, time, nmsu22_et_pheno_filtered) # data file name
    
    # Store heritabilities
    herit_results[i, paste0("Herit_", tolower(index))] <- result$h2[1]
    herit_results[i, paste0("Herit_", tolower(index), "_spa")] <- result$h2[2]
    
    # Store variance components
    if (!is.null(result$vc)) {
      result$vc$Index <- index
      result$vc$Time <- time
      varcomp_results <- bind_rows(varcomp_results, result$vc)
    }
  }
}

# Round heritability results to 2 decimals
herit_results_et <- herit_results %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

# Done!
print(head(herit_results_et))
print(head(varcomp_results))


```

```{r fix NA's}
# we have some time points that were not present in NDVI or NDRE but were in GNDVI. so those were kept above in the loop to make it easy to run the loop. 
# now lets remove those. The results of those will be NA in the herit_results. 

# NDVI 110, 124, 278
# For NDVI_110: set as NA, didn't converge with any
st_nmsu22_et_110 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_110 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_110)$varcomp
asreml::vpredict(st_nmsu22_et_110, heritA ~ V1 / (V1 + V5)) # 0.56



# For NDVI_124 : remove this
st_nmsu22_et_124 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_124 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_124)$varcomp


# For NDVI_278
st_nmsu22_et_278 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_278 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_278)$varcomp
asreml::vpredict(st_nmsu22_et_114, heritA ~ V3 / (V3 + V4)) # 0.35 with row only



# NDRE 90, 110, 124, 174, 256

# For NDRE_90
st_nmsu22_et_90 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_90 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_90)$varcomp

asreml::vpredict(st_nmsu22_et_90, heritA ~ V3 / (V3 + V4)) # 0.10 with row only


# For NDRE_110: remove this...
st_nmsu22_et_110 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_110 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_110)$varcomp

# For NDRE_174: remove this
st_nmsu22_et_174 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_174 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_174)$varcomp



# For NDRE_256: remove this
st_nmsu22_et_256 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_256 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_256)$varcomp

# filtering time

# remove time point 124 as well (none for ndvi and ndre, low for gndvi): for all
ncol(nmsu22_et_pheno) #106
nmsu22_et_pheno <- nmsu22_et_pheno %>%
  select(-NDVI_124, -GNDVI_124, - NDRE_124)
ncol(nmsu22_et_pheno) #103
# also from herit results
herit_results_et <- herit_results_et %>%
  filter(Time != 124) # keep all rows except 124 time point

# replace ndvi 278 with 0.35 (row) 
# replace ndre 90 with 0.10 (row),  110 (bound), 174 (bound) and 256 (bound)

# NDVI
herit_results_et$Herit_ndvi_spa[herit_results_et$Time == 278] <- 0
# NDRE
herit_results_et$Herit_ndre_spa[herit_results_et$Time == 90] <- 0.10
herit_results_et$Herit_ndre_spa[herit_results_et$Time == 110] <- 0
herit_results_et$Herit_ndre_spa[herit_results_et$Time == 174] <- 0
herit_results_et$Herit_ndre_spa[herit_results_et$Time == 256] <- 0



 #now export the file
write.csv(herit_results_et, file = "2022/MSI_data/output_files/Heritabilities_filtered_MSI_NMSU_ET_22.csv", row.names = F)

```





GNDVI and NDRE >> NDVI


Comparison box plots: NA's will be ignored by ggplot2

```{r to compare VI's after filtering time points}

herit_results_et_22 <- read.csv("2022/MSI_data/output_files/Heritabilities_filtered_MSI_NMSU_ET_22.csv", sep = ",") 


# Reshape all six heritability columns to long format
herit_long_all_et_22 <- herit_results_et_22 %>%
  select(Time, Herit_ndvi, Herit_ndvi_spa,
               Herit_gndvi, Herit_gndvi_spa,
               Herit_ndre, Herit_ndre_spa) %>%
  pivot_longer(cols = -Time, names_to = "Type", values_to = "Heritability") %>%
  mutate(
    Index = case_when(
  grepl("gndvi", Type) ~ "GNDVI",
  grepl("ndre", Type) ~ "NDRE",
  grepl("ndvi", Type) ~ "NDVI"
),
    Model = ifelse(grepl("_spa", Type), "Spatial", "Simple")
  )


herit_long_all_et_22_gg<- ggplot(herit_long_all_et_22, aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") + # Randomly "jitter" (spread out) the data points horizontally by up to ±0.2 units on the x-axis (i.e., within each box)
  # alternate geom_point() where all will be stacked in a vertical line so overalp issue.
  facet_wrap(~ Model) +
  labs(title = "Comparison of h² by VI and Model_type in ET 22 NMSU",
       x = "Vegetation Index",
       y = "Heritability (h²)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2") +
  coord_cartesian(ylim = c(0, 1))
herit_long_all_et_22_gg


ggsave(filename = "Boxplot_Heritability_h2_NarrowSense_Comparison_VI_nmsu_et_2022.png", plot = herit_long_all_et_22_gg, path = "2022/output_images", width = 8, height = 7,dpi = 400)

```


 

GNDVI_114  <->  yld1
GNDVI_150  <->  yld2
GNDVI_180  <->  yld3
yld4
GNDVI_292  <->  yld5
```{r ET 2022 Variance Components of Yield and last flight 06052026}
### =====================================================
### ET 22 Variance components
### =====================================================

### =====================================================
### Harvest 1
### =====================================================

# GNDVI_114
st_nmsu22_et_gndvi114 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_114 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_gndvi114)$varcomp

st_nmsu22_et_gndvi114_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_114 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_gndvi114_spa)$varcomp

# yld1
st_nmsu22_et_yld1 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld1 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld1)$varcomp

st_nmsu22_et_yld1_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld1 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld1_spa)$varcomp


### =====================================================
### Harvest 2: GNDVI_150
### =====================================================

# GNDVI_150
st_nmsu22_et_gndvi150 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_150 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_gndvi150)$varcomp

st_nmsu22_et_gndvi150_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_150 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_gndvi150_spa)$varcomp

# yld2
st_nmsu22_et_yld2 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld2 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld2)$varcomp

st_nmsu22_et_yld2_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld2 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld2_spa)$varcomp


### =====================================================
### Harvest 3: GNDVI_180 <-> yld3
### =====================================================

# GNDVI_180
st_nmsu22_et_gndvi180 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_180 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_gndvi180)$varcomp

st_nmsu22_et_gndvi180_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_180 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_gndvi180_spa)$varcomp

# yld3
st_nmsu22_et_yld3 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld3 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld3)$varcomp

st_nmsu22_et_yld3_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld3 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld3_spa)$varcomp


### =====================================================
### Harvest 4: yld4
### =====================================================

st_nmsu22_et_yld4 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld4 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld4)$varcomp

st_nmsu22_et_yld4_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld4 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld4_spa)$varcomp


### =====================================================
### Harvest 5: GNDVI_292 <-> yld5
### =====================================================

# GNDVI_292
st_nmsu22_et_gndvi292 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_292 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_gndvi292)$varcomp

st_nmsu22_et_gndvi292_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_292 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_gndvi292_spa)$varcomp

# yld5
st_nmsu22_et_yld5 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld5 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld5)$varcomp

st_nmsu22_et_yld5_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld5 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld5_spa)$varcomp


### =====================================================
### Variance component summary table
### =====================================================

varcomp_nmsu22_et <- data.frame(
  Trait = c("GNDVI_114","yld1",
            "GNDVI_150","yld2",
            "GNDVI_180","yld3",
            "yld4",
            "GNDVI_292","yld5"),
  Genetic = c(
    summary(st_nmsu22_et_gndvi114_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld1_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_gndvi150_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld2_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_gndvi180_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld3_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld4_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_gndvi292_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld5_spa)$varcomp["vm(Germplasm, grm_inverse)","component"]),
  Spatial = c(
    summary(st_nmsu22_et_gndvi114_spa)$varcomp["Row:Range","component"],
    summary(st_nmsu22_et_yld1_spa)$varcomp["Row:Range","component"],
    summary(st_nmsu22_et_gndvi150_spa)$varcomp["Row:Range","component"],
    summary(st_nmsu22_et_yld2_spa)$varcomp["Row:Range","component"],
    summary(st_nmsu22_et_gndvi180_spa)$varcomp["Row:Range","component"],
    summary(st_nmsu22_et_yld3_spa)$varcomp["Row:Range","component"],
    summary(st_nmsu22_et_yld4_spa)$varcomp["Row:Range","component"],
    summary(st_nmsu22_et_gndvi292_spa)$varcomp["Row:Range","component"],
    summary(st_nmsu22_et_yld5_spa)$varcomp["Row:Range","component"]),
  Residual = c(
    summary(st_nmsu22_et_gndvi114_spa)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld1_spa)$varcomp["units!R","component"],
    summary(st_nmsu22_et_gndvi150_spa)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld2_spa)$varcomp["units!R","component"],
    summary(st_nmsu22_et_gndvi180_spa)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld3_spa)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld4_spa)$varcomp["units!R","component"],
    summary(st_nmsu22_et_gndvi292_spa)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld5_spa)$varcomp["units!R","component"])
)

varcomp_nmsu22_et




### =====================================================
### Variance component summary table (Base model)
### =====================================================

varcomp_nmsu22_et_base <- data.frame(
  Trait = c("GNDVI_114","yld1",
            "GNDVI_150","yld2",
            "GNDVI_180","yld3",
            "yld4",
            "GNDVI_292","yld5"),
  Genetic = c(
    summary(st_nmsu22_et_gndvi114)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld1)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_gndvi150)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld2)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_gndvi180)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld3)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld4)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_gndvi292)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_nmsu22_et_yld5)$varcomp["vm(Germplasm, grm_inverse)","component"]),
  Residual = c(
    summary(st_nmsu22_et_gndvi114)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld1)$varcomp["units!R","component"],
    summary(st_nmsu22_et_gndvi150)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld2)$varcomp["units!R","component"],
    summary(st_nmsu22_et_gndvi180)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld3)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld4)$varcomp["units!R","component"],
    summary(st_nmsu22_et_gndvi292)$varcomp["units!R","component"],
    summary(st_nmsu22_et_yld5)$varcomp["units!R","component"])
)

varcomp_nmsu22_et_base

```




 
 