---
title: "Running multivariate models"
author: "Dinesh Ghimire"
date: "2025-08-22"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr","asreml", "readxl", "tidyr", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "GGally")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```

#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NMSU"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NMSU 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa_NMSU"
setwd("~/FFAR-Alfalfa-NMSU")
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is misssing please let me know. 
key<-read.csv("2022/meta_data/cultivar_list_key.csv") # contains both CU and NMSU cultivars here. 
# its same for 2022 and 2023 (cultivar list)
```

USE FILTERED FILES HERE... 



1. *First treatment NMSU_Normal 2022 *
with spatial in the model, random = ~ vm(Material_Id, ct_geno_inverse) + ar1(Row):ar1(Range),



*Normal irrigation treatment*
Loading and preparing data files
```{r loading files 08262025}
#import merged phenotype file 
nmsu22_nor_pheno <- read.csv("2022/MSI_data/Output_files/nmsu22_nor_combined_filtered_phenotypes_MSI.csv", sep = ",") %>% 
  select(-X) # remove index column
colnames(nmsu22_nor_pheno)
# Before running models, make sure the variables are in correct form. Need to convert into factors and numeric if needed.
  # converting all factorial variables into factor
nmsu22_nor_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu22_nor_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)

#import genotype file
grm_aa <- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1)
grm_aa <- as.matrix(grm_aa)
mean(diag(grm_aa)) # 1
hist(diag(grm_aa))

grm_agh <- grm_aa+diag(.0005,length(grm_aa[,1]),length(grm_aa[,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


Normally if the range of numbers vary a lot between vi's and yield, then we can scale yield to variance level of vi's. but here since yield is in lb and upto 2 digit, should have that issue. If got any, we can always use scaled yield.
```{r scaling yield }
# we only have to scale 6 harvest values based on GNDVI's value. Of course the variance between VI's differ but it should be okay to use Yield scaled onto GNDVI for analysis with NDRE and GGNDVI as well. 

# Define the mapping of yield traits to GNDVI columns
yield_traits <- paste0("yld", 1:6)
GNDVI_times <- c("GNDVI_114", "GNDVI_144", "GNDVI_180", "GNDVI_206", "GNDVI_236", "GNDVI_270") # selecting one of the time point in each harvest window. selecting based on higher variation time point from heatmap, not actually calculating all. 

# Loop over each yield trait
for (i in 1:length(yield_traits)) {
  
  trait <- yield_traits[i]
  GNDVI_col <- GNDVI_times[i]
  
  # Calculate variance of GNDVI and yield
  GNDVI_var <- var(nmsu22_nor_pheno[[GNDVI_col]], na.rm = TRUE)
  yld_var <- var(nmsu22_nor_pheno[[trait]], na.rm = TRUE)
  
  # Calculate scaling factor
  sf <- sqrt(GNDVI_var) / sqrt(yld_var)
  
  # Create new scaled yield column
  new_col <- paste0("sca_", trait)
  nmsu22_nor_pheno[[new_col]] <- nmsu22_nor_pheno[[trait]] * sf
  
  #  Print scaling factor
  cat("Scaling factor for", trait, "using", GNDVI_col, ":", sf, "\n")
}


# thus all scaled yields are added as a new column at last. 
```

one but not confirming now
```{r for checking outliers}
# Fit a linear model
model <- lm(yld1 ~ Rep, data = nmsu22_nor_pheno, na.action = na.exclude)

# Calculate residuals
nmsu22_nor_pheno$residuals <- residuals(model, na.action = na.exclude)
# Standardize the residuals
nmsu22_nor_pheno$residuals_standard <- scale(nmsu22_nor_pheno$residuals, center = T, scale = T) 
# Identify outliers
outliers <- nmsu22_nor_pheno %>% filter(abs(residuals_standard) > 3)
# Print outliers
View(outliers)
# one, not removing for now.

```


## 1.1. For NDVI 
- For Harvest window 1:  62, 76, 90, 104, 114 and yld1
- For Harvest window 2: "124", "130", "138", "144", "150" and yld2
- For Harvest window 3: "166", "174", "yld3"
- For Harvest window 4: "194", "200", "206", "yld4"
- For Harvest window 5: "222", "236", "240", "yld5"
- For Harvest window 6: "256", "264", "270",  "284", "292", "yld6"
= 29

### 1.1.1. Run models first
- Univariate without spatial
- Univariate with spatial
- Bivariate without spatial
- Bivariate with spatial

#### 1.1.1.1. Univariate for yield without spatial
h2 and SE
0.13	0.09
0.18	0.1
0.21	0.1
0.24	0.1
0.2	0.1
0.3	0.11
```{r No spatial }
## For yld1
st_nmsu22_nor_yld1 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld1)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld1, heritA ~ V1 / (V1 + V2)) 

## For yld2
st_nmsu22_nor_yld2 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld2)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld2, heritA ~ V1 / (V1 + V2)) 

## For yld3
st_nmsu22_nor_yld3 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld3)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld3, heritA ~ V1 / (V1 + V2)) 

## For yld4
st_nmsu22_nor_yld4 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld4)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld4, heritA ~ V1 / (V1 + V2)) 

## For yld5
st_nmsu22_nor_yld5 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld5)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld5, heritA ~ V1 / (V1 + V2)) 

## For yld6
st_nmsu22_nor_yld6 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld6 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld6)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld6, heritA ~ V1 / (V1 + V2)) 

```


#### 1.1.1.2. Univariate for yield with spatial

Yield	h2 (narrow)	SE
1	0.58	0.12
2	0.63	0.11
3	0.63	0.1
4	0.65	0.09
5	0.68	0.09
6	0.62	0.099
```{r From spatial BLUPS and Univariate models}
asreml.options(extra =200, maxit =200)
## For yld1 (however might have to scale cz the blups value are higher here)
st_nmsu22_nor_yld1 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld1)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld1, heritA ~ V1 / (V1 + V5)) 

## For yld2
st_nmsu22_nor_yld2 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld2)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld2, heritA ~ V1 / (V1 + V5)) 

## For yld3
st_nmsu22_nor_yld3 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld3)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld3, heritA ~ V1 / (V1 + V5)) 

## For yld4
st_nmsu22_nor_yld4 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld4)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld4, heritA ~ V1 / (V1 + V5)) 

## For yld5
st_nmsu22_nor_yld5 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld5)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld5, heritA ~ V1 / (V1 + V5)) 

## For yld6
st_nmsu22_nor_yld6 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld6 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_nor_pheno)}
summary(st_nmsu22_nor_yld6)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_nor_yld6, heritA ~ V1 / (V1 + V5)) 


```


#### 1.1.1.3. Bivariate for yield without spatial

for 1st window, none worked... genetic not converged.. 

```{r Bivariate models without Spatial based on NDVI}
colnames(nmsu22_nor_pheno)

## lets proceed with calculation for all harvest windows
# 1.  For harvest window 1 "62", "76", "90", "104", "114", "yld1"
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for NDVI_62:
mt_nmsu22_nor_ndvi_62 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_62) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_62)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
mt_nmsu22_nor_ndvi_62_gc <- asreml::vpredict(mt_nmsu22_nor_ndvi_62, gen.cor ~ (V2)/sqrt(V1*V3))
mt_nmsu22_nor_ndvi_62_gc # -0.99
mt_nmsu22_nor_ndvi_62_rc <- asreml::vpredict(mt_nmsu22_nor_ndvi_62, res.cor ~ (V6)/sqrt(V5*V7))
mt_nmsu22_nor_ndvi_62_rc # -0.19

 ## For NDVI_76
mt_nmsu22_nor_ndvi_76 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_76) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_76)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_ndvi_76, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.9950337
asreml::vpredict(mt_nmsu22_nor_ndvi_76, res.cor ~ (V6)/sqrt(V5*V7)) # - 0.29


 ## For NDVI_90
mt_nmsu22_nor_ndvi_90 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_90) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_90)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_ndvi_90, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.99
asreml::vpredict(mt_nmsu22_nor_ndvi_90, res.cor ~ (V6)/sqrt(V5*V7)) # -0.02

## For NDVI_104
mt_nmsu22_nor_ndvi_104 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_104) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_104)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_ndvi_104, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.76
asreml::vpredict(mt_nmsu22_nor_ndvi_104, res.cor ~ (V6)/sqrt(V5*V7)) # 0.75

## For NDVI_114
mt_nmsu22_nor_ndvi_114 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_114) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_114)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_114, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.999
asreml::vpredict(mt_nmsu22_nor_ndvi_114, res.cor ~ (V6)/sqrt(V5*V7)) # 0.93



# 2. For harvest window 2: "124", "130", "138", "144", "150", "yld2", 
## For NDVI_124
mt_nmsu22_nor_ndvi_124 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_124) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_124)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_124, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.93
asreml::vpredict(mt_nmsu22_nor_ndvi_124, res.cor ~ (V6)/sqrt(V5*V7)) # -0.50

## For NDVI_130
mt_nmsu22_nor_ndvi_130 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_130) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_130)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_130, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.97
asreml::vpredict(mt_nmsu22_nor_ndvi_130, res.cor ~ (V6)/sqrt(V5*V7)) # 0.74

## For NDVI_138
mt_nmsu22_nor_ndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_138)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.30
asreml::vpredict(mt_nmsu22_nor_ndvi_138, res.cor ~ (V6)/sqrt(V5*V7)) # 0.81

## For NDVI_144
mt_nmsu22_nor_ndvi_144 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_144) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_144)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_144, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.98
asreml::vpredict(mt_nmsu22_nor_ndvi_144, res.cor ~ (V6)/sqrt(V5*V7)) # 0.89

## For NDVI_150
mt_nmsu22_nor_ndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_150) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_150)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.97
asreml::vpredict(mt_nmsu22_nor_ndvi_150, res.cor ~ (V6)/sqrt(V5*V7)) # 0.91




# 3. For harvest window 3:  "166", "174", "yld3" 
## For NDVI_166
mt_nmsu22_nor_ndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_166) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_166)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.85
asreml::vpredict(mt_nmsu22_nor_ndvi_166, res.cor ~ (V6)/sqrt(V5*V7)) # 0.07

## For NDVI_174
mt_nmsu22_nor_ndvi_174 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_174) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_174)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_174, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.90
asreml::vpredict(mt_nmsu22_nor_ndvi_174, res.cor ~ (V6)/sqrt(V5*V7)) # 0.74

  
# 4. For harvest window 4: "194", "200", "206", "yld4",
## For NDVI_188: OUT

## For NDVI_194
mt_nmsu22_nor_ndvi_194 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, NDVI_194) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_194)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_194, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.99
asreml::vpredict(mt_nmsu22_nor_ndvi_194, res.cor ~ (V6)/sqrt(V5*V7)) # 0.64

## For NDVI_200
mt_nmsu22_nor_ndvi_200 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, NDVI_200) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_200)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_200, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.97
asreml::vpredict(mt_nmsu22_nor_ndvi_200, res.cor ~ (V6)/sqrt(V5*V7)) # 0.65

## For NDVI_206
mt_nmsu22_nor_ndvi_206 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, NDVI_206) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_206)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_206, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.96
asreml::vpredict(mt_nmsu22_nor_ndvi_206, res.cor ~ (V6)/sqrt(V5*V7)) # 0.86


# 5. For harvest window 5: "222",  "236", "240"

## For NDVI_222
mt_nmsu22_nor_ndvi_222 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_222) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_222)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_222, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.25
asreml::vpredict(mt_nmsu22_nor_ndvi_222, res.cor ~ (V6)/sqrt(V5*V7)) # 0.32

## For NDVI_236: didn't work
mt_nmsu22_nor_ndvi_236 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_236) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_236)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_236, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_236, res.cor ~ (V6)/sqrt(V5*V7)) # 

## For NDVI_240
mt_nmsu22_nor_ndvi_240 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_240) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_240)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_240, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.36
asreml::vpredict(mt_nmsu22_nor_ndvi_240, res.cor ~ (V6)/sqrt(V5*V7)) # 0.63

# 5. For harvest window 6: "256", "264", "270", "284", "292"

## For NDVI_256: didnt work
mt_nmsu22_nor_ndvi_256 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_256) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_256)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_256, gen.cor ~ (V2)/sqrt(V1*V3)) #
asreml::vpredict(mt_nmsu22_nor_ndvi_256, res.cor ~ (V6)/sqrt(V5*V7)) # 

## For NDVI_264
mt_nmsu22_nor_ndvi_264 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_264) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_264)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_264, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.94
asreml::vpredict(mt_nmsu22_nor_ndvi_264, res.cor ~ (V6)/sqrt(V5*V7)) # 0.70

## For NDVI_270
mt_nmsu22_nor_ndvi_270 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_270) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_270)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_270, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.97
asreml::vpredict(mt_nmsu22_nor_ndvi_270, res.cor ~ (V6)/sqrt(V5*V7)) # 0.77

## For NDVI_284
mt_nmsu22_nor_ndvi_284 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_284) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_284)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_284, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.93
asreml::vpredict(mt_nmsu22_nor_ndvi_284, res.cor ~ (V6)/sqrt(V5*V7)) # 0.35

## For NDVI_292
mt_nmsu22_nor_ndvi_292 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_292) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_292)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_292, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.97
asreml::vpredict(mt_nmsu22_nor_ndvi_292, res.cor ~ (V6)/sqrt(V5*V7)) # 0.50

```



#### 1.1.1.4. Bivariate for yield with spatial

run for with spatial and see if the correlations are still -vely high like above... if so might have to switch to random regression cz mt models not doing great for this smaller dataset (fewer observations)

```{r}
colnames(nmsu22_nor_pheno)

## lets proceed with calculation for all harvest windows
# 1.  For harvest window 1 "62", "76", "90", "104", "114", "yld1"
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for NDVI_62: bound
mt_nmsu22_nor_ndvi_62 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_62) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +diag(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_62)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu22_nor_ndvi_62, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_62, spa.cor ~ (V5)/sqrt(V4*V6)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_62, res.cor ~ (V11)/sqrt(V10*V12)) # 

 ## For NDVI_76 - didn't converge with full or diag residual or diag spatial...
mt_nmsu22_nor_ndvi_76 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_76) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + diag(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_76)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_ndvi_76, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_76, spa.cor ~ (V5)/sqrt(V4*V6)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_76, res.cor ~ (V11)/sqrt(V10*V12)) # 

 ## For NDVI_90: same didn't... 
mt_nmsu22_nor_ndvi_90 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_90) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_90)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_ndvi_90, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_90, spa.cor ~ (V5)/sqrt(V4*V6)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_90, res.cor ~ (V11)/sqrt(V10*V12)) # 

## For NDVI_104 - din't
mt_nmsu22_nor_ndvi_104 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_104) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_104)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_ndvi_104, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_104, spa.cor ~ (V5)/sqrt(V4*V6)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_104, res.cor ~ (V11)/sqrt(V10*V12)) # 

## For NDVI_114 - didn't
mt_nmsu22_nor_ndvi_114 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_114) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_114)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_114, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_114, spa.cor ~ (V5)/sqrt(V4*V6)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_114, res.cor ~ (V11)/sqrt(V10*V12)) # 

dsfd

# 2. For harvest window 2: "124", "130", "138", "144", "150", "yld2", 
## For NDVI_124
mt_nmsu22_nor_ndvi_124 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_124) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_124)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_124, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.29
asreml::vpredict(mt_nmsu22_nor_ndvi_124, spa.cor ~ (V5)/sqrt(V4*V6)) # -0.38
asreml::vpredict(mt_nmsu22_nor_ndvi_124, res.cor ~ (V11)/sqrt(V10*V12)) # 0.24
## For NDVI_130
mt_nmsu22_nor_ndvi_130 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_130) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_130)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_130, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.41
asreml::vpredict(mt_nmsu22_nor_ndvi_130, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.50
asreml::vpredict(mt_nmsu22_nor_ndvi_130, res.cor ~ (V11)/sqrt(V10*V12)) # 0.76

## For NDVI_138
mt_nmsu22_nor_ndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_138)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.88
asreml::vpredict(mt_nmsu22_nor_ndvi_138, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.24
asreml::vpredict(mt_nmsu22_nor_ndvi_138, res.cor ~ (V11)/sqrt(V10*V12)) # 0.63

## For NDVI_144
mt_nmsu22_nor_ndvi_144 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_144) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_144)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_144, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.88
asreml::vpredict(mt_nmsu22_nor_ndvi_144, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.64
asreml::vpredict(mt_nmsu22_nor_ndvi_144, res.cor ~ (V11)/sqrt(V10*V12)) # 0.86

## For NDVI_150
mt_nmsu22_nor_ndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_150) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_150)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.65
asreml::vpredict(mt_nmsu22_nor_ndvi_150, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.91
asreml::vpredict(mt_nmsu22_nor_ndvi_150, res.cor ~ (V11)/sqrt(V10*V12)) # 0.86




# 3. For harvest window 3:  "166", "174", "yld3" 
## For NDVI_166
mt_nmsu22_nor_ndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_166) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_166)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.85
asreml::vpredict(mt_nmsu22_nor_ndvi_166, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.15
asreml::vpredict(mt_nmsu22_nor_ndvi_166, res.cor ~ (V11)/sqrt(V10*V12)) # 0.21


## For NDVI_174
mt_nmsu22_nor_ndvi_174 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_174) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_174)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_174, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.90
asreml::vpredict(mt_nmsu22_nor_ndvi_174, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.38
asreml::vpredict(mt_nmsu22_nor_ndvi_174, res.cor ~ (V11)/sqrt(V10*V12)) # 0.65

  
# 4. For harvest window 4: "194", "200", "206", "yld4",

## For NDVI_194 - didn't work with full but with diag for residual it worked. 
mt_nmsu22_nor_ndvi_194 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, NDVI_194) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_194)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_194, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.90
asreml::vpredict(mt_nmsu22_nor_ndvi_194, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.49
asreml::vpredict(mt_nmsu22_nor_ndvi_194, res.cor ~ (V11)/sqrt(V10*V12)) # 

## For NDVI_200
mt_nmsu22_nor_ndvi_200 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, NDVI_200) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_200)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_200, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.90
asreml::vpredict(mt_nmsu22_nor_ndvi_200, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.47
asreml::vpredict(mt_nmsu22_nor_ndvi_200, res.cor ~ (V11)/sqrt(V10*V12)) # 0.55

## For NDVI_206
mt_nmsu22_nor_ndvi_206 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, NDVI_206) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_206)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_206, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.97
asreml::vpredict(mt_nmsu22_nor_ndvi_206, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.50
asreml::vpredict(mt_nmsu22_nor_ndvi_206, res.cor ~ (V11)/sqrt(V10*V12)) # 0.73


# 5. For harvest window 5: "222",  "236", "240"

## For NDVI_222
mt_nmsu22_nor_ndvi_222 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_222) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_222)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_222, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.73
asreml::vpredict(mt_nmsu22_nor_ndvi_222, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.36
asreml::vpredict(mt_nmsu22_nor_ndvi_222, res.cor ~ (V11)/sqrt(V10*V12)) # 0.15

## For NDVI_236: 
mt_nmsu22_nor_ndvi_236 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_236) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_236)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_236, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.93
asreml::vpredict(mt_nmsu22_nor_ndvi_236, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.38
asreml::vpredict(mt_nmsu22_nor_ndvi_236, res.cor ~ (V11)/sqrt(V10*V12)) # 0.47

## For NDVI_240
mt_nmsu22_nor_ndvi_240 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_240) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_240)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_240, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.76
asreml::vpredict(mt_nmsu22_nor_ndvi_240, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.48
asreml::vpredict(mt_nmsu22_nor_ndvi_240, res.cor ~ (V11)/sqrt(V10*V12)) # 0.33

# 5. For harvest window 6: "256", "264", "270", "284", "292"

## For NDVI_256: didn't work with diag for res or diag for spatial too.
mt_nmsu22_nor_ndvi_256 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_256) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_256)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_256, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_256, spa.cor ~ (V5)/sqrt(V4*V6)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_256, res.cor ~ (V11)/sqrt(V10*V12)) # 

## For NDVI_264
mt_nmsu22_nor_ndvi_264 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_264) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_264)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_264, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.90
asreml::vpredict(mt_nmsu22_nor_ndvi_264, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.11
asreml::vpredict(mt_nmsu22_nor_ndvi_264, res.cor ~ (V11)/sqrt(V10*V12)) # 0.42

## For NDVI_270
mt_nmsu22_nor_ndvi_270 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_270) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_270)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_270, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.95
asreml::vpredict(mt_nmsu22_nor_ndvi_270, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.70
asreml::vpredict(mt_nmsu22_nor_ndvi_270, res.cor ~ (V11)/sqrt(V10*V12)) # 0.58

## For NDVI_284: didn't work
mt_nmsu22_nor_ndvi_284 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_284) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_284)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_284, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_284, spa.cor ~ (V5)/sqrt(V4*V6)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_284, res.cor ~ (V11)/sqrt(V10*V12)) # 

## For NDVI_292: didn't work 
mt_nmsu22_nor_ndvi_292 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, NDVI_292) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_ndvi_292)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_ndvi_292, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_292, spa.cor ~ (V5)/sqrt(V4*V6)) # 
asreml::vpredict(mt_nmsu22_nor_ndvi_292, res.cor ~ (V11)/sqrt(V10*V12)) # 

```









## 1.2. For GNDVI 
- For Harvest window 1:  62, 76, 90, 96, 104,110, 114 and yld1
- For Harvest window 2:  "124", "130", "138", "144", "150" and yld2
- For Harvest window 3: "166", "174","180", "yld3"
- For Harvest window 4: "188", 194", "200", "206", "yld4"
- For Harvest window 5: "216", "222", "228", 236", "240", "yld5"
- For Harvest window 6: "264", "270", "278", "284", "292", "yld6"
= 35

### 1.2.1. Run models first
- Univariate without spatial
- Univariate with spatial
- Bivariate without spatial
- Bivariate with spatial

#### 1.2.1.1. Univariate for yield without spatial
SAME AS NDVI FOR YIELD (univariate)

#### 1.2.1.2. Univariate for yield with spatial
SAME AS NDVI FOR YIELD (univariate)


#### 1.2.1.3. Bivariate for yield without spatial



```{r Bivariate models without Spatial based on GNDVI}
colnames(nmsu22_nor_pheno)

## lets proceed with calculation for all harvest windows
# 1.  For harvest window 1 "62", "76", "90", "96", "104", "110", "114", "yld1"

asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for GNDVI_62:
mt_nmsu22_nor_gndvi_62 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_62) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_62)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu22_nor_gndvi_62, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.94
asreml::vpredict(mt_nmsu22_nor_gndvi_62, res.cor ~ (V6)/sqrt(V5*V7)) # 0.59

 ## For GNDVI_76
mt_nmsu22_nor_gndvi_76 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_76) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_76)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_76, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.95
asreml::vpredict(mt_nmsu22_nor_gndvi_76, res.cor ~ (V6)/sqrt(V5*V7)) # 0.84


 ## For GNDVI_90
mt_nmsu22_nor_gndvi_90 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_90) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_90)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_90, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.92
asreml::vpredict(mt_nmsu22_nor_gndvi_90, res.cor ~ (V6)/sqrt(V5*V7)) # 0.75

## For GNDVI_96
mt_nmsu22_nor_gndvi_96 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_96) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_96)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_96, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.99
asreml::vpredict(mt_nmsu22_nor_gndvi_96, res.cor ~ (V6)/sqrt(V5*V7)) # 0.95

## For GNDVI_104
mt_nmsu22_nor_gndvi_104 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_104) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_104)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_104, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.99
asreml::vpredict(mt_nmsu22_nor_gndvi_104, res.cor ~ (V6)/sqrt(V5*V7)) # 0.83

## For GNDVI_114
mt_nmsu22_nor_gndvi_114 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_114) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_114)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_114, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.99
asreml::vpredict(mt_nmsu22_nor_gndvi_114, res.cor ~ (V6)/sqrt(V5*V7)) # 0.82



# 2. For harvest window 2: "124", "130", "138", "144", "150", "yld2", 

## For GNDVI_124
mt_nmsu22_nor_gndvi_124 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_124) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_124)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_124, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.80
asreml::vpredict(mt_nmsu22_nor_gndvi_124, res.cor ~ (V6)/sqrt(V5*V7)) # 0.36

## For GNDVI_130
mt_nmsu22_nor_gndvi_130 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_130) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_130)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_130, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.24
asreml::vpredict(mt_nmsu22_nor_gndvi_130, res.cor ~ (V6)/sqrt(V5*V7)) # 0.48

## For GNDVI_138
mt_nmsu22_nor_gndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_138)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.26
asreml::vpredict(mt_nmsu22_nor_gndvi_138, res.cor ~ (V6)/sqrt(V5*V7)) # 0.50

## For GNDVI_144
mt_nmsu22_nor_gndvi_144 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_144) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_144)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_144, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.93
asreml::vpredict(mt_nmsu22_nor_gndvi_144, res.cor ~ (V6)/sqrt(V5*V7)) # 0.89

## For GNDVI_150
mt_nmsu22_nor_gndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_150) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_150)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.96
asreml::vpredict(mt_nmsu22_nor_gndvi_150, res.cor ~ (V6)/sqrt(V5*V7)) # 0.89




# 3. For harvest window 3:  "166", "174", "180", "yld3" 
## For GNDVI_166
mt_nmsu22_nor_gndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_166) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_166)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.03
asreml::vpredict(mt_nmsu22_nor_gndvi_166, res.cor ~ (V6)/sqrt(V5*V7)) # 0.51

## For GNDVI_174
mt_nmsu22_nor_gndvi_174 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_174) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_174)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_174, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.27
asreml::vpredict(mt_nmsu22_nor_gndvi_174, res.cor ~ (V6)/sqrt(V5*V7)) # 0.29

## For GNDVI_180
mt_nmsu22_nor_gndvi_180 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_180) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_180)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_180, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.68
asreml::vpredict(mt_nmsu22_nor_gndvi_180, res.cor ~ (V6)/sqrt(V5*V7)) # 0.82
  


# 4. For harvest window 4: "188", "194", "200", "206", "yld4",
## For GNDVI_188: 
mt_nmsu22_nor_gndvi_188 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_188) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_188)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_188, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.93
asreml::vpredict(mt_nmsu22_nor_gndvi_188, res.cor ~ (V6)/sqrt(V5*V7)) # -0.76

## For GNDVI_194
mt_nmsu22_nor_gndvi_194 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_194) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_194)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_194, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.74
asreml::vpredict(mt_nmsu22_nor_gndvi_194, res.cor ~ (V6)/sqrt(V5*V7)) # 0.78

## For GNDVI_200
mt_nmsu22_nor_gndvi_200 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_200) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_200)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_200, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.34
asreml::vpredict(mt_nmsu22_nor_gndvi_200, res.cor ~ (V6)/sqrt(V5*V7)) # -0.19

## For GNDVI_206
mt_nmsu22_nor_gndvi_206 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_206) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_206)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_206, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.28
asreml::vpredict(mt_nmsu22_nor_gndvi_206, res.cor ~ (V6)/sqrt(V5*V7)) # 0.79


# 5. For harvest window 5: "216", "222", "228", "236", "240",

## For GNDVI_216
mt_nmsu22_nor_gndvi_216 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_216) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_216)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_216, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.76
asreml::vpredict(mt_nmsu22_nor_gndvi_216, res.cor ~ (V6)/sqrt(V5*V7)) # -0.24

## For GNDVI_222
mt_nmsu22_nor_gndvi_222 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_222) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_222)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_222, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.85
asreml::vpredict(mt_nmsu22_nor_gndvi_222, res.cor ~ (V6)/sqrt(V5*V7)) # -0.39

## For GNDVI_228
mt_nmsu22_nor_gndvi_228 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_228) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_228)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_228, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.08
asreml::vpredict(mt_nmsu22_nor_gndvi_228, res.cor ~ (V6)/sqrt(V5*V7)) # 0.03

## For GNDVI_236: BOUND
mt_nmsu22_nor_gndvi_236 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_236) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_236)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_236, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_nor_gndvi_236, res.cor ~ (V6)/sqrt(V5*V7)) # 

## For GNDVI_240
mt_nmsu22_nor_gndvi_240 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_240) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_240)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_240, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.98
asreml::vpredict(mt_nmsu22_nor_gndvi_240, res.cor ~ (V6)/sqrt(V5*V7)) # 0.55

# 6. For harvest window 6: "264", "270", "278", "284", "292"

## For GNDVI_264
mt_nmsu22_nor_gndvi_264 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_264) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_264)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_264, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.07
asreml::vpredict(mt_nmsu22_nor_gndvi_264, res.cor ~ (V6)/sqrt(V5*V7)) # 0.43

## For GNDVI_270
mt_nmsu22_nor_gndvi_270 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_270) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_270)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_270, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.38
asreml::vpredict(mt_nmsu22_nor_gndvi_270, res.cor ~ (V6)/sqrt(V5*V7)) # 0.09

## For GNDVI_278
mt_nmsu22_nor_gndvi_278 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_278) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_278)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_278, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.83
asreml::vpredict(mt_nmsu22_nor_gndvi_278, res.cor ~ (V6)/sqrt(V5*V7)) # -0.035


## For GNDVI_284
mt_nmsu22_nor_gndvi_284 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_284) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_284)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_284, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.29
asreml::vpredict(mt_nmsu22_nor_gndvi_284, res.cor ~ (V6)/sqrt(V5*V7)) # 0.58

## For GNDVI_292
mt_nmsu22_nor_gndvi_292 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_292) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_292)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_292, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.56
asreml::vpredict(mt_nmsu22_nor_gndvi_292, res.cor ~ (V6)/sqrt(V5*V7)) # 0.29

```



#### 1.2.1.4. Bivariate for yield with spatial

```{r}
colnames(nmsu22_nor_pheno)

## lets proceed with calculation for all harvest windows
# 1.  For harvest window 1 "62", "76", "90", "96", "104", "110", "114", "yld1"
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for GNDVI_62:
mt_nmsu22_nor_gndvi_62 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_62) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_62)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu22_nor_gndvi_62, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.92
asreml::vpredict(mt_nmsu22_nor_gndvi_62, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.72
asreml::vpredict(mt_nmsu22_nor_gndvi_62, res.cor ~ (V11)/sqrt(V10*V12)) # 0.92

 ## For GNDVI_76 - 
mt_nmsu22_nor_gndvi_76 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_76) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_76)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_76, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.94 (with diag for residual)
asreml::vpredict(mt_nmsu22_nor_gndvi_76, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.89 (with diag for residual)
asreml::vpredict(mt_nmsu22_nor_gndvi_76, res.cor ~ (V11)/sqrt(V10*V12)) # diag

 ## For GNDVI_90: 
mt_nmsu22_nor_gndvi_90 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_90) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_90)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_90, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.91
asreml::vpredict(mt_nmsu22_nor_gndvi_90, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.90
asreml::vpredict(mt_nmsu22_nor_gndvi_90, res.cor ~ (V11)/sqrt(V10*V12)) # diag

## For GNDVI_96:  
mt_nmsu22_nor_gndvi_96 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_96) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_96)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_96, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.91
asreml::vpredict(mt_nmsu22_nor_gndvi_96, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.94
asreml::vpredict(mt_nmsu22_nor_gndvi_96, res.cor ~ (V11)/sqrt(V10*V12)) # 0.78

## For GNDVI_104
mt_nmsu22_nor_gndvi_104 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_104) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_104)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_104, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.88
asreml::vpredict(mt_nmsu22_nor_gndvi_104, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.87
asreml::vpredict(mt_nmsu22_nor_gndvi_104, res.cor ~ (V11)/sqrt(V10*V12)) # 0.82

## For GNDVI_114 -t
mt_nmsu22_nor_gndvi_114 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_114) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_114)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_114, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.89
asreml::vpredict(mt_nmsu22_nor_gndvi_114, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.86
asreml::vpredict(mt_nmsu22_nor_gndvi_114, res.cor ~ (V11)/sqrt(V10*V12)) # 0.57



# 2. For harvest window 2: "124", "130", "138", "144", "150", "yld2", 
## For GNDVI_124
mt_nmsu22_nor_gndvi_124 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_124) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_124)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_124, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.33
asreml::vpredict(mt_nmsu22_nor_gndvi_124, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.16
asreml::vpredict(mt_nmsu22_nor_gndvi_124, res.cor ~ (V11)/sqrt(V10*V12)) # -0.005
## For GNDVI_130
mt_nmsu22_nor_gndvi_130 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_130) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_130)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_130, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.82
asreml::vpredict(mt_nmsu22_nor_gndvi_130, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.52
asreml::vpredict(mt_nmsu22_nor_gndvi_130, res.cor ~ (V11)/sqrt(V10*V12)) # 0.82

## For GNDVI_138: didn't with all us, then didn't with diag for residual, didn't with diag for gen, worked with diag for spatial.
mt_nmsu22_nor_gndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + diag(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_138)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.85 (diag for spatial)
# spatial = diag
asreml::vpredict(mt_nmsu22_nor_gndvi_138, res.cor ~ (V10)/sqrt(V9*V11)) # 0.88

## For GNDVI_144
mt_nmsu22_nor_gndvi_144 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_144) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_144)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_144, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.89
asreml::vpredict(mt_nmsu22_nor_gndvi_144, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.80
asreml::vpredict(mt_nmsu22_nor_gndvi_144, res.cor ~ (V11)/sqrt(V10*V12)) # 0.76

## For GNDVI_150: for full, var st not positive definite
mt_nmsu22_nor_gndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_150) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_150)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.95
asreml::vpredict(mt_nmsu22_nor_gndvi_150, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.97
asreml::vpredict(mt_nmsu22_nor_gndvi_150, res.cor ~ (V11)/sqrt(V10*V12)) # diag




# 3. For harvest window 3:  "166", "174", "180", "yld3" 
## For GNDVI_166
mt_nmsu22_nor_gndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_166) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_166)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.82
asreml::vpredict(mt_nmsu22_nor_gndvi_166, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.20
asreml::vpredict(mt_nmsu22_nor_gndvi_166, res.cor ~ (V11)/sqrt(V10*V12)) # 0.84


## For GNDVI_174
mt_nmsu22_nor_gndvi_174 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_174) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_174)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_174, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.95
asreml::vpredict(mt_nmsu22_nor_gndvi_174, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.41
asreml::vpredict(mt_nmsu22_nor_gndvi_174, res.cor ~ (V11)/sqrt(V10*V12)) # 0.72

## For GNDVI_180:  
mt_nmsu22_nor_gndvi_180 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_180) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_180)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_180, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.92
asreml::vpredict(mt_nmsu22_nor_gndvi_180, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.81
asreml::vpredict(mt_nmsu22_nor_gndvi_180, res.cor ~ (V11)/sqrt(V10*V12)) # diag
  




# 4. For harvest window 4: "188", "194", "200", "206", "yld4",

## For GNDVI_188:  diag for gen didn't work but the model says no gen cor between 
mt_nmsu22_nor_gndvi_188 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_188) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_nor_pheno)}

summary(mt_nmsu22_nor_gndvi_188)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_188, gen.cor ~ (V2)/sqrt(V1*V3)) # ~ 0
asreml::vpredict(mt_nmsu22_nor_gndvi_188, spa.cor ~ (V5)/sqrt(V4*V6)) # -0.6
asreml::vpredict(mt_nmsu22_nor_gndvi_188, res.cor ~ (V11)/sqrt(V10*V12)) # diag

## For GNDVI_194 - 
mt_nmsu22_nor_gndvi_194 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_194) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_194)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_194, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.92
asreml::vpredict(mt_nmsu22_nor_gndvi_194, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.28
asreml::vpredict(mt_nmsu22_nor_gndvi_194, res.cor ~ (V11)/sqrt(V10*V12)) # 0.82

## For GNDVI_200 # didn't work
mt_nmsu22_nor_gndvi_200 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_200) ~ 1 + Rep, #rep effect as fixed
         random = ~ diag(trait):vm(Germplasm,  grm_inverse) + diag(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_200)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_200, gen.cor ~ (V2)/sqrt(V1*V3)) # NA
asreml::vpredict(mt_nmsu22_nor_gndvi_200, spa.cor ~ (V5)/sqrt(V4*V6)) # NA
asreml::vpredict(mt_nmsu22_nor_gndvi_200, res.cor ~ (V11)/sqrt(V10*V12)) # NA

## For GNDVI_206
mt_nmsu22_nor_gndvi_206 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_206) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_206)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_206, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.96
asreml::vpredict(mt_nmsu22_nor_gndvi_206, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.74
asreml::vpredict(mt_nmsu22_nor_gndvi_206, res.cor ~ (V11)/sqrt(V10*V12)) # 0.62


# 5. For harvest window 5: "216", "222", "228", "236", "240"
## For GNDVI_216:  
mt_nmsu22_nor_gndvi_216 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_216) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_216)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_216, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.46
asreml::vpredict(mt_nmsu22_nor_gndvi_216, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.08
asreml::vpredict(mt_nmsu22_nor_gndvi_216, res.cor ~ (V11)/sqrt(V10*V12)) # 0.06

## For GNDVI_222
mt_nmsu22_nor_gndvi_222 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_222) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_222)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_222, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.87
asreml::vpredict(mt_nmsu22_nor_gndvi_222, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.18
asreml::vpredict(mt_nmsu22_nor_gndvi_222, res.cor ~ (V11)/sqrt(V10*V12)) # 0.67

## For GNDVI_228:  
mt_nmsu22_nor_gndvi_228 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_228) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_228)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_228, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.77
asreml::vpredict(mt_nmsu22_nor_gndvi_228, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.30
asreml::vpredict(mt_nmsu22_nor_gndvi_228, res.cor ~ (V11)/sqrt(V10*V12)) # 0.74

## For GNDVI_236: 
mt_nmsu22_nor_gndvi_236 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_236) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_236)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_236, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.95
asreml::vpredict(mt_nmsu22_nor_gndvi_236, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.57
asreml::vpredict(mt_nmsu22_nor_gndvi_236, res.cor ~ (V11)/sqrt(V10*V12)) # 0.79

## For GNDVI_240
mt_nmsu22_nor_gndvi_240 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_240) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_240)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_240, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.90
asreml::vpredict(mt_nmsu22_nor_gndvi_240, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.69
asreml::vpredict(mt_nmsu22_nor_gndvi_240, res.cor ~ (V11)/sqrt(V10*V12)) # 0.70

# 5. For harvest window 6: "264", "270", "278", "284", "292"

## For GNDVI_264
mt_nmsu22_nor_gndvi_264 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_264) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_264)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_264, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.84
asreml::vpredict(mt_nmsu22_nor_gndvi_264, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.40
asreml::vpredict(mt_nmsu22_nor_gndvi_264, res.cor ~ (V11)/sqrt(V10*V12)) # 0.70

## For GNDVI_270
mt_nmsu22_nor_gndvi_270 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_270) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_270)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_270, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.87
asreml::vpredict(mt_nmsu22_nor_gndvi_270, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.25
asreml::vpredict(mt_nmsu22_nor_gndvi_270, res.cor ~ (V11)/sqrt(V10*V12)) # 0.78

## For GNDVI_278:  
mt_nmsu22_nor_gndvi_278 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_278) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_278)$varcomp # 
asreml::vpredict(mt_nmsu22_nor_gndvi_278, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.54
asreml::vpredict(mt_nmsu22_nor_gndvi_278, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.17
asreml::vpredict(mt_nmsu22_nor_gndvi_278, res.cor ~ (V11)/sqrt(V10*V12)) # 0.49

## For GNDVI_284: 
mt_nmsu22_nor_gndvi_284 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_284) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_284)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_284, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.45
asreml::vpredict(mt_nmsu22_nor_gndvi_284, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.54
asreml::vpredict(mt_nmsu22_nor_gndvi_284, res.cor ~ (V11)/sqrt(V10*V12)) # 0.50

## For GNDVI_292: 
mt_nmsu22_nor_gndvi_292 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_292) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_nor_pheno)}
summary(mt_nmsu22_nor_gndvi_292)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_nor_gndvi_292, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.13
asreml::vpredict(mt_nmsu22_nor_gndvi_292, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.50
asreml::vpredict(mt_nmsu22_nor_gndvi_292, res.cor ~ (V11)/sqrt(V10*V12)) # 0.31

```











################################# Early Termination ################################






2. *First treatment NMSU_Early_Termination 2022 *
with spatial in the model, random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),



*Early_Termination irrigation treatment*
Loading and preparing data files
```{r loading files 06272025}
#import merged phenotype file 
nmsu22_et_pheno <- read.csv("2022/MSI_data/Output_files/nmsu22_et_combined_filtered_phenotypes_MSI.csv", sep = ",") %>% 
  select(-X) # remove index column

# Before running models, make sure the variables are in correct form. Need to convert into factors and numeric if needed.
  # converting all factorial variables into factor
nmsu22_et_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu22_et_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
  ## make sure all variables are in correct form (factors or numeric)

#import genotype file
grm_aa <- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1)
grm_aa <- as.matrix(grm_aa)
mean(diag(grm_aa)) # 1
hist(diag(grm_aa))

grm_agh <- grm_aa+diag(.0005,length(grm_aa[,1]),length(grm_aa[,1]))
grm_inverse <- solve(grm_agh)

#now run the model for now without GRM but need to find GRM.

```


Normally if the range of numbers vary a lot between vi's and yield, then we can scale yield to variance level of vi's. but here since yield is in lb and upto 2 digit, should have that issue. If got any, we can always use scaled yield.
```{r scaling yield }
# we only have to scale 5 harvest values based on NDVI's value. Of course the variance between VI's differ but it should be okay to use Yield scaled onto NDVI for analysis with NDRE and GNDVI as well. 

# Define the mapping of yield traits to NDVI columns
yield_traits <- paste0("yld", 1:5)
ndvi_times <- c("NDVI_114", "NDVI_144", "NDVI_180", "NDVI_180", "NDVI_270") # selecting one of the time point in each harvest window. selecting based on higher variation time point from heatmap, not actually calculating all. # here since we filtered out NDVI 200 so using 180's scaling factor, it should be on same scale and similar.
 # yield 4 has smaller values cz after et period

# Loop over each yield trait
for (i in 1:length(yield_traits)) {
  
  trait <- yield_traits[i]
  ndvi_col <- ndvi_times[i]
  
  # Calculate variance of NDVI and yield
  ndvi_var <- var(nmsu22_et_pheno[[ndvi_col]], na.rm = TRUE)
  yld_var <- var(nmsu22_et_pheno[[trait]], na.rm = TRUE)
  
  # Calculate scaling factor
  sf <- sqrt(ndvi_var) / sqrt(yld_var)
  
  # Create new scaled yield column
  new_col <- paste0("sca_", trait)
  nmsu22_et_pheno[[new_col]] <- nmsu22_et_pheno[[trait]] * sf
  
  #  Print scaling factor
  cat("Scaling factor for", trait, "using", ndvi_col, ":", sf, "\n")
}

# thus all scaled yields are added as a new column at last. 
```


one on yield 5 but not removing as its close to 3. 
```{r for checking outliers}
# for all yields
library(dplyr)
library(purrr)


# Traits
yield_traits_et <- paste0("yld", 1:5)

# Function to fit model, calculate residuals, standardize, and extract outliers
process_outliers_et <- function(trait, data) {
  # Build formula dynamically
  formula <- as.formula(paste(trait, "~ Rep + Row + Range + Germplasm"))
  
  # Fit model
  model <- lm(formula, data = data, na.action = na.exclude)
  
  # Get residuals and standardized residuals
  data[[paste0(trait, "_resid")]] <- residuals(model, na.action = na.exclude)
  data[[paste0(trait, "_resid_std")]] <- scale(data[[paste0(trait, "_resid")]], center = TRUE, scale = TRUE)
  
  # Extract outliers
  outliers <- data %>% filter(abs(.data[[paste0(trait, "_resid_std")]]) > 3)
  
  return(outliers)
}

# Apply the function to all yield traits
outlier_list_et <- map(yield_traits_et, ~process_outliers(.x, nmsu22_et_combined))

# Name the list elements for clarity
names(outlier_list_et) <- yield_traits_et

# View outliers for each trait

# plot and standard residual value

View(outlier_list_et$yld1) # none
View(outlier_list_et$yld2) # none
View(outlier_list_et$yld3) # none
View(outlier_list_et$yld4) # none
View(outlier_list_et$yld5) # NM1703PAR germplasm, plot 316 with -3.18 (no need to remove)

```

## 2.1. For NDVI

### 2.1.1. Run models first
- Univariate without spatial
- Univariate with spatial
- Bivariate without spatial
- Bivariate with spatial

#### 2.1.1.1. Univariate for yield without spatial

```{r No spatial Univariate ET 2022 }
## For yld1
st_nmsu22_et_yld1 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld1)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld1, heritA ~ V1 / (V1 + V2)) 
# gen var:0.0004868955 , res. var: 0.0012569313, h2: 0.28 and SE: 0.10 

## For yld2
st_nmsu22_et_yld2 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld2)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld2, heritA ~ V1 / (V1 + V2)) 
# gen var:0.08621192 , res. var: 0.53078723, h2: 0.14 and SE: 0.09 

## For yld3
st_nmsu22_et_yld3 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld3)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld3, heritA ~ V1 / (V1 + V2)) 
# gen var:0.2386510 , res. var: 0.4470751, h2: 0.35 and SE: 0.11 

## For yld4
st_nmsu22_et_yld4 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld4)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld4, heritA ~ V1 / (V1 + V2)) 
# gen var:0.01470560 , res. var: 0.04819897, h2: 0.23 and SE: 0.10 

## For yld5
st_nmsu22_et_yld5 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld5)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld5, heritA ~ V1 / (V1 + V2)) 
# gen var:0.2819365 , res. var: 0.2623923, h2: 0.52 and SE: 0.10 

```


#### 2.1.1.2. Univariate for yield with spatial

```{r From spatial BLUPS and Univariate models ET 2022}
asreml.options(extra =200, maxit =200)
## For yld1 : full model bound with yield 1 and sca_yld1 too
st_nmsu22_et_yld1 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld1)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld1, heritA ~ V3 / (V3 + V4)) 
# gen var:0.23 , spa. var:  NA, res. var: 0.50, h2: 0.32 and SE: 0.11

## For yld2
st_nmsu22_et_yld2 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld2)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld2, heritA ~ V1 / (V1 + V5)) 
# gen var:0.12808683 , spa. var:  0.98321006, res. var: 0.08251119, h2: 0.61 and SE: 0.14

## For yld3
st_nmsu22_et_yld3 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld3)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld3, heritA ~ V1 / (V1 + V5)) 
# gen var:0.25969724 , spa. var:  0.67467433, res. var: 0.09610727, h2: 0.73 and SE: 0.10

## For yld4
st_nmsu22_et_yld4 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld4)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld4, heritA ~ V1 / (V1 + V5)) 
# gen var:0.02344976 , spa. var:  0.04140920, res. var: 0.02006766, h2: 0.54 and SE: 0.12

## For yld5
st_nmsu22_et_yld5 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu22_et_pheno)}
summary(st_nmsu22_et_yld5)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu22_et_yld5, heritA ~ V1 / (V1 + V5)) 
# gen var:0.29070261 , spa. var:  0.28910581, res. var: 0.05969409, h2: 0.83 and SE: 0.07


```

#### 2.1.1.3. Bivariate for yield without spatial

```{r Bivariate models without Spatial based on NDVI ET 2022}
colnames(nmsu22_et_pheno)

## lets proceed with calculation for all harvest windows
# 1.  For harvest window 1 "62", "76", "90", "104", "114", "yld1"
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for NDVI_62:
mt_nmsu22_et_ndvi_62 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_62) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_62)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu22_et_ndvi_62, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.9697431
asreml::vpredict(mt_nmsu22_et_ndvi_62, res.cor ~ (V6)/sqrt(V5*V7)) # 0.2287406

 ## For NDVI_76
mt_nmsu22_et_ndvi_76 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_76) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_76)$varcomp # 
asreml::vpredict(mt_nmsu22_et_ndvi_76, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.01898793
asreml::vpredict(mt_nmsu22_et_ndvi_76, res.cor ~ (V6)/sqrt(V5*V7)) # 0.1261295


 ## For NDVI_90: bound for genetic, only works with diag for genetic
mt_nmsu22_et_ndvi_90 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_90) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_90)$varcomp # 
asreml::vpredict(mt_nmsu22_et_ndvi_90, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_et_ndvi_90, res.cor ~ (V6)/sqrt(V5*V7)) # 

## For NDVI_104: bound for genetic, only works with diag for genetic
mt_nmsu22_et_ndvi_104 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_104) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_104)$varcomp # 
asreml::vpredict(mt_nmsu22_et_ndvi_104, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_et_ndvi_104, res.cor ~ (V6)/sqrt(V5*V7)) # 

## For NDVI_114: bound for genetic, only works with diag for genetic
mt_nmsu22_et_ndvi_114 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_114) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_114)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_114, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_et_ndvi_114, res.cor ~ (V6)/sqrt(V5*V7)) # 



# 2. For harvest window 2: "138", "144", "150", "yld2", 

## For NDVI_138
mt_nmsu22_et_ndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_138)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.934957
asreml::vpredict(mt_nmsu22_et_ndvi_138, res.cor ~ (V6)/sqrt(V5*V7)) # 0.4981678

## For NDVI_144
mt_nmsu22_et_ndvi_144 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_144) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_144)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_144, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.98
asreml::vpredict(mt_nmsu22_et_ndvi_144, res.cor ~ (V6)/sqrt(V5*V7)) # 0.70

## For NDVI_150 : 
mt_nmsu22_et_ndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_150) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):diag(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_150)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.05
asreml::vpredict(mt_nmsu22_et_ndvi_150, res.cor ~ (V6)/sqrt(V5*V7)) # diag




# 3. For harvest window 3:  "166", "174", "180", "yld3" 
## For NDVI_166
mt_nmsu22_et_ndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_166) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_166)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.9637036
asreml::vpredict(mt_nmsu22_et_ndvi_166, res.cor ~ (V6)/sqrt(V5*V7)) # -0.1909994

## For NDVI_174
mt_nmsu22_et_ndvi_174 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_174) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_174)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_174, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.9625614
asreml::vpredict(mt_nmsu22_et_ndvi_174, res.cor ~ (V6)/sqrt(V5*V7)) # 0.3331464

## For NDVI_180
mt_nmsu22_et_ndvi_180 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_180) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_180)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_180, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.9949798
asreml::vpredict(mt_nmsu22_et_ndvi_180, res.cor ~ (V6)/sqrt(V5*V7)) # 0.3614157
  


# 4. For harvest window 4: none


# 5. For harvest window 5: "256", "264", "270", "284", "292", yld 5

## For NDVI_256: 
mt_nmsu22_et_ndvi_256 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_256) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):diag(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_256)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_256, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.90
asreml::vpredict(mt_nmsu22_et_ndvi_256, res.cor ~ (V6)/sqrt(V5*V7)) # diag

## For NDVI_264
mt_nmsu22_et_ndvi_264 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_264) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_264)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_264, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.5286549
asreml::vpredict(mt_nmsu22_et_ndvi_264, res.cor ~ (V6)/sqrt(V5*V7)) # 0.38

## For NDVI_270
mt_nmsu22_et_ndvi_270 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_270) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_270)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_270, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.5574163
asreml::vpredict(mt_nmsu22_et_ndvi_270, res.cor ~ (V6)/sqrt(V5*V7)) # 0.7069418

## For NDVI_284
mt_nmsu22_et_ndvi_284 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_284) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_284)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_284, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.8958326
asreml::vpredict(mt_nmsu22_et_ndvi_284, res.cor ~ (V6)/sqrt(V5*V7)) # -0.1087692

## For NDVI_292
mt_nmsu22_et_ndvi_292 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_292) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_292)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_292, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.9309556
asreml::vpredict(mt_nmsu22_et_ndvi_292, res.cor ~ (V6)/sqrt(V5*V7)) # 0.18

```



#### 2.1.1.4. Bivariate for yield with spatial

```{r Bivariate with spatial NDVI ET 2022}
colnames(nmsu22_et_pheno)

## lets proceed with calculation for all harvest windows
# 1.  For harvest window 1 "62", "76", "90", "104", "114", "yld1"
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for NDVI_62:
mt_nmsu22_et_ndvi_62 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_62) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_62)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu22_et_ndvi_62, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.8341137
asreml::vpredict(mt_nmsu22_et_ndvi_62, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.2632507
asreml::vpredict(mt_nmsu22_et_ndvi_62, res.cor ~ (V11)/sqrt(V10*V12)) # -0.407691


 ## For NDVI_76 - 
mt_nmsu22_et_ndvi_76 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_76) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_76)$varcomp # 
asreml::vpredict(mt_nmsu22_et_ndvi_76, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.1209993
asreml::vpredict(mt_nmsu22_et_ndvi_76, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.313535
asreml::vpredict(mt_nmsu22_et_ndvi_76, res.cor ~ (V11)/sqrt(V10*V12)) # -0.3098076

 ## For NDVI_90: 
mt_nmsu22_et_ndvi_90 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_90) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_90)$varcomp # 
asreml::vpredict(mt_nmsu22_et_ndvi_90, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.5176705
asreml::vpredict(mt_nmsu22_et_ndvi_90, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.0305906
asreml::vpredict(mt_nmsu22_et_ndvi_90, res.cor ~ (V11)/sqrt(V10*V12)) # - 0.2763303

## For NDVI_104 
mt_nmsu22_et_ndvi_104 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_104) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_104)$varcomp # 
asreml::vpredict(mt_nmsu22_et_ndvi_104, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6468464
asreml::vpredict(mt_nmsu22_et_ndvi_104, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.08077386
asreml::vpredict(mt_nmsu22_et_ndvi_104, res.cor ~ (V11)/sqrt(V10*V12)) # 0.2337132

## For NDVI_114 
mt_nmsu22_et_ndvi_114 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, NDVI_114) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_114)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_114, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.83068
asreml::vpredict(mt_nmsu22_et_ndvi_114, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6600261
asreml::vpredict(mt_nmsu22_et_ndvi_114, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7001883


# 2. For harvest window 2: "138", "144", "150", "yld2", 
## For NDVI_138
mt_nmsu22_et_ndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_138)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9863851
asreml::vpredict(mt_nmsu22_et_ndvi_138, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5771157
asreml::vpredict(mt_nmsu22_et_ndvi_138, res.cor ~ (V11)/sqrt(V10*V12)) # 0.5961651

## For NDVI_144
mt_nmsu22_et_ndvi_144 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_144) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + diag(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_144)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_144, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.88
asreml::vpredict(mt_nmsu22_et_ndvi_144, spa.cor ~ (V5)/sqrt(V4*V6)) # diag
asreml::vpredict(mt_nmsu22_et_ndvi_144, res.cor ~ (V11)/sqrt(V10*V12)) # diag

## For NDVI_150
mt_nmsu22_et_ndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, NDVI_150) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_150)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.78
asreml::vpredict(mt_nmsu22_et_ndvi_150, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.98
asreml::vpredict(mt_nmsu22_et_ndvi_150, res.cor ~ (V11)/sqrt(V10*V12)) # diag



# 3. For harvest window 3:  "166", "174","180", "yld3" 
## For NDVI_166
mt_nmsu22_et_ndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_166) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_166)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.89
asreml::vpredict(mt_nmsu22_et_ndvi_166, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.08
asreml::vpredict(mt_nmsu22_et_ndvi_166, res.cor ~ (V11)/sqrt(V10*V12)) # diag


## For NDVI_174
mt_nmsu22_et_ndvi_174 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_174) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_174)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_174, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.9742073
asreml::vpredict(mt_nmsu22_et_ndvi_174, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4727956
asreml::vpredict(mt_nmsu22_et_ndvi_174, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7338408

## For NDVI_180
mt_nmsu22_et_ndvi_180 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, NDVI_180) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + diag(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_180)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_180, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.34
asreml::vpredict(mt_nmsu22_et_ndvi_180, spa.cor ~ (V5)/sqrt(V4*V6)) # diag
asreml::vpredict(mt_nmsu22_et_ndvi_180, res.cor ~ (V11)/sqrt(V10*V12)) # 0.56
5.051309e-05/(sqrt(8.596032e-05 * 9.484102e-05))
  
# 4. For harvest window 4: none, out bcz of ET period

# 5. For harvest window 5:  "256", "264", "270", "284", "292"

## For NDVI_256: 
mt_nmsu22_et_ndvi_256 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_256) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_256)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_256, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.2015048
asreml::vpredict(mt_nmsu22_et_ndvi_256, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4760932
asreml::vpredict(mt_nmsu22_et_ndvi_256, res.cor ~ (V11)/sqrt(V10*V12)) # -0.714309

## For NDVI_264
mt_nmsu22_et_ndvi_264 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_264) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_264)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_264, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.94804
asreml::vpredict(mt_nmsu22_et_ndvi_264, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4502897
asreml::vpredict(mt_nmsu22_et_ndvi_264, res.cor ~ (V11)/sqrt(V10*V12)) # 0.2068542

## For NDVI_270
mt_nmsu22_et_ndvi_270 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_270) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_270)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_270, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.5757416
asreml::vpredict(mt_nmsu22_et_ndvi_270, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8299576
asreml::vpredict(mt_nmsu22_et_ndvi_270, res.cor ~ (V11)/sqrt(V10*V12)) # 0.4554774

## For NDVI_284: 
mt_nmsu22_et_ndvi_284 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_284) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_284)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_284, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6118516
asreml::vpredict(mt_nmsu22_et_ndvi_284, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.3176299
asreml::vpredict(mt_nmsu22_et_ndvi_284, res.cor ~ (V11)/sqrt(V10*V12)) # 0.4687594

## For NDVI_292: 
mt_nmsu22_et_ndvi_292 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, NDVI_292) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_ndvi_292)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_ndvi_292, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.3655154
asreml::vpredict(mt_nmsu22_et_ndvi_292, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.100122
asreml::vpredict(mt_nmsu22_et_ndvi_292, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6140479

```







## 2.2. For GNDVI 
#### 2.2.1.1. Univariate for yield without spatial
#### 2.2.1.2. Univariate for yield with spatial
  ### univariate are same so not needed here, same as NDVI section above as it's for yield.


- For Harvest window 1:  62, 76, 90, 96, 104,110, 114 and yld1
- For Harvest window 2: "130", "138", "144", "150" and yld2
- For Harvest window 3: "166", "174","180", "yld3"
- For Harvest window 4: "yld4"
- For Harvest window 5: "256", "264", "270", "278", "284", "292", "yld5"
 = 25 total traits 

sdfjhksdhfdskf copy from ndvi and paste as et, then double check the time points.,, 



#### 2.2.1.3. Bivariate for yield without spatial


```{r Bivariate models without Spatial based on GNDVI ET 2022}
colnames(nmsu22_et_pheno)

## lets proceed with calculation for all harvest windows
# 1.  For harvest window 1 "62", "76", "90", "96", "104", "110", "114", "yld1"
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for GNDVI_62:
mt_nmsu22_et_gndvi_62 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_62) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_62)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu22_et_gndvi_62, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.234622
asreml::vpredict(mt_nmsu22_et_gndvi_62, res.cor ~ (V6)/sqrt(V5*V7)) # 0.3662845

 ## For GNDVI_76
mt_nmsu22_et_gndvi_76 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_76) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_76)$varcomp # 
asreml::vpredict(mt_nmsu22_et_gndvi_76, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8963954
asreml::vpredict(mt_nmsu22_et_gndvi_76, res.cor ~ (V6)/sqrt(V5*V7)) # 0.3874995


 ## For GNDVI_90: 
mt_nmsu22_et_gndvi_90 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_90) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_90)$varcomp # 
asreml::vpredict(mt_nmsu22_et_gndvi_90, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.9605303
asreml::vpredict(mt_nmsu22_et_gndvi_90, res.cor ~ (V6)/sqrt(V5*V7)) # - 0.09

## For GNDVI_96: 
mt_nmsu22_et_gndvi_96 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_96) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_96)$varcomp # 
asreml::vpredict(mt_nmsu22_et_gndvi_96, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.7340523
asreml::vpredict(mt_nmsu22_et_gndvi_96, res.cor ~ (V6)/sqrt(V5*V7)) #  0.07746338

## For GNDVI_104: bound
mt_nmsu22_et_gndvi_104 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_104) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_104)$varcomp # 
asreml::vpredict(mt_nmsu22_et_gndvi_104, gen.cor ~ (V2)/sqrt(V1*V3)) # 
asreml::vpredict(mt_nmsu22_et_gndvi_104, res.cor ~ (V6)/sqrt(V5*V7)) # 



## For GNDVI_114: 
mt_nmsu22_et_gndvi_114 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_114) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_114)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_114, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.01853662
asreml::vpredict(mt_nmsu22_et_gndvi_114, res.cor ~ (V6)/sqrt(V5*V7)) # 0.5049902



# 2. For harvest window 2: "130", "138", "144", "150", "yld2", 

## For GNDVI_130
mt_nmsu22_et_gndvi_130 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_130) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_130)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_130, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6943274
asreml::vpredict(mt_nmsu22_et_gndvi_130, res.cor ~ (V6)/sqrt(V5*V7)) # 0.487057

## For GNDVI_138
mt_nmsu22_et_gndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_138)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.5056137
asreml::vpredict(mt_nmsu22_et_gndvi_138, res.cor ~ (V6)/sqrt(V5*V7)) # 0.3810523

## For GNDVI_144
mt_nmsu22_et_gndvi_144 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_144) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_144)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_144, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.834969
asreml::vpredict(mt_nmsu22_et_gndvi_144, res.cor ~ (V6)/sqrt(V5*V7)) # 0.6881815

## For GNDVI_150 : 
mt_nmsu22_et_gndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_150) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_150)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9756707
asreml::vpredict(mt_nmsu22_et_gndvi_150, res.cor ~ (V6)/sqrt(V5*V7)) # 0.8491808




# 3. For harvest window 3:  "166", "174", "180", "yld3" 
## For GNDVI_166
mt_nmsu22_et_gndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_166) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_166)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8046159
asreml::vpredict(mt_nmsu22_et_gndvi_166, res.cor ~ (V6)/sqrt(V5*V7)) # 0.6444093

## For GNDVI_174
mt_nmsu22_et_gndvi_174 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_174) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_174)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_174, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.1167405
asreml::vpredict(mt_nmsu22_et_gndvi_174, res.cor ~ (V6)/sqrt(V5*V7)) # 0.231189

## For GNDVI_180
mt_nmsu22_et_gndvi_180 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_180) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_180)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_180, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.8683234
asreml::vpredict(mt_nmsu22_et_gndvi_180, res.cor ~ (V6)/sqrt(V5*V7)) # 0.099586
  


# 4. For harvest window 4: none


# 5. For harvest window 5: "256", "264", "270", "278", "284", "292", yld 5

## For GNDVI_256: 
mt_nmsu22_et_gndvi_256 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_256) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_256)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_256, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.02620773
asreml::vpredict(mt_nmsu22_et_gndvi_256, res.cor ~ (V6)/sqrt(V5*V7)) # 0.3242831

## For GNDVI_264
mt_nmsu22_et_gndvi_264 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_264) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_264)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_264, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7618024
asreml::vpredict(mt_nmsu22_et_gndvi_264, res.cor ~ (V6)/sqrt(V5*V7)) # 0.6343272

## For GNDVI_270
mt_nmsu22_et_gndvi_270 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_270) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_270)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_270, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9991839
asreml::vpredict(mt_nmsu22_et_gndvi_270, res.cor ~ (V6)/sqrt(V5*V7)) # 0.7530617

## For GNDVI_278
mt_nmsu22_et_gndvi_278 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_278) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_278)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_278, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.4892682
asreml::vpredict(mt_nmsu22_et_gndvi_278, res.cor ~ (V6)/sqrt(V5*V7)) # 0.1597266

## For GNDVI_284
mt_nmsu22_et_gndvi_284 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_284) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_284)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_284, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.002764574
asreml::vpredict(mt_nmsu22_et_gndvi_284, res.cor ~ (V6)/sqrt(V5*V7)) # 0.4285541

## For GNDVI_292
mt_nmsu22_et_gndvi_292 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_292) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_292)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_292, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.1029625
asreml::vpredict(mt_nmsu22_et_gndvi_292, res.cor ~ (V6)/sqrt(V5*V7)) # 0.07906954

```



#### 2.2.1.4. Bivariate for yield with spatial


```{r Bivariate with spatial GNDVI ET 2022}
colnames(nmsu22_et_pheno)

## lets proceed with calculation for all harvest windows
# 1.  For harvest window 1 "62", "76", "90", "96", "104", "110", "114", "yld1"
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events

## for GNDVI_62:
mt_nmsu22_et_gndvi_62 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_62) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_62)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu22_et_gndvi_62, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9380776
asreml::vpredict(mt_nmsu22_et_gndvi_62, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8390832
asreml::vpredict(mt_nmsu22_et_gndvi_62, res.cor ~ (V11)/sqrt(V10*V12)) # diag


 ## For GNDVI_76 - 
mt_nmsu22_et_gndvi_76 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_76) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_76)$varcomp # 
asreml::vpredict(mt_nmsu22_et_gndvi_76, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9976949
asreml::vpredict(mt_nmsu22_et_gndvi_76, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4974418
asreml::vpredict(mt_nmsu22_et_gndvi_76, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8692595

 ## For GNDVI_90: 
mt_nmsu22_et_gndvi_90 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_90) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_90)$varcomp # 
asreml::vpredict(mt_nmsu22_et_gndvi_90, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8629566
asreml::vpredict(mt_nmsu22_et_gndvi_90, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6602697
asreml::vpredict(mt_nmsu22_et_gndvi_90, res.cor ~ (V11)/sqrt(V10*V12)) # diag

## For GNDVI_96: 
mt_nmsu22_et_gndvi_96 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_96) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_96)$varcomp # 
asreml::vpredict(mt_nmsu22_et_gndvi_96, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9190446
asreml::vpredict(mt_nmsu22_et_gndvi_96, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.2292429
asreml::vpredict(mt_nmsu22_et_gndvi_96, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7449104

## For GNDVI_104 
mt_nmsu22_et_gndvi_104 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_104) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_104)$varcomp # 
asreml::vpredict(mt_nmsu22_et_gndvi_104, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7354527
asreml::vpredict(mt_nmsu22_et_gndvi_104, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.134075
asreml::vpredict(mt_nmsu22_et_gndvi_104, res.cor ~ (V11)/sqrt(V10*V12)) # 0.798134


## For GNDVI_114 
mt_nmsu22_et_gndvi_114 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_114) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_114)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_114, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.5130809
asreml::vpredict(mt_nmsu22_et_gndvi_114, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.3738768
asreml::vpredict(mt_nmsu22_et_gndvi_114, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7252035


# 2. For harvest window 2: "130", "138", "144", "150", "yld2", 
## For GNDVI_130
mt_nmsu22_et_gndvi_130 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_130) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_130)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_130, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7317891
asreml::vpredict(mt_nmsu22_et_gndvi_130, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.409049
asreml::vpredict(mt_nmsu22_et_gndvi_130, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7662119


## For GNDVI_138
mt_nmsu22_et_gndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_138)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7129305
asreml::vpredict(mt_nmsu22_et_gndvi_138, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.3684452
asreml::vpredict(mt_nmsu22_et_gndvi_138, res.cor ~ (V11)/sqrt(V10*V12)) # 0.9706936

## For GNDVI_144
mt_nmsu22_et_gndvi_144 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_144) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_144)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_144, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.806035
asreml::vpredict(mt_nmsu22_et_gndvi_144, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7161231
asreml::vpredict(mt_nmsu22_et_gndvi_144, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6335536

## For GNDVI_150
mt_nmsu22_et_gndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_150) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_150)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8288367
asreml::vpredict(mt_nmsu22_et_gndvi_150, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.9341437
asreml::vpredict(mt_nmsu22_et_gndvi_150, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8500274



# 3. For harvest window 3:  "166", "174","180", "yld3" 
## For GNDVI_166
mt_nmsu22_et_gndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_166) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_166)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8965227
asreml::vpredict(mt_nmsu22_et_gndvi_166, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6120084
asreml::vpredict(mt_nmsu22_et_gndvi_166, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6203659


## For GNDVI_174
mt_nmsu22_et_gndvi_174 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_174) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_174)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_174, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9088766
asreml::vpredict(mt_nmsu22_et_gndvi_174, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6975339
asreml::vpredict(mt_nmsu22_et_gndvi_174, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7713746

## For GNDVI_180
mt_nmsu22_et_gndvi_180 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_180) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_180)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_180, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8424085
asreml::vpredict(mt_nmsu22_et_gndvi_180, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4517871
asreml::vpredict(mt_nmsu22_et_gndvi_180, res.cor ~ (V11)/sqrt(V10*V12)) # diag

# 4. For harvest window 4: none, out bcz of ET period

# 5. For harvest window 5:  "256", "264", "270", "278", "284", "292"

## For GNDVI_256: 
mt_nmsu22_et_gndvi_256 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_256) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_256)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_256, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9125113
asreml::vpredict(mt_nmsu22_et_gndvi_256, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.616979
asreml::vpredict(mt_nmsu22_et_gndvi_256, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6435723

## For GNDVI_264
mt_nmsu22_et_gndvi_264 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_264) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_264)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_264, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8853239
asreml::vpredict(mt_nmsu22_et_gndvi_264, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7526155
asreml::vpredict(mt_nmsu22_et_gndvi_264, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6545729

## For GNDVI_270
mt_nmsu22_et_gndvi_270 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_270) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_270)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_270, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9537701
asreml::vpredict(mt_nmsu22_et_gndvi_270, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8215292
asreml::vpredict(mt_nmsu22_et_gndvi_270, res.cor ~ (V11)/sqrt(V10*V12)) # 0.5296318

## For GNDVI_278
mt_nmsu22_et_gndvi_278 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_278) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_278)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_278, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7990656
asreml::vpredict(mt_nmsu22_et_gndvi_278, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4593878
asreml::vpredict(mt_nmsu22_et_gndvi_278, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7383811

## For GNDVI_284: 
mt_nmsu22_et_gndvi_284 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_284) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_284)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_284, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.5044502
asreml::vpredict(mt_nmsu22_et_gndvi_284, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5552077
asreml::vpredict(mt_nmsu22_et_gndvi_284, res.cor ~ (V11)/sqrt(V10*V12)) # 0.5655953

## For GNDVI_292: 
mt_nmsu22_et_gndvi_292 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_292) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu22_et_pheno)}
summary(mt_nmsu22_et_gndvi_292)$varcomp # to check if the model converged well
asreml::vpredict(mt_nmsu22_et_gndvi_292, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.2747943
asreml::vpredict(mt_nmsu22_et_gndvi_292, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.237456
asreml::vpredict(mt_nmsu22_et_gndvi_292, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6360765

```



 
 