---
title: "Calculating pheno correlations (raw data) 2022 NY"
author: "Dinesh Ghimire"
date: "2026-05-06"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr", "tidyr", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "GGally")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/dg663/FFAR-Alfalfa-NY-DG")
```


#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NY-DG"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NY-DG
#below is what I gsny when I call gsnywd. Just check to make sure that the root for this script is the base of the github project, somsnyimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa-NY-DG"
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If somsnyhing is misssing please lsny me know. 
key<-read.csv("2022/meta_data/cultivar_list_key.csv") # contains both CU and NY cultivars here. 
# its same for 2022 and 2023 (cultivar list)
```




#NY 22


################################# Musgrave Trial ################################



1. *NY Musgrave 2022 *

Loading and preparing data files
```{r loading files 05062026}
#import merged phenotype file 
ny_mus_pheno_22 <- read.csv("2022/MSI_data/output_files/Final_NY22_mus_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_mus_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_mus_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_mus_pheno_22) 

```


# Goal here: Place each harvest (yield) at the end of its respective window and visually separate windows in heatmaps.

```{r Pheno HeatMaps Correlation VI's and Yield Mus 2022 }
## ============================================================
## PHENO HEATMAPS: VI × YIELD (WINDOW-ALIGNED)
## NY Mus 2022
## ============================================================

## ------------------------------------------------------------
## Step 1: Identify columns
## ------------------------------------------------------------

ndvi_cols  <- grep("^NDVI_",  colnames(ny_mus_pheno_22), value = TRUE)
gndvi_cols <- grep("^GNDVI_", colnames(ny_mus_pheno_22), value = TRUE)
ndre_cols  <- grep("^NDRE_",  colnames(ny_mus_pheno_22), value = TRUE)

yield_cols <- c("yld1", "yld2", "yld3")


## ------------------------------------------------------------
## Step 2: Extract DOYs and define windows (data-driven)
## ------------------------------------------------------------

ndvi_doy <- as.numeric(sub("NDVI_", "", ndvi_cols))
ndvi_doy <- sort(ndvi_doy)

# Define windows based on your harvest structure !!!!!
window1 <- ndvi_doy[ndvi_doy <= 154]
window2 <- ndvi_doy[ndvi_doy > 154 & ndvi_doy <= 203]
window3 <- ndvi_doy[ndvi_doy > 203 & ndvi_doy <= 263]


## ------------------------------------------------------------
## Step 3: Build ordered column vectors
## ------------------------------------------------------------

full_ndvi_order <- c(
  paste0("NDVI_", window1), "yld1",
  paste0("NDVI_", window2), "yld2",
  paste0("NDVI_", window3), "yld3"
)

full_gndvi_order <- gsub("NDVI_", "GNDVI_", full_ndvi_order)
full_ndre_order  <- gsub("NDVI_", "NDRE_",  full_ndvi_order)
full_ndre_order

## ------------------------------------------------------------
## Step 4: Correlation matrices
## ------------------------------------------------------------

get_cor_matrix <- function(data, vi_cols, yld_cols) {
  cor(data[, c(vi_cols, yld_cols)], use = "pairwise.complete.obs")
}

ndvi_cor  <- get_cor_matrix(ny_mus_pheno_22, ndvi_cols, yield_cols)
gndvi_cor <- get_cor_matrix(ny_mus_pheno_22, gndvi_cols, yield_cols)
ndre_cor  <- get_cor_matrix(ny_mus_pheno_22, ndre_cols, yield_cols)


## ------------------------------------------------------------
## Step 5: Define window boundaries (for boxes)
## ------------------------------------------------------------

window_bounds <- list(
  window1 = c(1, length(window1)),
  window2 = c(length(window1)+2, length(window1)+1+length(window2)),
  window3 = c(length(window1)+length(window2)+3,
              length(window1)+length(window2)+2+length(window3))
)


## ------------------------------------------------------------
## Step 6: Plot function
## ------------------------------------------------------------

plot_heatmap_with_boxes <- function(cor_matrix, order_vec, title_text) {
  
  # keep only valid columns (prevents crash)
  valid_order <- intersect(order_vec, rownames(cor_matrix))
  
  cor_matrix_ordered <- cor_matrix[valid_order, valid_order]
  
  cor_long <- reshape2::melt(cor_matrix_ordered)
  
  p <- ggplot(cor_long, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile() +
    
    scale_fill_gradient2(
      low = "blue", mid = "white", high = "red",
      midpoint = 0, limits = c(-1, 1)
    ) +
    
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(hjust = 0.5, face = "bold")
    ) +
    
    labs(title = title_text, x = "Traits", y = "Traits")
  
  # add window boxes
  for (bounds in window_bounds) {
    p <- p + geom_rect(
      xmin = bounds[1] - 0.5,
      xmax = bounds[2] + 0.5,
      ymin = bounds[1] - 0.5,
      ymax = bounds[2] + 0.5,
      color = "black",
      fill = NA,
      linewidth = 0.8
    )
  }
  
  return(p)
}


## ------------------------------------------------------------
## Step 7: Generate heatmaps
## ------------------------------------------------------------

ndvi_heatmap_mus_22  <- plot_heatmap_with_boxes(
  ndvi_cor, full_ndvi_order,
  "NDVI vs Yield Correlation (NY Mus 2022)")
ndvi_heatmap_mus_22
gndvi_heatmap_mus_22 <- plot_heatmap_with_boxes(
  gndvi_cor, full_gndvi_order,
  "GNDVI vs Yield Correlation (NY Mus 2022)")
gndvi_heatmap_mus_22
ndre_heatmap_mus_22  <- plot_heatmap_with_boxes(
  ndre_cor, full_ndre_order,
  "NDRE vs Yield Correlation (NY Mus 2022)")


## ------------------------------------------------------------
## Step 8: Save
## ------------------------------------------------------------

ggsave("NDVI_Pheno_cor_heatmap_NY_Mus_2022.png",
       ndvi_heatmap_mus_22,
       path = "2022/output_images",
       width = 10, height = 8, dpi = 400)

ggsave("GNDVI_Pheno_cor_heatmap_NY_Mus_2022.png",
       gndvi_heatmap_mus_22,
       path = "2022/output_images",
       width = 10, height = 8, dpi = 400)

ggsave("NDRE_Pheno_cor_heatmap_NY_Mus_2022.png",
       ndre_heatmap_mus_22,
       path = "2022/output_images",
       width = 10, height = 8, dpi = 400)

```













################################# Snyder Trial ################################



Loading and preparing data files
```{r loading files 05062026}
#import merged phenotype file 
ny_sny_pheno_22 <- read.csv("2022/MSI_data/output_files/Final_NY22_sny_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_sny_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_sny_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_sny_pheno_22) 

```


# Goal here: Place each harvest (yield) at the end of its respective window and visually separate windows in heatmaps.

```{r Pheno HeatMaps Correlation VI's and Yield Sny 2022 }
## ============================================================
## PHENO HEATMAPS: VI × YIELD (WINDOW-ALIGNED)
## NY Sny 2022
## ============================================================

## ------------------------------------------------------------
## Step 1: Identify columns
## ------------------------------------------------------------

ndvi_cols  <- grep("^NDVI_",  colnames(ny_sny_pheno_22), value = TRUE)
gndvi_cols <- grep("^GNDVI_", colnames(ny_sny_pheno_22), value = TRUE)
ndre_cols  <- grep("^NDRE_",  colnames(ny_sny_pheno_22), value = TRUE)

yield_cols <- c("yld1", "yld2", "yld3")


## ------------------------------------------------------------
## Step 2: Extract DOYs and define windows (data-driven)
## ------------------------------------------------------------

ndvi_doy <- as.numeric(sub("NDVI_", "", ndvi_cols))
ndvi_doy <- sort(ndvi_doy)

# Define windows based on your harvest structure !!!
window1 <- ndvi_doy[ndvi_doy <= 165]
window2 <- ndvi_doy[ndvi_doy > 165 & ndvi_doy <= 208]
window3 <- ndvi_doy[ndvi_doy > 208 & ndvi_doy <= 259 ]


## ------------------------------------------------------------
## Step 3: Build ordered column vectors
## ------------------------------------------------------------

full_ndvi_order <- c(
  paste0("NDVI_", window1), "yld1",
  paste0("NDVI_", window2), "yld2",
  paste0("NDVI_", window3), "yld3"
)

full_gndvi_order <- gsub("NDVI_", "GNDVI_", full_ndvi_order)
full_ndre_order  <- gsub("NDVI_", "NDRE_",  full_ndvi_order)
full_ndre_order

## ------------------------------------------------------------
## Step 4: Correlation matrices
## ------------------------------------------------------------

get_cor_matrix <- function(data, vi_cols, yld_cols) {
  cor(data[, c(vi_cols, yld_cols)], use = "pairwise.complete.obs")
}

ndvi_cor  <- get_cor_matrix(ny_sny_pheno_22, ndvi_cols, yield_cols)
gndvi_cor <- get_cor_matrix(ny_sny_pheno_22, gndvi_cols, yield_cols)
ndre_cor  <- get_cor_matrix(ny_sny_pheno_22, ndre_cols, yield_cols)


## ------------------------------------------------------------
## Step 5: Define window boundaries (for boxes)
## ------------------------------------------------------------

window_bounds <- list(
  window1 = c(1, length(window1)),
  window2 = c(length(window1)+2, length(window1)+1+length(window2)),
  window3 = c(length(window1)+length(window2)+3,
              length(window1)+length(window2)+2+length(window3))
)


## ------------------------------------------------------------
## Step 6: Plot function
## ------------------------------------------------------------

plot_heatmap_with_boxes <- function(cor_matrix, order_vec, title_text) {
  
  # keep only valid columns (prevents crash)
  valid_order <- intersect(order_vec, rownames(cor_matrix))
  
  cor_matrix_ordered <- cor_matrix[valid_order, valid_order]
  
  cor_long <- reshape2::melt(cor_matrix_ordered)
  
  p <- ggplot(cor_long, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile() +
    
    scale_fill_gradient2(
      low = "blue", mid = "white", high = "red",
      midpoint = 0, limits = c(-1, 1)
    ) +
    
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(hjust = 0.5, face = "bold")
    ) +
    
    labs(title = title_text, x = "Traits", y = "Traits")
  
  # add window boxes
  for (bounds in window_bounds) {
    p <- p + geom_rect(
      xmin = bounds[1] - 0.5,
      xmax = bounds[2] + 0.5,
      ymin = bounds[1] - 0.5,
      ymax = bounds[2] + 0.5,
      color = "black",
      fill = NA,
      linewidth = 0.8
    )
  }
  
  return(p)
}


## ------------------------------------------------------------
## Step 7: Generate heatmaps
## ------------------------------------------------------------

ndvi_heatmap_sny_22  <- plot_heatmap_with_boxes(
  ndvi_cor, full_ndvi_order,
  "NDVI vs Yield Correlation (NY Sny 2022)")
ndvi_heatmap_sny_22
gndvi_heatmap_sny_22 <- plot_heatmap_with_boxes(
  gndvi_cor, full_gndvi_order,
  "GNDVI vs Yield Correlation (NY Sny 2022)")
gndvi_heatmap_sny_22
ndre_heatmap_sny_22  <- plot_heatmap_with_boxes(
  ndre_cor, full_ndre_order,
  "NDRE vs Yield Correlation (NY Sny 2022)")


## ------------------------------------------------------------
## Step 8: Save
## ------------------------------------------------------------

ggsave("NDVI_Pheno_cor_heatmap_ny_sny_2022.png",
       ndvi_heatmap_sny_22,
       path = "2022/output_images",
       width = 10, height = 8, dpi = 400)

ggsave("GNDVI_Pheno_cor_heatmap_ny_sny_2022.png",
       gndvi_heatmap_sny_22,
       path = "2022/output_images",
       width = 10, height = 8, dpi = 400)

ggsave("NDRE_Pheno_cor_heatmap_ny_sny_2022.png",
       ndre_heatmap_sny_22,
       path = "2022/output_images",
       width = 10, height = 8, dpi = 400)

```


 