---
title: "Running univariate models with GRM and spatial to calculate Heritability_ASReml in NY 2022 Trials"
author: "Dinesh Ghimire"
date: "2026-05-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr","asreml", "readxl", "tidyr", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "GGally")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```

#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NY-DG"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NY-DG 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa-NY-DG"
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is misssing please let me know. 
key<-read.csv("2022/meta_data/cultivar_list_key.csv") # contains both CU and NY cultivars here. 
# its same for 2022 and 2023 (cultivar list)
```




#NY 22


################################# Musgrave Trial ################################



1. *NY Musgrave 2022 *

Loading and preparing data files
```{r loading files 05072026}
#import merged phenotype file 
ny_mus_pheno_22 <- read.csv("2022/MSI_data/output_files/Final_NY22_mus_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_mus_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_mus_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_mus_pheno_22) 
str(ny_mus_pheno_22)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))

grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


- Scaling and outlier check on yield already done in script1, so we can jump into running models


Run the models

```{r run univariate models with spatial and GRM 05072026}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
colnames(ny_mus_pheno_22) 

# first for NDVI
# we have 22 imaging events so 22 heritabilities here. 
# NDVI_131, NDVI_145, NDVI_154, NDVI_159, NDVI_162, NDVI_164, NDVI_175, NDVI_178, NDVI_181, NDVI_187, NDVI_189, NDVI_192, NDVI_201, NDVI_203, NDVI_210, NDVI_215, NDVI_223, NDVI_236, NDVI_238, NDVI_243, NDVI_259, NDVI_263

#  For NDVI_131
st_ny22_mus_ndvi_131 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_131 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndvi_131)$varcomp

asreml::vpredict(st_ny22_mus_ndvi_131, heritA ~ V1 / (V1 + V5)) # 0.7031229 with 0.1662979
# without grm 0.64

```



For yield 

```{r Heritability of yield with GRM 05072026}

asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
# for yield 1
st_ny22_mus_yld1 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) + ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_yld1)$varcomp
asreml::vpredict(st_ny22_mus_yld1, heritA ~ V1 / (V1 + V5)) # 0.99 but bcz with full model, not converged
# with only range, 
asreml::vpredict(st_ny22_mus_yld1, heritA ~ V1 / (V1 + V4)) # 0.221098


# for yield 2
 # asreml.options(extra=1000,maxit =1000) # to setup more iterations for successful burn-in or convergence

st_ny22_mus_yld2 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_yld2)$varcomp
asreml::vpredict(st_ny22_mus_yld2, heritA ~ V1 / (V1 + V5)) # 0.9020296 with 0.48 SE


# for yield 3
st_ny22_mus_yld3 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_yld3)$varcomp
asreml::vpredict(st_ny22_mus_yld3, heritA ~ V1 / (V1 + V4)) # 0.3410772 with 0.19 SE

```





# Run for all VI's
### STORE BOTH HERITABILITY AND VARIANCE COMPONENTS

 To check variance components as well: store VC in another data frame
```{r For all VI's 05112026}
asreml.options(extra=300,maxit =300) # to setup more iterations for successful burn-in or convergence
# 131, 145, 154, 159, 162, 164, 175, 178, 181, 187, 189, 192, 201, 203, 210, 215, 223, 236, 238, 243, 259, 263
# Define time points
times_mus_22 <- c(131, 145, 154, 159, 162, 164, 175, 178, 181, 187, 189, 192, 201, 203, 210, 215, 223, 236, 238, 243, 259, 263)

# Initialize heritability results
herit_results <- data.frame(
  Time = times_mus_22,
  Herit_ndvi = NA, Herit_ndvi_spa = NA,
  Herit_gndvi = NA, Herit_gndvi_spa = NA,
  Herit_ndre = NA, Herit_ndre_spa = NA)

# Initialize variance component storage
varcomp_results <- data.frame(
  Time = numeric(0),
  Index = character(0),
  Model = character(0),
  Component = character(0),
  Estimate = numeric(0),
  stringsAsFactors = FALSE)

# Function to fit model and return heritability and variance components
fit_heritability <- function(index_prefix, time_point, data) {
  col_name <- paste0(index_prefix, "_", time_point)
  
  if (!col_name %in% colnames(data)) {
    warning(paste("Column", col_name, "not found. Skipping."))
    return(list(h2 = c(NA, NA), vc = NULL))
  }
  
  # --- Simple Model ---
  mod_simple <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Simple model failed for", col_name))
    return(NULL)
  })
  
  h2_simple <- NA
  vc_simple <- NULL
  if (!is.null(mod_simple)) {
    h2_simple <- tryCatch({
      vpredict(mod_simple, herit ~ V1 / (V1 + V2))$Estimate
    }, error = function(e) NA)
    
    vc_simple <- tryCatch({
      vc <- summary(mod_simple)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Simple")
    }, error = function(e) NULL)
  }
  
  # --- Spatial Model ---
  mod_spatial <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 ")),
           random = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Spatial model failed for", col_name))
    return(NULL)
  })
  
  h2_spatial <- NA
  vc_spatial <- NULL
  if (!is.null(mod_spatial)) {
    h2_spatial <- tryCatch({
      vpredict(mod_spatial, herit_spatial ~ V1 / (V1 + V5))$Estimate
    }, error = function(e) NA)
    
    vc_spatial <- tryCatch({
      vc <- summary(mod_spatial)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Spatial")
    }, error = function(e) NULL)
  }

  return(list(h2 = c(h2_simple, h2_spatial),
              vc = rbind(vc_simple, vc_spatial)))
}

# Loop through all time points and indices
for (i in seq_along(times_mus_22)) {
  time <- times_mus_22[i]
  cat("\nProcessing time point:", time, "\n")
  
  for (index in c("NDVI", "GNDVI", "NDRE")) {
    result <- fit_heritability(index, time, ny_mus_pheno_22)
    
    # Store heritabilities
    herit_results[i, paste0("Herit_", tolower(index))] <- result$h2[1]
    herit_results[i, paste0("Herit_", tolower(index), "_spa")] <- result$h2[2]
    
    # Store variance components
    if (!is.null(result$vc)) {
      result$vc$Index <- index
      result$vc$Time <- time
      varcomp_results <- bind_rows(varcomp_results, result$vc)
    }
  }
}

# Round heritability results to 2 decimals
herit_results_mus_22 <- herit_results %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

# Done!
print(head(herit_results_mus_22))
# print(head(varcomp_results))
View(herit_results_mus_22)
```


check for NA's or 1
- NDVI with spatial 189, 192, 201
- GNDVI with spatial: 154, 159, 189, 201, 210, 238, 243, 259, 263 
- NDRE with spatial: 159, 162, 164, 187, 201, 210, 236, 238, 243, 259, 263

Common to remove: 189

```{r NDVI Checking, 05072026}
# NDVI: 189, almost none
st_ny22_mus_ndvi_189 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_189 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndvi_189)$varcomp
 # tried all three combn
asreml::vpredict(st_ny22_mus_ndvi_189, heritA ~ V3 / (V3 + V4)) # 0.42 with row only.. restrained with range


# NDVI_192 : 
st_ny22_mus_ndvi_192 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_192 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndvi_192)$varcomp
asreml::vpredict(st_ny22_mus_ndvi_192, heritA ~ V1 / (V1 + V5)) # 0.99 

# NDVI_201 : 
st_ny22_mus_ndvi_201 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_201 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndvi_201)$varcomp
asreml::vpredict(st_ny22_mus_ndvi_201, heritA ~ V1 / (V1 + V5)) # 0.99

```



- GNDVI with spatial: 154, 159, 189, 201, 210, 238, 243, 259, 263 
```{r GNDVI Checking for cleaning, 05072026}
# GNDVI with spatial: 154, 159, 189, 201, 210, 238., 243, 259, 263 

# GNDVI_154 : 
st_ny22_mus_gndvi_154 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_154 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_gndvi_154)$varcomp
asreml::vpredict(st_ny22_mus_gndvi_154, heritA ~ V1 / (V1 + V5)) # 0.99 (spatial borrowed from residual)

# GNDVI_159 
st_ny22_mus_gndvi_159 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_159 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_gndvi_159)$varcomp
asreml::vpredict(st_ny22_mus_gndvi_159, heritA ~ V1 / (V1 + V5)) # 0.99 (spatial borrowed from residual)


# GNDVI_189: bad flight REMOVE after cross checking correlation
st_ny22_mus_gndvi_189 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_189 ~ 1, 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_gndvi_189)$varcomp
asreml::vpredict(st_ny22_mus_gndvi_189, heritA ~ V3 / (V3 + V4)) # ~ 0 with row only, but bound F.. restrained with row
asreml::vpredict(st_ny22_mus_gndvi_189, heritA ~ V1 / (V1 + V4)) # ~ 0 with row only, but bound F.. restrained with row


# GNDVI_201 (lets keep and look cz final flight of harvest 2
st_ny22_mus_gndvi_201 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_201 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_gndvi_201)$varcomp
asreml::vpredict(st_ny22_mus_gndvi_201, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI_210 : 0.32 (row)
st_ny22_mus_gndvi_210 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_210 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_gndvi_210)$varcomp
asreml::vpredict(st_ny22_mus_gndvi_210, heritA ~ V1 / (V1 + V4)) 
# doesn't run with full
asreml::vpredict(st_ny22_mus_gndvi_210, heritA ~ V3 / (V3 + V4)) 


# GNDVI_238 : ~ 0.16 (row)
st_ny22_mus_gndvi_238 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_238 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_gndvi_238)$varcomp
asreml::vpredict(st_ny22_mus_gndvi_238, heritA ~ V1 / (V1 + V4)) 
# witrh range only , 0.1 and with row only 0.16.Both didn't converge
asreml::vpredict(st_ny22_mus_gndvi_238, heritA ~ V3 / (V3 + V4)) 


# GNDVI_243 : 0.14 (row)
st_ny22_mus_gndvi_243 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_243 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) + ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_gndvi_243)$varcomp
asreml::vpredict(st_ny22_mus_gndvi_243, heritA ~ V1 / (V1 + V5)) 
asreml::vpredict(st_ny22_mus_gndvi_243, heritA ~ V3 / (V3 + V4)) 


# GNDVI_259 : ~ 0.99
st_ny22_mus_gndvi_259 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_259 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_gndvi_259)$varcomp
asreml::vpredict(st_ny22_mus_gndvi_259, heritA ~ V1 / (V1 + V5)) 
asreml::vpredict(st_ny22_mus_gndvi_259, heritA ~ V3 / (V3 + V4)) 

# GNDVI_263 : ~ 1
st_ny22_mus_gndvi_263 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_263 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_gndvi_263)$varcomp
asreml::vpredict(st_ny22_mus_gndvi_263, heritA ~ V1 / (V1 + V5)) 


```



- NDRE with spatial: 159, 162, 164, 187, 201, 210, 236, 238, 243, 259, 263
```{r NDRE Checking, 05112026}
# NDRE_159 : 0.99
st_ny22_mus_ndre_159 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_159 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_159)$varcomp
asreml::vpredict(st_ny22_mus_ndre_159, heritA ~ V1 / (V1 + V5)) # 0.99 (spatial borrowed from residual)

# NDRE_162 : 0.99
st_ny22_mus_ndre_162 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_162 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_162)$varcomp
asreml::vpredict(st_ny22_mus_ndre_162, heritA ~ V1 / (V1 + V5)) # 0.99 (spatial borrowed from residual)

# NDRE_164 : 0.99
st_ny22_mus_ndre_164 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_164 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_164)$varcomp
asreml::vpredict(st_ny22_mus_ndre_164, heritA ~ V1 / (V1 + V5)) # 0.99 (spatial borrowed from residual)

# NDRE_187 : ~ 0.1
st_ny22_mus_ndre_187 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_187 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_187)$varcomp
asreml::vpredict(st_ny22_mus_ndre_187, heritA ~ V3 / (V3 + V4)) # ~ 0.1

# NDRE_201 : ~ 0.1
st_ny22_mus_ndre_201 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_201 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_201)$varcomp
asreml::vpredict(st_ny22_mus_ndre_201, heritA ~ V3 / (V3 + V4)) # ~ 0.08

# NDRE_203 : 0.03
st_ny22_mus_ndre_203 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_203 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_203)$varcomp
asreml::vpredict(st_ny22_mus_ndre_203, heritA ~ V3 / (V3 + V4)) # 0.03

# NDRE_210 : 0.22
st_ny22_mus_ndre_210 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_210 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_210)$varcomp
asreml::vpredict(st_ny22_mus_ndre_210, heritA ~ V3 / (V3 + V4)) # ~ 0.22

# NDRE_236 : 0.13
st_ny22_mus_ndre_236 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_236 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_236)$varcomp
asreml::vpredict(st_ny22_mus_ndre_236, heritA ~ V3 / (V3 + V4)) # ~ 0.13


# NDRE_238 : 0.1
st_ny22_mus_ndre_238 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_238 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_238)$varcomp
asreml::vpredict(st_ny22_mus_ndre_238, heritA ~ V1 / (V1 + V4)) # 0.1 

# NDRE_243 : 0.1
st_ny22_mus_ndre_243 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_243 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_243)$varcomp
asreml::vpredict(st_ny22_mus_ndre_243, heritA ~ V3 / (V3 + V4)) # 0.1

# NDRE_259 : 0.99
st_ny22_mus_ndre_259 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_259 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_259)$varcomp
asreml::vpredict(st_ny22_mus_ndre_259, heritA ~ V1 / (V1 + V5)) # 0.99 (spatial borrowed from residual)


# NDRE_263 : 0.1
st_ny22_mus_ndre_263 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_263 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(st_ny22_mus_ndre_263)$varcomp
asreml::vpredict(st_ny22_mus_ndre_263, heritA ~ V3 / (V3 + V4)) # 0.1 
```





make box plot to visulaize and compare 

```{r Box plot to visualize heritability, Mus_22}
herit_results_mus_22 <- read.csv("2022/MSI_data/output_files/Heritabilities_withGRM_NY_Musgrave_22.csv", sep = ",") 
colnames(herit_results_mus_22)

# Reshape all six heritability columns to long format
herit_results_mus_22_long_all <- herit_results_mus_22 %>%
  select(Time, h2_ndvi, h2_ndvi_spa,
               h2_gndvi, h2_gndvi_spa,
               h2_ndre, h2_ndre_spa) %>%
  pivot_longer(cols = -Time, names_to = "Type", values_to = "Heritability") %>%
  mutate(
    Index = case_when(
  grepl("gndvi", Type) ~ "GNDVI",
  grepl("ndvi", Type) ~ "NDVI",
  grepl("ndre", Type) ~ "NDRE"
),
    Model = ifelse(grepl("_spa", Type), "Spatial", "Simple"))


gg_mus_22 <- ggplot(herit_results_mus_22_long_all, aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") + # Randomly "jitter" (spread out) the data points horizontally by up to ±0.2 units on the x-axis (i.e., within each box)
  # alternate geom_point() where all will be stacked in a vertical line so overalp issue.
  facet_wrap(~ Model) +
  labs(title = "h² by VI and Model Type in Musgrave NY 2022",
       x = "Vegetation Index",
       y = "Heritability (h²)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(c(0,1)) +
  scale_fill_brewer(palette = "Set2")

gg_mus_22

ggsave(filename = "Boxplot_Heritability_h2_VI_NY_Musgrave_2022.png", plot = gg_mus_22, path = "2022/output_images", width = 8, height = 7,dpi = 400)


############# Only simple h2 #######

# make boxplot of h2_ndvi, h2_gndvi, h2_ndre ( not spatial ones)

# Reshape only non-spatial heritability columns
herit_results_mus_22_long_simple <- herit_results_mus_22 %>%
  select(Time, h2_ndvi, h2_gndvi, h2_ndre) %>%
  pivot_longer(
    cols = -Time,
    names_to = "Type",
    values_to = "Heritability") %>%
  mutate(
    Index = case_when(
      Type == "h2_ndvi"  ~ "NDVI",
      Type == "h2_gndvi" ~ "GNDVI",
      Type == "h2_ndre"  ~ "NDRE"))

# Boxplot
gg_mus_22_simple <- ggplot(
  herit_results_mus_22_long_simple,
  aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +
  labs(
    title = "h² by Vegetation Index in Musgrave NY 2022",
    x = "Vegetation Index",
    y = "Heritability (h²)" ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(0, 1) +
  scale_fill_brewer(palette = "Set2")

gg_mus_22_simple

ggsave(filename = "Boxplot_Heritability_h2_VI_NY_Musgrave_2022_Simple_only.png",plot = gg_mus_22_simple,path = "2022/output_images",width = 7,height = 6,dpi = 400)

```







################################# Snyder Trial ################################




1. *NY Snyder 2022 *

Loading and preparing data files
```{r loading files 05122026}
#import merged phenotype file 
ny_sny_pheno_22 <- read.csv("2022/MSI_data/output_files/Final_NY22_Sny_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_sny_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_sny_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_sny_pheno_22) 
str(ny_sny_pheno_22)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))

grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


- Scaling and outlier check on yield already done in script1, so we can jump into running models


Run the models

```{r run univariate models with spatial and GRM 05122026}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
colnames(ny_sny_pheno_22) 

# first for NDVI
# we have 23 imaging events so 23 heritabilities here. 
# 133, 159, 164, 168, 171, 175, 178, 181, 187, 189, 192, 194, 196, 207, 223, 231, 236, 238, 243, 245, 252, 258, 259

#  For NDVI_133
st_ny22_sny_ndvi_133 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_133 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_ndvi_133)$varcomp

asreml::vpredict(st_ny22_sny_ndvi_133, heritA ~ V1 / (V1 + V5)) # 0.7958171 with 0.10

```



For yield 

```{r Heritability of yield with GRM 05122026}

asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
# for yield 1
st_ny22_sny_yld1 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_yld1)$varcomp
asreml::vpredict(st_ny22_sny_yld1, heritA ~ V1 / (V1 + V5)) # 0.6928387 with 0.08 se


# for yield 2
st_ny22_sny_yld2 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_yld2)$varcomp
asreml::vpredict(st_ny22_sny_yld2, heritA ~ V1 / (V1 + V5)) # 0.7565235 with 0.11 SE


# for yield 3
st_ny22_sny_yld3 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_yld3)$varcomp
asreml::vpredict(st_ny22_sny_yld3, heritA ~ V1 / (V1 + V5)) # 0.760 with 0.07 SE
```





# Run for all VI's
### STORE BOTH HERITABILITY AND VARIANCE COMPONENTS

 To check variance components as well: store VC in another data frame
```{r For all VI's 05122026}
asreml.options(extra=300,maxit =300) # to setup more iterations for successful burn-in or convergence
# 133, 159, 164, 168, 171, 175, 178, 181, 187, 189, 192, 194, 196, 207, 223, 231, 236, 238, 243, 245, 252, 258, 259
# Define time points
times_sny_22 <- c(133, 159, 164, 168, 171, 175, 178, 181, 187, 189, 192, 194, 196, 207, 223, 231, 236, 238, 243, 245, 252, 258, 259)

# Initialize heritability results
herit_results <- data.frame(
  Time = times_sny_22,
  Herit_ndvi = NA, Herit_ndvi_spa = NA,
  Herit_gndvi = NA, Herit_gndvi_spa = NA,
  Herit_ndre = NA, Herit_ndre_spa = NA)

# Initialize variance component storage
varcomp_results <- data.frame(
  Time = numeric(0),
  Index = character(0),
  Model = character(0),
  Component = character(0),
  Estimate = numeric(0),
  stringsAsFactors = FALSE)

# Function to fit model and return heritability and variance components
fit_heritability <- function(index_prefix, time_point, data) {
  col_name <- paste0(index_prefix, "_", time_point)
  
  if (!col_name %in% colnames(data)) {
    warning(paste("Column", col_name, "not found. Skipping."))
    return(list(h2 = c(NA, NA), vc = NULL))
  }
  
  # --- Simple Model ---
  mod_simple <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep ")),
           random = ~ vm(Germplasm, grm_inverse),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Simple model failed for", col_name))
    return(NULL)
  })
  
  h2_simple <- NA
  vc_simple <- NULL
  if (!is.null(mod_simple)) {
    h2_simple <- tryCatch({
      vpredict(mod_simple, herit ~ V1 / (V1 + V2))$Estimate
    }, error = function(e) NA)
    
    vc_simple <- tryCatch({
      vc <- summary(mod_simple)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Simple")
    }, error = function(e) NULL)
  }
  
  # --- Spatial Model ---
  mod_spatial <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 ")),
           random = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Spatial model failed for", col_name))
    return(NULL)
  })
  
  h2_spatial <- NA
  vc_spatial <- NULL
  if (!is.null(mod_spatial)) {
    h2_spatial <- tryCatch({
      vpredict(mod_spatial, herit_spatial ~ V1 / (V1 + V5))$Estimate
    }, error = function(e) NA)
    
    vc_spatial <- tryCatch({
      vc <- summary(mod_spatial)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Spatial")
    }, error = function(e) NULL)
  }

  return(list(h2 = c(h2_simple, h2_spatial),
              vc = rbind(vc_simple, vc_spatial)))
}

# Loop through all time points and indices
for (i in seq_along(times_sny_22)) {
  time <- times_sny_22[i]
  cat("\nProcessing time point:", time, "\n")
  
  for (index in c("NDVI", "GNDVI", "NDRE")) {
    result <- fit_heritability(index, time, ny_sny_pheno_22)
    
    # Store heritabilities
    herit_results[i, paste0("Herit_", tolower(index))] <- result$h2[1]
    herit_results[i, paste0("Herit_", tolower(index), "_spa")] <- result$h2[2]
    
    # Store variance components
    if (!is.null(result$vc)) {
      result$vc$Index <- index
      result$vc$Time <- time
      varcomp_results <- bind_rows(varcomp_results, result$vc)
    }
  }
}

# Round heritability results to 2 decimals
herit_results_sny_22 <- herit_results %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

# Done!
print(head(herit_results_sny_22))
print(head(varcomp_results))
View(herit_results_sny_22)

```



check for NA's or 1
- NDVI with spatial 192, 238, 
- GNDVI with spatial: 168, 171, 175, 181, 189, 192, 194, 196, 223, 231, 238, 245, 252
- NDRE with spatial: 171, 175, 178, 187, 194

### NDVI with spatial 192, 238, 
```{r Checking NDVI flights, 05122026}
# NDVI: 192, almost none
st_ny22_sny_ndvi_192 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_192 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_ndvi_192)$varcomp
 # tried all three combn
asreml::vpredict(st_ny22_sny_ndvi_192, heritA ~ V1 / (V1 + V5)) 


# NDVI: 238, 0.99 (spatial steals all residuals) but with only row, almost none... 
st_ny22_sny_ndvi_238 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_238 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_ndvi_238)$varcomp
 # tried all three combn
asreml::vpredict(st_ny22_sny_ndvi_238, heritA ~ V1 / (V1 + V5)) 

```


### GNDVI with spatial: 168, 171, 175, 181, 189, 192, 194, 196, 223, 231, 238, 245, 252
```{r Checking GNDVI flights, 05122026}
# GNDVI: 168, 0.99
st_ny22_sny_gndvi_168 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_168 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_168)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_168, heritA ~ V1 / (V1 + V5)) 

# GNDVI: 171, 0.99
st_ny22_sny_gndvi_171 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_171 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_171)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_171, heritA ~ V1 / (V1 + V5))

# GNDVI: 175, 0.99
st_ny22_sny_gndvi_175 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_175 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_175)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_175, heritA ~ V1 / (V1 + V5))

# GNDVI: 181, 0.99 (but difficulty in converging)
st_ny22_sny_gndvi_181 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_181 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_181)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_181, heritA ~ V1 / (V1 + V5))


# GNDVI: 189, 0.99
st_ny22_sny_gndvi_189 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_189 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_189)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_189, heritA ~ V1 / (V1 + V5))


# GNDVI: 192, almost none with only row or range. But 0.99 (spatial stealing all of residual). need to confirm this flight's use with mt model.. REMOVE it
st_ny22_sny_gndvi_192 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_192 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_192)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_192, heritA ~ V1 / (V1 + V5)) 

# GNDVI: 194, 0.99 (spatial steals residuals)
st_ny22_sny_gndvi_194 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_194 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_194)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_194, heritA ~ V1 / (V1 + V5)) 

# GNDVI: 196, didn't ran with grm and full spatial so ran without grm and got 0.999, but confirm with mt model
st_ny22_sny_gndvi_196 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_196 ~ 1 , 
         random = ~ id(Germplasm) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_196)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_196, heritA ~ V1 / (V1 + V5)) # 0.999

# GNDVI: 223, 0.999
st_ny22_sny_gndvi_223 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_223 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_223)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_223, heritA ~ V1 / (V1 + V5)) 

# GNDVI: 231, 0.999
st_ny22_sny_gndvi_231 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_231 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_231)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_231, heritA ~ V1 / (V1 + V5)) 

# GNDVI: 238, 0.999
st_ny22_sny_gndvi_238 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_238 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_238)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_238, heritA ~ V1 / (V1 + V5)) 

# GNDVI: 245, 0.53 with row only
st_ny22_sny_gndvi_245 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_245 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_245)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_245, heritA ~ V3 / (V3 + V4)) # 0.53

# GNDVI: 252, 0.35 with row
st_ny22_sny_gndvi_252 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_252 ~ 1 , 
         random = ~ id(Germplasm) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_gndvi_252)$varcomp
asreml::vpredict(st_ny22_sny_gndvi_252, heritA ~ V1 / (V1 + V5)) # 0.35 with row
```



### NDRE with spatial: 171, 175, 178, 187, 194

```{r Checking NDRE flights, 05122026}
# NDRE: 171, 0.99 (spatial steals residual)
st_ny22_sny_ndre_171 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_171 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_ndre_171)$varcomp
asreml::vpredict(st_ny22_sny_ndre_171, heritA ~ V1 / (V1 + V5)) 

# NDRE: 175, 0.99 (spatial steals residual)
st_ny22_sny_ndre_175 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_175 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_ndre_175)$varcomp
asreml::vpredict(st_ny22_sny_ndre_175, heritA ~ V1 / (V1 + V5)) 

# NDRE: 178, almost none (0) in all three combinations
st_ny22_sny_ndre_178 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_178 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_ndre_178)$varcomp
asreml::vpredict(st_ny22_sny_ndre_178, heritA ~ V1 / (V1 + V5)) 

# NDRE: 187, 0.55 
st_ny22_sny_ndre_187 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_187 ~ 1 , 
         random = ~ id(Germplasm) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_ndre_187)$varcomp
asreml::vpredict(st_ny22_sny_ndre_187, heritA ~ V1 / (V1 + V5)) # 0.55

# NDRE: 194, 0.99 (spatial steals residual)
st_ny22_sny_ndre_194 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_194 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(st_ny22_sny_ndre_194)$varcomp
asreml::vpredict(st_ny22_sny_ndre_194, heritA ~ V1 / (V1 + V5)) 
```



make box plot to visulaize and compare 

```{r Box plot to visualize heritability, Sny_22}
herit_results_sny_22 <- read.csv("2022/MSI_data/output_files/Heritabilities_withGRM_NY_Snyder_22.csv", sep = ",") 
colnames(herit_results_sny_22)

# Reshape all six heritability columns to long format
herit_results_sny_22_long_all <- herit_results_sny_22 %>%
  select(Time, h2_ndvi, h2_ndvi_spa,
               h2_gndvi, h2_gndvi_spa,
               h2_ndre, h2_ndre_spa) %>%
  pivot_longer(cols = -Time, names_to = "Type", values_to = "Heritability") %>%
  mutate(
    Index = case_when(
  grepl("gndvi", Type) ~ "GNDVI",
  grepl("ndvi", Type) ~ "NDVI",
  grepl("ndre", Type) ~ "NDRE"
),
    Model = ifelse(grepl("_spa", Type), "Spatial", "Simple"))


gg_sny_22 <- ggplot(herit_results_sny_22_long_all, aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") + # Randomly "jitter" (spread out) the data points horizontally by up to ±0.2 units on the x-axis (i.e., within each box)
  # alternate geom_point() where all will be stacked in a vertical line so overalp issue.
  facet_wrap(~ Model) +
  labs(title = "h² by VI and Model Type in Snyder NY 2022",
       x = "Vegetation Index",
       y = "Heritability (h²)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(c(0,1)) +
  scale_fill_brewer(palette = "Set2")

gg_sny_22

ggsave(filename = "Boxplot_Heritability_h2_VI_NY_snygrave_2022.png", plot = gg_sny_22, path = "2022/output_images", width = 8, height = 7,dpi = 400)


############# Only simple h2 #######
# make boxplot of h2_ndvi, h2_gndvi, h2_ndre ( not spatial ones)

# Reshape only non-spatial heritability columns
herit_results_sny_22_long_simple <- herit_results_sny_22 %>%
  select(Time, h2_ndvi, h2_gndvi, h2_ndre) %>%
  pivot_longer(
    cols = -Time,
    names_to = "Type",
    values_to = "Heritability") %>%
  mutate(
    Index = case_when(
      Type == "h2_ndvi"  ~ "NDVI",
      Type == "h2_gndvi" ~ "GNDVI",
      Type == "h2_ndre"  ~ "NDRE"))

# Boxplot
gg_sny_22_simple <- ggplot(
  herit_results_sny_22_long_simple,
  aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +
  labs(
    title = "h² by Vegetation Index in Snyder NY 2022",
    x = "Vegetation Index",
    y = "Heritability (h²)" ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(0, 1) +
  scale_fill_brewer(palette = "Set2")

gg_sny_22_simple

ggsave(filename = "Boxplot_Heritability_h2_VI_NY_Snyder_2022_Simple_only.png",plot = gg_sny_22_simple,path = "2022/output_images",width = 7,height = 6,dpi = 400)
```

 
 