---
title: "Running bivariate models (mt models) with GRM and spatial to calculate genetic and spatial correlations"
author: "Dinesh Ghimire"
date: "2026-05-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr","asreml", "readxl", "tidyr", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "GGally")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```

#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NY-DG"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NY-DG 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa-NY-DG"
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is misssing please let me know. 
key<-read.csv("2022/meta_data/cultivar_list_key.csv") # contains both CU and NY cultivars here. 
# its same for 2022 and 2023 (cultivar list)
```




#NY 22


################################# Musgrave Trial ################################



1. *NY Musgrave 2022 *

Loading and preparing data files
```{r loading files 05072026}
#import merged phenotype file 
ny_mus_pheno_22 <- read.csv("2022/MSI_data/output_files/Final_NY22_mus_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_mus_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_mus_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_mus_pheno_22) 
str(ny_mus_pheno_22)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))


grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


- Scaling and outlier check on yield already done in script1, so we can jump into running models

RUN for all... 

## 1.2. For GNDVI 
- For Harvest window 1:  131, 145, 154, and yld1
- For Harvest window 2:  159, 162, 164, 175, 178, 181, 187, 189, 192, 201, 203, and yld2
- For Harvest window 3: 210, 215, 223, 236, 238, 243, 259, 263, yld3


## only running bivariate for yield with ar spatial to see how MSI and corresponding yield are 
# Bivariate for yield with spatial
- if doesn't run, try sca_yld OR simpler model structure like diag instead of unstructured or one dimensional spatial
- always double check correlation value formula (shortcut used like V10 or V11 being used in the equation)

```{r Multitrait model, GNDVI All flights, 05112026}
### Multitrait model
asreml.options(extra=200, maxit =200) 
unique(ny_mus_pheno_22$Germplasm)
##################### Harvest 1 Window ####################
 # - For Harvest window 1:  131, 145, 154, and yld1

# GNDVI_131 : 
mt_mus_pheno_22_gndvi_131 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_131) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # diag resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_131)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_131, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6881653
asreml::vpredict(mt_mus_pheno_22_gndvi_131, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5979523
asreml::vpredict(mt_mus_pheno_22_gndvi_131, res.cor ~ (V11)/sqrt(V10*V12)) 




# GNDVI_145 
mt_mus_pheno_22_gndvi_145 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_145) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_145)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_145, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6569631
asreml::vpredict(mt_mus_pheno_22_gndvi_145, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.3435204
asreml::vpredict(mt_mus_pheno_22_gndvi_145, res.cor ~ (V11)/sqrt(V10*V12)) 

# GNDVI_154 
mt_mus_pheno_22_gndvi_154 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_154) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_154)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_154, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7982475
asreml::vpredict(mt_mus_pheno_22_gndvi_154, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.551793
asreml::vpredict(mt_mus_pheno_22_gndvi_154, res.cor ~ (V11)/sqrt(V10*V12)) 





##################### Harvest 2 Window ####################
# For Harvest window 2:  159, 162, 164, 175, 178, 181, 187, 189, 192, 201, 203, and yld2

# GNDVI_159 
mt_mus_pheno_22_gndvi_159 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_159) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_159)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_159, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6224571
asreml::vpredict(mt_mus_pheno_22_gndvi_159, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.04083438
asreml::vpredict(mt_mus_pheno_22_gndvi_159, res.cor ~ (V11)/sqrt(V10*V12)) 

# GNDVI_162 
mt_mus_pheno_22_gndvi_162 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_162) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_162)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_162, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6498479
asreml::vpredict(mt_mus_pheno_22_gndvi_162, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.1260206
asreml::vpredict(mt_mus_pheno_22_gndvi_162, res.cor ~ (V11)/sqrt(V10*V12)) 


# GNDVI_164 (worked iwth yld, not sca_yld)
mt_mus_pheno_22_gndvi_164 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld2, GNDVI_164) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_164)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_164, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6372299
asreml::vpredict(mt_mus_pheno_22_gndvi_164, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1630968
asreml::vpredict(mt_mus_pheno_22_gndvi_164, res.cor ~ (V11)/sqrt(V10*V12)) # 


# GNDVI_175 
mt_mus_pheno_22_gndvi_175 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_175) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_175)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_175, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8049411
asreml::vpredict(mt_mus_pheno_22_gndvi_175, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.2569012
asreml::vpredict(mt_mus_pheno_22_gndvi_175, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_178 
mt_mus_pheno_22_gndvi_178 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_178) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_178)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_178, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7540542
asreml::vpredict(mt_mus_pheno_22_gndvi_178, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.3804054
asreml::vpredict(mt_mus_pheno_22_gndvi_178, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_181 (worked with yld2)
mt_mus_pheno_22_gndvi_181 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld2, GNDVI_181) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_181)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_181, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7670361
asreml::vpredict(mt_mus_pheno_22_gndvi_181, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.3409831
asreml::vpredict(mt_mus_pheno_22_gndvi_181, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_187 
mt_mus_pheno_22_gndvi_187 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_187) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_187)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_187, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8549017
asreml::vpredict(mt_mus_pheno_22_gndvi_187, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4881311
asreml::vpredict(mt_mus_pheno_22_gndvi_187, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_189 
mt_mus_pheno_22_gndvi_189 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_189) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_189)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_189, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7816049
asreml::vpredict(mt_mus_pheno_22_gndvi_189, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6029039
asreml::vpredict(mt_mus_pheno_22_gndvi_189, res.cor ~ (V11)/sqrt(V10*V12)) 


# GNDVI_192 (worked iwth yld2)
mt_mus_pheno_22_gndvi_192 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld2, GNDVI_192) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_192)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_192, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6509408
asreml::vpredict(mt_mus_pheno_22_gndvi_192, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6777008
asreml::vpredict(mt_mus_pheno_22_gndvi_192, res.cor ~ (V11)/sqrt(V10*V12)) # 
(2.716697e-05)/ sqrt(1.087104e-03)*sqrt(5.875462e-06)

# GNDVI_201 
mt_mus_pheno_22_gndvi_201 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_201) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_201)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_201, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6827731
asreml::vpredict(mt_mus_pheno_22_gndvi_201, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6031344
asreml::vpredict(mt_mus_pheno_22_gndvi_201, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_203 
mt_mus_pheno_22_gndvi_203 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_203) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_203)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_203, gen.cor ~ (V2)/sqrt(V1*V3)) # -0.06414355
asreml::vpredict(mt_mus_pheno_22_gndvi_203, spa.cor ~ (V5)/sqrt(V4*V6)) # -0.02408801
asreml::vpredict(mt_mus_pheno_22_gndvi_203, res.cor ~ (V11)/sqrt(V10*V12)) # 





##################### Harvest 3 Window ####################
#For Harvest window 3: 210, 215, 223, 236, 238, 243, 259, 263, yld3

# GNDVI_210 : bad (remove)
mt_mus_pheno_22_gndvi_210 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_210) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_210)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_210, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.41
asreml::vpredict(mt_mus_pheno_22_gndvi_210, spa.cor ~ (V5)/sqrt(V4*V6)) # 0
asreml::vpredict(mt_mus_pheno_22_gndvi_210, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_215 : bad remove
mt_mus_pheno_22_gndvi_215 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_215) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_215)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_215, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.1789935
asreml::vpredict(mt_mus_pheno_22_gndvi_215, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4212035
asreml::vpredict(mt_mus_pheno_22_gndvi_215, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_223 
mt_mus_pheno_22_gndvi_223 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_223) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_223)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_223, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.1501679
asreml::vpredict(mt_mus_pheno_22_gndvi_223, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4342083
asreml::vpredict(mt_mus_pheno_22_gndvi_223, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_236 : 
mt_mus_pheno_22_gndvi_236 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_236) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_236)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_236, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.2810241
asreml::vpredict(mt_mus_pheno_22_gndvi_236, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5123481
asreml::vpredict(mt_mus_pheno_22_gndvi_236, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_238 : 
mt_mus_pheno_22_gndvi_238 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_238) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_238)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_238, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.307748
asreml::vpredict(mt_mus_pheno_22_gndvi_238, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5657587
asreml::vpredict(mt_mus_pheno_22_gndvi_238, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_243 : 
mt_mus_pheno_22_gndvi_243 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_243) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_243)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_243, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.4380544
asreml::vpredict(mt_mus_pheno_22_gndvi_243, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6155311
asreml::vpredict(mt_mus_pheno_22_gndvi_243, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_259 : 
mt_mus_pheno_22_gndvi_259 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_259) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_259)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_259, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.3762865
asreml::vpredict(mt_mus_pheno_22_gndvi_259, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5476225
asreml::vpredict(mt_mus_pheno_22_gndvi_259, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_263 : 
mt_mus_pheno_22_gndvi_263 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_263) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # us resid
         data = ny_mus_pheno_22)}
summary(mt_mus_pheno_22_gndvi_263)$varcomp
asreml::vpredict(mt_mus_pheno_22_gndvi_263, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.2855455
asreml::vpredict(mt_mus_pheno_22_gndvi_263, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6006065
asreml::vpredict(mt_mus_pheno_22_gndvi_263, res.cor ~ (V11)/sqrt(V10*V12)) # 

```


Based on assessment of flight quality, I removed flights 203, 210, 215, 223


## Graphs: Genetic and Spatial trend throughout the season
```{r Mt models graphs GNDVI Musgrave 2022 05122026}
mt_models_gndvi_mus_22 <- read.csv("2022/MSI_data/output_files/MT_models_GC_SC_values_GNDVI_Musgrave_2022.csv")
colnames(mt_models_gndvi_mus_22)
# Use your dataframe
df_gndvi_mus_22 <- mt_models_gndvi_mus_22
# Preserve order of Trait
df_gndvi_mus_22$Trait <- factor(df_gndvi_mus_22$Trait, levels = df_gndvi_mus_22$Trait)

# ---- Graph 1: Genetic1 only ----
df_gndvi_mus_22_no_spa <- df_gndvi_mus_22 %>%
  select(Trait, Genetic) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value")

gndvi_mus_22_p1 <- ggplot(df_gndvi_mus_22_no_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI without Spatial  mus 2022",
       y = "Correlation")
gndvi_mus_22_p1
# ---- Graph 2: Genetic, & Spatial  ----
df_gndvi_mus_22_spa <- df_gndvi_mus_22 %>%
  select(Trait, Genetic, Spatial) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value") %>%
  filter(!is.na(Value))

gndvi_mus_22_p2 <- ggplot(df_gndvi_mus_22_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Genetic" = "forestgreen", 
                                "Spatial" = "darkorange")) +
  scale_linetype_manual(values = c("Genetic" = "solid", 
                                   "Spatial" = "solid")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "GNDVI Trend over Musgrave 2022",
       y = "Correlation")

# Show both plots
gndvi_mus_22_p1 # only genetic
gndvi_mus_22_p2 # genetic and spatial

# ggsave
ggsave("2022/output_images/Genetic_Spatial_Correlations_NY_Musgrave_2022.png", gndvi_mus_22_p2, width = 8, height = 6, dpi = 400)
```








################################# Snyder Trial ################################

1. *NY Snyder 2022 *

Loading and preparing data files
```{r loading files 05122026}
#import merged phenotype file 
ny_sny_pheno_22 <- read.csv("2022/MSI_data/output_files/Final_NY22_sny_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_sny_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_sny_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_sny_pheno_22) 
str(ny_sny_pheno_22)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))


grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


- Scaling and outlier check on yield already done in script1, so we can jump into running models

RUN for all... 

## 1.2. For GNDVI 
- For Harvest window 1:  133, 159, 164, and yld1
- For Harvest window 2:  168, 171, 175, 178, 181, 187, 189, 192, 194, 196, 207, and yld2
- For Harvest window 3: 223, 231, 236, 238, 243, 245, 252, 258, 259, yld3
= total 23 flights


## only running bivariate for yield with ar spatial to see how MSI and corresponding yield are 
# Bivariate for yield with spatial
- if doesn't run, try sca_yld OR simpler model structure like diag instead of unstructured or one dimensional spatial


```{r Multitrait model, GNDVI All flights, Snyder 05122026}
### Multitrait model
asreml.options(extra=200, maxit =200) 
unique(ny_sny_pheno_22$Germplasm)
##################### Harvest 1 Window ####################
 # - For Harvest window 1:  133, 159, 164, and yld1

# GNDVI_133 : 
mt_sny_pheno_22_gndvi_133 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_133) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_133)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_133, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.4901762
asreml::vpredict(mt_sny_pheno_22_gndvi_133, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.3388202
asreml::vpredict(mt_sny_pheno_22_gndvi_133, res.cor ~ (V11)/sqrt(V10*V12)) # 0.9639333



# GNDVI_159 
mt_sny_pheno_22_gndvi_159 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_159) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_159)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_159, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.307729
asreml::vpredict(mt_sny_pheno_22_gndvi_159, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.1083884
asreml::vpredict(mt_sny_pheno_22_gndvi_159, res.cor ~ (V11)/sqrt(V10*V12)) # 0.1920479

# GNDVI_164 
mt_sny_pheno_22_gndvi_164 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_164) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_164)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_164, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.03
asreml::vpredict(mt_sny_pheno_22_gndvi_164, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.17
asreml::vpredict(mt_sny_pheno_22_gndvi_164, res.cor ~ (V11)/sqrt(V10*V12)) # diag





##################### Harvest 2 Window ####################
# For Harvest window 2:  168, 171, 175, 178, 181, 187, 189, 192, 194, 196, 207, and yld2

# GNDVI_168 : bad flight
mt_sny_pheno_22_gndvi_168 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_168) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_168)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_168, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.05907064 
asreml::vpredict(mt_sny_pheno_22_gndvi_168, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.06646539
asreml::vpredict(mt_sny_pheno_22_gndvi_168, res.cor ~ (V11)/sqrt(V10*V12)) # - 0.645327

# GNDVI_171 #
mt_sny_pheno_22_gndvi_171 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(yld2, GNDVI_171) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):diag(trait),# diag residual
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_171)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_171, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.28
asreml::vpredict(mt_sny_pheno_22_gndvi_171, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.05
asreml::vpredict(mt_sny_pheno_22_gndvi_171, res.cor ~ (V11)/sqrt(V10*V12)) # 0.74

# GNDVI_175 #
mt_sny_pheno_22_gndvi_175 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(yld2, GNDVI_175) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):diag(trait), # diag residual
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_175)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_175, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.5373959
asreml::vpredict(mt_sny_pheno_22_gndvi_175, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.2384265
asreml::vpredict(mt_sny_pheno_22_gndvi_175, res.cor ~ (V11)/sqrt(V10*V12)) # diag


# GNDVI_178 #
mt_sny_pheno_22_gndvi_178 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(yld2, GNDVI_178) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):diag(trait), # diag residual
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_178)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_178, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.5556746
asreml::vpredict(mt_sny_pheno_22_gndvi_178, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4159851
asreml::vpredict(mt_sny_pheno_22_gndvi_178, res.cor ~ (V11)/sqrt(V10*V12)) # diag residual


# GNDVI_181 #
mt_sny_pheno_22_gndvi_181 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(yld2, GNDVI_181) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):diag(trait),
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_181)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_181, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.52
asreml::vpredict(mt_sny_pheno_22_gndvi_181, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.24
asreml::vpredict(mt_sny_pheno_22_gndvi_181, res.cor ~ (V11)/sqrt(V10*V12)) # diag residual


# GNDVI_187 #
mt_sny_pheno_22_gndvi_187 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(yld2, GNDVI_187) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):us(trait),
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_187)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_187, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.3298013
asreml::vpredict(mt_sny_pheno_22_gndvi_187, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.17
asreml::vpredict(mt_sny_pheno_22_gndvi_187, res.cor ~ (V11)/sqrt(V10*V12)) # 0.19


# GNDVI_189 #
mt_sny_pheno_22_gndvi_189 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(yld2, GNDVI_189) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):diag(trait), # diag residual
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_189)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_189, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.35
asreml::vpredict(mt_sny_pheno_22_gndvi_189, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.20
asreml::vpredict(mt_sny_pheno_22_gndvi_189, res.cor ~ (V11)/sqrt(V10*V12)) # diag residual


# GNDVI_192 # bad flight
mt_sny_pheno_22_gndvi_192 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(sca_yld2, GNDVI_192) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):diag(trait),
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_192)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_192, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.32
asreml::vpredict(mt_sny_pheno_22_gndvi_192, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1
asreml::vpredict(mt_sny_pheno_22_gndvi_192, res.cor ~ (V11)/sqrt(V10*V12)) # diag


# GNDVI_194 #
mt_sny_pheno_22_gndvi_194 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(yld2, GNDVI_194) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):us(trait),
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_194)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_194, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.27
asreml::vpredict(mt_sny_pheno_22_gndvi_194, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.21
asreml::vpredict(mt_sny_pheno_22_gndvi_194, res.cor ~ (V11)/sqrt(V10*V12)) # 0.45


# GNDVI_196 #
mt_sny_pheno_22_gndvi_196 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(yld2, GNDVI_196) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):us(trait),
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_196)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_196, gen.cor ~ (V2)/sqrt(V1*V3)) #0.30
asreml::vpredict(mt_sny_pheno_22_gndvi_196, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.41
asreml::vpredict(mt_sny_pheno_22_gndvi_196, res.cor ~ (V11)/sqrt(V10*V12)) # 0.32


# GNDVI_207 #
mt_sny_pheno_22_gndvi_207 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = cbind(yld2, GNDVI_207) ~ trait,
    random = ~ us(trait):vm(Germplasm, grm_inverse) +
      us(trait):ar1(Row):ar1(Range),
    residual = ~ id(units):us(trait),
    data = ny_sny_pheno_22)}

summary(mt_sny_pheno_22_gndvi_207)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_207, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.32
asreml::vpredict(mt_sny_pheno_22_gndvi_207, spa.cor ~ (V5)/sqrt(V4*V6)) # -0.14
asreml::vpredict(mt_sny_pheno_22_gndvi_207, res.cor ~ (V11)/sqrt(V10*V12)) # 0.15





##################### Harvest 3 Window ####################
#For Harvest window 3: 223, 231, 236, 238, 243, 245, 252, 258, 259, yld3

# GNDVI_223 # bad flight here(could be informative, but not in my study)
mt_sny_pheno_22_gndvi_223 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_223) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_223)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_223, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.4682964
asreml::vpredict(mt_sny_pheno_22_gndvi_223, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.4912361
asreml::vpredict(mt_sny_pheno_22_gndvi_223, res.cor ~ (V11)/sqrt(V10*V12)) # 0.96

# GNDVI_231 : bad flight here(could be informative, but not in my study)
mt_sny_pheno_22_gndvi_231 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_231) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_231)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_231, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.5063595
asreml::vpredict(mt_sny_pheno_22_gndvi_231, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.438223
asreml::vpredict(mt_sny_pheno_22_gndvi_231, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6696606

# GNDVI_236 : bad flight here(could be informative, but not in my study)
mt_sny_pheno_22_gndvi_236 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_236) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_236)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_236, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.4771722
asreml::vpredict(mt_sny_pheno_22_gndvi_236, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1728919
asreml::vpredict(mt_sny_pheno_22_gndvi_236, res.cor ~ (V11)/sqrt(V10*V12)) # 0.3186256

# GNDVI_238 : bad flight here(could be informative, but not in my study)
mt_sny_pheno_22_gndvi_238 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_238) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_238)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_238, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.5063141
asreml::vpredict(mt_sny_pheno_22_gndvi_238, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.14
asreml::vpredict(mt_sny_pheno_22_gndvi_238, res.cor ~ (V11)/sqrt(V10*V12)) # 0.46

# GNDVI_243 : bad flight here(could be informative, but not in my study)
mt_sny_pheno_22_gndvi_243 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_243) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_243)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_243, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.3506548
asreml::vpredict(mt_sny_pheno_22_gndvi_243, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1466783
asreml::vpredict(mt_sny_pheno_22_gndvi_243, res.cor ~ (V11)/sqrt(V10*V12)) # 0.2959028

# GNDVI_245 : bad flight here(could be informative, but not in my study)
mt_sny_pheno_22_gndvi_245 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_245) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_245)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_245, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.3580192
asreml::vpredict(mt_sny_pheno_22_gndvi_245, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.13
asreml::vpredict(mt_sny_pheno_22_gndvi_245, res.cor ~ (V11)/sqrt(V10*V12)) # -0.12

# GNDVI_252 : bad flight here(could be informative, but not in my study)
mt_sny_pheno_22_gndvi_252 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_252) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_252)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_252, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.1
asreml::vpredict(mt_sny_pheno_22_gndvi_252, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.23
asreml::vpredict(mt_sny_pheno_22_gndvi_252, res.cor ~ (V11)/sqrt(V10*V12)) # - 0.003

# GNDVI_258 
mt_sny_pheno_22_gndvi_258 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_258) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_258)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_258, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.2401172
asreml::vpredict(mt_sny_pheno_22_gndvi_258, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.3642449
asreml::vpredict(mt_sny_pheno_22_gndvi_258, res.cor ~ (V11)/sqrt(V10*V12)) # - 0.0020418


# GNDVI_259 : 
mt_sny_pheno_22_gndvi_259 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_259) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_22)}
summary(mt_sny_pheno_22_gndvi_259)$varcomp
asreml::vpredict(mt_sny_pheno_22_gndvi_259, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.05040666
asreml::vpredict(mt_sny_pheno_22_gndvi_259, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.480287
asreml::vpredict(mt_sny_pheno_22_gndvi_259, res.cor ~ (V11)/sqrt(V10*V12)) # 0.1749894

```




## Graphs: Genetic and Spatial trend throughout the season
```{r Mt models graphs GNDVI Snyder 2022 05122026}
mt_models_gndvi_sny_22 <- read.csv("2022/MSI_data/output_files/MT_models_GC_SC_values_GNDVI_Snyder_2022.csv")
colnames(mt_models_gndvi_sny_22)
# Use your dataframe
df_gndvi_sny_22 <- mt_models_gndvi_sny_22
# Preserve order of Trait
df_gndvi_sny_22$Trait <- factor(df_gndvi_sny_22$Trait, levels = df_gndvi_sny_22$Trait)

# ---- Graph 1: Genetic1 only ----
df_gndvi_sny_22_no_spa <- df_gndvi_sny_22 %>%
  select(Trait, Genetic) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value")

gndvi_sny_22_p1 <- ggplot(df_gndvi_sny_22_no_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI without Spatial  Snyder 2022",
       y = "Correlation")
gndvi_sny_22_p1
# ---- Graph 2: Genetic, & Spatial  ----
df_gndvi_sny_22_spa <- df_gndvi_sny_22 %>%
  select(Trait, Genetic, Spatial) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value") %>%
  filter(!is.na(Value))

gndvi_sny_22_p2 <- ggplot(df_gndvi_sny_22_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Genetic" = "forestgreen", 
                                "Spatial" = "darkorange")) +
  scale_linetype_manual(values = c("Genetic" = "solid", 
                                   "Spatial" = "solid")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "GNDVI Trend over Snyder 2022",
       y = "Correlation")

# Show both plots
gndvi_sny_22_p1 # only genetic
gndvi_sny_22_p2 # genetic and spatial

# ggsave
ggsave("2022/output_images/Genetic_Spatial_Correlations_NY_Snyder_2022.png", gndvi_sny_22_p2, width = 8, height = 6, dpi = 400)
```




 
 