---
title: "Running RR models for 2022 NY trials" 
author: "Dinesh Ghimire"
date: "2026-05-15"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr","asreml", "readxl", "tidyr",  "broom", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "lattice","GGally", "tibble", "lme4")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```


#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NY-DG"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NY-DG 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa-NY-DG"
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is misssing please let me know. 
key<-read.csv("2022/meta_data/cultivar_list_key.csv") # contains both CU and NY cultivars here. 
# its same for 2022 and 2023 (cultivar list)
```




#NY 22

#################################### MUSGRAVE ####################################
#################################### MUSGRAVE ####################################
#################################### MUSGRAVE ####################################

Note: With Yield sclaed to last GNDVI in that window and included in RR model



# DG notes:
 #1. stdtime function:
 # Rescale time variable t to [-1, 1], suitable for Legendre polynomials.
 # Create matrix M of raw polynomial terms:
 #   - Column 1: constant 1
 #   - Column 2: standardized time
 #   - Column 3: squared standardized time
 #   - ... up to degree n
  # Note: this is the *raw polynomial basis (not yet orthogonalized)*.

 # 2. Legendre function:
 # Create matrix of Legendre polynomials of order n:
 #   - Uses a recursive formula to build P0, P1, P2, ...
 # here, lambda[i, j] tells you: how much of raw polynomial i goes into Legendre polynomial j.
 #   - Normalizes for numerical stability
 #   - Optionally applies Gengler (1999) scaling for comparability (used in animal breeding). It scales between +1,1 for order 0. 
 FOR MORE DETAILS: Gengler (1999), Estimation of (Co)variance Function Coefficients for Test Day Yield with a Expectation-Maximization Restricted Maximum Likelihood Algorithm1
 
 Eg. 
### With Gengler scaling (x ∈ [−1,1] ), the first two basis functions become:
Order 0: ϕ_0(x)=1
Order 1: ϕ_1(x) = sqrt(3)*x i.e. in our linear case, this becomes sqrt(3) = 1.732 Thus scaled between 1.732 and -1.732
	
| Quantity                              | Range        |
| ------------------------------------- | ------------ |
| **Input variable (x)**                | ([-1,1])     |
| **Legendre basis values (\phi_j(x))** | Can exceed 1 like our case |


# Example usage:
unique_gdd <- sort(unique(nmsu22_nor_pheno_gndvi_only_long$GDD))
M <- stdtime(unique_gdd, 2)    # *raw polynomial terms*
L <- Legendre(2)               # *Legendre transformation matrix*
Phi <- M %*% t(L)              # orthogonal Legendre polynomials at each GDD

 # Longitudinal random regression model
Order0 (baseline growth) =  LP term (intercept) - IN ASREML, leg(GDD,0)
Order1 (growth rate) = the linear Legendre polynomial term (slope) - IN ASREML, leg(GDD, 1)
Order2 (acceleration or deceleration of growth) = the quadratic Legendre polynomial term (curvature)- IN ASREML, leg(GDD, 2)
Order 3 (acceleration or deceleration of growth): the cubic LP term (inflection in growth trajectory): useful in non linear more complex patterns such as growth that accelerates then decelerates and vice versa.


lp_mod_nor_22_h1 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI ~ 1 + at(GDD_fac):Rep, # at(GDD):Rep for nesting Rep within each time points or GDD
  random = ~ str(~ leg(GDD, 1):Germplasm, ~ us(2):vm(Germplasm, grm_inverse)) + 
    str (~ leg(GDD,1):Plot, ~us(2):id(Plot)),
  # leg(GDD,t) meaning t = degree of polynomial, here 1 meaning linear
  residual = ~ dsum(~ units | GDD_fac),    # since we have different harvest data in a column, need nesting       
  data = nmsu22_nor_gdd_harvest1 )}
Here we have:  
# 1. Fixed effect:
Rep effects nested within time:
 - at(GDD_fac):Rep = This fits replication effects separately at each harvest/GDD period. (need GDD as factor here)
 - Rep effects are NOT assumed constant across time. Because different harvests may be in different seasons, weather, irrigation patterns → rep influence changes.

#2. Random Effects: 
## Block A
Genomic random effects: 
   - str(~ leg(GDD, 1):Germplasm,  ~ us(2):vm(Germplasm, grm_inverse))
   - most imp part where it fits genotype growth curves with genomic relationships.
   - leg(GDD,1) = Legendre polynomial of degree 1 → linear growth curve. ( because we saw linear type growth)
   - for each germplasm (genotype), the model estimates: 1. genetic intercept and 2. genetic slope ( rate of VI change per GDD)
   - meaning every genotype has its own genetic growth trajectory.
   - variance structure part of us(2) for intercept and slope ( in linear leg of 1): Allows correlation between genetic level and genetic rate of increase.
   - 

## Block B: 
plot longitudinal effects   
 -   str (~ leg(GDD,1):Plot, ~us(2):id(Plot))
 - purpose: plots have environmental time trends ( soil, irrigation difference)
 - random intercept and slope for each plot
 - variance structure: us(2):id(plot) = each plot has its own independent unstructured 2*2 covaraince and no correlation among plots...
 - this captures non-genetic longitudinal patterns
 
 
## Residual structure:
nested residuals for each GDD or flights:
 - residual = ~ dsum(~ units | GDD_fac)
 - each GDD or harvest has its own residual variance ( here need GDD as factor)
 - This is needed because variance of NDVI changes with growth stage (early = small variance, late = large variance).

# Model output: variance component
- leg(GDD,1):Germplasm!us(2)_1:1   = 5.362902e-04   → Genetic intercept variance
- leg(GDD,1):Germplasm!us(2)_2:1   = -1.543664e-04  → Cov(intercept, slope)
- leg(GDD,1):Germplasm!us(2)_2:2   = 4.837571e-05   → Genetic slope variance
 similarly for plot level output too... 1st is plot intercept variance, plot intercept-slope cov, & plot slope var
# Model output: random effects
 - Positive order1 → steeper response to GDD → faster growth.
 - Negative order1 → slower growth.
 - Positive order2 → upward curve (accelerating growth).
 - Negative order2 → downward curve (plateauing).
Overall, This is how random regression models identify genotype × time patterns.

 
### PEE from manual script based model
Legendre polynomials are the values of the Legendre polynomials at specific time points (or GDDs).

- PEE = Legendre Polynomial Intercept ( which is 1 from Phi matrix) * Plot Level Coefficients for Intercept   + Legendre Polynomial Slope/Linear for harvest date or last imaging event ( which is some number from Phi matrix) * Plot Level Coefficients for slope/Linear

- i.e. PEE = LP intercept × plot intercept coefficient + LP slope × plot slope coefficient [for linear curve]

Random Regression models estimate a covariance matrix of random regression coefficients (K matrix). This matrix must be positive definite (all eigenvalues > 0). Assumes change in covariance pattern across time (opposed to Repeatability model with constant variance)

Repeatability model assumption: same genetic or plot effect at all time points, i.e. constant variance ( cor between time points = 1)




############################  Permanent Environmental Effect ############################      
########   05152026

- Lets use PEE as regressor and compare with no spatial and spatial (autoregressive) model
- now lets use above extracted blups spatial matrix as PEE and fit as a regressor in the model
PEE = intercept + GDD * linear
Eg. 
Phi in harvest 1.
    LPInt      LPLin
62      1 -1.7320508
76      1 -0.7994081
90      1  0.1332347
96      1  0.5329387
104     1  1.0658774
114     1  1.7320508

Then PEE for harvest = 1* Intercept from spatial blups + 1.732 (from 114th GDD) * Linear from spatial blups



#### Running Models




1. *NY Musgrave 2022 *

Loading and preparing data files
```{r loading files 05192026}
#import merged phenotype file 
ny_mus_pheno_22 <- read.csv("2022/MSI_data/output_files/Final_filtered_NY22_Musgrave_phenotypes_MSI.csv") # %>% # new filtered file used
  # select(-X) # remove index column

# need to convert into factors
ny_mus_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_mus_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_mus_pheno_22) 
str(ny_mus_pheno_22)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))

grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


```{r Run this, Adding scaled Yield as input to GNDVI, Mus22, 05192026}
# ============================================================
# STEP 0: ADD SCALED YIELD AS TERMINAL GNDVI PER HARVEST
# ============================================================

# Define mapping: yield → terminal GDD
yld_traits_mus22  <- paste0("sca_yld", 1:3)
gdd_targets_mus22 <- c(157, 202, 264) # add gdd of harvest dates, double check these are greater than or equal to last day of flight and less than next flight in next harvest window. LAST as 264 bcz there is already a flight of 263

# Add new GNDVI columns
for(i in 1:3){
  
  new_col <- paste0("GNDVI_", gdd_targets_mus22[i])
  
  # overwrite / create terminal GNDVI using scaled yield
  ny_mus_pheno_22[[new_col]] <- ny_mus_pheno_22[[yld_traits_mus22[i]]]
  
  cat("Assigned", yld_traits_mus22[i], "→", new_col, "\n")
}


# ============================================================
# STEP 1: KEEP GNDVI + PIVOT LONG
# ============================================================

ny_mus_pheno_22_gndvi_only <- ny_mus_pheno_22 %>%
  dplyr::select(-starts_with("NDVI_")) %>%
  dplyr::select(-starts_with("NDRE_"))

ny_mus_pheno_22_gndvi_only_long <- pivot_longer(
  ny_mus_pheno_22_gndvi_only,
  cols = starts_with("GNDVI_"),
  names_to = "GDD",
  names_prefix = "GNDVI_",
  values_to = "GNDVI")

ny_mus_pheno_22_gndvi_only_long$GDD <- 
  as.numeric(ny_mus_pheno_22_gndvi_only_long$GDD)


# ============================================================
# STEP 2: ASSIGN HARVEST WINDOWS (UNCHANGED)
# ============================================================

ny_mus_pheno_22_gndvi_only_long_harvestwise <- 
  ny_mus_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 157 ~ "Harvest1",
    GDD >= 158 & GDD <= 202 ~ "Harvest2",
    GDD >= 203 & GDD <= 264 ~ "Harvest3")) # 263 flight and 264 last yield cz they are on same date so to differentiate, made this diff.
```


```{r separate out each harvest window and then fit quadratic regression line 05192025}

# NOw lets plot
# First, assign harvest based on GDD
ny_mus_pheno_22_gndvi_only_long_harvestwise <- 
  ny_mus_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 157 ~ "Harvest1",
    GDD >= 158 & GDD <= 202 ~ "Harvest2",
    GDD >= 203 & GDD <= 264 ~ "Harvest3")) # 263 flight and 264 last yield cz they are on same date so to differentiate, made this diff.

# Plot with quadratic fit using geom_smooth
ggplot_quad_nor_22 <- ggplot(ny_mus_pheno_22_gndvi_only_long_harvestwise, aes(x = GDD, y = GNDVI)) + # set up plot with GDD on x and GNDVi on y axis
  geom_point(alpha = 0.5) + # add points, alpha = 0.5 makes semitransparent points to see overlapping points
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), se = FALSE, color = "red") + # adds a quadratic regression line to the points. se = false removes shaded confidence interval around the line. 
  facet_wrap(~Harvest, scales = "free_x") + # separate panels for each harvest
  labs(x = "GDD", y = "GNDVI", title = "Quadratic Fit of GNDVI vs GDD by Harvest in Mus 22") +
  theme_bw()

ggplot_quad_nor_22
```


############################ Harvest1: ############################ 
 - 1. Using Random regression
# combined ( in a single chunk) Harvest 1 comparison across models
- quad
```{r Combined chunk for harvest 1 using Random Regression 05192026}
asreml.options(extra = 200, maxit = 200)
# ---------- PART 0: Manange data file -----------

#  Manage data file: First, assign harvest based on GDD
# first make harvest 2 only data file
ny_mus_pheno_22_gndvi_only_long_harvestwise <- 
  ny_mus_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 157 ~ "Harvest1",
    GDD >= 158 & GDD <= 202 ~ "Harvest2",
    GDD >= 203 & GDD <= 264 ~ "Harvest3")) # 263 flight and 264 last yield cz they are on same date so to differentiate, made this diff.
# Subset harvest 1
ny22_mus_gdd_harvest1 <- ny_mus_pheno_22_gndvi_only_long_harvestwise %>%
  filter(Harvest == "Harvest1") # !!! capital H

# create factor + numeric GDD
ny22_mus_gdd_harvest1$GDD_fac <- as.factor(ny22_mus_gdd_harvest1$GDD)
ny22_mus_gdd_harvest1 <- ny22_mus_gdd_harvest1[order(ny22_mus_gdd_harvest1$GDD_fac),]

unique(ny22_mus_gdd_harvest1$GDD_fac) # 131 145 154 157(for yld1)

# Build Legendre basis
unique_gdd_yld1 <- sort(unique(ny22_mus_gdd_harvest1$GDD))
unique_gdd_yld1 <- as.numeric(unique_gdd_yld1)

 ######## fitting quadratic equation
M_yld1 <- stdtime(unique_gdd_yld1, 2) # for quad, 2
lambda_yld1 <- Legendre(2, gengler=TRUE) # for quad, 2

Phi_yld1 <- M_yld1 %*% t(lambda_yld1)
colnames(Phi_yld1) <- c("LPInt", "LPLin", "LPQuad")
rownames(Phi_yld1) <- unique_gdd_yld1
Phi_yld1
ordLP_yld1 <- match(ny22_mus_gdd_harvest1$GDD, rownames(Phi_yld1))
ny22_mus_gdd_harvest1$LPInt_yld1 <- Phi_yld1[ordLP_yld1, "LPInt"]
ny22_mus_gdd_harvest1$LPLin_yld1 <- Phi_yld1[ordLP_yld1, "LPLin"]
ny22_mus_gdd_harvest1$LPQuad_yld1 <- Phi_yld1[ordLP_yld1, "LPQuad"]
ny22_mus_gdd_harvest1[1, 15:19]

# Fit RR model (harvest 1)
asreml.options(extra = 200, maxit = 200)

lp_mod_mus_22_yld1 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI ~ 1 + at(GDD_fac):Rep , # fitting Rep effect
    random = ~ str(~ Germplasm:LPInt_yld1 + Germplasm:LPLin_yld1 + Germplasm:LPQuad_yld1, ~ us(3):vm(Germplasm, grm_inverse)) + str(~ Plot:LPInt_yld1 + Plot:LPLin_yld1 + Plot:LPQuad_yld1, ~ us(3):id(Plot)), # # for quadratic, us(3)
    residual = ~ dsum(~ units | GDD_fac),
    data = ny22_mus_gdd_harvest1)}

summary(lp_mod_mus_22_yld1)$varcomp
lp_mod_mus_22_yld1 <- update.asreml(lp_mod_mus_22_yld1, start = lp_mod_mus_22_yld1$gammas)

# Save random effects (BLUPs)
blups_yld1 <- lp_mod_mus_22_yld1$coefficients$random
blups_yld1


# ============================================================
# Extract spatial BLUPs from quadratic random regression
# ============================================================

# --- Convert BLUPs to data frame ---
blups_df_yld1 <- as.data.frame(blups_yld1) %>%
  rename(solution = 1)

# Simplify term names
blups_df_yld1$term <- gsub(":.*", "", rownames(blups_df_yld1))


# ============================================================
#  Extract plot-level random regression coefficients
# ============================================================

### Intercept BLUPs
BLUP_spa_int_yld1 <- blups_df_yld1 %>%
  filter(grepl("^Plot_.*LPInt_yld1$", rownames(blups_df_yld1)))

### Linear BLUPs
BLUP_spa_lin_yld1 <- blups_df_yld1 %>%
  filter(grepl("^Plot_.*LPLin_yld1$", rownames(blups_df_yld1)))

### Quadratic BLUPs
BLUP_spa_quad_yld1 <- blups_df_yld1 %>%
  filter(grepl("^Plot_.*LPQuad_yld1$", rownames(blups_df_yld1)))


# ============================================================
# 2. Combine spatial coefficients
# ============================================================

spa_BLUP_matrix_yld1 <- cbind(
  BLUP_spa_int_yld1$solution,
  BLUP_spa_lin_yld1$solution,
  BLUP_spa_quad_yld1$solution)

colnames(spa_BLUP_matrix_yld1) <- c(
  "spa_int_yld1",
  "spa_lin_yld1",
  "spa_quad_yld1")

spa_BLUP_matrix_yld1


# ============================================================
# 3. Convert to data frame
# ============================================================

ny22_mus_spa_yld1 <- as.data.frame(spa_BLUP_matrix_yld1)

# ============================================================
# 4. Add spatial coefficients to phenotype data
# ============================================================

ny_mus_pheno_22$spa_int_yld1  <- ny22_mus_spa_yld1$spa_int_yld1
ny_mus_pheno_22$spa_lin_yld1  <- ny22_mus_spa_yld1$spa_lin_yld1
ny_mus_pheno_22$spa_quad_yld1 <- ny22_mus_spa_yld1$spa_quad_yld1


# ============================================================
# 5. Construct Permanent Environmental Effect (PEE)
# ============================================================

### Extract Legendre coefficients for target GDD/time
Phi_yld1

### Example:
### If final GDD corresponds to:
### LPInt  = 1
### LPLin  = 1.7320508
### LPQuad = 2.2360679
###
### then:

ny_mus_pheno_22$PEE_157 <-
    1 * ny_mus_pheno_22$spa_int_yld1 +
    1.7320508 * ny_mus_pheno_22$spa_lin_yld1 +
    2.2360679 * ny_mus_pheno_22$spa_quad_yld1


# ---------- Step 2: Fit PEE's as fixed regressor on a yield model with spatial correction ----------

# Fit single-trait model for yield (yld1)
# Spatial effects are included as fixed covariates (PEE correction)
uni_ny22_mus_yld1_pee <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed   = yld1 ~ 1 + Row + PEE_157,   # PEE as fixed regressors
    random  = ~ vm(Germplasm, grm_inverse),         # genomic effect
    residual= ~ id(units),
    data    = ny_mus_pheno_22)}


# Extract variance components
summary(uni_ny22_mus_yld1_pee)$varcomp




# 1st model: yield alone genetic model without spatial (autoregressive)
uni_ny22_mus_yld1 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 + Row, #row effect as fixed
         random = ~ vm(Germplasm, grm_inverse),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(uni_ny22_mus_yld1)$varcomp



# 2nd model: yield alone genetic model with spatial (autoregressive)
## Univariate with autoregressive spatial correction
asreml.options(extra = 200, maxit = 200)

uni_ny22_mus_yld1_spa <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 , #row effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row), # only row converged
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(uni_ny22_mus_yld1_spa)$varcomp






# compare

# variance components
summary(uni_ny22_mus_yld1)$varcomp
summary(uni_ny22_mus_yld1_spa)$varcomp
summary(uni_ny22_mus_yld1_pee)$varcomp

# heritability
h2_no_spa  <- asreml::vpredict(uni_ny22_mus_yld1, heritA ~ V1 / (V1 + V2))
h2_no_spa # 0.3363689
h2_AR1     <- asreml::vpredict(uni_ny22_mus_yld1_spa, heritA ~ V3 / (V3 + V4))
h2_AR1 # 0.3364333
# asreml::vpredict(uni_ny22_mus_yld1_spa, heritA ~ V1 / (V1 + V5)) # if both
# asreml::vpredict(uni_ny22_mus_yld1_spa, heritA ~ V3 / (V3 + V4)) # if row only
h2_PEE     <-asreml::vpredict(uni_ny22_mus_yld1_pee, heritA ~ V1 / (V1 + V2))
h2_PEE # 0.44 up from 0.38 compared to no yiedl information in RR model
 # 0.99


# visualize PEE too
 ## just PEE
ggplot(ny_mus_pheno_22, aes(x=Row, y=Range, fill=ny_mus_pheno_22$PEE_157)) + # !!!
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title=" PEE", fill="PEE") +
  theme_minimal()

# Visualize Residual - PEE to see what;s the pattern left. For that need to rescale residuals to PEE and then visualize these... (SAME SCALE NEEDED)

uni_ny22_mus_yld1_res <- residuals(uni_ny22_mus_yld1)
uni_ny22_mus_yld1_res
uni_ny22_mus_yld1_res <- scale(uni_ny22_mus_yld1_res)
PEE_157 <- scale(ny_mus_pheno_22$PEE_157)
plot(PEE_157)
uni_ny22_mus_yld1_res_pee_scaled <- uni_ny22_mus_yld1_res - PEE_157
uni_ny22_mus_yld1_res_pee_scaled
  # ggplot
ggplot(ny_mus_pheno_22, aes(x=Row, y=Range, fill=uni_ny22_mus_yld1_res_pee_scaled)) +
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title="Residuals - PEE", fill="Residual") +
  theme_minimal()





# Get log-likelihood, AIC, BIC
summary(uni_ny22_mus_yld1)$loglik # 
summary(uni_ny22_mus_yld1)$aic
summary(uni_ny22_mus_yld1)$aic

```


### 2. GEER: Function to run for all reps and models 

0.54, 0.60, 0.99
```{r GEER between two blocks (rep12 and rep34) for yld1 (only yield data) mus 2022, 05192026}
### =========================================================
### GEER using row-block subsets
### Musgrave NY 2022
### Harvest 1 (yld1)
###
### Strategy:
###   Since Rep is unavailable as an independent factor,
###   create pseudo-replicates using adjacent row blocks.
###
###   Block 1 = Rows 1 + 2
###   Block 2 = Rows 3 + 4
###
###   Fit models separately within each block:
###     1. Base
###     2. Spatial
###     3. PEE
###
###   Then correlate genomic BLUPs between blocks.
### =========================================================


### ---------------------------------------------------------
### 1. Create row-block subsets
### ---------------------------------------------------------

mus22_yld1_row12 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(1, 2))

mus22_yld1_row34 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(3, 4))


### Check dimensions
dim(mus22_yld1_row12)
dim(mus22_yld1_row34)


### ---------------------------------------------------------
### 2. Ensure GRM inverse
### ---------------------------------------------------------

attr(grm_inverse, "INVERSE") <- TRUE


### =========================================================
### 3. BASE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

base_mus22_yld1_row12 <- asreml(
  fixed    = yld1 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld1_row12)

summary(base_mus22_yld1_row12)$varcomp


### Extract BLUPs
blup_base_mus22_yld1_row12 <- coef(base_mus22_yld1_row12)$random

blup_base_mus22_yld1_row12 <- blup_base_mus22_yld1_row12[
  grep("Germplasm", rownames(blup_base_mus22_yld1_row12)),
  ,
  drop = FALSE]

g_base_mus22_yld1_row12 <- blup_base_mus22_yld1_row12[,1]
names(g_base_mus22_yld1_row12) <- rownames(blup_base_mus22_yld1_row12)



### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

base_mus22_yld1_row34 <- asreml(
  fixed    = yld1 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld1_row34)

summary(base_mus22_yld1_row34)$varcomp


### Extract BLUPs
blup_base_mus22_yld1_row34 <- coef(base_mus22_yld1_row34)$random

blup_base_mus22_yld1_row34 <- blup_base_mus22_yld1_row34[
  grep("Germplasm", rownames(blup_base_mus22_yld1_row34)),
  ,
  drop = FALSE]

g_base_mus22_yld1_row34 <- blup_base_mus22_yld1_row34[,1]
names(g_base_mus22_yld1_row34) <- rownames(blup_base_mus22_yld1_row34)


### ---------------------------------------------------------
### GEER: Base model
### ---------------------------------------------------------

common_base_mus22_yld1 <- intersect(names(g_base_mus22_yld1_row12), names(g_base_mus22_yld1_row34))
common_base_mus22_yld1
cor(g_base_mus22_yld1_row12[common_base_mus22_yld1], g_base_mus22_yld1_row34[common_base_mus22_yld1])




### =========================================================
### 4. SPATIAL MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

spa_mus22_yld1_row12 <- asreml(
  fixed    = yld1 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range),
  residual = ~ id(units),
  data     = mus22_yld1_row12)

summary(spa_mus22_yld1_row12)$varcomp


### Extract BLUPs
blup_spa_mus22_yld1_row12 <- coef(spa_mus22_yld1_row12)$random
blup_spa_mus22_yld1_row12
blup_spa_mus22_yld1_row12 <- blup_spa_mus22_yld1_row12[
  grep("Germplasm", rownames(blup_spa_mus22_yld1_row12)),
  ,
  drop = FALSE
]

g_spa_mus22_yld1_row12 <- blup_spa_mus22_yld1_row12[,1]
names(g_spa_mus22_yld1_row12) <- rownames(blup_spa_mus22_yld1_row12)
g_spa_mus22_yld1_row12


### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

spa_mus22_yld1_row34 <- asreml(
  fixed    = yld1 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range),
  residual = ~ id(units),
  data     = mus22_yld1_row34)

summary(spa_mus22_yld1_row34)$varcomp


### Extract BLUPs
blup_spa_mus22_yld1_row34 <- coef(spa_mus22_yld1_row34)$random

blup_spa_mus22_yld1_row34 <- blup_spa_mus22_yld1_row34[
  grep("Germplasm", rownames(blup_spa_mus22_yld1_row34)),
  ,
  drop = FALSE]

g_spa_mus22_yld1_row34 <- blup_spa_mus22_yld1_row34[,1]
names(g_spa_mus22_yld1_row34) <- rownames(blup_spa_mus22_yld1_row34)



### ---------------------------------------------------------
### GEER: Spatial model
### ---------------------------------------------------------

common_spa_mus22_yld1 <- intersect(names(g_spa_mus22_yld1_row12), names(g_spa_mus22_yld1_row34))

cor(g_spa_mus22_yld1_row12[common_spa_mus22_yld1], g_spa_mus22_yld1_row34[common_spa_mus22_yld1])


### =========================================================
### 5. PEE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

pee_mus22_yld1_row12 <- asreml(
  fixed    = yld1 ~ 1  + Rep +PEE_157,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld1_row12)

summary(pee_mus22_yld1_row12)$varcomp


### Extract BLUPs
blup_pee_mus22_yld1_row12 <- coef(pee_mus22_yld1_row12)$random

blup_pee_mus22_yld1_row12 <- blup_pee_mus22_yld1_row12[
  grep("Germplasm", rownames(blup_pee_mus22_yld1_row12)),
  ,
  drop = FALSE]

g_pee_mus22_yld1_row12 <- blup_pee_mus22_yld1_row12[,1]
names(g_pee_mus22_yld1_row12) <- rownames(blup_pee_mus22_yld1_row12)
g_pee_mus22_yld1_row12


### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

pee_mus22_yld1_row34 <- asreml(
  fixed    = yld1 ~ 1  + Rep + PEE_157,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld1_row34)

summary(pee_mus22_yld1_row34)$varcomp


### Extract BLUPs
blup_pee_mus22_yld1_row34 <- coef(pee_mus22_yld1_row34)$random

blup_pee_mus22_yld1_row34 <- blup_pee_mus22_yld1_row34[
  grep("Germplasm", rownames(blup_pee_mus22_yld1_row34)),
  ,
  drop = FALSE]

g_pee_mus22_yld1_row34 <- blup_pee_mus22_yld1_row34[,1]
names(g_pee_mus22_yld1_row34) <- rownames(blup_pee_mus22_yld1_row34)



### ---------------------------------------------------------
### GEER: PEE model
### ---------------------------------------------------------

common_pee_mus22_yld1 <- intersect(names(g_pee_mus22_yld1_row12), names(g_pee_mus22_yld1_row34))

cor( g_pee_mus22_yld1_row12[common_pee_mus22_yld1], g_pee_mus22_yld1_row34[common_pee_mus22_yld1])




### =========================================================
### 6. Final GEER summary table
### =========================================================

GEER_rowblock_results_mus22_yld1 <- data.frame(

  Model = c("Base", "Spatial", "PEE"),

  Pearson = c(
    cor(g_base_mus22_yld1_row12[common_base_mus22_yld1],
        g_base_mus22_yld1_row34[common_base_mus22_yld1]),

    cor(g_spa_mus22_yld1_row12[common_spa_mus22_yld1],
        g_spa_mus22_yld1_row34[common_spa_mus22_yld1]),

    cor(g_pee_mus22_yld1_row12[common_pee_mus22_yld1],
        g_pee_mus22_yld1_row34[common_pee_mus22_yld1])))

GEER_rowblock_results_mus22_yld1
```

### 3. Validity of GEER Results (block wise results)

updated with addition of base model
0.82&0.83, 0.91&0.85, 0.99 & 0.99

```{r Validation of GEER results using all data base model, mus22 yld1, 05192026}

### =========================================================
### VALIDATION OF GEER RESULTS
### Musgrave NY 2022
### Harvest 1 (yld1)
###
### Goal:
###   Compare genomic BLUPs from:
###     1. Base model using all data
###     2. Spatial model using each block
###     3. PEE model using each block
###
### Validation:
###   Correlate block-specific BLUPs against
###   baseline genomic BLUPs from full dataset
### =========================================================


### ---------------------------------------------------------
### 1. Ensure inverse GRM
### ---------------------------------------------------------

asreml.options(extra = 200, maxit = 200)

attr(grm_inverse, "INVERSE") <- TRUE



### ---------------------------------------------------------
### 2. Create row-block subsets
### ---------------------------------------------------------

mus22_yld1_row12 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(1, 2))

mus22_yld1_row34 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(3, 4))



### =========================================================
### 3. BASE MODEL USING ALL DATA
### =========================================================

mus22_yld1_base_all <- asreml(
  fixed    = yld1 ~ 1,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = ny_mus_pheno_22)

summary(mus22_yld1_base_all)$varcomp


### Extract genomic BLUPs
mus22_yld1_gblup_base_all <- coef(mus22_yld1_base_all)$random

mus22_yld1_gblup_base_all <- mus22_yld1_gblup_base_all[
  grep("Germplasm", rownames(mus22_yld1_gblup_base_all)),
  ,
  drop = FALSE]

mus22_yld1_gblup_base_all <- mus22_yld1_gblup_base_all[,1]

names(mus22_yld1_gblup_base_all) <- 
  rownames(mus22_yld1_gblup_base_all)
mus22_yld1_gblup_base_all

##### part of spatial model and PEE model in above chunk...

### =========================================================
### 4. VALIDATION CORRELATIONS
### =========================================================

### Base correlations
mus22_yld1_cor_base <- c(
  cor(g_base_mus22_yld1_row12, mus22_yld1_gblup_base_all),
  cor(g_base_mus22_yld1_row34, mus22_yld1_gblup_base_all))

mus22_yld1_cor_base


### Spatial correlations (Ar)
mus22_yld1_cor_spa <- c(
  cor(g_spa_mus22_yld1_row12, mus22_yld1_gblup_base_all),
  cor(g_spa_mus22_yld1_row34, mus22_yld1_gblup_base_all))

mus22_yld1_cor_spa



### PEE correlations
mus22_yld1_cor_pee <- c(
  cor(g_pee_mus22_yld1_row12, mus22_yld1_gblup_base_all),
  cor(
    g_pee_mus22_yld1_row34, mus22_yld1_gblup_base_all))
mus22_yld1_cor_pee



### =========================================================
### 5. Summary table
### =========================================================

mus22_yld1_validation_results <- data.frame(

  Model = c("Base_Row12", "Base_Row34", 
            "Spatial_Row12",
            "Spatial_Row34",
            "PEE_Row12",
            "PEE_Row34"),

  Pearson = c(
    mus22_yld1_cor_base[1],
    mus22_yld1_cor_base[2],
    mus22_yld1_cor_spa[1],
    mus22_yld1_cor_spa[2],
    mus22_yld1_cor_pee[1],
    mus22_yld1_cor_pee[2]))

mus22_yld1_validation_results

```








############################ Harvest 2 ############################ 


 - 1. Using Random regression
# combined ( in a single chunk) Harvest 1 comparison across models
 - quad
```{r Combined chunk for harvest2 using Quad Random Regression 05202026}
asreml.options(extra = 200, maxit = 200)
# ---------- PART 0: Manange data file -----------

#  Manage data file: First, assign harvest based on GDD
# first make harvest 2 only data file
ny_mus_pheno_22_gndvi_only_long_harvestwise <- 
  ny_mus_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 157 ~ "Harvest1",
    GDD >= 158 & GDD <= 202 ~ "Harvest2",
    GDD >= 203 & GDD <= 264 ~ "Harvest3")) # 263 flight and 264 last yield cz they are on same date so to differentiate, made this diff.
# Subset harvest 2 !!!!!
ny22_mus_gdd_harvest2 <- ny_mus_pheno_22_gndvi_only_long_harvestwise %>%
  filter(Harvest == "Harvest2") # !!! capital H

# create factor + numeric GDD
ny22_mus_gdd_harvest2$GDD_fac <- as.factor(ny22_mus_gdd_harvest2$GDD)
ny22_mus_gdd_harvest2 <- ny22_mus_gdd_harvest2[order(ny22_mus_gdd_harvest2$GDD_fac),]

unique(ny22_mus_gdd_harvest2$GDD_fac) # 162 164 175 178 181 187 192 201 202(for yld2)

# Build Legendre basis
unique_gdd_yld2 <- sort(unique(ny22_mus_gdd_harvest2$GDD))
unique_gdd_yld2 <- as.numeric(unique_gdd_yld2)

 ######## fitting quadratic equation
M_yld2 <- stdtime(unique_gdd_yld2, 2) # for quad, 2
lambda_yld2 <- Legendre(2, gengler=TRUE) # for quad, 2

Phi_yld2 <- M_yld2 %*% t(lambda_yld2)
colnames(Phi_yld2) <- c("LPInt", "LPLin", "LPQuad")
rownames(Phi_yld2) <- unique_gdd_yld2
Phi_yld2
ordLP_yld2 <- match(ny22_mus_gdd_harvest2$GDD, rownames(Phi_yld2))
ny22_mus_gdd_harvest2$LPInt_yld2 <- Phi_yld2[ordLP_yld2, "LPInt"]
ny22_mus_gdd_harvest2$LPLin_yld2 <- Phi_yld2[ordLP_yld2, "LPLin"]
ny22_mus_gdd_harvest2$LPQuad_yld2 <- Phi_yld2[ordLP_yld2, "LPQuad"]
ny22_mus_gdd_harvest2[1, 15:19]

# Fit RR model (harvest 1)
asreml.options(extra = 200, maxit = 200)

lp_mod_mus_22_yld2 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI ~ 1 + at(GDD_fac):Rep , # fitting Rep effect
    random = ~ str(~ Germplasm:LPInt_yld2 + Germplasm:LPLin_yld2 + Germplasm:LPQuad_yld2, ~ us(3):vm(Germplasm, grm_inverse)) + str(~ Plot:LPInt_yld2 + Plot:LPLin_yld2 + Plot:LPQuad_yld2, ~ us(3):id(Plot)), # # for quadratic, us(3)
    residual = ~ dsum(~ units | GDD_fac),
    data = ny22_mus_gdd_harvest2)}

summary(lp_mod_mus_22_yld2)$varcomp
lp_mod_mus_22_yld2 <- update.asreml(lp_mod_mus_22_yld2, start = lp_mod_mus_22_yld2$gammas)

# Save random effects (BLUPs)
blups_yld2 <- lp_mod_mus_22_yld2$coefficients$random
blups_yld2


# ============================================================
# Extract spatial BLUPs from quadratic random regression
# ============================================================

# --- Convert BLUPs to data frame ---
blups_df_yld2 <- as.data.frame(blups_yld2) %>%
  rename(solution = 1)

# Simplify term names
blups_df_yld2$term <- gsub(":.*", "", rownames(blups_df_yld2))


# ============================================================
#  Extract plot-level random regression coefficients
# ============================================================

### Intercept BLUPs
BLUP_spa_int_yld2 <- blups_df_yld2 %>%
  filter(grepl("^Plot_.*LPInt_yld2$", rownames(blups_df_yld2)))

### Linear BLUPs
BLUP_spa_lin_yld2 <- blups_df_yld2 %>%
  filter(grepl("^Plot_.*LPLin_yld2$", rownames(blups_df_yld2)))

### Quadratic BLUPs
BLUP_spa_quad_yld2 <- blups_df_yld2 %>%
  filter(grepl("^Plot_.*LPQuad_yld2$", rownames(blups_df_yld2)))


# ============================================================
# 2. Combine spatial coefficients
# ============================================================

spa_BLUP_matrix_yld2 <- cbind(
  BLUP_spa_int_yld2$solution,
  BLUP_spa_lin_yld2$solution,
  BLUP_spa_quad_yld2$solution)

colnames(spa_BLUP_matrix_yld2) <- c(
  "spa_int_yld2",
  "spa_lin_yld2",
  "spa_quad_yld2")

spa_BLUP_matrix_yld2


# ============================================================
# 3. Convert to data frame
# ============================================================

ny22_mus_spa_yld2 <- as.data.frame(spa_BLUP_matrix_yld2)

# ============================================================
# 4. Add spatial coefficients to phenotype data
# ============================================================

ny_mus_pheno_22$spa_int_yld2  <- ny22_mus_spa_yld2$spa_int_yld2
ny_mus_pheno_22$spa_lin_yld2  <- ny22_mus_spa_yld2$spa_lin_yld2
ny_mus_pheno_22$spa_quad_yld2 <- ny22_mus_spa_yld2$spa_quad_yld2


# ============================================================
# 5. Construct Permanent Environmental Effect (PEE)
# ============================================================

### Extract Legendre coefficients for target GDD/time
Phi_yld2

### Example:
### If final GDD corresponds to:
### LPInt  = 1
### LPLin  = 1.7320508
### LPQuad = 2.2360679
###
### then:

ny_mus_pheno_22$PEE_202 <-
    1 * ny_mus_pheno_22$spa_int_yld2 +
    1.7320508 * ny_mus_pheno_22$spa_lin_yld2 +
    2.2360679 * ny_mus_pheno_22$spa_quad_yld2


# Inspect
head(ny_mus_pheno_22$PEE_202)


# ---------- Step 2: Fit PEE's as fixed regressor on a yield model with spatial correction ----------

# Fit single-trait model for yield (yld2)
# Spatial effects are included as fixed covariates (PEE correction)
uni_ny22_mus_yld2_pee <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed   = yld2 ~ 1 + Row + PEE_202,   # PEE as fixed regressors
    random  = ~ vm(Germplasm, grm_inverse),         # genomic effect
    residual= ~ id(units),
    data    = ny_mus_pheno_22)}


# Extract variance components
summary(uni_ny22_mus_yld2_pee)$varcomp




# 1st model: yield alone genetic model without spatial (autoregressive)
uni_ny22_mus_yld2 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Row, #row effect as fixed
         random = ~ vm(Germplasm, grm_inverse),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(uni_ny22_mus_yld2)$varcomp



# 2nd model: yield alone genetic model with spatial (autoregressive)
## Univariate with autoregressive spatial correction
asreml.options(extra = 200, maxit = 200)

uni_ny22_mus_yld2_spa <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 ,
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range), 
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(uni_ny22_mus_yld2_spa)$varcomp






# compare

# variance components
summary(uni_ny22_mus_yld2)$varcomp
summary(uni_ny22_mus_yld2_spa)$varcomp
summary(uni_ny22_mus_yld2_pee)$varcomp

# heritability
h2_no_spa  <- asreml::vpredict(uni_ny22_mus_yld2, heritA ~ V1 / (V1 + V2))
h2_no_spa # 0.1412076
h2_AR1     <- asreml::vpredict(uni_ny22_mus_yld2_spa, heritA ~ V1 / (V1 + V5))
h2_AR1 # 0.9020297
# asreml::vpredict(uni_ny22_mus_yld2_spa, heritA ~ V1 / (V1 + V5)) # if both
# asreml::vpredict(uni_ny22_mus_yld2_spa, heritA ~ V3 / (V3 + V4)) # if row only
h2_PEE     <-asreml::vpredict(uni_ny22_mus_yld2_pee, heritA ~ V1 / (V1 + V2))
h2_PEE # 0.10 


# visualize PEE too
 ## just PEE
ggplot(ny_mus_pheno_22, aes(x=Row, y=Range, fill=ny_mus_pheno_22$PEE_202)) + # !!!
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title=" PEE", fill="PEE") +
  theme_minimal()

# Visualize Residual - PEE to see what;s the pattern left. For that need to rescale residuals to PEE and then visualize these... (SAME SCALE NEEDED)

uni_ny22_mus_yld2_res <- residuals(uni_ny22_mus_yld2)
uni_ny22_mus_yld2_res
uni_ny22_mus_yld2_res <- scale(uni_ny22_mus_yld2_res)
PEE_202 <- scale(ny_mus_pheno_22$PEE_202)
plot(PEE_202)
uni_ny22_mus_yld2_res_pee_scaled <- uni_ny22_mus_yld2_res - PEE_202
uni_ny22_mus_yld2_res_pee_scaled
  # ggplot
ggplot(ny_mus_pheno_22, aes(x=Row, y=Range, fill=uni_ny22_mus_yld2_res_pee_scaled)) +
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title="Residuals - PEE", fill="Residual") +
  theme_minimal()





# Get log-likelihood, AIC, BIC
summary(uni_ny22_mus_yld2)$loglik # 
summary(uni_ny22_mus_yld2)$aic
summary(uni_ny22_mus_yld2)$aic
```


```{r plot spatial from AR model, Yld2, Mus22, 05222026}
### =========================================================
### Plot spatial effects surface
### Musgrave NY 2022
### yld2 spatial model
### =========================================================

library(dplyr)
library(ggplot2)


### ---------------------------------------------------------
### 1. Extract spatial effects only
### ---------------------------------------------------------

spa_effects_mus22_yld2 <- 
  uni_ny22_mus_yld2_spa$coefficients$random

spa_effects_mus22_yld2 <- spa_effects_mus22_yld2[
  grep("^Row_", rownames(spa_effects_mus22_yld2)),
  ,
  drop = FALSE]


### ---------------------------------------------------------
### 2. Convert to data frame
### ---------------------------------------------------------

spa_effects_mus22_yld2_df <- data.frame(
  Term   = rownames(spa_effects_mus22_yld2),
  Effect = spa_effects_mus22_yld2[,1])

head(spa_effects_mus22_yld2_df)



### ---------------------------------------------------------
### 3. Extract Row and Range numbers
### ---------------------------------------------------------

spa_effects_mus22_yld2_df$Row <- as.numeric(
  sub("Row_([0-9]+):Range_([0-9]+)", "\\1",
      spa_effects_mus22_yld2_df$Term))

spa_effects_mus22_yld2_df$Range <- as.numeric(
  sub("Row_([0-9]+):Range_([0-9]+)", "\\2",
      spa_effects_mus22_yld2_df$Term))

head(spa_effects_mus22_yld2_df)



### ---------------------------------------------------------
### 4. Plot spatial surface
### ---------------------------------------------------------

ggplot(
  spa_effects_mus22_yld2_df,
  aes(x = Row,
      y = Range,
      fill = Effect)) +

  geom_tile() +

  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0) +

  labs(
    title = "Musgrave2022_Yld2_AR_model_spatial_effects",
    fill = "Spatial\nEffect") +

  theme_minimal()
```

### 2. GEER: Function to run for all reps and models 

```{r GEER between two blocks mus 2022}
# rep 23 vs 14
### =========================================================
### GEER using alternating row blocks
### Musgrave NY 2022
### Harvest2 (yld2)
###
### Block 1 = Rows 2 + 3
### Block 2 = Rows 1 + 4
###
### Goal:
###   Reduce adjacency between pseudo-replicates
### =========================================================


### ---------------------------------------------------------
### 1. Create row-block subsets
### ---------------------------------------------------------

mus22_yld2_row23 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(2, 3))

mus22_yld2_row14 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(1, 4))

dim(mus22_yld2_row23)
dim(mus22_yld2_row14)


### ---------------------------------------------------------
### 2. Ensure GRM inverse
### ---------------------------------------------------------

attr(grm_inverse, "INVERSE") <- TRUE



### =========================================================
### 3. BASE MODEL
### =========================================================

### Block 1: Rows 2 + 3
base_mus22_yld2_row23 <- asreml(
  fixed    = yld2 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld2_row23)

summary(base_mus22_yld2_row23)$varcomp


### Extract genomic BLUPs
blup_base_mus22_yld2_row23 <- coef(base_mus22_yld2_row23)$random

blup_base_mus22_yld2_row23 <- blup_base_mus22_yld2_row23[
  grep("Germplasm", rownames(blup_base_mus22_yld2_row23)),
  ,
  drop = FALSE]

g_base_mus22_yld2_row23 <- blup_base_mus22_yld2_row23[,1]

names(g_base_mus22_yld2_row23) <- 
  rownames(blup_base_mus22_yld2_row23)



### Block 2: Rows 1 + 4
base_mus22_yld2_row14 <- asreml(
  fixed    = yld2 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld2_row14)

summary(base_mus22_yld2_row14)$varcomp


### Extract genomic BLUPs
blup_base_mus22_yld2_row14 <- coef(base_mus22_yld2_row14)$random

blup_base_mus22_yld2_row14 <- blup_base_mus22_yld2_row14[
  grep("Germplasm", rownames(blup_base_mus22_yld2_row14)),
  ,
  drop = FALSE]

g_base_mus22_yld2_row14 <- blup_base_mus22_yld2_row14[,1]

names(g_base_mus22_yld2_row14) <- 
  rownames(blup_base_mus22_yld2_row14)


### GEER: Base
common_base_mus22_yld2 <- intersect(
  names(g_base_mus22_yld2_row23),
  names(g_base_mus22_yld2_row14))

cor(
  g_base_mus22_yld2_row23[common_base_mus22_yld2],
  g_base_mus22_yld2_row14[common_base_mus22_yld2])



### =========================================================
### 4. SPATIAL MODEL
### =========================================================

### Block 1: Rows 2 + 3
spa_mus22_yld2_row23 <- asreml(
  fixed    = yld2 ~ 1 ,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
  residual = ~ id(units),
  data     = mus22_yld2_row23)

summary(spa_mus22_yld2_row23)$varcomp


### Extract genomic BLUPs
blup_spa_mus22_yld2_row23 <- coef(spa_mus22_yld2_row23)$random

blup_spa_mus22_yld2_row23 <- blup_spa_mus22_yld2_row23[
  grep("Germplasm", rownames(blup_spa_mus22_yld2_row23)),
  ,
  drop = FALSE]

g_spa_mus22_yld2_row23 <- blup_spa_mus22_yld2_row23[,1]

names(g_spa_mus22_yld2_row23) <- 
  rownames(blup_spa_mus22_yld2_row23)



### Block 2: Rows 1 + 4
spa_mus22_yld2_row14 <- asreml(
  fixed    = yld2 ~ 1 ,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
  residual = ~ id(units),
  data     = mus22_yld2_row14)

summary(spa_mus22_yld2_row14)$varcomp


### Extract genomic BLUPs
blup_spa_mus22_yld2_row14 <- coef(spa_mus22_yld2_row14)$random

blup_spa_mus22_yld2_row14 <- blup_spa_mus22_yld2_row14[
  grep("Germplasm", rownames(blup_spa_mus22_yld2_row14)),
  ,
  drop = FALSE]

g_spa_mus22_yld2_row14 <- blup_spa_mus22_yld2_row14[,1]

names(g_spa_mus22_yld2_row14) <- 
  rownames(blup_spa_mus22_yld2_row14)


### GEER: Spatial
common_spa_mus22_yld2 <- intersect(
  names(g_spa_mus22_yld2_row23),
  names(g_spa_mus22_yld2_row14))

cor(
  g_spa_mus22_yld2_row23[common_spa_mus22_yld2],
  g_spa_mus22_yld2_row14[common_spa_mus22_yld2])



### =========================================================
### 5. PEE MODEL
### =========================================================

### Block 1: Rows 2 + 3
pee_mus22_yld2_row23 <- asreml(
  fixed    = yld2 ~ 1 + Rep + PEE_202,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld2_row23)

summary(pee_mus22_yld2_row23)$varcomp


### Extract genomic BLUPs
blup_pee_mus22_yld2_row23 <- coef(pee_mus22_yld2_row23)$random

blup_pee_mus22_yld2_row23 <- blup_pee_mus22_yld2_row23[
  grep("Germplasm", rownames(blup_pee_mus22_yld2_row23)),
  ,
  drop = FALSE]

g_pee_mus22_yld2_row23 <- blup_pee_mus22_yld2_row23[,1]

names(g_pee_mus22_yld2_row23) <- 
  rownames(blup_pee_mus22_yld2_row23)



### Block 2: Rows 1 + 4
pee_mus22_yld2_row14 <- asreml(
  fixed    = yld2 ~ 1 + Rep + PEE_202,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld2_row14)

summary(pee_mus22_yld2_row14)$varcomp


### Extract genomic BLUPs
blup_pee_mus22_yld2_row14 <- coef(pee_mus22_yld2_row14)$random

blup_pee_mus22_yld2_row14 <- blup_pee_mus22_yld2_row14[
  grep("Germplasm", rownames(blup_pee_mus22_yld2_row14)),
  ,
  drop = FALSE]

g_pee_mus22_yld2_row14 <- blup_pee_mus22_yld2_row14[,1]

names(g_pee_mus22_yld2_row14) <- 
  rownames(blup_pee_mus22_yld2_row14)


### GEER: PEE
common_pee_mus22_yld2 <- intersect(
  names(g_pee_mus22_yld2_row23),
  names(g_pee_mus22_yld2_row14))

cor(
  g_pee_mus22_yld2_row23[common_pee_mus22_yld2],
  g_pee_mus22_yld2_row14[common_pee_mus22_yld2])



### =========================================================
### 6. Summary table
### =========================================================

GEER_rowblock_results_mus22_yld2_alt <- data.frame(

  Model = c("Base", "Spatial", "PEE"),

  Pearson = c(

    cor(
      g_base_mus22_yld2_row23[common_base_mus22_yld2],
      g_base_mus22_yld2_row14[common_base_mus22_yld2]),

    cor(
      g_spa_mus22_yld2_row23[common_spa_mus22_yld2],
      g_spa_mus22_yld2_row14[common_spa_mus22_yld2]),

    cor(
      g_pee_mus22_yld2_row23[common_pee_mus22_yld2],
      g_pee_mus22_yld2_row14[common_pee_mus22_yld2])))

GEER_rowblock_results_mus22_yld2_alt
```


### 3. Validity of GEER Results (block wise results)

0.76&0.78, 0.72&0.54, 0.81&0.67
```{r rep23 vs 14}

### =========================================================
### VALIDATION OF GEER RESULTS
### Musgrave NY 2022
### Harvest 2 (yld2)
###
### Goal:
###   Compare genomic BLUPs from:
###     1. Base model using all data
###     2. Spatial model using each block
###     3. PEE model using each block
###
### Validation:
###   Correlate block-specific BLUPs against
###   baseline genomic BLUPs from full dataset
###
### Row block strategy:
###   Block 1 = Rows 2 + 3
###   Block 2 = Rows 1 + 4
### =========================================================


### ---------------------------------------------------------
### 1. Ensure inverse GRM
### ---------------------------------------------------------

asreml.options(extra = 200, maxit = 200)

attr(grm_inverse, "INVERSE") <- TRUE



### ---------------------------------------------------------
### 2. Create row-block subsets
### ---------------------------------------------------------

mus22_yld2_row23 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(2, 3))

mus22_yld2_row14 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(1, 4))



### =========================================================
### 3. BASE MODEL USING ALL DATA
### =========================================================

mus22_yld2_base_all <- asreml(
  fixed    = yld2 ~ 1,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = ny_mus_pheno_22)

summary(mus22_yld2_base_all)$varcomp


### Extract genomic BLUPs
mus22_yld2_gblup_base_all <- coef(mus22_yld2_base_all)$random

mus22_yld2_gblup_base_all <- mus22_yld2_gblup_base_all[
  grep("Germplasm", rownames(mus22_yld2_gblup_base_all)),
  ,
  drop = FALSE]

mus22_yld2_gblup_base_all <- mus22_yld2_gblup_base_all[,1]

names(mus22_yld2_gblup_base_all) <- 
  rownames(mus22_yld2_gblup_base_all)

mus22_yld2_gblup_base_all



### =========================================================
### 4. VALIDATION CORRELATIONS
### =========================================================

### base correlations
mus22_yld2_cor_base <- c(

  cor(
    g_base_mus22_yld2_row23,
    mus22_yld2_gblup_base_all),

  cor(
    g_base_mus22_yld2_row14,
    mus22_yld2_gblup_base_all))


### Spatial correlations
mus22_yld2_cor_spa <- c(

  cor(
    g_spa_mus22_yld2_row23,
    mus22_yld2_gblup_base_all),

  cor(
    g_spa_mus22_yld2_row14,
    mus22_yld2_gblup_base_all))

mus22_yld2_cor_spa



### PEE correlations
mus22_yld2_cor_pee <- c(

  cor(
    g_pee_mus22_yld2_row23,
    mus22_yld2_gblup_base_all),

  cor(
    g_pee_mus22_yld2_row14,
    mus22_yld2_gblup_base_all))

mus22_yld2_cor_pee



### =========================================================
### 5. Summary table
### =========================================================

mus22_yld2_validation_results_alt <- data.frame(

  Model = c("Base_Row12", "Base_Row12", 
            "Spatial_Row23",
            "Spatial_Row14",
            "PEE_Row23",
            "PEE_Row14"),

  Pearson = c(
    mus22_yld2_cor_base[1],
    mus22_yld2_cor_base[2],
    mus22_yld2_cor_spa[1],
    mus22_yld2_cor_spa[2],
    mus22_yld2_cor_pee[1],
    mus22_yld2_cor_pee[2]))

mus22_yld2_validation_results_alt

```






############################ Harvest3: ############################ 


### 1. Using Random regression

```{r Combined chunk for harvest3 using Quad Random Regression 05202026}
asreml.options(extra = 200, maxit = 200)
# ---------- PART 0: Manange data file -----------

#  Manage data file: First, assign harvest based on GDD
# first make harvest 2 only data file
ny_mus_pheno_22_gndvi_only_long_harvestwise <- 
  ny_mus_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 157 ~ "Harvest1",
    GDD >= 158 & GDD <= 202 ~ "Harvest2",
    GDD >= 203 & GDD <= 264 ~ "Harvest3")) # 263 flight and 264 last yield cz they are on same date so to differentiate, made this diff.
# Subset harvest 2 !!!!!
ny22_mus_gdd_harvest3 <- ny_mus_pheno_22_gndvi_only_long_harvestwise %>%
  filter(Harvest == "Harvest3") # !!! capital H

# create factor + numeric GDD
ny22_mus_gdd_harvest3$GDD_fac <- as.factor(ny22_mus_gdd_harvest3$GDD)
ny22_mus_gdd_harvest3 <- ny22_mus_gdd_harvest3[order(ny22_mus_gdd_harvest3$GDD_fac),]

unique(ny22_mus_gdd_harvest3$GDD_fac) # 210 215 223 236 238 243 259 263 264(for yld3)

# Build Legendre basis
unique_gdd_yld3 <- sort(unique(ny22_mus_gdd_harvest3$GDD))
unique_gdd_yld3 <- as.numeric(unique_gdd_yld3)

 ######## fitting quadratic equation
M_yld3 <- stdtime(unique_gdd_yld3, 2) # for quad, 2
lambda_yld3 <- Legendre(2, gengler=TRUE) # for quad, 2

Phi_yld3 <- M_yld3 %*% t(lambda_yld3)
colnames(Phi_yld3) <- c("LPInt", "LPLin", "LPQuad")
rownames(Phi_yld3) <- unique_gdd_yld3
Phi_yld3
ordLP_yld3 <- match(ny22_mus_gdd_harvest3$GDD, rownames(Phi_yld3))
ny22_mus_gdd_harvest3$LPInt_yld3 <- Phi_yld3[ordLP_yld3, "LPInt"]
ny22_mus_gdd_harvest3$LPLin_yld3 <- Phi_yld3[ordLP_yld3, "LPLin"]
ny22_mus_gdd_harvest3$LPQuad_yld3 <- Phi_yld3[ordLP_yld3, "LPQuad"]
ny22_mus_gdd_harvest3[1, 15:19]

# Fit RR model (harvest 1)
asreml.options(extra = 200, maxit = 200)

lp_mod_mus_22_yld3 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI ~ 1 + at(GDD_fac):Rep , # fitting Rep effect
    random = ~ str(~ Germplasm:LPInt_yld3 + Germplasm:LPLin_yld3 + Germplasm:LPQuad_yld3, ~ us(3):vm(Germplasm, grm_inverse)) + str(~ Plot:LPInt_yld3 + Plot:LPLin_yld3 + Plot:LPQuad_yld3, ~ us(3):id(Plot)), # # for quadratic, us(3)
    residual = ~ dsum(~ units | GDD_fac),
    data = ny22_mus_gdd_harvest3)}

summary(lp_mod_mus_22_yld3)$varcomp
lp_mod_mus_22_yld3 <- update.asreml(lp_mod_mus_22_yld3, start = lp_mod_mus_22_yld3$gammas)

# Save random effects (BLUPs)
blups_yld3 <- lp_mod_mus_22_yld3$coefficients$random
blups_yld3


# ============================================================
# Extract spatial BLUPs from quadratic random regression
# ============================================================

# --- Convert BLUPs to data frame ---
blups_df_yld3 <- as.data.frame(blups_yld3) %>%
  rename(solution = 1)

# Simplify term names
blups_df_yld3$term <- gsub(":.*", "", rownames(blups_df_yld3))


# ============================================================
#  Extract plot-level random regression coefficients
# ============================================================

### Intercept BLUPs
BLUP_spa_int_yld3 <- blups_df_yld3 %>%
  filter(grepl("^Plot_.*LPInt_yld3$", rownames(blups_df_yld3)))

### Linear BLUPs
BLUP_spa_lin_yld3 <- blups_df_yld3 %>%
  filter(grepl("^Plot_.*LPLin_yld3$", rownames(blups_df_yld3)))

### Quadratic BLUPs
BLUP_spa_quad_yld3 <- blups_df_yld3 %>%
  filter(grepl("^Plot_.*LPQuad_yld3$", rownames(blups_df_yld3)))


# ============================================================
# 2. Combine spatial coefficients
# ============================================================

spa_BLUP_matrix_yld3 <- cbind(
  BLUP_spa_int_yld3$solution,
  BLUP_spa_lin_yld3$solution,
  BLUP_spa_quad_yld3$solution)

colnames(spa_BLUP_matrix_yld3) <- c(
  "spa_int_yld3",
  "spa_lin_yld3",
  "spa_quad_yld3")

spa_BLUP_matrix_yld3


# ============================================================
# 3. Convert to data frame
# ============================================================

ny22_mus_spa_yld3 <- as.data.frame(spa_BLUP_matrix_yld3)

# ============================================================
# 4. Add spatial coefficients to phenotype data
# ============================================================

ny_mus_pheno_22$spa_int_yld3  <- ny22_mus_spa_yld3$spa_int_yld3
ny_mus_pheno_22$spa_lin_yld3  <- ny22_mus_spa_yld3$spa_lin_yld3
ny_mus_pheno_22$spa_quad_yld3 <- ny22_mus_spa_yld3$spa_quad_yld3


# ============================================================
# 5. Construct Permanent Environmental Effect (PEE)
# ============================================================

### Extract Legendre coefficients for target GDD/time
Phi_yld3

### Example:
### If final GDD corresponds to:
### LPInt  = 1
### LPLin  = 1.7320508
### LPQuad = 2.2360679
###
### then:

ny_mus_pheno_22$PEE_264 <-
    1 * ny_mus_pheno_22$spa_int_yld3 +
    1.7320508 * ny_mus_pheno_22$spa_lin_yld3 +
    2.2360679 * ny_mus_pheno_22$spa_quad_yld3


# Inspect
head(ny_mus_pheno_22$PEE_264)


# ---------- Step 2: Fit PEE's as fixed regressor on a yield model with spatial correction ----------

# Fit single-trait model for yield (yld3)
# Spatial effects are included as fixed covariates (PEE correction)
uni_ny22_mus_yld3_pee <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed   = yld3 ~ 1 + Row + PEE_264,   # PEE as fixed regressors
    random  = ~ vm(Germplasm, grm_inverse),         # genomic effect
    residual= ~ id(units),
    data    = ny_mus_pheno_22)}


# Extract variance components
summary(uni_ny22_mus_yld3_pee)$varcomp




# 1st model: yield alone genetic model without spatial (autoregressive)
uni_ny22_mus_yld3 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Row, #row effect as fixed
         random = ~ vm(Germplasm, grm_inverse),
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(uni_ny22_mus_yld3)$varcomp



# 2nd model: yield alone genetic model with spatial (autoregressive)
## Univariate with autoregressive spatial correction
asreml.options(extra = 200, maxit = 200)

uni_ny22_mus_yld3_spa <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 ,
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range), 
         residual = ~ id(units),
         data = ny_mus_pheno_22)}
summary(uni_ny22_mus_yld3_spa)$varcomp






# compare

# variance components
summary(uni_ny22_mus_yld3)$varcomp
summary(uni_ny22_mus_yld3_spa)$varcomp
summary(uni_ny22_mus_yld3_pee)$varcomp

# heritability
h2_no_spa  <- asreml::vpredict(uni_ny22_mus_yld3, heritA ~ V1 / (V1 + V2))
h2_no_spa # 0.1070108
h2_AR1     <- asreml::vpredict(uni_ny22_mus_yld3_spa, heritA ~ V1 / (V1 + V5))
h2_AR1 # 0.99
# asreml::vpredict(uni_ny22_mus_yld3_spa, heritA ~ V1 / (V1 + V5)) # if both
# asreml::vpredict(uni_ny22_mus_yld3_spa, heritA ~ V3 / (V3 + V4)) # if row only
h2_PEE     <-asreml::vpredict(uni_ny22_mus_yld3_pee, heritA ~ V1 / (V1 + V2))
h2_PEE # 0.31 


# visualize PEE too
 ## just PEE
ggplot(ny_mus_pheno_22, aes(x=Row, y=Range, fill=ny_mus_pheno_22$PEE_264)) + # !!!
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title=" PEE", fill="PEE") +
  theme_minimal()

# Visualize Residual - PEE to see what;s the pattern left. For that need to rescale residuals to PEE and then visualize these... (SAME SCALE NEEDED)

uni_ny22_mus_yld3_res <- residuals(uni_ny22_mus_yld3)
uni_ny22_mus_yld3_res
uni_ny22_mus_yld3_res <- scale(uni_ny22_mus_yld3_res)
PEE_264 <- scale(ny_mus_pheno_22$PEE_264)
plot(PEE_264)
uni_ny22_mus_yld3_res_pee_scaled <- uni_ny22_mus_yld3_res - PEE_264
uni_ny22_mus_yld3_res_pee_scaled
  # ggplot
ggplot(ny_mus_pheno_22, aes(x=Row, y=Range, fill=uni_ny22_mus_yld3_res_pee_scaled)) +
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title="Residuals - PEE", fill="Residual") +
  theme_minimal()





# Get log-likelihood, AIC, BIC
summary(uni_ny22_mus_yld3)$loglik # 
summary(uni_ny22_mus_yld3)$aic
summary(uni_ny22_mus_yld3)$aic
```

### 2. GEER: Function to run for all reps and models 

0.36, 0.34, 0.54
```{r GEER between two blocks (rep12 and rep34) for yld3 (only yield data) mus 2022, 05202026}
### =========================================================
### GEER using row-block subsets
### Musgrave NY 2022
### harvest3 (yld3)
###
### Strategy:
###   Since Rep is unavailable as an independent factor,
###   create pseudo-replicates using adjacent row blocks.
###
###   Block 1 = Rows 1 + 2
###   Block 2 = Rows 3 + 4
###
###   Fit models separately within each block:
###     1. Base
###     2. Spatial
###     3. PEE
###
###   Then correlate genomic BLUPs between blocks.
### =========================================================


### ---------------------------------------------------------
### 1. Create row-block subsets
### ---------------------------------------------------------

mus22_yld3_row12 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(1, 2))

mus22_yld3_row34 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(3, 4))


### Check dimensions
dim(mus22_yld3_row12)
dim(mus22_yld3_row34)


### ---------------------------------------------------------
### 2. Ensure GRM inverse
### ---------------------------------------------------------

attr(grm_inverse, "INVERSE") <- TRUE


### =========================================================
### 3. BASE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

base_mus22_yld3_row12 <- asreml(
  fixed    = yld3 ~ 1 ,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld3_row12)

summary(base_mus22_yld3_row12)$varcomp


### Extract BLUPs
blup_base_mus22_yld3_row12 <- coef(base_mus22_yld3_row12)$random

blup_base_mus22_yld3_row12 <- blup_base_mus22_yld3_row12[
  grep("Germplasm", rownames(blup_base_mus22_yld3_row12)),
  ,
  drop = FALSE]

g_base_mus22_yld3_row12 <- blup_base_mus22_yld3_row12[,1]
names(g_base_mus22_yld3_row12) <- rownames(blup_base_mus22_yld3_row12)



### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

base_mus22_yld3_row34 <- asreml(
  fixed    = yld3 ~ 1  ,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld3_row34)

summary(base_mus22_yld3_row34)$varcomp


### Extract BLUPs
blup_base_mus22_yld3_row34 <- coef(base_mus22_yld3_row34)$random

blup_base_mus22_yld3_row34 <- blup_base_mus22_yld3_row34[
  grep("Germplasm", rownames(blup_base_mus22_yld3_row34)),
  ,
  drop = FALSE]

g_base_mus22_yld3_row34 <- blup_base_mus22_yld3_row34[,1]
names(g_base_mus22_yld3_row34) <- rownames(blup_base_mus22_yld3_row34)


### ---------------------------------------------------------
### GEER: Base model
### ---------------------------------------------------------

common_base_mus22_yld3 <- intersect(names(g_base_mus22_yld3_row12), names(g_base_mus22_yld3_row34))
common_base_mus22_yld3
cor(g_base_mus22_yld3_row12[common_base_mus22_yld3], g_base_mus22_yld3_row34[common_base_mus22_yld3])




### =========================================================
### 4. SPATIAL MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

spa_mus22_yld3_row12 <- asreml(
  fixed    = yld3 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range), # usually doesn't converge two dimenional autoregressive
  residual = ~ id(units),
  data     = mus22_yld3_row12)

summary(spa_mus22_yld3_row12)$varcomp


### Extract BLUPs
blup_spa_mus22_yld3_row12 <- coef(spa_mus22_yld3_row12)$random
blup_spa_mus22_yld3_row12
blup_spa_mus22_yld3_row12 <- blup_spa_mus22_yld3_row12[
  grep("Germplasm", rownames(blup_spa_mus22_yld3_row12)),
  ,
  drop = FALSE
]

g_spa_mus22_yld3_row12 <- blup_spa_mus22_yld3_row12[,1]
names(g_spa_mus22_yld3_row12) <- rownames(blup_spa_mus22_yld3_row12)
g_spa_mus22_yld3_row12


### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

spa_mus22_yld3_row34 <- asreml(
  fixed    = yld3 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range),
  residual = ~ id(units),
  data     = mus22_yld3_row34)

summary(spa_mus22_yld3_row34)$varcomp


### Extract BLUPs
blup_spa_mus22_yld3_row34 <- coef(spa_mus22_yld3_row34)$random

blup_spa_mus22_yld3_row34 <- blup_spa_mus22_yld3_row34[
  grep("Germplasm", rownames(blup_spa_mus22_yld3_row34)),
  ,
  drop = FALSE]

g_spa_mus22_yld3_row34 <- blup_spa_mus22_yld3_row34[,1]
names(g_spa_mus22_yld3_row34) <- rownames(blup_spa_mus22_yld3_row34)



### ---------------------------------------------------------
### GEER: Spatial model
### ---------------------------------------------------------

common_spa_mus22_yld3 <- intersect(names(g_spa_mus22_yld3_row12), names(g_spa_mus22_yld3_row34))

cor(g_spa_mus22_yld3_row12[common_spa_mus22_yld3], g_spa_mus22_yld3_row34[common_spa_mus22_yld3])


### =========================================================
### 5. PEE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

pee_mus22_yld3_row12 <- asreml(
  fixed    = yld3 ~ 1 + Rep + PEE_264,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld3_row12)

summary(pee_mus22_yld3_row12)$varcomp


### Extract BLUPs
blup_pee_mus22_yld3_row12 <- coef(pee_mus22_yld3_row12)$random

blup_pee_mus22_yld3_row12 <- blup_pee_mus22_yld3_row12[
  grep("Germplasm", rownames(blup_pee_mus22_yld3_row12)),
  ,
  drop = FALSE]

g_pee_mus22_yld3_row12 <- blup_pee_mus22_yld3_row12[,1]
names(g_pee_mus22_yld3_row12) <- rownames(blup_pee_mus22_yld3_row12)



### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

pee_mus22_yld3_row34 <- asreml(
  fixed    = yld3 ~ 1 + Rep + PEE_264,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = mus22_yld3_row34)

summary(pee_mus22_yld3_row34)$varcomp


### Extract BLUPs
blup_pee_mus22_yld3_row34 <- coef(pee_mus22_yld3_row34)$random

blup_pee_mus22_yld3_row34 <- blup_pee_mus22_yld3_row34[
  grep("Germplasm", rownames(blup_pee_mus22_yld3_row34)),
  ,
  drop = FALSE]

g_pee_mus22_yld3_row34 <- blup_pee_mus22_yld3_row34[,1]
names(g_pee_mus22_yld3_row34) <- rownames(blup_pee_mus22_yld3_row34)
blup_pee_mus22_yld3_row34


### ---------------------------------------------------------
### GEER: PEE model
### ---------------------------------------------------------

common_pee_mus22_yld3 <- intersect(names(g_pee_mus22_yld3_row12), names(g_pee_mus22_yld3_row34))
common_pee_mus22_yld3
cor( g_pee_mus22_yld3_row12[common_pee_mus22_yld3], g_pee_mus22_yld3_row34[common_pee_mus22_yld3])
# cor(blup_pee_mus22_yld3_row34,  blup_pee_mus22_yld3_row12)


### =========================================================
### 6. Final GEER summary table
### =========================================================

GEER_rowblock_results_mus22_yld3 <- data.frame(

  Model = c("Base", "Spatial", "PEE"),

  Pearson = c(
    cor(g_base_mus22_yld3_row12[common_base_mus22_yld3],
        g_base_mus22_yld3_row34[common_base_mus22_yld3]),

    cor(g_spa_mus22_yld3_row12[common_spa_mus22_yld3],
        g_spa_mus22_yld3_row34[common_spa_mus22_yld3]),

    cor(g_pee_mus22_yld3_row12[common_pee_mus22_yld3],
        g_pee_mus22_yld3_row34[common_pee_mus22_yld3])))

GEER_rowblock_results_mus22_yld3
```

### 3. Validity of GEER Results (block wise results)

updated with addition of base model
0.78&0.75, 0.76 and 0.69, 0.72 and 0.77
```{r Validation of GEER results using all data base model, mus22 yld3, 06072026}

### =========================================================
### VALIDATION OF GEER RESULTS
### Musgrave NY 2022
### Harvest 1 (yld3)
###
### Goal:
###   Compare genomic BLUPs from:
###     1. Base model using all data
###     2. Spatial model using each block
###     3. PEE model using each block
###
### Validation:
###   Correlate block-specific BLUPs against
###   baseline genomic BLUPs from full dataset
### =========================================================


### ---------------------------------------------------------
### 1. Ensure inverse GRM
### ---------------------------------------------------------

asreml.options(extra = 200, maxit = 200)

attr(grm_inverse, "INVERSE") <- TRUE



### ---------------------------------------------------------
### 2. Create row-block subsets
### ---------------------------------------------------------

mus22_yld3_row12 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(1, 2))

mus22_yld3_row34 <- ny_mus_pheno_22 %>%
  filter(Row %in% c(3, 4))



### =========================================================
### 3. BASE MODEL USING ALL DATA
### =========================================================

mus22_yld3_base_all <- asreml(
  fixed    = yld3 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = ny_mus_pheno_22)

summary(mus22_yld3_base_all)$varcomp


### Extract genomic BLUPs
mus22_yld3_gblup_base_all <- coef(mus22_yld3_base_all)$random

mus22_yld3_gblup_base_all <- mus22_yld3_gblup_base_all[
  grep("Germplasm", rownames(mus22_yld3_gblup_base_all)),
  ,
  drop = FALSE]

mus22_yld3_gblup_base_all <- mus22_yld3_gblup_base_all[,1]

names(mus22_yld3_gblup_base_all) <- 
  rownames(mus22_yld3_gblup_base_all)
mus22_yld3_gblup_base_all

##### part of spatial model and PEE model in above chunk...

### =========================================================
### 4. VALIDATION CORRELATIONS
### =========================================================

### Base correlations
mus22_yld3_cor_base <- c(
  cor(g_base_mus22_yld3_row12, mus22_yld3_gblup_base_all),
  cor(g_base_mus22_yld3_row34, mus22_yld3_gblup_base_all))

mus22_yld3_cor_base

### Spatial correlations (Ar)
mus22_yld3_cor_spa <- c(
  cor(g_spa_mus22_yld3_row12, mus22_yld3_gblup_base_all),
  cor(g_spa_mus22_yld3_row34, mus22_yld3_gblup_base_all))

mus22_yld3_cor_spa



### PEE correlations
mus22_yld3_cor_pee <- c(
  cor(g_pee_mus22_yld3_row12, mus22_yld3_gblup_base_all),
  cor(
    g_pee_mus22_yld3_row34, mus22_yld3_gblup_base_all))
mus22_yld3_cor_pee



### =========================================================
### 5. Summary table
### =========================================================

mus22_yld3_validation_results <- data.frame(

  Model = c("Base_Row12", "Base_Row12", 
            "Spatial_Row12",
            "Spatial_Row34",
            "PEE_Row12",
            "PEE_Row34"),

  Pearson = c(
    mus22_yld3_cor_base[1],
    mus22_yld3_cor_base[2],
    mus22_yld3_cor_spa[1],
    mus22_yld3_cor_spa[2],
    mus22_yld3_cor_pee[1],
    mus22_yld3_cor_pee[2]))

mus22_yld3_validation_results

```






### Plots

# Metrics as Baseline deviation (for cor, GEER, validation of GEER:
## 1. Metrics as Baseline deviation (ΔG):

### Why deviation metric: 
 - Absolute GEER values are not directly comparable across harvests because each harvest has a different baseline level of predictability. 
 - Expressing model performance as deviation from the baseline (ΔG) removes this baseline effect and isolates the marginal contribution of the spatial model.  
 - This allows results from multiple harvests to be combined and compared on a common scale.

### What is deviaiton? 
Baseline = Base model (no spatial correction)
Models = Spa (AR1), PEE (RR)

So for each replicate-pair GEER:
 1. ΔG_Spa = GEER_Spa − GEER_Base
 2. ΔG_PEE = GEER_PEE − GEER_Base

### What it tells?
 - how much does each model improve over baseline?
 - Isolates gain from spatial correction like AR or RR(PEE). 

GEER as deviation, ΔGEER for all harvests


```{r GEER as deviation for all harvests Mus22, 06072026}

### =========================================================
### Boxplot of GEER deviation across harvests
### Musgrave NY 2022
### =========================================================

### NOTE:
### Deviation calculated as:
###
###   Model GEER - Base GEER
###
### Positive values indicate improvement over base model.
### =========================================================



### ---------------------------------------------------------
### 1. Create data frame
### ---------------------------------------------------------

GEER_deviation_mus22_allharvests <- data.frame(
  Harvest = c("Harvest1", "Harvest2","Harvest3"),
  Spatial = c(0.06, 0.14, -0.02),
  PEE = c( 0.45, 0.17, 0.18))

GEER_deviation_mus22_allharvests


### ---------------------------------------------------------
### 2. Convert to long format
### ---------------------------------------------------------

GEER_deviation_mus22_allharvests_long <- 
  GEER_deviation_mus22_allharvests %>%
  
  pivot_longer(
    cols = c(Spatial, PEE),
    names_to = "Model",
    values_to = "Deviation")

GEER_deviation_mus22_allharvests_long



### ---------------------------------------------------------
### 3. Set order of boxplots
### ---------------------------------------------------------

GEER_deviation_mus22_allharvests_long$Model <- factor(
  GEER_deviation_mus22_allharvests_long$Model,
  levels = c("Spatial", "PEE"))



### ---------------------------------------------------------
### 4. Boxplot
### ---------------------------------------------------------

gg_GEER_mus22 <- ggplot(GEER_deviation_mus22_allharvests_long, aes(x = Model, y = Deviation, fill = Model)) +
  geom_boxplot(width = 0.3, alpha = 0.9,outlier.shape = NA) +
  scale_fill_manual(values = c(
    "Base" = "gray",
    "Spatial" = "lightblue",
    "PEE"     = "lightgreen")) +
  labs(
    title = "MUS22 All Harvests: GEER Deviation from Base Model",
    x = "Model",
    y = "GEER Deviation") +
  theme_bw(base_size = 13) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2)) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5),
    legend.position = "right")
gg_GEER_mus22

ggsave(filename = "GEER_boxplots_Musgrave_2022_NMSU.png", plot = gg_GEER_mus22, path = "2022/output_images", width = 10, height = 10,dpi = 400) 
    
    
```























....



################################# Snyder Trial ################################
################################# Snyder Trial ################################
################################# Snyder Trial ################################





Loading and preparing data files
```{r loading files 05192026}
#import merged phenotype file 
ny_sny_pheno_22 <- read.csv("2022/MSI_data/output_files/Final_filtered_NY22_Snyder_phenotypes_MSI.csv") # %>% # new filtered file used
  # select(-X) # remove index column

# need to convert into factors
ny_sny_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_sny_pheno_22[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_sny_pheno_22) 
str(ny_sny_pheno_22)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))

grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


```{r Run this, Adding scaled Yield as input to GNDVI, sny22, 05192026}
# ============================================================
# STEP 0: ADD SCALED YIELD AS TERMINAL GNDVI PER HARVEST
# ============================================================

# Define mapping: yield → terminal GDD
yld_traits_sny22  <- paste0("sca_yld", 1:3)
gdd_targets_sny22 <- c(165, 208, 260) # add gdd of harvest dates, double check these are greater than or equal to last day of flight and less than next flight in next harvest window. LAST as 260 bcz there is already a flight of 259

# Add new GNDVI columns
for(i in 1:3){
  
  new_col <- paste0("GNDVI_", gdd_targets_sny22[i])
  
  # overwrite / create terminal GNDVI using scaled yield
  ny_sny_pheno_22[[new_col]] <- ny_sny_pheno_22[[yld_traits_sny22[i]]]
  
  cat("Assigned", yld_traits_sny22[i], "→", new_col, "\n")
}


# ============================================================
# STEP 1: KEEP GNDVI + PIVOT LONG
# ============================================================

ny_sny_pheno_22_gndvi_only <- ny_sny_pheno_22 %>%
  dplyr::select(-starts_with("NDVI_")) %>%
  dplyr::select(-starts_with("NDRE_"))

ny_sny_pheno_22_gndvi_only_long <- pivot_longer(
  ny_sny_pheno_22_gndvi_only,
  cols = starts_with("GNDVI_"),
  names_to = "GDD",
  names_prefix = "GNDVI_",
  values_to = "GNDVI")

ny_sny_pheno_22_gndvi_only_long$GDD <- 
  as.numeric(ny_sny_pheno_22_gndvi_only_long$GDD)


# ============================================================
# STEP 2: ASSIGN HARVEST WINDOWS (UNCHANGED)
# ============================================================

ny_sny_pheno_22_gndvi_only_long_harvestwise <- 
  ny_sny_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 165 ~ "Harvest1",
    GDD >= 166 & GDD <= 208 ~ "Harvest2",
    GDD >= 209 & GDD <= 260 ~ "Harvest3")) # 259 flight and 260 last yield (actually 259) cz they are on same date so to differentiate, made this diff.
# Subset harvest1
```


```{r separate out each harvest window and then fit quadratic regression line 05192025}

# NOw lets plot
# First, assign harvest based on GDD
ny_sny_pheno_22_gndvi_only_long_harvestwise <- 
  ny_sny_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 165 ~ "Harvest1",
    GDD >= 166 & GDD <= 208 ~ "Harvest2",
    GDD >= 209 & GDD <= 260 ~ "Harvest3")) # 259 flight and 260 last yield (actually 259) cz they are on same date so to differentiate, made this diff.
# Subset harvest 1
# Plot with quadratic fit using geom_smooth
ggplot_quad_sny_22 <- ggplot(ny_sny_pheno_22_gndvi_only_long_harvestwise, aes(x = GDD, y = GNDVI)) + # set up plot with GDD on x and GNDVi on y axis
  geom_point(alpha = 0.5) + # add points, alpha = 0.5 makes semitransparent points to see overlapping points
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), se = FALSE, color = "red") + # adds a quadratic regression line to the points. se = false removes shaded confidence interval around the line. 
  facet_wrap(~Harvest, scales = "free_x") + # separate panels for each harvest
  labs(x = "GDD", y = "GNDVI", title = "Quadratic Fit of GNDVI vs GDD by Harvest in Sny 22") +
  theme_bw()

ggplot_quad_sny_22
```


############################ Harvest1: ############################ 

### 1. Using Random regression
```{r Combined chunk for harvest 1 using Random Regression 05192026}
asreml.options(extra = 200, maxit = 200)
# ---------- PART 0: Manange data file -----------

#  Manage data file: First, assign harvest based on GDD
# first make harvest 2 only data file
ny_sny_pheno_22_gndvi_only_long_harvestwise <- 
  ny_sny_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 165 ~ "Harvest1",
    GDD >= 166 & GDD <= 208 ~ "Harvest2",
    GDD >= 209 & GDD <= 260 ~ "Harvest3")) # 259 flight and 260 last yield (actually 259) cz they are on same date so to differentiate, made this diff.
# Subset harvest 1
ny22_sny_gdd_harvest1 <- ny_sny_pheno_22_gndvi_only_long_harvestwise %>%
  filter(Harvest == "Harvest1") # !!! capital H

# create factor + numeric GDD
ny22_sny_gdd_harvest1$GDD_fac <- as.factor(ny22_sny_gdd_harvest1$GDD)
ny22_sny_gdd_harvest1 <- ny22_sny_gdd_harvest1[order(ny22_sny_gdd_harvest1$GDD_fac),]

unique(ny22_sny_gdd_harvest1$GDD_fac) # 133 159 164 165(for yld1)

# Build Legendre basis
unique_gdd_yld1 <- sort(unique(ny22_sny_gdd_harvest1$GDD))
unique_gdd_yld1 <- as.numeric(unique_gdd_yld1)

 ######## fitting quadratic equation
M_yld1 <- stdtime(unique_gdd_yld1, 2) # for quad, 2
lambda_yld1 <- Legendre(2, gengler=TRUE) # for quad, 2

Phi_yld1 <- M_yld1 %*% t(lambda_yld1)
colnames(Phi_yld1) <- c("LPInt", "LPLin", "LPQuad")
rownames(Phi_yld1) <- unique_gdd_yld1
Phi_yld1
ordLP_yld1 <- match(ny22_sny_gdd_harvest1$GDD, rownames(Phi_yld1))
ny22_sny_gdd_harvest1$LPInt_yld1 <- Phi_yld1[ordLP_yld1, "LPInt"]
ny22_sny_gdd_harvest1$LPLin_yld1 <- Phi_yld1[ordLP_yld1, "LPLin"]
ny22_sny_gdd_harvest1$LPQuad_yld1 <- Phi_yld1[ordLP_yld1, "LPQuad"]
ny22_sny_gdd_harvest1[1, 15:19]

# Fit RR model (harvest 1)
asreml.options(extra = 200, maxit = 200)

lp_mod_sny_22_yld1 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI ~ 1 + at(GDD_fac):Rep , # fitting Rep effect
    random = ~ str(~ Germplasm:LPInt_yld1 + Germplasm:LPLin_yld1 + Germplasm:LPQuad_yld1, ~ us(3):vm(Germplasm, grm_inverse)) + str(~ Plot:LPInt_yld1 + Plot:LPLin_yld1 + Plot:LPQuad_yld1, ~ us(3):id(Plot)), # # for quadratic, us(3)
    residual = ~ dsum(~ units | GDD_fac),
    data = ny22_sny_gdd_harvest1)}

summary(lp_mod_sny_22_yld1)$varcomp
lp_mod_sny_22_yld1 <- update.asreml(lp_mod_sny_22_yld1, start = lp_mod_sny_22_yld1$gammas)

# Save random effects (BLUPs)
blups_yld1 <- lp_mod_sny_22_yld1$coefficients$random
blups_yld1


# ============================================================
# Extract spatial BLUPs from quadratic random regression
# ============================================================

# --- Convert BLUPs to data frame ---
blups_df_yld1 <- as.data.frame(blups_yld1) %>%
  rename(solution = 1)

# Simplify term names
blups_df_yld1$term <- gsub(":.*", "", rownames(blups_df_yld1))


# ============================================================
#  Extract plot-level random regression coefficients
# ============================================================

### Intercept BLUPs
BLUP_spa_int_yld1 <- blups_df_yld1 %>%
  filter(grepl("^Plot_.*LPInt_yld1$", rownames(blups_df_yld1)))

### Linear BLUPs
BLUP_spa_lin_yld1 <- blups_df_yld1 %>%
  filter(grepl("^Plot_.*LPLin_yld1$", rownames(blups_df_yld1)))

### Quadratic BLUPs
BLUP_spa_quad_yld1 <- blups_df_yld1 %>%
  filter(grepl("^Plot_.*LPQuad_yld1$", rownames(blups_df_yld1)))


# ============================================================
# 2. Combine spatial coefficients
# ============================================================

spa_BLUP_matrix_yld1 <- cbind(
  BLUP_spa_int_yld1$solution,
  BLUP_spa_lin_yld1$solution,
  BLUP_spa_quad_yld1$solution)

colnames(spa_BLUP_matrix_yld1) <- c(
  "spa_int_yld1",
  "spa_lin_yld1",
  "spa_quad_yld1")

spa_BLUP_matrix_yld1


# ============================================================
# 3. Convert to data frame
# ============================================================

ny22_sny_spa_yld1 <- as.data.frame(spa_BLUP_matrix_yld1)

# ============================================================
# 4. Add spatial coefficients to phenotype data
# ============================================================

ny_sny_pheno_22$spa_int_yld1  <- ny22_sny_spa_yld1$spa_int_yld1
ny_sny_pheno_22$spa_lin_yld1  <- ny22_sny_spa_yld1$spa_lin_yld1
ny_sny_pheno_22$spa_quad_yld1 <- ny22_sny_spa_yld1$spa_quad_yld1


# ============================================================
# 5. Construct Permanent Environmental Effect (PEE)
# ============================================================

### Extract Legendre coefficients for target GDD/time
Phi_yld1

### Example:
### If final GDD corresponds to:
### LPInt  = 1
### LPLin  = 1.7320508
### LPQuad = 2.2360679
###
### then:

ny_sny_pheno_22$PEE_165 <-
    1 * ny_sny_pheno_22$spa_int_yld1 +
    1.7320508 * ny_sny_pheno_22$spa_lin_yld1 +
    2.2360679 * ny_sny_pheno_22$spa_quad_yld1


# ---------- Step 2: Fit PEE's as fixed regressor on a yield model with spatial correction ----------

# Fit single-trait model for yield (yld1)
# Spatial effects are included as fixed covariates (PEE correction)
uni_ny22_sny_yld1_pee <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed   = yld1 ~ 1 + Row + PEE_165,   # PEE as fixed regressors
    random  = ~ vm(Germplasm, grm_inverse),         # genomic effect
    residual= ~ id(units),
    data    = ny_sny_pheno_22)}


# Extract variance components
summary(uni_ny22_sny_yld1_pee)$varcomp




# 1st model: yield alone genetic model without spatial (autoregressive)
uni_ny22_sny_yld1 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 + Row, #row effect as fixed
         random = ~ vm(Germplasm, grm_inverse),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(uni_ny22_sny_yld1)$varcomp



# 2nd model: yield alone genetic model with spatial (autoregressive)
## Univariate with autoregressive spatial correction
asreml.options(extra = 200, maxit = 200)

uni_ny22_sny_yld1_spa <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 , #row effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range), # only row converged
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(uni_ny22_sny_yld1_spa)$varcomp






# compare

# variance components
summary(uni_ny22_sny_yld1)$varcomp
summary(uni_ny22_sny_yld1_spa)$varcomp
summary(uni_ny22_sny_yld1_pee)$varcomp

# heritability
h2_no_spa  <- asreml::vpredict(uni_ny22_sny_yld1, heritA ~ V1 / (V1 + V2))
h2_no_spa # 0.4931953
h2_AR1     <- asreml::vpredict(uni_ny22_sny_yld1_spa, heritA ~ V1 / (V1 + V5))
h2_AR1 # 0.6928387
# asreml::vpredict(uni_ny22_sny_yld1_spa, heritA ~ V1 / (V1 + V5)) # if both
# asreml::vpredict(uni_ny22_sny_yld1_spa, heritA ~ V3 / (V3 + V4)) # if row only
h2_PEE     <-asreml::vpredict(uni_ny22_sny_yld1_pee, heritA ~ V1 / (V1 + V2))
h2_PEE # 0.50 


# visualize PEE too
 ## just PEE
ggplot(ny_sny_pheno_22, aes(x=Row, y=Range, fill=ny_sny_pheno_22$PEE_165)) + # !!!
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title=" PEE", fill="PEE") +
  theme_minimal()

# Visualize Residual - PEE to see what;s the pattern left. For that need to rescale residuals to PEE and then visualize these... (SAME SCALE NEEDED)

uni_ny22_sny_yld1_res <- residuals(uni_ny22_sny_yld1)
uni_ny22_sny_yld1_res
uni_ny22_sny_yld1_res <- scale(uni_ny22_sny_yld1_res)
PEE_165 <- scale(ny_sny_pheno_22$PEE_165)
plot(PEE_165)
uni_ny22_sny_yld1_res_pee_scaled <- uni_ny22_sny_yld1_res - PEE_165
uni_ny22_sny_yld1_res_pee_scaled
  # ggplot
ggplot(ny_sny_pheno_22, aes(x=Row, y=Range, fill=uni_ny22_sny_yld1_res_pee_scaled)) +
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title="Residuals - PEE", fill="Residual") +
  theme_minimal()





# Get log-likelihood, AIC, BIC
summary(uni_ny22_sny_yld1)$loglik # 
summary(uni_ny22_sny_yld1)$aic
summary(uni_ny22_sny_yld1)$aic

```


### 2. GEER: Function to run for all reps and models 
0.70, 0.72, 0.73
```{r GEER between two blocks (rep12 and rep34) for yld1 (only yield data) sny 2022, 05192026}
### =========================================================
### GEER using row-block subsets
### Snyder NY 2022
### Harvest 1 (yld1)
###
### Strategy:
###   Since Rep is unavailable as an independent factor,
###   create pseudo-replicates using adjacent row blocks.
###
###   Block 1 = Rows 1 + 2
###   Block 2 = Rows 3 + 4
###
###   Fit models separately within each block:
###     1. Base
###     2. Spatial
###     3. PEE
###
###   Then correlate genomic BLUPs between blocks.
### =========================================================


### ---------------------------------------------------------
### 1. Create row-block subsets
### ---------------------------------------------------------

sny22_yld1_row12 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(1, 2))

sny22_yld1_row34 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(3, 4))


### Check dimensions
dim(sny22_yld1_row12)
dim(sny22_yld1_row34)


### ---------------------------------------------------------
### 2. Ensure GRM inverse
### ---------------------------------------------------------

attr(grm_inverse, "INVERSE") <- TRUE


### =========================================================
### 3. BASE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

base_sny22_yld1_row12 <- asreml(
  fixed    = yld1 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld1_row12)

summary(base_sny22_yld1_row12)$varcomp


### Extract BLUPs
blup_base_sny22_yld1_row12 <- coef(base_sny22_yld1_row12)$random

blup_base_sny22_yld1_row12 <- blup_base_sny22_yld1_row12[
  grep("Germplasm", rownames(blup_base_sny22_yld1_row12)),
  ,
  drop = FALSE]

g_base_sny22_yld1_row12 <- blup_base_sny22_yld1_row12[,1]
names(g_base_sny22_yld1_row12) <- rownames(blup_base_sny22_yld1_row12)



### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

base_sny22_yld1_row34 <- asreml(
  fixed    = yld1 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld1_row34)

summary(base_sny22_yld1_row34)$varcomp


### Extract BLUPs
blup_base_sny22_yld1_row34 <- coef(base_sny22_yld1_row34)$random

blup_base_sny22_yld1_row34 <- blup_base_sny22_yld1_row34[
  grep("Germplasm", rownames(blup_base_sny22_yld1_row34)),
  ,
  drop = FALSE]

g_base_sny22_yld1_row34 <- blup_base_sny22_yld1_row34[,1]
names(g_base_sny22_yld1_row34) <- rownames(blup_base_sny22_yld1_row34)


### ---------------------------------------------------------
### GEER: Base model
### ---------------------------------------------------------

common_base_sny22_yld1 <- intersect(names(g_base_sny22_yld1_row12), names(g_base_sny22_yld1_row34))
common_base_sny22_yld1
cor(g_base_sny22_yld1_row12[common_base_sny22_yld1], g_base_sny22_yld1_row34[common_base_sny22_yld1])




### =========================================================
### 4. SPATIAL MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

spa_sny22_yld1_row12 <- asreml(
  fixed    = yld1 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range),
  residual = ~ id(units),
  data     = sny22_yld1_row12)

summary(spa_sny22_yld1_row12)$varcomp


### Extract BLUPs
blup_spa_sny22_yld1_row12 <- coef(spa_sny22_yld1_row12)$random
blup_spa_sny22_yld1_row12
blup_spa_sny22_yld1_row12 <- blup_spa_sny22_yld1_row12[
  grep("Germplasm", rownames(blup_spa_sny22_yld1_row12)),
  ,
  drop = FALSE
]

g_spa_sny22_yld1_row12 <- blup_spa_sny22_yld1_row12[,1]
names(g_spa_sny22_yld1_row12) <- rownames(blup_spa_sny22_yld1_row12)
g_spa_sny22_yld1_row12


### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

spa_sny22_yld1_row34 <- asreml(
  fixed    = yld1 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range),
  residual = ~ id(units),
  data     = sny22_yld1_row34)

summary(spa_sny22_yld1_row34)$varcomp


### Extract BLUPs
blup_spa_sny22_yld1_row34 <- coef(spa_sny22_yld1_row34)$random

blup_spa_sny22_yld1_row34 <- blup_spa_sny22_yld1_row34[
  grep("Germplasm", rownames(blup_spa_sny22_yld1_row34)),
  ,
  drop = FALSE]

g_spa_sny22_yld1_row34 <- blup_spa_sny22_yld1_row34[,1]
names(g_spa_sny22_yld1_row34) <- rownames(blup_spa_sny22_yld1_row34)



### ---------------------------------------------------------
### GEER: Spatial model
### ---------------------------------------------------------

common_spa_sny22_yld1 <- intersect(names(g_spa_sny22_yld1_row12), names(g_spa_sny22_yld1_row34))

cor(g_spa_sny22_yld1_row12[common_spa_sny22_yld1], g_spa_sny22_yld1_row34[common_spa_sny22_yld1])


### =========================================================
### 5. PEE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

pee_sny22_yld1_row12 <- asreml(
  fixed    = yld1 ~ 1  + Rep +PEE_165,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld1_row12)

summary(pee_sny22_yld1_row12)$varcomp


### Extract BLUPs
blup_pee_sny22_yld1_row12 <- coef(pee_sny22_yld1_row12)$random

blup_pee_sny22_yld1_row12 <- blup_pee_sny22_yld1_row12[
  grep("Germplasm", rownames(blup_pee_sny22_yld1_row12)),
  ,
  drop = FALSE]

g_pee_sny22_yld1_row12 <- blup_pee_sny22_yld1_row12[,1]
names(g_pee_sny22_yld1_row12) <- rownames(blup_pee_sny22_yld1_row12)
g_pee_sny22_yld1_row12


### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

pee_sny22_yld1_row34 <- asreml(
  fixed    = yld1 ~ 1  + Rep + PEE_165,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld1_row34)

summary(pee_sny22_yld1_row34)$varcomp


### Extract BLUPs
blup_pee_sny22_yld1_row34 <- coef(pee_sny22_yld1_row34)$random

blup_pee_sny22_yld1_row34 <- blup_pee_sny22_yld1_row34[
  grep("Germplasm", rownames(blup_pee_sny22_yld1_row34)),
  ,
  drop = FALSE]

g_pee_sny22_yld1_row34 <- blup_pee_sny22_yld1_row34[,1]
names(g_pee_sny22_yld1_row34) <- rownames(blup_pee_sny22_yld1_row34)



### ---------------------------------------------------------
### GEER: PEE model
### ---------------------------------------------------------

common_pee_sny22_yld1 <- intersect(names(g_pee_sny22_yld1_row12), names(g_pee_sny22_yld1_row34))

cor( g_pee_sny22_yld1_row12[common_pee_sny22_yld1], g_pee_sny22_yld1_row34[common_pee_sny22_yld1])




### =========================================================
### 6. Final GEER summary table
### =========================================================

GEER_rowblock_results_sny22_yld1 <- data.frame(

  Model = c("Base", "Spatial", "PEE"),

  Pearson = c(
    cor(g_base_sny22_yld1_row12[common_base_sny22_yld1],
        g_base_sny22_yld1_row34[common_base_sny22_yld1]),

    cor(g_spa_sny22_yld1_row12[common_spa_sny22_yld1],
        g_spa_sny22_yld1_row34[common_spa_sny22_yld1]),

    cor(g_pee_sny22_yld1_row12[common_pee_sny22_yld1],
        g_pee_sny22_yld1_row34[common_pee_sny22_yld1])))

GEER_rowblock_results_sny22_yld1
```


### 3. Validity of GEER Results (block wise results)

0.93&0.89, 0.92 and 0.90 , and 0.93 and 0.89
```{r Validation of GEER results using all data base model, sny22 yld1, 06072026}

### =========================================================
### VALIDATION OF GEER RESULTS
### Snyder NY 2022
### Harvest 1 (yld1)
###
### Goal:
###   Compare genomic BLUPs from:
###     1. Base model using all data
###     2. Spatial model using each block
###     3. PEE model using each block
###
### Validation:
###   Correlate block-specific BLUPs against
###   baseline genomic BLUPs from full dataset
### =========================================================


### ---------------------------------------------------------
### 1. Ensure inverse GRM
### ---------------------------------------------------------

asreml.options(extra = 200, maxit = 200)

attr(grm_inverse, "INVERSE") <- TRUE



### ---------------------------------------------------------
### 2. Create row-block subsets
### ---------------------------------------------------------

sny22_yld1_row12 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(1, 2))

sny22_yld1_row34 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(3, 4))



### =========================================================
### 3. BASE MODEL USING ALL DATA
### =========================================================

sny22_yld1_base_all <- asreml(
  fixed    = yld1 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = ny_sny_pheno_22)

summary(sny22_yld1_base_all)$varcomp


### Extract genomic BLUPs
sny22_yld1_gblup_base_all <- coef(sny22_yld1_base_all)$random

sny22_yld1_gblup_base_all <- sny22_yld1_gblup_base_all[
  grep("Germplasm", rownames(sny22_yld1_gblup_base_all)),
  ,
  drop = FALSE]

sny22_yld1_gblup_base_all <- sny22_yld1_gblup_base_all[,1]

names(sny22_yld1_gblup_base_all) <- 
  rownames(sny22_yld1_gblup_base_all)
sny22_yld1_gblup_base_all

##### part of spatial model and PEE model in above chunk...

### =========================================================
### 4. VALIDATION CORRELATIONS
### =========================================================

### Base correlations
sny22_yld1_cor_base <- c(
  cor(g_base_sny22_yld1_row12, sny22_yld1_gblup_base_all),
  cor(g_base_sny22_yld1_row34, sny22_yld1_gblup_base_all))

sny22_yld1_cor_base

### Spatial correlations (Ar)
sny22_yld1_cor_spa <- c(
  cor(g_spa_sny22_yld1_row12, sny22_yld1_gblup_base_all),
  cor(g_spa_sny22_yld1_row34, sny22_yld1_gblup_base_all))

sny22_yld1_cor_spa



### PEE correlations
sny22_yld1_cor_pee <- c(
  cor(g_pee_sny22_yld1_row12, sny22_yld1_gblup_base_all),
  cor(
    g_pee_sny22_yld1_row34, sny22_yld1_gblup_base_all))
sny22_yld1_cor_pee



### =========================================================
### 5. Summary table
### =========================================================

sny22_yld1_validation_results <- data.frame(

  Model = c("Base_Row12", "Base_Row12", 
            "Spatial_Row12",
            "Spatial_Row34",
            "PEE_Row12",
            "PEE_Row34"),

  Pearson = c(
    sny22_yld1_cor_base[1],
    sny22_yld1_cor_base[2],
    sny22_yld1_cor_spa[1],
    sny22_yld1_cor_spa[2],
    sny22_yld1_cor_pee[1],
    sny22_yld1_cor_pee[2]))

sny22_yld1_validation_results

```






############################ Harvest2: ############################ 

### 1. Using Random regression
- linear
```{r Combined chunk for harvest2 using Linear Random Regression 05202026}
asreml.options(extra = 200, maxit = 200)
# ---------- PART 0: Manange data file -----------

#  Manage data file: First, assign harvest based on GDD
# first make harvest 2 only data file
ny_sny_pheno_22_gndvi_only_long_harvestwise <- 
  ny_sny_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 165 ~ "Harvest1",
    GDD >= 166 & GDD <= 208 ~ "Harvest2",
    GDD >= 209 & GDD <= 260 ~ "Harvest3")) # 259 flight and 260 last yield (actually 259) cz they are on same date so to differentiate, made this diff.
# Subset harvest 2 !!!!!
ny22_sny_gdd_harvest2 <- ny_sny_pheno_22_gndvi_only_long_harvestwise %>%
  filter(Harvest == "Harvest2") # !!! capital H

# create factor + numeric GDD
ny22_sny_gdd_harvest2$GDD_fac <- as.factor(ny22_sny_gdd_harvest2$GDD)
ny22_sny_gdd_harvest2 <- ny22_sny_gdd_harvest2[order(ny22_sny_gdd_harvest2$GDD_fac),]

unique(ny22_sny_gdd_harvest2$GDD_fac) # 168 171 175 178 181 187 189 194 196 207 208(for yld2)

# Build Legendre basis
unique_gdd_yld2 <- sort(unique(ny22_sny_gdd_harvest2$GDD))
unique_gdd_yld2 <- as.numeric(unique_gdd_yld2)

######## fitting linear equation
M_yld2 <- stdtime(unique_gdd_yld2, 1) # for linear, 1
lambda_yld2 <- Legendre(1, gengler=TRUE) # for quad, 2

Phi_yld2 <- M_yld2 %*% t(lambda_yld2)
colnames(Phi_yld2) <- c("LPInt", "LPLin")
rownames(Phi_yld2) <- unique_gdd_yld2
Phi_yld2
ordLP_yld2 <- match(ny22_sny_gdd_harvest2$GDD, rownames(Phi_yld2))
ny22_sny_gdd_harvest2$LPInt_yld2 <- Phi_yld2[ordLP_yld2, "LPInt"]
ny22_sny_gdd_harvest2$LPLin_yld2 <- Phi_yld2[ordLP_yld2, "LPLin"]

# Fit RR model (harvest 1)
asreml.options(extra = 200, maxit = 200)

lp_mod_sny_22_yld2 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI ~ 1 + at(GDD_fac):Rep , # fitting Rep effect
    random = ~ str(~ Germplasm:LPInt_yld2 + Germplasm:LPLin_yld2, ~ us(2):vm(Germplasm, grm_inverse)) + str(~ Plot:LPInt_yld2 + Plot:LPLin_yld2, ~ us(2):id(Plot)), # # for quadratic, us(3)
    residual = ~ dsum(~ units | GDD_fac),
    data = ny22_sny_gdd_harvest2)}

summary(lp_mod_sny_22_yld2)$varcomp
lp_mod_sny_22_yld2 <- update.asreml(lp_mod_sny_22_yld2, start = lp_mod_sny_22_yld2$gammas)

# Save random effects (BLUPs)
blups_yld2 <- lp_mod_sny_22_yld2$coefficients$random
blups_yld2

# --- Convert full BLUPs to a data frame ---
blups_df_yld2 <- as.data.frame(blups_yld2) %>% rename(solution = 1)
blups_df_yld2
# Add term names (simplify)
blups_df_yld2$term <- gsub(":.*", "", rownames(blups_df_yld2))

# --- Filter subsets ---
# BLUP_germplasm_yld2 <- blups_df_yld2 %>% filter(grepl("^Germplasm", rownames(blups_df_yld2)))
BLUP_spa_int_yld2 <- blups_df_yld2 %>% 
  filter(grepl("^Plot_.*LPInt_yld2$", rownames(blups_df_yld2)))

BLUP_spa_slp_yld2 <- blups_df_yld2 %>% 
  filter(grepl("^Plot_.*LPLin_yld2$", rownames(blups_df_yld2)))

# --- Spatial BLUPs ---
BLUP_spa_int_yld2$solution   # Intercepts
BLUP_spa_slp_yld2$solution   # Slopes

# --- Combine intercept and slope into a matrix ---
spa_BLUP_matrix_yld2 <- cbind(
  BLUP_spa_int_yld2$solution, 
  BLUP_spa_slp_yld2$solution)
spa_BLUP_matrix_yld2
# ---------- Extract spatial BLUPs as PEE (Permanent Environmental Effects) ----------

# Convert the matrix into a data frame for including it in the model below cz my model expects columns in the data frame
ny22_sny_spa_yld2 <- as.data.frame(spa_BLUP_matrix_yld2)
# add these as columns in my data file
ny_sny_pheno_22$spa_int_yld2 <- ny22_sny_spa_yld2[,1]
ny_sny_pheno_22$spa_slp_yld2 <- ny22_sny_spa_yld2[,2]

# !!!!!!!!!!!!!!! Permanent Environmental Effect !!!!!!!!!!!!! 
Phi_yld2
# here intercept and slope for GDD on PEE_157 are 1 and 1.7320508 


 ### NOTE: the final LPLin for final GDD is always 1.732... the value is same as PEE_208

# PEE_208 = 1 * spa_int_yld2 + 1.7320508 * spa_slp_yld2
ny_sny_pheno_22$PEE_208 <- 
  1 * ny_sny_pheno_22$spa_int_yld2 +
  1.7320508 * ny_sny_pheno_22$spa_slp_yld2


# ---------- Step 2: Fit PEE's as fixed regressor on a yield model with spatial correction ----------

# Fit single-trait model for yield (yld2)
# Spatial effects are included as fixed covariates (PEE correction)
uni_ny22_sny_yld2_pee <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed   = yld2 ~ 1 + Row + PEE_208,   # PEE as fixed regressors
    random  = ~ vm(Germplasm, grm_inverse),         # genomic effect
    residual= ~ id(units),
    data    = ny_sny_pheno_22)}


# Extract variance components
summary(uni_ny22_sny_yld2_pee)$varcomp




# 1st model: yield alone genetic model without spatial (autoregressive)
uni_ny22_sny_yld2 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Row, #row effect as fixed
         random = ~ vm(Germplasm, grm_inverse),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(uni_ny22_sny_yld2)$varcomp



# 2nd model: yield alone genetic model with spatial (autoregressive)
## Univariate with autoregressive spatial correction
asreml.options(extra = 200, maxit = 200)

uni_ny22_sny_yld2_spa <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 ,
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range), 
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(uni_ny22_sny_yld2_spa)$varcomp






# compare

# variance components
summary(uni_ny22_sny_yld2)$varcomp
summary(uni_ny22_sny_yld2_spa)$varcomp
summary(uni_ny22_sny_yld2_pee)$varcomp

# heritability
h2_no_spa  <- asreml::vpredict(uni_ny22_sny_yld2, heritA ~ V1 / (V1 + V2))
h2_no_spa # 0.3749863
h2_AR1     <- asreml::vpredict(uni_ny22_sny_yld2_spa, heritA ~ V1 / (V1 + V5))
h2_AR1 # 0.75
# asreml::vpredict(uni_ny22_sny_yld2_spa, heritA ~ V1 / (V1 + V5)) # if both
# asreml::vpredict(uni_ny22_sny_yld2_spa, heritA ~ V3 / (V3 + V4)) # if row only
h2_PEE     <-asreml::vpredict(uni_ny22_sny_yld2_pee, heritA ~ V1 / (V1 + V2))
h2_PEE # 0.51 


# visualize PEE too
 ## just PEE
ggplot(ny_sny_pheno_22, aes(x=Row, y=Range, fill=ny_sny_pheno_22$PEE_208)) + # !!!
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title=" PEE", fill="PEE") +
  theme_minimal()

# Visualize Residual - PEE to see what;s the pattern left. For that need to rescale residuals to PEE and then visualize these... (SAME SCALE NEEDED)

uni_ny22_sny_yld2_res <- residuals(uni_ny22_sny_yld2)
uni_ny22_sny_yld2_res
uni_ny22_sny_yld2_res <- scale(uni_ny22_sny_yld2_res)
PEE_208 <- scale(ny_sny_pheno_22$PEE_208)
plot(PEE_208)
uni_ny22_sny_yld2_res_pee_scaled <- uni_ny22_sny_yld2_res - PEE_208
uni_ny22_sny_yld2_res_pee_scaled
  # ggplot
ggplot(ny_sny_pheno_22, aes(x=Row, y=Range, fill=uni_ny22_sny_yld2_res_pee_scaled)) +
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title="Residuals - PEE", fill="Residual") +
  theme_minimal()





# Get log-likelihood, AIC, BIC
summary(uni_ny22_sny_yld2)$loglik # 
summary(uni_ny22_sny_yld2)$aic
summary(uni_ny22_sny_yld2)$aic
```


### 2. GEER: Function to run for all reps and models 
0.65, 0.76, 0.78
```{r GEER between two blocks (rep12 and rep34) for yld2 (only yield data) sny 2022, 05202026}
### =========================================================
### GEER using row-block subsets
### Snyder NY 2022
### Harvest2 (yld2)
###
### Strategy:
###   Since Rep is unavailable as an independent factor,
###   create pseudo-replicates using adjacent row blocks.
###
###   Block 1 = Rows 1 + 2
###   Block 2 = Rows 3 + 4
###
###   Fit models separately within each block:
###     1. Base
###     2. Spatial
###     3. PEE
###
###   Then correlate genomic BLUPs between blocks.
### =========================================================


### ---------------------------------------------------------
### 1. Create row-block subsets
### ---------------------------------------------------------

sny22_yld2_row12 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(1, 2))

sny22_yld2_row34 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(3, 4))


### Check dimensions
dim(sny22_yld2_row12)
dim(sny22_yld2_row34)


### ---------------------------------------------------------
### 2. Ensure GRM inverse
### ---------------------------------------------------------

attr(grm_inverse, "INVERSE") <- TRUE


### =========================================================
### 3. BASE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

base_sny22_yld2_row12 <- asreml(
  fixed    = yld2 ~ 1 ,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld2_row12)

summary(base_sny22_yld2_row12)$varcomp


### Extract BLUPs
blup_base_sny22_yld2_row12 <- coef(base_sny22_yld2_row12)$random

blup_base_sny22_yld2_row12 <- blup_base_sny22_yld2_row12[
  grep("Germplasm", rownames(blup_base_sny22_yld2_row12)),
  ,
  drop = FALSE]

g_base_sny22_yld2_row12 <- blup_base_sny22_yld2_row12[,1]
names(g_base_sny22_yld2_row12) <- rownames(blup_base_sny22_yld2_row12)



### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

base_sny22_yld2_row34 <- asreml(
  fixed    = yld2 ~ 1 ,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld2_row34)

summary(base_sny22_yld2_row34)$varcomp


### Extract BLUPs
blup_base_sny22_yld2_row34 <- coef(base_sny22_yld2_row34)$random

blup_base_sny22_yld2_row34 <- blup_base_sny22_yld2_row34[
  grep("Germplasm", rownames(blup_base_sny22_yld2_row34)),
  ,
  drop = FALSE]

g_base_sny22_yld2_row34 <- blup_base_sny22_yld2_row34[,1]
names(g_base_sny22_yld2_row34) <- rownames(blup_base_sny22_yld2_row34)


### ---------------------------------------------------------
### GEER: Base model
### ---------------------------------------------------------

common_base_sny22_yld2 <- intersect(names(g_base_sny22_yld2_row12), names(g_base_sny22_yld2_row34))
common_base_sny22_yld2
cor(g_base_sny22_yld2_row12[common_base_sny22_yld2], g_base_sny22_yld2_row34[common_base_sny22_yld2])




### =========================================================
### 4. SPATIAL MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

spa_sny22_yld2_row12 <- asreml(
  fixed    = yld2 ~ 1 ,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
  residual = ~ id(units),
  data     = sny22_yld2_row12)

summary(spa_sny22_yld2_row12)$varcomp


### Extract BLUPs
blup_spa_sny22_yld2_row12 <- coef(spa_sny22_yld2_row12)$random
blup_spa_sny22_yld2_row12
blup_spa_sny22_yld2_row12 <- blup_spa_sny22_yld2_row12[
  grep("Germplasm", rownames(blup_spa_sny22_yld2_row12)),
  ,
  drop = FALSE
]

g_spa_sny22_yld2_row12 <- blup_spa_sny22_yld2_row12[,1]
names(g_spa_sny22_yld2_row12) <- rownames(blup_spa_sny22_yld2_row12)
g_spa_sny22_yld2_row12


### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

spa_sny22_yld2_row34 <- asreml(
  fixed    = yld2 ~ 1,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
  residual = ~ id(units),
  data     = sny22_yld2_row34)

summary(spa_sny22_yld2_row34)$varcomp


### Extract BLUPs
blup_spa_sny22_yld2_row34 <- coef(spa_sny22_yld2_row34)$random

blup_spa_sny22_yld2_row34 <- blup_spa_sny22_yld2_row34[
  grep("Germplasm", rownames(blup_spa_sny22_yld2_row34)),
  ,
  drop = FALSE]

g_spa_sny22_yld2_row34 <- blup_spa_sny22_yld2_row34[,1]
names(g_spa_sny22_yld2_row34) <- rownames(blup_spa_sny22_yld2_row34)



### ---------------------------------------------------------
### GEER: Spatial model
### ---------------------------------------------------------

common_spa_sny22_yld2 <- intersect(names(g_spa_sny22_yld2_row12), names(g_spa_sny22_yld2_row34))

cor(g_spa_sny22_yld2_row12[common_spa_sny22_yld2], g_spa_sny22_yld2_row34[common_spa_sny22_yld2])


### =========================================================
### 5. PEE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

pee_sny22_yld2_row12 <- asreml(
  fixed    = yld2 ~ 1 + PEE_208,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld2_row12)

summary(pee_sny22_yld2_row12)$varcomp


### Extract BLUPs
blup_pee_sny22_yld2_row12 <- coef(pee_sny22_yld2_row12)$random

blup_pee_sny22_yld2_row12 <- blup_pee_sny22_yld2_row12[
  grep("Germplasm", rownames(blup_pee_sny22_yld2_row12)),
  ,
  drop = FALSE]

g_pee_sny22_yld2_row12 <- blup_pee_sny22_yld2_row12[,1]
names(g_pee_sny22_yld2_row12) <- rownames(blup_pee_sny22_yld2_row12)



### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

pee_sny22_yld2_row34 <- asreml(
  fixed    = yld2 ~ 1 + PEE_208,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld2_row34)

summary(pee_sny22_yld2_row34)$varcomp


### Extract BLUPs
blup_pee_sny22_yld2_row34 <- coef(pee_sny22_yld2_row34)$random

blup_pee_sny22_yld2_row34 <- blup_pee_sny22_yld2_row34[
  grep("Germplasm", rownames(blup_pee_sny22_yld2_row34)),
  ,
  drop = FALSE]

g_pee_sny22_yld2_row34 <- blup_pee_sny22_yld2_row34[,1]
names(g_pee_sny22_yld2_row34) <- rownames(blup_pee_sny22_yld2_row34)
blup_pee_sny22_yld2_row34


### ---------------------------------------------------------
### GEER: PEE model
### ---------------------------------------------------------

common_pee_sny22_yld2 <- intersect(names(g_pee_sny22_yld2_row12), names(g_pee_sny22_yld2_row34))
common_pee_sny22_yld2
cor( g_pee_sny22_yld2_row12[common_pee_sny22_yld2], g_pee_sny22_yld2_row34[common_pee_sny22_yld2])
# cor(blup_pee_sny22_yld2_row34,  blup_pee_sny22_yld2_row12)


### =========================================================
### 6. Final GEER summary table
### =========================================================

GEER_rowblock_results_sny22_yld2 <- data.frame(

  Model = c("Base", "Spatial", "PEE"),

  Pearson = c(
    cor(g_base_sny22_yld2_row12[common_base_sny22_yld2],
        g_base_sny22_yld2_row34[common_base_sny22_yld2]),

    cor(g_spa_sny22_yld2_row12[common_spa_sny22_yld2],
        g_spa_sny22_yld2_row34[common_spa_sny22_yld2]),

    cor(g_pee_sny22_yld2_row12[common_pee_sny22_yld2],
        g_pee_sny22_yld2_row34[common_pee_sny22_yld2])))

GEER_rowblock_results_sny22_yld2
```


### 3. Validity of GEER Results (block wise results)
0.92&0.92, and 0.92&0.91
```{r Validation of GEER results using all data base model, sny22 yld2, 06072026}

### =========================================================
### VALIDATION OF GEER RESULTS
### Snyder NY 2022
### Harvest 1 (yld2)
###
### Goal:
###   Compare genomic BLUPs from:
###     1. Base model using all data
###     2. Spatial model using each block
###     3. PEE model using each block
###
### Validation:
###   Correlate block-specific BLUPs against
###   baseline genomic BLUPs from full dataset
### =========================================================


### ---------------------------------------------------------
### 1. Ensure inverse GRM
### ---------------------------------------------------------

asreml.options(extra = 200, maxit = 200)

attr(grm_inverse, "INVERSE") <- TRUE



### ---------------------------------------------------------
### 2. Create row-block subsets
### ---------------------------------------------------------

sny22_yld2_row12 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(1, 2))

sny22_yld2_row34 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(3, 4))



### =========================================================
### 3. BASE MODEL USING ALL DATA
### =========================================================

sny22_yld2_base_all <- asreml(
  fixed    = yld2 ~ 1 ,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = ny_sny_pheno_22)

summary(sny22_yld2_base_all)$varcomp


### Extract genomic BLUPs
sny22_yld2_gblup_base_all <- coef(sny22_yld2_base_all)$random

sny22_yld2_gblup_base_all <- sny22_yld2_gblup_base_all[
  grep("Germplasm", rownames(sny22_yld2_gblup_base_all)),
  ,
  drop = FALSE]

sny22_yld2_gblup_base_all <- sny22_yld2_gblup_base_all[,1]

names(sny22_yld2_gblup_base_all) <- 
  rownames(sny22_yld2_gblup_base_all)
sny22_yld2_gblup_base_all

##### part of spatial model and PEE model in above chunk...

### =========================================================
### 4. VALIDATION CORRELATIONS
### =========================================================

### Base correlations (Ar)
sny22_yld2_cor_base <- c(
  cor(g_base_sny22_yld2_row12, sny22_yld2_gblup_base_all),
  cor(g_base_sny22_yld2_row34, sny22_yld2_gblup_base_all))
sny22_yld2_cor_base

### Spatial correlations (Ar)
sny22_yld2_cor_spa <- c(
  cor(g_spa_sny22_yld2_row12, sny22_yld2_gblup_base_all),
  cor(g_spa_sny22_yld2_row34, sny22_yld2_gblup_base_all))

sny22_yld2_cor_spa



### PEE correlations
sny22_yld2_cor_pee <- c(
  cor(g_pee_sny22_yld2_row12, sny22_yld2_gblup_base_all),
  cor(
    g_pee_sny22_yld2_row34, sny22_yld2_gblup_base_all))
sny22_yld2_cor_pee



### =========================================================
### 5. Summary table
### =========================================================

sny22_yld2_validation_results <- data.frame(

  Model = c("Base_Row12", "Base_Row12", 
            "Spatial_Row12",
            "Spatial_Row34",
            "PEE_Row12",
            "PEE_Row34"),

  Pearson = c(
    sny22_yld2_cor_base[1],
    sny22_yld2_cor_base[2],
    sny22_yld2_cor_spa[1],
    sny22_yld2_cor_spa[2],
    sny22_yld2_cor_pee[1],
    sny22_yld2_cor_pee[2]))

sny22_yld2_validation_results

```









############################ Harvest3: ############################ 

### 1. Using Random regression
 linear
```{r Combined chunk for harvest2 using Linear Random Regression 05202026}
asreml.options(extra = 200, maxit = 200)
# ---------- PART 0: Manange data file -----------

#  Manage data file: First, assign harvest based on GDD
# first make harvest 2 only data file
ny_sny_pheno_22_gndvi_only_long_harvestwise <- 
  ny_sny_pheno_22_gndvi_only_long %>%
  mutate(Harvest = case_when(
    GDD >= 60  & GDD <= 165 ~ "Harvest1",
    GDD >= 166 & GDD <= 208 ~ "Harvest2",
    GDD >= 209 & GDD <= 260 ~ "Harvest3")) # 259 flight and 260 last yield (actually 259) cz they are on same date so to differentiate, made this diff.
# Subset harvest 2 !!!!!
ny22_sny_gdd_harvest3 <- ny_sny_pheno_22_gndvi_only_long_harvestwise %>%
  filter(Harvest == "Harvest3") # !!! capital H

# create factor + numeric GDD
ny22_sny_gdd_harvest3$GDD_fac <- as.factor(ny22_sny_gdd_harvest3$GDD)
ny22_sny_gdd_harvest3 <- ny22_sny_gdd_harvest3[order(ny22_sny_gdd_harvest3$GDD_fac),]

unique(ny22_sny_gdd_harvest3$GDD_fac) # 223 231 236 238 243 245 252 258 259 260(for yld3)

# Build Legendre basis
unique_gdd_yld3 <- sort(unique(ny22_sny_gdd_harvest3$GDD))
unique_gdd_yld3 <- as.numeric(unique_gdd_yld3)

######## fitting linear equation
M_yld3 <- stdtime(unique_gdd_yld3, 1) # for linear, 1
lambda_yld3 <- Legendre(1, gengler=TRUE) # for quad, 2

Phi_yld3 <- M_yld3 %*% t(lambda_yld3)
colnames(Phi_yld3) <- c("LPInt", "LPLin")
rownames(Phi_yld3) <- unique_gdd_yld3
Phi_yld3
ordLP_yld3 <- match(ny22_sny_gdd_harvest3$GDD, rownames(Phi_yld3))
ny22_sny_gdd_harvest3$LPInt_yld3 <- Phi_yld3[ordLP_yld3, "LPInt"]
ny22_sny_gdd_harvest3$LPLin_yld3 <- Phi_yld3[ordLP_yld3, "LPLin"]

# Fit RR model (harvest 1)
asreml.options(extra = 200, maxit = 200)

lp_mod_sny_22_yld3 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI ~ 1 + at(GDD_fac):Rep , # fitting Rep effect
    random = ~ str(~ Germplasm:LPInt_yld3 + Germplasm:LPLin_yld3, ~ us(2):vm(Germplasm, grm_inverse)) + str(~ Plot:LPInt_yld3 + Plot:LPLin_yld3, ~ us(2):id(Plot)), # # for quadratic, us(3)
    residual = ~ dsum(~ units | GDD_fac),
    data = ny22_sny_gdd_harvest3)}

summary(lp_mod_sny_22_yld3)$varcomp
lp_mod_sny_22_yld3 <- update.asreml(lp_mod_sny_22_yld3, start = lp_mod_sny_22_yld3$gammas)

# Save random effects (BLUPs)
blups_yld3 <- lp_mod_sny_22_yld3$coefficients$random
blups_yld3

# --- Convert full BLUPs to a data frame ---
blups_df_yld3 <- as.data.frame(blups_yld3) %>% rename(solution = 1)
blups_df_yld3
# Add term names (simplify)
blups_df_yld3$term <- gsub(":.*", "", rownames(blups_df_yld3))

# --- Filter subsets ---
# BLUP_germplasm_yld3 <- blups_df_yld3 %>% filter(grepl("^Germplasm", rownames(blups_df_yld3)))
BLUP_spa_int_yld3 <- blups_df_yld3 %>% 
  filter(grepl("^Plot_.*LPInt_yld3$", rownames(blups_df_yld3)))

BLUP_spa_slp_yld3 <- blups_df_yld3 %>% 
  filter(grepl("^Plot_.*LPLin_yld3$", rownames(blups_df_yld3)))

# --- Spatial BLUPs ---
BLUP_spa_int_yld3$solution   # Intercepts
BLUP_spa_slp_yld3$solution   # Slopes

# --- Combine intercept and slope into a matrix ---
spa_BLUP_matrix_yld3 <- cbind(
  BLUP_spa_int_yld3$solution, 
  BLUP_spa_slp_yld3$solution)
spa_BLUP_matrix_yld3
# ---------- Extract spatial BLUPs as PEE (Permanent Environmental Effects) ----------

# Convert the matrix into a data frame for including it in the model below cz my model expects columns in the data frame
ny22_sny_spa_yld3 <- as.data.frame(spa_BLUP_matrix_yld3)
# add these as columns in my data file
ny_sny_pheno_22$spa_int_yld3 <- ny22_sny_spa_yld3[,1]
ny_sny_pheno_22$spa_slp_yld3 <- ny22_sny_spa_yld3[,2]

# !!!!!!!!!!!!!!! Permanent Environmental Effect !!!!!!!!!!!!! 
Phi_yld3
# here intercept and slope for GDD on PEE_157 are 1 and 1.7320508 


 ### NOTE: the final LPLin for final GDD is always 1.732... the value is same as PEE_260

# PEE_260 = 1 * spa_int_yld3 + 1.7320508 * spa_slp_yld3
ny_sny_pheno_22$PEE_260 <- 
  1 * ny_sny_pheno_22$spa_int_yld3 +
  1.7320508 * ny_sny_pheno_22$spa_slp_yld3



# ---------- Step 2: Fit PEE's as fixed regressor on a yield model with spatial correction ----------

# Fit single-trait model for yield (yld3)
# Spatial effects are included as fixed covariates (PEE correction)
uni_ny22_sny_yld3_pee <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed   = yld3 ~ 1 + Row + PEE_260,   # PEE as fixed regressors
    random  = ~ vm(Germplasm, grm_inverse),         # genomic effect
    residual= ~ id(units),
    data    = ny_sny_pheno_22)}


# Extract variance components
summary(uni_ny22_sny_yld3_pee)$varcomp




# 1st model: yield alone genetic model without spatial (autoregressive)
uni_ny22_sny_yld3 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Row, #row effect as fixed
         random = ~ vm(Germplasm, grm_inverse),
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(uni_ny22_sny_yld3)$varcomp



# 2nd model: yield alone genetic model with spatial (autoregressive)
## Univariate with autoregressive spatial correction
asreml.options(extra = 200, maxit = 200)

uni_ny22_sny_yld3_spa <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 ,
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range), 
         residual = ~ id(units),
         data = ny_sny_pheno_22)}
summary(uni_ny22_sny_yld3_spa)$varcomp






# compare

# variance components
summary(uni_ny22_sny_yld3)$varcomp
summary(uni_ny22_sny_yld3_spa)$varcomp
summary(uni_ny22_sny_yld3_pee)$varcomp

# heritability
h2_no_spa  <- asreml::vpredict(uni_ny22_sny_yld3, heritA ~ V1 / (V1 + V2))
h2_no_spa # 0.52
h2_AR1     <- asreml::vpredict(uni_ny22_sny_yld3_spa, heritA ~ V1 / (V1 + V5))
h2_AR1 # 0.76
# asreml::vpredict(uni_ny22_sny_yld3_spa, heritA ~ V1 / (V1 + V5)) # if both
# asreml::vpredict(uni_ny22_sny_yld3_spa, heritA ~ V3 / (V3 + V4)) # if row only
h2_PEE     <-asreml::vpredict(uni_ny22_sny_yld3_pee, heritA ~ V1 / (V1 + V2))
h2_PEE # 0.59 


# visualize PEE too
 ## just PEE
ggplot(ny_sny_pheno_22, aes(x=Row, y=Range, fill=ny_sny_pheno_22$PEE_260)) + # !!!
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title=" PEE", fill="PEE") +
  theme_minimal()

# Visualize Residual - PEE to see what;s the pattern left. For that need to rescale residuals to PEE and then visualize these... (SAME SCALE NEEDED)

uni_ny22_sny_yld3_res <- residuals(uni_ny22_sny_yld3)
uni_ny22_sny_yld3_res
uni_ny22_sny_yld3_res <- scale(uni_ny22_sny_yld3_res)
PEE_260 <- scale(ny_sny_pheno_22$PEE_260)
plot(PEE_260)
uni_ny22_sny_yld3_res_pee_scaled <- uni_ny22_sny_yld3_res - PEE_260
uni_ny22_sny_yld3_res_pee_scaled
  # ggplot
ggplot(ny_sny_pheno_22, aes(x=Row, y=Range, fill=uni_ny22_sny_yld3_res_pee_scaled)) +
  geom_tile() +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  labs(title="Residuals - PEE", fill="Residual") +
  theme_minimal()





# Get log-likelihood, AIC, BIC
summary(uni_ny22_sny_yld3)$loglik # 
summary(uni_ny22_sny_yld3)$aic
summary(uni_ny22_sny_yld3)$aic
```


### 2. GEER: Function to run for all reps and models 
-  0.70, 0.77, 0.76
```{r GEER between two blocks (rep12 and rep34) for yld3 (only yield data) sny 2022, 05202026}
### =========================================================
### GEER using row-block subsets
### Snyder NY 2022
### harvest3 (yld3)
###
### Strategy:
###   Since Rep is unavailable as an independent factor,
###   create pseudo-replicates using adjacent row blocks.
###
###   Block 1 = Rows 1 + 2
###   Block 2 = Rows 3 + 4
###
###   Fit models separately within each block:
###     1. Base
###     2. Spatial
###     3. PEE
###
###   Then correlate genomic BLUPs between blocks.
### =========================================================


### ---------------------------------------------------------
### 1. Create row-block subsets
### ---------------------------------------------------------

sny22_yld3_row12 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(1, 2))

sny22_yld3_row34 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(3, 4))


### Check dimensions
dim(sny22_yld3_row12)
dim(sny22_yld3_row34)


### ---------------------------------------------------------
### 2. Ensure GRM inverse
### ---------------------------------------------------------

attr(grm_inverse, "INVERSE") <- TRUE


### =========================================================
### 3. BASE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

base_sny22_yld3_row12 <- asreml(
  fixed    = yld3 ~ 1 + Rep ,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld3_row12)

summary(base_sny22_yld3_row12)$varcomp


### Extract BLUPs
blup_base_sny22_yld3_row12 <- coef(base_sny22_yld3_row12)$random

blup_base_sny22_yld3_row12 <- blup_base_sny22_yld3_row12[
  grep("Germplasm", rownames(blup_base_sny22_yld3_row12)),
  ,
  drop = FALSE]

g_base_sny22_yld3_row12 <- blup_base_sny22_yld3_row12[,1]
names(g_base_sny22_yld3_row12) <- rownames(blup_base_sny22_yld3_row12)



### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

base_sny22_yld3_row34 <- asreml(
  fixed    = yld3 ~ 1 + Rep ,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld3_row34)

summary(base_sny22_yld3_row34)$varcomp


### Extract BLUPs
blup_base_sny22_yld3_row34 <- coef(base_sny22_yld3_row34)$random

blup_base_sny22_yld3_row34 <- blup_base_sny22_yld3_row34[
  grep("Germplasm", rownames(blup_base_sny22_yld3_row34)),
  ,
  drop = FALSE]

g_base_sny22_yld3_row34 <- blup_base_sny22_yld3_row34[,1]
names(g_base_sny22_yld3_row34) <- rownames(blup_base_sny22_yld3_row34)


### ---------------------------------------------------------
### GEER: Base model
### ---------------------------------------------------------

common_base_sny22_yld3 <- intersect(names(g_base_sny22_yld3_row12), names(g_base_sny22_yld3_row34))
common_base_sny22_yld3
cor(g_base_sny22_yld3_row12[common_base_sny22_yld3], g_base_sny22_yld3_row34[common_base_sny22_yld3])




### =========================================================
### 4. SPATIAL MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

spa_sny22_yld3_row12 <- asreml(
  fixed    = yld3 ~ 1 + Rep ,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range),
  residual = ~ id(units),
  data     = sny22_yld3_row12)

summary(spa_sny22_yld3_row12)$varcomp


### Extract BLUPs
blup_spa_sny22_yld3_row12 <- coef(spa_sny22_yld3_row12)$random
blup_spa_sny22_yld3_row12
blup_spa_sny22_yld3_row12 <- blup_spa_sny22_yld3_row12[
  grep("Germplasm", rownames(blup_spa_sny22_yld3_row12)),
  ,
  drop = FALSE
]

g_spa_sny22_yld3_row12 <- blup_spa_sny22_yld3_row12[,1]
names(g_spa_sny22_yld3_row12) <- rownames(blup_spa_sny22_yld3_row12)
g_spa_sny22_yld3_row12


### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

spa_sny22_yld3_row34 <- asreml(
  fixed    = yld3 ~ 1,
  random   = ~ vm(Germplasm, grm_inverse) + ar1(Range), #:ar1(Row)
  residual = ~ id(units),
  data     = sny22_yld3_row34)

summary(spa_sny22_yld3_row34)$varcomp


### Extract BLUPs
blup_spa_sny22_yld3_row34 <- coef(spa_sny22_yld3_row34)$random

blup_spa_sny22_yld3_row34 <- blup_spa_sny22_yld3_row34[
  grep("Germplasm", rownames(blup_spa_sny22_yld3_row34)),
  ,
  drop = FALSE]

g_spa_sny22_yld3_row34 <- blup_spa_sny22_yld3_row34[,1]
names(g_spa_sny22_yld3_row34) <- rownames(blup_spa_sny22_yld3_row34)



### ---------------------------------------------------------
### GEER: Spatial model
### ---------------------------------------------------------

common_spa_sny22_yld3 <- intersect(names(g_spa_sny22_yld3_row12), names(g_spa_sny22_yld3_row34))

cor(g_spa_sny22_yld3_row12[common_spa_sny22_yld3], g_spa_sny22_yld3_row34[common_spa_sny22_yld3])


### =========================================================
### 5. PEE MODEL
### =========================================================

### ---------------------------------------------------------
### Block 1 (Rows 1 + 2)
### ---------------------------------------------------------

pee_sny22_yld3_row12 <- asreml(
  fixed    = yld3 ~ 1 + Rep + PEE_260,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld3_row12)

summary(pee_sny22_yld3_row12)$varcomp


### Extract BLUPs
blup_pee_sny22_yld3_row12 <- coef(pee_sny22_yld3_row12)$random

blup_pee_sny22_yld3_row12 <- blup_pee_sny22_yld3_row12[
  grep("Germplasm", rownames(blup_pee_sny22_yld3_row12)),
  ,
  drop = FALSE]

g_pee_sny22_yld3_row12 <- blup_pee_sny22_yld3_row12[,1]
names(g_pee_sny22_yld3_row12) <- rownames(blup_pee_sny22_yld3_row12)



### ---------------------------------------------------------
### Block 2 (Rows 3 + 4)
### ---------------------------------------------------------

pee_sny22_yld3_row34 <- asreml(
  fixed    = yld3 ~ 1 + Rep + PEE_260,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = sny22_yld3_row34)

summary(pee_sny22_yld3_row34)$varcomp


### Extract BLUPs
blup_pee_sny22_yld3_row34 <- coef(pee_sny22_yld3_row34)$random

blup_pee_sny22_yld3_row34 <- blup_pee_sny22_yld3_row34[
  grep("Germplasm", rownames(blup_pee_sny22_yld3_row34)),
  ,
  drop = FALSE]

g_pee_sny22_yld3_row34 <- blup_pee_sny22_yld3_row34[,1]
names(g_pee_sny22_yld3_row34) <- rownames(blup_pee_sny22_yld3_row34)
blup_pee_sny22_yld3_row34


### ---------------------------------------------------------
### GEER: PEE model
### ---------------------------------------------------------

common_pee_sny22_yld3 <- intersect(names(g_pee_sny22_yld3_row12), names(g_pee_sny22_yld3_row34))
common_pee_sny22_yld3
cor( g_pee_sny22_yld3_row12[common_pee_sny22_yld3], g_pee_sny22_yld3_row34[common_pee_sny22_yld3])
# cor(blup_pee_sny22_yld3_row34,  blup_pee_sny22_yld3_row12)


### =========================================================
### 6. Final GEER summary table
### =========================================================

GEER_rowblock_results_sny22_yld3 <- data.frame(

  Model = c("Base", "Spatial", "PEE"),

  Pearson = c(
    cor(g_base_sny22_yld3_row12[common_base_sny22_yld3],
        g_base_sny22_yld3_row34[common_base_sny22_yld3]),

    cor(g_spa_sny22_yld3_row12[common_spa_sny22_yld3],
        g_spa_sny22_yld3_row34[common_spa_sny22_yld3]),

    cor(g_pee_sny22_yld3_row12[common_pee_sny22_yld3],
        g_pee_sny22_yld3_row34[common_pee_sny22_yld3])))

GEER_rowblock_results_sny22_yld3
```

### 3. Validity of GEER Results (block wise results)
0.93, 0.9, 0.95 & 0.91 and 0.94 & 0.92

```{r Validation of GEER results using all data base model, sny22 yld3, 06072026}

### =========================================================
### VALIDATION OF GEER RESULTS
### Snyder NY 2022
### Harvest 1 (yld3)
###
### Goal:
###   Compare genomic BLUPs from:
###     1. Base model using all data
###     2. Spatial model using each block
###     3. PEE model using each block
###
### Validation:
###   Correlate block-specific BLUPs against
###   baseline genomic BLUPs from full dataset
### =========================================================


### ---------------------------------------------------------
### 1. Ensure inverse GRM
### ---------------------------------------------------------

asreml.options(extra = 200, maxit = 200)

attr(grm_inverse, "INVERSE") <- TRUE



### ---------------------------------------------------------
### 2. Create row-block subsets
### ---------------------------------------------------------

sny22_yld3_row12 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(1, 2))

sny22_yld3_row34 <- ny_sny_pheno_22 %>%
  filter(Row %in% c(3, 4))



### =========================================================
### 3. BASE MODEL USING ALL DATA
### =========================================================

sny22_yld3_base_all <- asreml(
  fixed    = yld3 ~ 1 + Rep,
  random   = ~ vm(Germplasm, grm_inverse),
  residual = ~ id(units),
  data     = ny_sny_pheno_22)

summary(sny22_yld3_base_all)$varcomp


### Extract genomic BLUPs
sny22_yld3_gblup_base_all <- coef(sny22_yld3_base_all)$random

sny22_yld3_gblup_base_all <- sny22_yld3_gblup_base_all[
  grep("Germplasm", rownames(sny22_yld3_gblup_base_all)),
  ,
  drop = FALSE]

sny22_yld3_gblup_base_all <- sny22_yld3_gblup_base_all[,1]

names(sny22_yld3_gblup_base_all) <- 
  rownames(sny22_yld3_gblup_base_all)
sny22_yld3_gblup_base_all

##### part of spatial model and PEE model in above chunk...

### =========================================================
### 4. VALIDATION CORRELATIONS
### =========================================================

### Base correlations 
sny22_yld3_cor_base <- c(
  cor(g_base_sny22_yld3_row12, sny22_yld3_gblup_base_all),
  cor(g_base_sny22_yld3_row34, sny22_yld3_gblup_base_all))
sny22_yld3_cor_base

### Spatial correlations (Ar)
sny22_yld3_cor_spa <- c(
  cor(g_spa_sny22_yld3_row12, sny22_yld3_gblup_base_all),
  cor(g_spa_sny22_yld3_row34, sny22_yld3_gblup_base_all))

sny22_yld3_cor_spa



### PEE correlations
sny22_yld3_cor_pee <- c(
  cor(g_pee_sny22_yld3_row12, sny22_yld3_gblup_base_all),
  cor(
    g_pee_sny22_yld3_row34, sny22_yld3_gblup_base_all))
sny22_yld3_cor_pee



### =========================================================
### 5. Summary table
### =========================================================

sny22_yld3_validation_results <- data.frame(

  Model = c("Base_Row12", "Base_Row12", 
            "Spatial_Row12",
            "Spatial_Row34",
            "PEE_Row12",
            "PEE_Row34"),

  Pearson = c(
    sny22_yld3_cor_base[1],
    sny22_yld3_cor_base[2],
    sny22_yld3_cor_spa[1],
    sny22_yld3_cor_spa[2],
    sny22_yld3_cor_pee[1],
    sny22_yld3_cor_pee[2]))

sny22_yld3_validation_results

```




#### PLOTS 
### 1. GEER for all harvest (Raw)

```{r GEER all harvests Sny22, 05222026}
### =========================================================
### Boxplot of GEER across harvests
### Snyder NY 2022
### =========================================================

### ---------------------------------------------------------
### 1. Create data frame
### ---------------------------------------------------------

GEER_sny22_allharvests <- data.frame(
  Harvest = c("Harvest1", "Harvest2", "Harvest3"),
  Base = c(0.7, 0.65, 0.7),
  Spatial = c(0.72, 0.76, 0.77),
  PEE = c(0.73, 0.78, 0.76))
GEER_sny22_allharvests



### ---------------------------------------------------------
### 2. Convert to long format
### ---------------------------------------------------------

GEER_sny22_allharvests_long <- 
  GEER_sny22_allharvests %>%
  
  pivot_longer(
    cols = c(Base, Spatial, PEE),
    names_to = "Model",
    values_to = "GEER")

GEER_sny22_allharvests_long

### Set order of boxplots
GEER_sny22_allharvests_long$Model <- factor(
  GEER_sny22_allharvests_long$Model,
  levels = c("Base", "Spatial", "PEE"))

### ---------------------------------------------------------
### 3. Boxplot
### ---------------------------------------------------------

ggplot(
  GEER_sny22_allharvests_long,
  aes(x = Model,
      y = GEER,
      fill = Model)) +

  geom_boxplot(alpha = 0.7) +

  geom_point(
    color = "red",
    shape = 18,
    size = 2,
    position = position_jitter(width = 0.05)) +

  scale_fill_manual(values = c(
    "Base"    = "grey80",
    "Spatial" = "lightblue3",
    "PEE"     = "lightgreen")) +

  labs(
    title = "Snyder22 All Harvests: Model Comparison using GEER",
    x = "Model",
    y = "GEER (Pearson correlation)") +

  theme_bw(base_size = 12) +

  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5),

    legend.position = "none")
```

### 2. GEER as deviation, ΔGEER for all harvests

```{r GEER as deviation for all harvests sny22, 05222026}

### =========================================================
### Boxplot of GEER deviation across harvests
### Snyder NY 2022
### =========================================================

### NOTE:
### Deviation calculated as:
###
###   Model GEER - Base GEER
###
### Positive values indicate improvement over base model.
### =========================================================



### ---------------------------------------------------------
### 1. Create data frame
### ---------------------------------------------------------

GEER_deviation_sny22_allharvests <- data.frame(
  Harvest = c("Harvest1", "Harvest2","Harvest3"),
  Spatial = c(0.02, 0.11, 0.07), #!!!!!!!!!!!!!
  PEE = c( 0.03, 0.13, 0.06)) #!!!!!!!!!!!!!

GEER_deviation_sny22_allharvests


### ---------------------------------------------------------
### 2. Convert to long format
### ---------------------------------------------------------

GEER_deviation_sny22_allharvests_long <- 
  GEER_deviation_sny22_allharvests %>%
  
  pivot_longer(
    cols = c(Spatial, PEE),
    names_to = "Model",
    values_to = "Deviation")

GEER_deviation_sny22_allharvests_long



### ---------------------------------------------------------
### 3. Set order of boxplots
### ---------------------------------------------------------

GEER_deviation_sny22_allharvests_long$Model <- factor(
  GEER_deviation_sny22_allharvests_long$Model,
  levels = c("Spatial", "PEE"))



### ---------------------------------------------------------
### 4. Boxplot
### ---------------------------------------------------------

ggplot(
  GEER_deviation_sny22_allharvests_long,
  aes(x = Model,
      y = Deviation,
      fill = Model)) +
  geom_boxplot(alpha = 0.7) +
  geom_point(
    color = "red",
    shape = 18,
    size = 2,
    position = position_jitter(width = 0.05)) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed") +
  scale_fill_manual(values = c(
    "Spatial" = "lightblue3",
    "PEE"     = "lightgreen")) +
  labs(
    title = "Snyder22 All Harvests: GEER Deviation from Base Model",
    x = "Model",
    y = "GEER Deviation") +
  theme_bw(base_size = 12) +

  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5),
    legend.position = "none")

```

For paper, making more consistent GEER as deviation (only ggplot changed)

```{r GEER as deviation for all harvests sny22, 06072026}

### =========================================================
### Boxplot of GEER deviation across harvests
### Snyder NY 2022
### =========================================================

### NOTE:
### Deviation calculated as:
###
###   Model GEER - Base GEER
###
### Positive values indicate improvement over base model.
### =========================================================



### ---------------------------------------------------------
### 1. Create data frame
### ---------------------------------------------------------

GEER_deviation_sny22_allharvests <- data.frame(
  Harvest = c("Harvest1", "Harvest2","Harvest3"),
  Spatial = c(0.02, 0.11, 0.07), #!!!!!!!!!!!!!
  PEE = c( 0.03, 0.13, 0.06)) #!!!!!!!!!!!!!

GEER_deviation_sny22_allharvests


### ---------------------------------------------------------
### 2. Convert to long format
### ---------------------------------------------------------

GEER_deviation_sny22_allharvests_long <- 
  GEER_deviation_sny22_allharvests %>%
  
  pivot_longer(
    cols = c(Spatial, PEE),
    names_to = "Model",
    values_to = "Deviation")

GEER_deviation_sny22_allharvests_long



### ---------------------------------------------------------
### 3. Set order of boxplots
### ---------------------------------------------------------

GEER_deviation_sny22_allharvests_long$Model <- factor(
  GEER_deviation_sny22_allharvests_long$Model,
  levels = c("Spatial", "PEE"))



### ---------------------------------------------------------
### 4. Boxplot
### ---------------------------------------------------------

gg_GEER_sny22 <- ggplot(GEER_deviation_sny22_allharvests_long, aes(x = Model, y = Deviation, fill = Model)) +
  geom_boxplot(width = 0.3, alpha = 0.9,outlier.shape = NA) +
  scale_fill_manual(values = c(
    "Base" = "gray",
    "Spatial" = "lightblue",
    "PEE"     = "lightgreen")) +
  labs(
    title = "Sny22 All Harvests: GEER Deviation from Base Model",
    x = "Model",
    y = "GEER Deviation") +
  theme_bw(base_size = 13) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2)) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5),
    legend.position = "right")
gg_GEER_sny22

ggsave(filename = "GEER_boxplots_Snyder_2022_NMSU.png", plot = gg_GEER_sny22, path = "2022/output_images", width = 10, height = 10,dpi = 400) 
```


### 3. Validation of GEER by checking with base model gblups
```{r Validation of GEER all harvests sny22, 06072026}
### =========================================================
### Boxplot of Validation GEER across harvests
### Snyder NY 2022
### =========================================================

### NOTE:
### These values represent correlations between:
###   1. Spatial model BLUPs vs Base model BLUPs
###   2. PEE model BLUPs vs Base model BLUPs
###
### Therefore only Spatial and PEE are plotted.
### =========================================================



### ---------------------------------------------------------
### 1. Create data frame
### ---------------------------------------------------------

ValGEER_sny22_allharvests <- data.frame(
  Harvest = c("Harvest1",
              "Harvest2",
              "Harvest3"),
  Base = c(
    mean(c(0.93, 0.89)),
    mean(c(0.9, 0.9)),
    mean(c(0.93, 0.9))),
  Spatial = c(
    mean(c(0.92, 0.9)),
    mean(c(0.92, 0.92)),
    mean(c(0.95, 0.91))),
  PEE = c(
    mean(c(0.93, 0.89)),
    mean(c(0.92, 0.91)),
    mean(c(0.94, 0.92))))

ValGEER_sny22_allharvests



### ---------------------------------------------------------
### 2. Convert to long format
### ---------------------------------------------------------

ValGEER_sny22_allharvests_long <- 
  ValGEER_sny22_allharvests %>%
  
  pivot_longer(
    cols = c(Base, Spatial, PEE),
    names_to = "Model",
    values_to = "ValGEER")

ValGEER_sny22_allharvests_long



### ---------------------------------------------------------
### 3. Set order of boxplots
### ---------------------------------------------------------

ValGEER_sny22_allharvests_long$Model <- factor(
  ValGEER_sny22_allharvests_long$Model,
  levels = c("Base", "Spatial", "PEE"))



### ---------------------------------------------------------
### 4. Boxplot
### ---------------------------------------------------------

gg_ValGEER_sny22 <- ggplot(ValGEER_sny22_allharvests_long, aes(x = Model, y = ValGEER, fill = Model)) +
  geom_boxplot(width = 0.3, alpha = 0.9,outlier.shape = NA) +
  scale_fill_manual(values = c(
    "Base" = "gray",
    "Spatial" = "lightblue",
    "PEE"     = "lightgreen")) +
  labs(
    title = "SNY22 All Harvests: Validation GEER",
    x = "Model",
    y = "Validation GEER (correlation)") +
  theme_bw(base_size = 13) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2)) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5),
    legend.position = "right")
gg_ValGEER_sny22

ggsave(filename = "ValGEER_boxplots_Snyder_2022_NMSU.png", plot = gg_ValGEER_sny22, path = "2022/output_images", width = 10, height = 10,dpi = 400) 
    
```






####### For Manuscript  #######

####### MAKE COMBINED PLOTS FOR RESULTS of NY from both years...  #######

each trial AR and RR together bar graph, then another trial, like wise 8 pairs for 8 trials

1. GEER

```{r Combine all GEER deviations from 4 trials of NY, 06172026}
# combine data first
geer_all_ny <- data.frame(
  Trial = c(rep("MUS22",6),
            rep("SNY22",6),
            rep("MUS23",6),
            rep("SNY23",6)),
  Harvest = rep(c("Harvest1","Harvest2","Harvest3"), 8),
  Model = rep(c("AR","RR"), each = 3, times = 4),
  Delta = c(
    # MUS22
    0.06, 0.14, -0.02,
    0.45, 0.17, 0.18,

    # SNY22
    0.02, 0.11, 0.07,
    0.03, 0.13, 0.06,

    # MUS23
    0.05, 0.03, 0.04,
    0.04, 0.00, 0.16,

    # SNY23
    0.15, -0.02, 0.03,
    0.15, -0.01, 0.09 ))

# GGPLOT
geer_all_ny$Model <- factor(
  geer_all_ny$Model,
  levels = c("AR","RR"))

geer_all_ny$Trial <- factor(
  geer_all_ny$Trial,
  levels = c("MUS22","SNY22","MUS23","SNY23"))


gg_GEER_deviation_all_ny <- ggplot(geer_all_ny,
                                   aes(x=Trial,y=Delta,fill=Model)) +
  geom_boxplot(width=0.5,alpha=0.9,outlier.shape=NA) +
  scale_fill_manual(values=c(
    "AR"="lightblue",
    "RR"="lightgreen")) +
  labs(
    title="GEER Deviation from Base Model in NY Trials",
    x="Trials",
    y=expression(Delta*"GEER")) +
  theme_bw(base_size=17) +
  scale_y_continuous(
    limits=c(-0.1,0.5),
    breaks=seq(-0.1,0.5,by=0.1)) +
  theme(
    plot.title=element_text(face="bold",hjust=0.5,size=20),
    axis.text=element_text(size=18,color="black"),
    axis.title=element_text(size=20,face="bold"),
    legend.title=element_text(size=18,face="bold"),
    legend.text=element_text(size=16),
    legend.position="right" )

gg_GEER_deviation_all_ny

ggsave(filename = "Combined_GEER_deviation_boxplots_All_Trials_NY.png", plot = gg_GEER_deviation_all_ny, path = "2022/output_images", width = 10, height = 10,dpi = 400) 

```





2. ValGEER

```{r Combine all valGEER deviations from 4 trials of NY, 06172026}
### =========================================================
### Combine ValGEER from all NY trials
### =========================================================

valGEER_all_ny <- data.frame(

  Trial = c(rep("MUS22",18),
            rep("SNY22",18),
            rep("MUS23",18),
            rep("SNY23",18)),

  Harvest = rep(c("Harvest1","Harvest2","Harvest3"), each = 6, times = 4),

  Model = rep(c("Base","Base","AR","AR","RR","RR"), times = 12),

  Value = c(

    # =====================
    # MUS22
    # =====================
    0.82,0.83,0.91,0.85,0.99,0.99,
    0.76,0.78,0.72,0.54,0.81,0.69,
    0.78,0.75,0.76,0.69,0.72,0.77,

    # =====================
    # SNY22
    # =====================
    0.93,0.89,0.92,0.90,0.93,0.90,
    0.90,0.90,0.91,0.92,0.92,0.91,
    0.93,0.90,0.95,0.91,0.94,0.92,

    # =====================
    # MUS23
    # =====================
    0.94,0.83,0.88,0.84,0.83,0.92,
    0.93,0.90,0.93,0.91,0.94,0.86,
    0.88,0.86,0.88,0.86,0.93,0.85,

    # =====================
    # SNY23
    # =====================
    0.88,0.84,0.82,0.85,0.92,0.87,
    0.94,0.91,0.94,0.90,0.92,0.92,
    0.89,0.92,0.90,0.89,0.89,0.94 ))

valGEER_all_ny$Model <- factor(
  valGEER_all_ny$Model,
  levels = c("Base","AR","RR"))

valGEER_all_ny$Trial <- factor(
  valGEER_all_ny$Trial,
  levels = c("MUS22","SNY22","MUS23","SNY23"))

# GGPLOT

gg_ValGEER_all_ny <- ggplot(valGEER_all_ny, aes(x = Trial, y = Value, fill = Model)) +
geom_boxplot(width = 0.55, alpha = 0.9, outlier.shape = NA) +
  scale_fill_manual(
    values = c(
      "Base" = "lightgray",
      "AR"   = "lightblue",
      "RR"   = "lightgreen")) +
  labs(
    title = "Validation GEER in NY Trials",
    x = "Trials",
    y = "Validation GEER") +
  theme_bw(base_size = 17) +
  scale_y_continuous(
    limits = c(0.2, 1.0),
    breaks = seq(0.2, 1.0, by = 0.2)) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 20),
    axis.text = element_text(size = 18, color = "black"),
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 16),
    legend.position = "right")

gg_valGEER_all_ny

ggsave(filename = "Combined_valGEER_deviation_boxplots_All_Trials_NY.png", plot = gg_valGEER_all_ny, path = "2022/output_images", width = 12, height = 10,dpi = 400) 

```




