---
title: "Plots"
author: "Dinesh Ghimire"
date: "2025-07-31"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familar with those before using this script
# List of required packages
required_packages <- c("dplyr", "lme4", "tidyr", "ggplot2", "readxl", "viridis","grid", "gridExtra")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```

#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NMSU"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NMSU 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa_NMSU"
setwd("/Users/dg663/FFAR-Alfalfa_NMSU")
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is missing please let me know. 
key<-read.csv("2023/meta_data/cultivar_list_key.csv") # contains both CU and NMSU cultivars here. 
# its same for 2022 and 2023 (cultivar list)
```


#NMSU 23


# NMSU 23 Normal Irrigation
## Now lets process "Normal" treatment

################################# Normal Irrigation Treatment ################################




Before starting, lets look at how VI's (1st chunk) and yield (2nd chunk) traits look like. 
```{r visualizing VI's in heatmaps}
# read the final combined version of file and drop 1st unnecessary index column
nmsu_nor_pheno_23 <- read.csv("2023/MSI_data/output_files/nmsu23_nor_combined_phenotypes_MSI.csv")  %>%
  dplyr::select(-X)
colnames(nmsu_nor_pheno_23)
# need to convert into factors
nmsu_nor_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu_nor_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(nmsu_nor_pheno_23) # to ensure
# now we need to make heatmaps of each NDVI, NDRE, GNDVI separately. For that, at first, need to change into long format
#first need to change into long format so that we can use it for ggplot2 and also select VI's

nmsu_nor_long_all <- nmsu_nor_pheno_23 %>%
  pivot_longer(cols = matches("^(NDVI|NDRE|GNDVI)_\\d+$"),
               names_to = "Index_Time",
               values_to = "Value") %>%
  separate(Index_Time, into = c("Index", "Time"), sep = "_") %>%
  mutate(Time = as.numeric(Time))  # Ensure Time is numeric for sorting

# then since we need to create heatmaps for each VI's separately to observe each heatmaps with more resolution, lets separate out for each VI and do all
# Subset each index
ndvi_nor_data <- nmsu_nor_long_all %>% filter(Index == "NDVI")
gndvi_nor_data <- nmsu_nor_long_all %>% filter(Index == "GNDVI")
ndre_nor_data <- nmsu_nor_long_all %>% filter(Index == "NDRE")
  # all three have same rows and checked back to merged file, the values look good.

# Plot faceted heatmaps
 # 1. for NDVI
gg_ndvi_nor <- ggplot(ndvi_nor_data, aes(x = Row, y = Range, fill = Value)) +
  geom_tile() +
  scale_fill_viridis(option = "D", na.value = "grey90") +
  facet_wrap(~Time, ncol = 8) +
  labs(title = "NDVI Heatmaps Across Time NMSU NOR 2023",
       x = "Row", y = "Range", fill = "NDVI") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        plot.title = element_text(hjust = 0.5, face = "bold"))

gg_ndvi_nor
# save the file
ggsave(filename = "Raw_NDVI_Heatmap_nmsu_nor_2023.png", plot = gg_ndvi_nor, path = "2023/output_images", width = 15, height = 12,dpi = 400)

# 2. for GNDVI
gg_gndvi_nor <- ggplot(gndvi_nor_data, aes(x = Row, y = Range, fill = Value)) +
  geom_tile() +
  scale_fill_viridis(option = "D", na.value = "grey90") +
  facet_wrap(~Time, ncol = 8) +
  labs(title = "GNDVI Heatmaps Across Time NMSU NOR 2023",
       x = "Row", y = "Range", fill = "GNDVI") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        plot.title = element_text(hjust = 0.5, face = "bold"))

gg_gndvi_nor
# save the file
ggsave(filename = "Raw_GNDVI_Heatmap_nmsu_nor_2023.png", plot = gg_gndvi_nor, path = "2023/output_images", width = 15, height = 12,dpi = 400)


# 3. for NDRE
gg_ndre_nor <- ggplot(ndre_nor_data, aes(x = Row, y = Range, fill = Value)) +
  geom_tile() +
  scale_fill_viridis(option = "D", na.value = "grey90") +
  facet_wrap(~Time, ncol = 8) +
  labs(title = "NDRE Heatmaps Across Time NMSU NOR 2023",
       x = "Row", y = "Range", fill = "NDRE") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        plot.title = element_text(hjust = 0.5, face = "bold"))

gg_ndre_nor
# save the file
ggsave(filename = "Raw_NDRE_Heatmap_nmsu_nor_2023.png", plot = gg_ndre_nor, path = "2023/output_images", width = 15, height = 12,dpi = 400)

```


*Plot the agronomic traits value i.e. yield here*

```{r heatmap for yield data}
# its easy to use combined yield file and not combined all pheno file (i can load all combined and select only yld part too)

#for yield data
nmsu_nor_23_yld <- read.csv("2023/GroundTruth_data/Normal_Irrigation/nmsu23_nor_yield_merged.csv") %>%
  dplyr::select((-X))
colnames(nmsu_nor_23_yld)
# rename
nmsu_nor_23_yld <- nmsu_nor_23_yld %>% 
  rename( Row = row, Range = range)
# Convert plot info to factors 
nmsu_nor_23_yld[, c("Range", "Rep", "Plot", "Row")] <- 
  lapply(nmsu_nor_23_yld[, c("Range", "Rep", "Plot", "Row")], as.factor)

#lets make this wide format to longer format
nmsu_nor_23_yld_long <- gather(nmsu_nor_23_yld, harvest, value, yld1:yld6)

ggplot(nmsu_nor_23_yld_long, aes(x=harvest, y = value)) +
  geom_point()

# now lets make complete heatmap
gg_yld_heatmap_nor <- ggplot(nmsu_nor_23_yld_long, aes(x = Row, y = Range, fill = value)) +
  geom_tile() +
  scale_fill_viridis(option = "C", name = "Yield", na.value = "grey80") +
  facet_wrap(~harvest, ncol = 6) +
  labs(title = "NMSU 23 Nor Yield Heatmaps (YLD1 to YLD6)", x = "Row", y = "Range") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6))

gg_yld_heatmap_nor
# save the file
ggsave(filename = "Raw_Yields_Heatmap_nmsu_2023.png", plot = gg_yld_heatmap_nor, path = "2023/output_images", width = 15, height = 12,dpi = 400)

```





difference between NDVI, GNDVI, and NDRE mean values
```{r difference between NDVI, GNDVI, and NDRE mean values}

# Calculate mean differences between vegetation indices (NDVI, NDRE, GNDVI) over time
mean_diffs_nor <- nmsu_nor_long_all %>%
  filter(Index %in% c("NDVI", "NDRE", "GNDVI")) %>% # filter three indices
  group_by(Time, Index) %>% # groyup by time and index
  summarise(mean_value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Index, values_from = mean_value) %>% # convert from long to wide format
  mutate(
    NDVI_minus_NDRE = NDVI - NDRE,
    NDVI_minus_GNDVI = NDVI - GNDVI,
    GNDVI_minus_NDRE = GNDVI - NDRE
  ) %>%
  arrange(Time)
# Convert wide to long for plotting all 3 comparisons
mean_diffs_nor_long <- mean_diffs_nor %>%
  dplyr::select(Time, NDVI_minus_NDRE, NDVI_minus_GNDVI, GNDVI_minus_NDRE) %>%
  pivot_longer(cols = -Time, names_to = "Comparison", values_to = "Difference")

# Plot
ggplot(mean_diffs_nor_long, aes(x = Time, y = Difference, color = Comparison)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Mean Differences Between Vegetation Indices Over Time",
       x = "Time (DOY or DAP)", y = "Mean Difference") +
  theme_minimal() +
  scale_color_manual(values = c(
    "NDVI_minus_NDRE" = "green",
    "NDVI_minus_GNDVI" = "blue",
    "GNDVI_minus_NDRE" = "orange"
  )) +
  theme(legend.title = element_blank())

```



### 09052025
# Summarizing univaraite models (h2) and mt models (GC, SC, RC) in graphs
 ## Univariate: plot heritabilities
```{r 2023 Normal Trend Graphs, 09052025}
# unfiltered read data
herit_results_nor_23 <- read.csv("2023/MSI_data/output_files/Heritabilities_MSI_filtered_aa_nor_23.csv")

df_long_nor_23 <- herit_results_nor_23 %>%
  pivot_longer(
    cols = -Time,
    names_to = "Index_Model",
    values_to = "Heritability"
  ) %>%
  mutate(
    Model = ifelse(grepl("_spa$", Index_Model), "With Spatial", "No Spatial"),
    Index = case_when(
      grepl("gndvi", Index_Model, ignore.case = TRUE) ~ "GNDVI",
      grepl("ndvi", Index_Model, ignore.case = TRUE) ~ "NDVI",
      grepl("ndre", Index_Model, ignore.case = TRUE) ~ "NDRE"
    )
  )

# plot
ggplot_nor_23 <- ggplot(df_long_nor_23, aes(x = Time, y = Heritability, color = Model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~Index, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c("No Spatial" = "red", "With Spatial" = "blue")) +
  theme_minimal(base_size = 14) +
  labs(x = "GDD (Time)",
       y = expression(h^2~" (Heritability)"),
       color = "Model",
       title = "Heritability Trajectories (2023, Normal Irrigation)")
ggplot_nor_23
# ggsave
ggsave("2023/output_images/Heritability_Trends_All_Indices_NMSU_Nor_2023.png", ggplot_nor_23, width = 14, height = 10, dpi = 400)

```


### Mt models graphs showing genetic correlation, spatial correlation (if) and residual correlation

```{r Mt models graphs GNDVI NOR 2023 09052025}
mt_models_gndvi_nor_23 <- read.csv("2023/MSI_data/output_files/MT_models_GC_SC_RC_values_GNDVI_nor_23.csv")

# Use your dataframe
df_gndvi_nor_23 <- mt_models_gndvi_nor_23
# Preserve order of Trait
df_gndvi_nor_23$Trait <- factor(df_gndvi_nor_23$Trait, levels = df_gndvi_nor_23$Trait)

# ---- Graph 1: Genetic1 & Residual.2 ----
df_gndvi_nor_23_no_spa <- df_gndvi_nor_23 %>%
  select(Trait, Genetic1, Residual.2) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value")

p1 <- ggplot(df_gndvi_nor_23_no_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI without Spatial NOR 2023",
       y = "Correlation")

# ---- Graph 2: Genetic, Spatial & Residual ----
df_gndvi_nor_23_spa <- df_gndvi_nor_23 %>%
  select(Trait, Genetic, Spatial, Residual) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value") %>%
  filter(!is.na(Value))

p2 <- ggplot(df_gndvi_nor_23_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI with Spatial NOR 2023",
       y = "Correlation")

# Show both plots
p1
p2

```




## Graphs: Genetic and Spatial trend throughout the season
```{r Mt models graphs GNDVI Normal 2023 06042026}
mt_models_gndvi_nor_23 <- read.csv("2023/MSI_data/output_files/MT_models_GC_SC_values_GNDVI_Normal_2023.csv")
colnames(mt_models_gndvi_nor_23)
# Use your dataframe
df_gndvi_nor_23 <- mt_models_gndvi_nor_23
# Preserve order of Trait
df_gndvi_nor_23$Trait <- factor(df_gndvi_nor_23$Trait, levels = df_gndvi_nor_23$Trait)

# ---- Graph 1: Genetic1 only ----
df_gndvi_nor_23_no_spa <- df_gndvi_nor_23 %>%
  select(Trait, Genetic) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value")

gndvi_nor_23_p1 <- ggplot(df_gndvi_nor_23_no_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI without Spatial Normal 2023",
       y = "Correlation")
gndvi_nor_23_p1
# ---- Graph 2: Genetic, & Spatial  ----
df_gndvi_nor_23_spa <- df_gndvi_nor_23 %>%
  select(Trait, Genetic, Spatial) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value") %>%
  filter(!is.na(Value))

gndvi_nor_23_p2 <- ggplot(df_gndvi_nor_23_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Genetic" = "forestgreen", 
                                "Spatial" = "darkorange")) +
  scale_linetype_manual(values = c("Genetic" = "solid", 
                                   "Spatial" = "solid")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "GNDVI Trend over Normal 2023",
       y = "Correlation")

# Show both plots
gndvi_nor_23_p1 # only genetic
gndvi_nor_23_p2 # genetic and spatial

# ggsave
ggsave("2023/output_images/Genetic_Spatial_Correlations_NMSU_Normal_2023.png", gndvi_nor_23_p2, width = 8, height = 6, dpi = 400)
```















################################# Early Termination ################################





# 2. Now lets process "Early Termination" treatment

## NMSU ET Irrigation 2023




Before starting, lets look at how VI's (1st chunk) and yield (2nd chunk) traits look like. 

```{r visualizing VI's in heatmaps NMSU_ET_2023 062625}
# read the final combined version of file and drop 1st unnecessary index column
nmsu_et_pheno_23 <- read.csv("2023/MSI_data/output_files/nmsu23_et_combined_phenotypes_MSI.csv")  %>%
  dplyr::select(-X)
# need to convert into factors
nmsu_et_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu_et_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(nmsu_et_pheno_23) # to ensure
# now we need to make heatmaps of each NDVI, NDRE, GNDVI separately. For that, at first, need to change into long format
#first need to change into long format so that we can use it for ggplot2 and also select VI's

nmsu_et_long_all <- nmsu_et_pheno_23 %>%
  pivot_longer(cols = matches("^(NDVI|NDRE|GNDVI)_\\d+$"),
               names_to = "Index_Time",
               values_to = "Value") %>%
  separate(Index_Time, into = c("Index", "Time"), sep = "_") %>%
  mutate(Time = as.numeric(Time))  # Ensure Time is numeric for sorting

# then since we need to create heatmaps for each VI's separately to observe each heatmaps with more resolution, lets separate out for each VI and do all
# Subset each index
ndvi_et_data <- nmsu_et_long_all %>% filter(Index == "NDVI")
gndvi_et_data <- nmsu_et_long_all %>% filter(Index == "GNDVI")
ndre_et_data <- nmsu_et_long_all %>% filter(Index == "NDRE")
  # all three have same rows and checked back to merged file, the values look good.

# Plot faceted heatmaps
 # 1. for NDVI
gg_ndvi_et <- ggplot(ndvi_et_data, aes(x = Row, y = Range, fill = Value)) +
  geom_tile() +
  scale_fill_viridis(option = "D", na.value = "grey90") +
  facet_wrap(~Time, ncol = 8) +
  labs(title = "NDVI Heatmaps Across Time NMSU ET 2023",
       x = "Row", y = "Range", fill = "NDVI") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        plot.title = element_text(hjust = 0.5, face = "bold"))

gg_ndvi_et
# save the file
ggsave(filename = "Raw_NDVI_Heatmap_nmsu_et_2023.png", plot = gg_ndvi_et, path = "2023/output_images", width = 15, height = 12,dpi = 400)

# 2. for GNDVI
gg_gndvi_et <- ggplot(gndvi_et_data, aes(x = Row, y = Range, fill = Value)) +
  geom_tile() +
  scale_fill_viridis(option = "D", na.value = "grey90") +
  facet_wrap(~Time, ncol = 8) +
  labs(title = "GNDVI Heatmaps Across Time NMSU ET 2023",
       x = "Row", y = "Range", fill = "GNDVI") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        plot.title = element_text(hjust = 0.5, face = "bold"))

gg_gndvi_et
# save the file
ggsave(filename = "Raw_GNDVI_Heatmap_nmsu_et_2023.png", plot = gg_gndvi_et, path = "2023/output_images", width = 15, height = 12,dpi = 400)


# 3. for NDRE
gg_ndre_et <- ggplot(ndre_et_data, aes(x = Row, y = Range, fill = Value)) +
  geom_tile() +
  scale_fill_viridis(option = "D", na.value = "grey90") +
  facet_wrap(~Time, ncol = 8) +
  labs(title = "NDRE Heatmaps Across Time NMSU ET 2023",
       x = "Row", y = "Range", fill = "NDRE") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        plot.title = element_text(hjust = 0.5, face = "bold"))

gg_ndre_et
# save the file
ggsave(filename = "Raw_NDRE_Heatmap_nmsu_et_2023.png", plot = gg_ndre_et, path = "2023/output_images", width = 15, height = 12,dpi = 400)

```


*Plot the agronomic traits value i.e. yield here*

```{r heatmap for yield data}
#for yield data
nmsu_et_23_yld <- read.csv("2023/GroundTruth_data/Early_Termination/nmsu23_et_yield_merged.csv") %>%
  dplyr::select(-X)
colnames(nmsu_et_23_yld)
# Convert plot info to factors 
nmsu_et_23_yld[, c("Range", "Rep", "Plot", "Row")] <- 
  lapply(nmsu_et_23_yld[, c("Range", "Rep", "Plot", "Row")], as.factor)

#lets make this wide format to longer format
nmsu_et_23_yld_long <- gather(nmsu_et_23_yld, harvest, value, yld1:yld5) # specify how many yields

# now lets make complete heatmap
gg_yld_heatmap_et <- ggplot(nmsu_et_23_yld_long, aes(x = Row, y = Range, fill = value)) +
  geom_tile() +
  scale_fill_viridis(option = "C", name = "Yield", na.value = "grey80") +
  facet_wrap(~harvest, ncol = 5) +
  labs(title = "NMSU 23 ET Yield Heatmaps (YLD1 to YLD6)", x = "Row", y = "Range") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6))

gg_yld_heatmap_et
# save the file
ggsave(filename = "Raw_Yields_Heatmap_nmsu_et_2023.png", plot = gg_yld_heatmap_et, path = "2023/output_images", width = 15, height = 12,dpi = 400)

```





difference between NDVI, GNDVI, and NDRE mean values
```{r difference between NDVI, GNDVI, and NDRE mean values}
# Calculate mean differences between vegetation indices (NDVI, NDRE, GNDVI) over time
mean_diffs_et <- nmsu_et_long_all %>%
  filter(Index %in% c("NDVI", "NDRE", "GNDVI")) %>% # filter three indices
  group_by(Time, Index) %>% # groyup by time and index
  summarise(mean_value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Index, values_from = mean_value) %>% # convert from long to wide format
  mutate(
    NDVI_minus_NDRE = NDVI - NDRE,
    NDVI_minus_GNDVI = NDVI - GNDVI,
    GNDVI_minus_NDRE = GNDVI - NDRE
  ) %>%
  arrange(Time)
mean_diffs_et
# Convert wide to long for plotting all 3 comparisons
mean_diffs_et_long <- mean_diffs_et %>%
  dplyr::select(Time, NDVI_minus_NDRE, NDVI_minus_GNDVI, GNDVI_minus_NDRE) %>%
  pivot_longer(cols = -Time, names_to = "Comparison", values_to = "Difference")

# Plot
ggplot(mean_diffs_et_long, aes(x = Time, y = Difference, color = Comparison)) +
  geom_line(size = 1.2) +
  labs(title = "Mean Differences Between Vegetation Indices Over Time NMSU ET 2023",
       x = "Time (DOY or DAP)", y = "Mean Difference") +
  theme_minimal() +
  scale_color_manual(values = c(
    "NDVI_minus_NDRE" = "green",
    "NDVI_minus_GNDVI" = "blue",
    "GNDVI_minus_NDRE" = "orange"
  )) +
  theme(legend.title = element_blank())

```


### 09052025
# Summarizing univaraite models (h2) and mt models (GC, SC, RC) in graphs
 ## Univariate: plot heritabilities
```{r 2023 ET Trend Graphs, 09052025}
# unfiltered read data
herit_results_et_23 <- read.csv("2023/MSI_data/output_files/Heritabilities_MSI_filtered_aa_et_23.csv")

df_long_et_23 <- herit_results_et_23 %>%
  pivot_longer(
    cols = -Time,
    names_to = "Index_Model",
    values_to = "Heritability"
  ) %>%
  mutate(
    Model = ifelse(grepl("_spa$", Index_Model), "With Spatial", "No Spatial"),
    Index = case_when(
      grepl("gndvi", Index_Model, ignore.case = TRUE) ~ "GNDVI",
      grepl("ndvi", Index_Model, ignore.case = TRUE) ~ "NDVI",
      grepl("ndre", Index_Model, ignore.case = TRUE) ~ "NDRE"
    )
  )

# plot
ggplot_et_23 <- ggplot(df_long_et_23, aes(x = Time, y = Heritability, color = Model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~Index, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c("No Spatial" = "red", "With Spatial" = "blue")) +
  theme_minimal(base_size = 14) +
  labs(x = "GDD (Time)",
       y = expression(h^2~" (Heritability)"),
       color = "Model",
       title = "Heritability Trajectories (2023, ET Irrigation)")
ggplot_et_23
# ggsave
ggsave("2023/output_images/Heritability_Trends_All_Indices_NMSU_ET_2023.png", ggplot_et_23, width = 14, height = 10, dpi = 400)

```

### Mt models graphs showing genetic correlation, spatial correlation (if) and residual correlation

```{r Mt models graphs GNDVI ET 2023 09052025}
mt_models_gndvi_et_23 <- read.csv("2023/MSI_data/output_files/MT_models_GC_SC_RC_values_GNDVI_et_23.csv")

# Use your dataframe
df_gndvi_et_23 <- mt_models_gndvi_et_23
# Preserve order of Trait
df_gndvi_et_23$Trait <- factor(df_gndvi_et_23$Trait, levels = df_gndvi_et_23$Trait)

# ---- Graph 1: Genetic1 & Residual.2 ----
df_gndvi_et_23_no_spa <- df_gndvi_et_23 %>%
  select(Trait, Genetic1, Residual.2) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value")

p1 <- ggplot(df_gndvi_et_23_no_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI without Spatial ET 2023",
       y = "Correlation")

# ---- Graph 2: Genetic, Spatial & Residual ----
df_gndvi_et_23_spa <- df_gndvi_et_23 %>%
  select(Trait, Genetic, Spatial, Residual) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value") %>%
  filter(!is.na(Value))

p2 <- ggplot(df_gndvi_et_23_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI with Spatial ET 2023",
       y = "Correlation")

# Show both plots
p1
p2

```

 
 
## Graphs: Genetic and Spatial trend throughout the season
```{r Mt models graphs GNDVI ET 2023 06042026}
mt_models_gndvi_et_23 <- read.csv("2023/MSI_data/output_files/MT_models_GC_SC_values_GNDVI_ET_2023.csv")
colnames(mt_models_gndvi_et_23)
# Use your dataframe
df_gndvi_et_23 <- mt_models_gndvi_et_23
# Preserve order of Trait
df_gndvi_et_23$Trait <- factor(df_gndvi_et_23$Trait, levels = df_gndvi_et_23$Trait)

# ---- Graph 1: Genetic1 only ----
df_gndvi_et_23_no_spa <- df_gndvi_et_23 %>%
  select(Trait, Genetic) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value")

gndvi_et_23_p1 <- ggplot(df_gndvi_et_23_no_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI without Spatial ET 2023",
       y = "Correlation")
gndvi_et_23_p1
# ---- Graph 2: Genetic, & Spatial  ----
df_gndvi_et_23_spa <- df_gndvi_et_23 %>%
  select(Trait, Genetic, Spatial) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value") %>%
  filter(!is.na(Value))

gndvi_et_23_p2 <- ggplot(df_gndvi_et_23_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Genetic" = "forestgreen", 
                                "Spatial" = "darkorange")) +
  scale_linetype_manual(values = c("Genetic" = "solid", 
                                   "Spatial" = "solid")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "GNDVI Trend over ET 2023",
       y = "Correlation")

# Show both plots
gndvi_et_23_p1 # only genetic
gndvi_et_23_p2 # genetic and spatial

# ggsave
ggsave("2023/output_images/Genetic_Spatial_Correlations_NMSU_ET_2023.png", gndvi_et_23_p2, width = 8, height = 6, dpi = 400)
```




