---
title: "Running univariate models to calculate Heritability_ASReml NMSU 2023"
author: "Dinesh Ghimire"
date: "2025-08-01"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr","asreml", "tidyr", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "GGally")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```

#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NMSU"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NMSU 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa_NMSU"
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is misssing please let me know. 
key<-read.csv("2023/meta_data/cultivar_list_key.csv") # contains both CU and NMSU cultivars here. 
# its same for 2022 and 2023 (cultivar list)
```




1. *First treatment NMSU_Normal 2023 *
with spatial in the model, random = ~ vm(Material_Id, ct_geno_inverse) + ar1(Row):ar1(Range),



*Normal irrigation treatment*
Loading and preparing data files
```{r loading files 08012025}
#import merged phenotype file 
nmsu23_nor_pheno <- read.csv("2023/MSI_data/Output_files/nmsu23_nor_combined_phenotypes_MSI.csv", sep = ",") %>% 
  dplyr::select(-X) # remove index column

# Before running models, make sure the variables are in correct form. Need to convert into factors and numeric if needed.
  # converting all factorial variables into factor
nmsu23_nor_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu23_nor_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)



#import genotype file (same for 2022 and 2023)
grm_aa <- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1)
grm_aa <- as.matrix(grm_aa)
mean(diag(grm_aa)) # 1
hist(diag(grm_aa))

grm_agh <- grm_aa+diag(.0005,length(grm_aa[,1]),length(grm_aa[,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

#now run the model for now without GRM but need to find GRM.

```



Normally if the range of numbers vary a lot between vi's and yield, then we can scale yield to variance level of vi's. but here since yield is in lb and upto 2 digit, should have that issue. If got any, we can always use scaled yield.
```{r scaling yield }
# we only have to scale 6 harvest values based on NDVI's value. Of course the variance between VI's differ but it should be okay to use Yield scaled onto NDVI for analysis with NDRE and GNDVI as well. 

# Define the mapping of yield traits to NDVI columns
yield_traits <- paste0("yld", 2:7)
ndvi_times <- c("NDVI_145", "NDVI_165", "NDVI_201", "NDVI_228", "NDVI_270", "NDVI_299") # selecting one of the time point in each harvest window. selecting based on higher variation time point from heatmap, not actually calculating all. FOR this check the harvest windows or pheno cor heatmaps to see which time point falls in which window

# Loop over each yield trait
for (i in 1:length(yield_traits)) {
  
  trait <- yield_traits[i]
  ndvi_col <- ndvi_times[i]
  
  # Calculate variance of NDVI and yield
  ndvi_var <- var(nmsu23_nor_pheno[[ndvi_col]], na.rm = TRUE)
  yld_var <- var(nmsu23_nor_pheno[[trait]], na.rm = TRUE)
  
  # Calculate scaling factor
  sf <- sqrt(ndvi_var) / sqrt(yld_var)
  
  # Create new scaled yield column
  new_col <- paste0("sca_", trait)
  nmsu23_nor_pheno[[new_col]] <- nmsu23_nor_pheno[[trait]] * sf
  
  #  Print scaling factor
  cat("Scaling factor for", trait, "using", ndvi_col, ":", sf, "\n")
}


# thus all scaled yields are added as a new column at last. 
```

one but not confirming now
```{r for checking outliers GZA_F}
# Fit a linear model
model <- lm(yld2 ~ Rep, data = nmsu23_nor_pheno, na.action = na.exclude)

# Calculate residuals
nmsu23_nor_pheno$residuals <- residuals(model, na.action = na.exclude)
# Standardize the residuals
nmsu23_nor_pheno$residuals_standard <- scale(nmsu23_nor_pheno$residuals, center = T, scale = T) 
# Identify outliers
outliers <- nmsu23_nor_pheno %>% filter(abs(residuals_standard) > 3)
# Print outliers
View(outliers)
# three in the range of 3.5, so not removing for now.

```


Before running models, I will be keeping only the imaging events that overlap with available yield data i.e. remove time points before 1st harvest cz we don't have data on that. And for multi-trait models later on, I will need harvest data to correspond to. Thus remove imaging events 73.. 117. starting from 131.

```{r run univariate models with spatial 08152025}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
colnames(nmsu23_nor_pheno)
# first for NDVI
# we have 32 imaging events 
# Days: 73, 88, 95, 101, 107, 117, 131, 138, 145, 152, 160, 161, 165, 173, 179, 187, 192, 201, 206, 215, 223, 228, 237, 242, 257, 263, 270, 278, 283, 293, 299, 307
# filtered days: 32-6 = 26 h2.
#131, 138, 145, 152, 160, 161, 165, 173, 179, 187, 192, 201, 206, 215, 223, 228, 237, 242, 257, 263, 270, 278, 283, 293, 299, 307

#  For NDVI_152
st_nmsu23_nor_152 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_152 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_152)$varcomp
asreml::vpredict(st_nmsu23_nor_152, heritA ~ V1 / (V1 + V5)) # 0.80


# run for all time points
```


 The fact that h2 > H2 here isn’t an error — it’s just that the models are not nested, so the two are capturing variance differently.

```{r Heritability of yield (Narrow Sense as we don't have GRM) date}

asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence

# for yield 2
 # asreml.options(extra=1000,maxit =1000) # to setup more iterations for successful burn-in or convergence

st_nmsu23_nor_yld2 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld2)$varcomp
asreml::vpredict(st_nmsu23_nor_yld2, heritA ~ V1 / (V1 + V5)) # 0.58 with 0.11 SE
# slightly higher than broad sense (without GRM): but that's possible bcz of how we are calculating narrow sense heritability. it's not nested within broad sense here. 
# in this case, looking at var comp, additive genetic component pulled some component of spatial when using GRM. 
st_nmsu23_nor_yld2_no_grm <- { 
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ id(Germplasm) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld2_no_grm)$varcomp
asreml::vpredict(st_nmsu23_nor_yld2_no_grm, heritA ~ V1 / (V1 + V5)) #0.56


# for yield 3
st_nmsu23_nor_yld3 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld3)$varcomp
asreml::vpredict(st_nmsu23_nor_yld3, heritA ~ V1 / (V1 + V5)) # 0.70 with 0.086 SE


# for yield 4
st_nmsu23_nor_yld4 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld4)$varcomp
asreml::vpredict(st_nmsu23_nor_yld4, heritA ~ V1 / (V1 + V5)) # 0.68 with 0.089 SE


# for yield 5
st_nmsu23_nor_yld5 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld5)$varcomp
asreml::vpredict(st_nmsu23_nor_yld5, heritA ~ V1 / (V1 + V5)) # 0.69 with 0.09 SE


# for yield 6
st_nmsu23_nor_yld6 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld6 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld6)$varcomp
asreml::vpredict(st_nmsu23_nor_yld6, heritA ~ V1 / (V1 + V5)) # 0.66 with 0.09 SE


# for yield 7
st_nmsu23_nor_yld7 <- { 
    attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld7 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld7)$varcomp
asreml::vpredict(st_nmsu23_nor_yld7, heritA ~ V1 / (V1 + V5)) # 0.74 with 0.075 SE

```


# Run for all VI's
### STORE BOTH HERITABILITY AND VARIANCE COMPONENTS

 To check variance components as well: store VC in another data frame
 # Days: 131, 138, 145, 152, 160, 161, 165, 173, 179, 187, 192, 201, 206, 215, 223, 228, 237, 242, 257, 263, 270, 278, 283, 293, 299, 307 (CZ REMOVED TIME POINTS BEFORE 1st harvest)


### For modeling, check experimental design first, Rep>Range>Row here
```{r For all VI's 08012025}
# Define time points
times <- c(131, 138, 145, 152, 160, 161, 165, 173, 179, 187, 192, 201, 206, 215, 223, 228, 237, 242, 257, 263, 270, 278, 283, 293, 299, 307)

# Initialize heritability results
herit_results_nor_23 <- data.frame(
  Time = times,
  Herit_ndvi = NA, Herit_ndvi_spa = NA,
  Herit_gndvi = NA, Herit_gndvi_spa = NA,
  Herit_ndre = NA, Herit_ndre_spa = NA
)

# Initialize variance component storage
varcomp_results <- data.frame(
  Time = numeric(0),
  Index = character(0),
  Model = character(0),
  Component = character(0),
  Estimate = numeric(0),
  stringsAsFactors = FALSE
)

# Function to fit model and return heritability and variance components
fit_heritability <- function(index_prefix, time_point, data) {
  col_name <- paste0(index_prefix, "_", time_point)
  
  if (!col_name %in% colnames(data)) {
    warning(paste("Column", col_name, "not found. Skipping."))
    return(list(h2 = c(NA, NA), vc = NULL))
  }
  
  # --- Simple Model ---
  mod_simple <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE # add grm here
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Simple model failed for", col_name))
    return(NULL)
  })
  
  h2_simple <- NA
  vc_simple <- NULL
  if (!is.null(mod_simple)) {
    h2_simple <- tryCatch({
      vpredict(mod_simple, herit ~ V1 / (V1 + V2))$Estimate
    }, error = function(e) NA)
    
    vc_simple <- tryCatch({
      vc <- summary(mod_simple)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Simple")
    }, error = function(e) NULL)
  }
  
  # --- Spatial Model --- 
  mod_spatial <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE # add grm here
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Spatial model failed for", col_name))
    return(NULL)
  })
  
  h2_spatial <- NA
  vc_spatial <- NULL
  if (!is.null(mod_spatial)) {
    h2_spatial <- tryCatch({
      vpredict(mod_spatial, herit_spatial ~ V1 / (V1 + V5))$Estimate
    }, error = function(e) NA)
    
    vc_spatial <- tryCatch({
      vc <- summary(mod_spatial)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Spatial")
    }, error = function(e) NULL)
  }

  return(list(h2 = c(h2_simple, h2_spatial),
              vc = rbind(vc_simple, vc_spatial)))
}

# Loop through all time points and indices
for (i in seq_along(times)) {
  time <- times[i]
  cat("\nProcessing time point:", time, "\n")
  
  for (index in c("NDVI", "GNDVI", "NDRE")) {
    result <- fit_heritability(index, time, nmsu23_nor_pheno)
    
    # Store heritabilities
    herit_results_nor_23[i, paste0("Herit_", tolower(index))] <- result$h2[1]
    herit_results_nor_23[i, paste0("Herit_", tolower(index), "_spa")] <- result$h2[2]
    
    # Store variance components
    if (!is.null(result$vc)) {
      result$vc$Index <- index
      result$vc$Time <- time
      varcomp_results <- bind_rows(varcomp_results, result$vc)
    }
  }
}

# Round heritability results to 2 decimals
herit_results_nor_23 <- herit_results_nor_23 %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

# Done!
print(head(herit_results_nor_23))
print(head(varcomp_results))


```

Checking some NA's or very high heritabilities (eg. 0.87 in GNDVI 215 but 0 in NDVI, also 1 herit of ndvi 223) or suspiciously differnet estimate among VI's to make sure the realibility of the estimates.

```{r}
# PART 1: First lets deal with NA's :for those that didn't run
# NDVI: 283: 0.11 with row only
# 1. for 283 NDVI with spatial, its NA on Herit with spatial
st_nmsu23_nor_ndvi_283 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_283 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_ndvi_283)$varcomp
asreml::vpredict(st_nmsu23_nor_ndvi_283, heritA ~ V3 / (V3 + V4)) # Row: 0.11 with 0.11 SE (bound with range)
  

# GNDVI: 131: set to 0 and then remove 113 for all three
# 1. for 131 GNDVI with spatial, its NA on Herit with spatial
st_nmsu23_nor_131_gndvi <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_131 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_131_gndvi)$varcomp
asreml::vpredict(st_nmsu23_nor_ndvi_283, heritA ~ V3 / (V3 + V4)) # almost 0 cz bound

# NDRE: 283- set to 0
st_nmsu23_nor_283_ndre <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_283 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_283_ndre)$varcomp
asreml::vpredict(st_nmsu23_nor_283_ndre, heritA ~ V3 / (V3 + V4)) # Row: 0.028 with 0.09 SE (almost 0)
asreml::vpredict(st_nmsu23_nor_161_gndvi, heritA ~ V3 / (V3 + V4)) # Range: ~ 0 and bound for genetic and spatial


# PART 2: too high values
 # NDVI: 187 and 192 (too low than other two), 223 (1 heritability)
  ## NDVI 187: all looks good, still got smaller value, 0.15
st_nmsu23_nor_ndvi_187 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_187 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_ndvi_187)$varcomp
asreml::vpredict(st_nmsu23_nor_ndvi_187, heritA ~ V1 / (V1 + V5)) #  0.15 

  ## NDVI 192: its 0 (all good), with both full model and sparse (either row or range)
st_nmsu23_nor_ndvi_192 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_192 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_ndvi_192)$varcomp
asreml::vpredict(st_nmsu23_nor_ndvi_192, heritA ~ V1 / (V1 + V5)) #  0.15 

  ## NDVI 223: change to 0.47 with row
st_nmsu23_nor_ndvi_223 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_223 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_ndvi_223)$varcomp
asreml::vpredict(st_nmsu23_nor_ndvi_223, heritA ~ V3 / (V3 + V4)) #  0.47 with row... bound with both so 1 


 # GNDVI: 187(much than ndvi), 206 (too low than other two), 215 (much higher than other two), 257 (much higher than other two),  283 (other two indices have almost 0 heritability), 

## GNDVI 187:all looks good: still got higher value, 0.82 and everything looks good. genetic v.c. higher than ndvi
st_nmsu23_nor_gndvi_187 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_187 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_gndvi_187)$varcomp
asreml::vpredict(st_nmsu23_nor_gndvi_187, heritA ~ V1 / (V1 + V5)) #  0.15 

  ## GNDVI 206: replace with 0.60 (Row)
    ## still 0.31 (smaller than ndvi and ndre), but bound for spatial. so only did with row and got 0.60 (looks good too)
st_nmsu23_nor_gndvi_206 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_206 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_gndvi_206)$varcomp
asreml::vpredict(st_nmsu23_nor_gndvi_206, heritA ~ V3 / (V3 + V4)) #  0.60 with row and 0.30 with range

  ## GNDVI 215: all looks good
st_nmsu23_nor_gndvi_215 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_215 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_gndvi_215)$varcomp
asreml::vpredict(st_nmsu23_nor_gndvi_215, heritA ~ V1 / (V1 + V5)) #  0.87 (all looks good)

  ## GNDVI 257: all good
st_nmsu23_nor_gndvi_257 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_257 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_gndvi_257)$varcomp
asreml::vpredict(st_nmsu23_nor_gndvi_257, heritA ~ V1 / (V1 + V5)) #  0.83 (all looks good)

  ## GNDVI 283: false, bound with full model, h2 = 0.1, lets set up as 0 and remove this
st_nmsu23_nor_gndvi_283 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_283 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_gndvi_283)$varcomp
asreml::vpredict(st_nmsu23_nor_gndvi_283, heritA ~ V3 / (V3 + V4)) #  0.09 with row 



# now lets fill in this vacant NA in the dataframe with the value we calculated here
# four values to add: 223 (0.47), 283(0.11) NDVI with spatial, 131(0), 206(0.6) & 283(0) GNDVI with spatial, 283(0.1) NDRE with spatial

  #  add row or range based spatial values in the dataframe
# NDVI_223 — use Row-based model (h2 = 0.46)
herit_results_nor_23$Herit_ndvi_spa[herit_results_nor_23$Time == 223] <- 0.46

# NDVI_283 — use Row-based model (h2 = 0.13)
herit_results_nor_23$Herit_ndvi_spa[herit_results_nor_23$Time == 283] <- 0.13

# GNDVI_131 —  (h2 = 0)
herit_results_nor_23$Herit_gndvi_spa[herit_results_nor_23$Time == 131] <- 0

# GNDVI_206 — use Row-based model (h2 = 0.6)
herit_results_nor_23$Herit_gndvi_spa[herit_results_nor_23$Time == 206] <- 0.6

# GNDVI_283 — (h2 = 0)
herit_results_nor_23$Herit_gndvi_spa[herit_results_nor_23$Time == 283] <- 0

# NDRE_283 — use Row-based model (h2 = 0.1)
herit_results_nor_23$Herit_ndre_spa[herit_results_nor_23$Time == 283] <- 0.1


# part 3: remove time points with all ~ 0 heritability ( all <0.1)
# 131, 161, 270, 283

# now remove time points
head(herit_results_nor_23)
dim(herit_results_nor_23) # 26*7
bad_times <- c(131,161, 270, 283)
herit_results_nor_23_clean <- herit_results_nor_23[!herit_results_nor_23$Time %in% bad_times, ]
dim(herit_results_nor_23_clean) # 22*7, thus 4 time points removed


# manually checked and all values are in right place, now lets export this file for future use. 

write.csv(herit_results_nor_23_clean, file = "2023/MSI_data/output_files/Heritabilities_MSI_filtered_aa_nor_23.csv", row.names = F)
```



Comparison box plots

```{r}

herit_results_filtered_nor_23 <- read.csv("2023/MSI_data/output_files/Heritabilities_MSI_filtered_aa_nor_23.csv", sep = ",") 


# Reshape all six heritability columns to long format
herit_results_filtered_nor_23_long_all <- herit_results_filtered_nor_23 %>%
  dplyr::select(Time, Herit_ndvi, Herit_ndvi_spa,
               Herit_gndvi, Herit_gndvi_spa,
               Herit_ndre, Herit_ndre_spa) %>%
  pivot_longer(cols = -Time, names_to = "Type", values_to = "Heritability") %>%
  mutate(
    Index = case_when(
  grepl("gndvi", Type) ~ "GNDVI",
  grepl("ndre", Type) ~ "NDRE",
  grepl("ndvi", Type) ~ "NDVI"
),
    Model = ifelse(grepl("_spa", Type), "Spatial", "Simple")
  )


ggplot_herit_results_filtered_nor_23_long_all <- ggplot(herit_results_filtered_nor_23_long_all, aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") + # Randomly "jitter" (spread out) the data points horizontally by up to ±0.2 units on the x-axis (i.e., within each box)
  # alternate geom_point() where all will be stacked in a vertical line so overlap issue.
  facet_wrap(~ Model) +
  labs(title = "Comparison of Heritability (h²) by VI and Model Type Normal 2023 NMSU",
       x = "Vegetation Index",
       y = "Heritability (h²)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")+
  ylim(c(0,1))
ggplot_herit_results_filtered_nor_23_long_all
ggsave(filename = "Boxplot_Heritability_Comparison_VI_nmsu_nor_2023.png", plot = ggplot_herit_results_filtered_nor_23_long_all, path = "2023/output_images", width = 10, height = 8,dpi = 400)

```













################################# Early Termination ################################






2. *First treatment NMSU_Early_Termination 2023 *
with spatial in the model, random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),



*Early_Termination irrigation treatment*
Loading and preparing data files
```{r loading files 08152025}
#import merged phenotype file 
nmsu23_et_pheno <- read.csv("2023/MSI_data/Output_files/nmsu23_et_combined_phenotypes_MSI.csv", sep = ",") %>% 
  select(-X) # remove index column

# Before running models, make sure the variables are in correct form. Need to convert into factors and numeric if needed.
  # converting all factorial variables into factor
nmsu23_et_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu23_et_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
  ## make sure all variables are in correct form (factors or numeric)

#import genotype file (same for 2022 and 2023)
grm_aa <- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1)
grm_aa <- as.matrix(grm_aa)
mean(diag(grm_aa)) # 1
hist(diag(grm_aa))

grm_agh <- grm_aa+diag(.0005,length(grm_aa[,1]),length(grm_aa[,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

#now run the model for now without GRM but need to find GRM.

```




Normally if the range of numbers vary a lot between vi's and yield, then we can scale yield to variance level of vi's. but here since yield is in lb and upto 2 digit, should have that issue. If got any, we can always use scaled yield.
```{r scaling yield }
# we only have to scale 6 harvest values based on NDVI's value. Of course the variance between VI's differ but it should be okay to use Yield scaled onto NDVI for analysis with NDRE and GNDVI as well. 

# Define the mapping of yield traits to NDVI columns
yield_traits <- paste0("yld", 2:6)
ndvi_times <- c("NDVI_145", "NDVI_165", "NDVI_201", "NDVI_228", "NDVI_270", "NDVI_299") # selecting one of the time point in each harvest window. selecting based on higher variation time point from heatmap, not actually calculating all. FOR this check the harvest windows or pheno cor heatmaps to see which time point falls in which window

# Loop over each yield trait
for (i in 1:length(yield_traits)) {
  
  trait <- yield_traits[i]
  ndvi_col <- ndvi_times[i]
  
  # Calculate variance of NDVI and yield
  ndvi_var <- var(nmsu23_et_pheno[[ndvi_col]], na.rm = TRUE)
  yld_var <- var(nmsu23_et_pheno[[trait]], na.rm = TRUE)
  
  # Calculate scaling factor
  sf <- sqrt(ndvi_var) / sqrt(yld_var)
  
  # Create new scaled yield column
  new_col <- paste0("sca_", trait)
  nmsu23_et_pheno[[new_col]] <- nmsu23_et_pheno[[trait]] * sf
  
  #  Print scaling factor
  cat("Scaling factor for", trait, "using", ndvi_col, ":", sf, "\n")
}


# thus all scaled yields are added as a new column at last. 
```

```{r for checking outliers GZA_F}
# Fit a linear model
model <- lm(yld2 ~ Rep, data = nmsu23_et_pheno, na.action = na.exclude)

# Calculate residuals
nmsu23_et_pheno$residuals <- residuals(model, na.action = na.exclude)
# Standardize the residuals
nmsu23_et_pheno$residuals_standard <- scale(nmsu23_et_pheno$residuals, center = T, scale = T) 
# Identify outliers
outliers <- nmsu23_et_pheno %>% filter(abs(residuals_standard) > 3)
# Print outliers
# View(outliers)
# none

```


Before running models, I will be keeping only the imaging events that overlap with available yield data i.e. remove time points before 1st harvest cz we don't have data on that. And for multi-trait models later on, I will need harvest data to correspond to. Thus remove imaging events 73.. 117. starting from 131.

- Actual order of image collection and harvest: 
"73", "88", "95", "101", "107", "117", "yld1",
  "131", "138", "145", "152",  "yld2",
  "160", "165", "173", "179",  "yld3",
  "187", "192", "201", "206",  "yld4",
  "215", "223", "228", "237", "242", "257", "263", "270", "278",  "yld5",
  "283", "293", "299", "307", "yld6"
  
```{r run univariate models with spatial 08152025}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
colnames(nmsu23_et_pheno)
# first for NDVI
# we have 31 imaging events  in ET treatment. 
# Days: 73, 88, 95, 101, 107, 117, 131, 138, 145, 152, 160, 165, 173, 179, 187, 192, 201, 206, 215, 223, 228, 237, 242, 257, 263, 270, 278, 283, 293, 299, 307
# however first 6 are before first harvest and we don't have data for that, so remove those
# filtered days: 31-6 = 25 h2.
# 131, 138, 145, 152, 160, 165, 173, 179, 187, 192, 201, 206, 215, 223, 228, 237, 242, 257, 263, 270, 278, 283, 293, 299, 307

#  For NDVI_152
st_nmsu23_et_152 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_152 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_152)$varcomp
asreml::vpredict(st_nmsu23_et_152, heritA ~ V1 / (V1 + V5)) # 0.71

# run for all time points
```


## Run for all time points and all yields
 The fact that h2 > H2 here isn’t an error — it’s just that the models are not nested, so the two are capturing variance differently.

```{r Heritability of yield (Narrow Sense as we have GRM) 08152025}

asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence

# for yield 2
 # asreml.options(extra=1000,maxit =1000) # to setup more iterations for successful burn-in or convergence

st_nmsu23_et_yld2 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld2)$varcomp
asreml::vpredict(st_nmsu23_et_yld2, heritA ~ V1 / (V1 + V5)) # 0.67 with 0.10 SE


# for yield 3
st_nmsu23_et_yld3 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld3)$varcomp
asreml::vpredict(st_nmsu23_et_yld3, heritA ~ V1 / (V1 + V5)) # 0.85 with 0.09 SE


# for yield 4
st_nmsu23_et_yld4 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld4)$varcomp
asreml::vpredict(st_nmsu23_et_yld4, heritA ~ V1 / (V1 + V5)) # 0.67 with 0.12 SE


# for yield 5
st_nmsu23_et_yld5 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld5)$varcomp
asreml::vpredict(st_nmsu23_et_yld5, heritA ~ V1 / (V1 + V5)) # 0.88 with 0.05 SE


# for yield 6
asreml.options(extra=500,maxit =500) # to setup more iterations for successful burn-in or convergence
st_nmsu23_et_yld6 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld6 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld6)$varcomp
asreml::vpredict(st_nmsu23_et_yld6, heritA ~ V1 / (V1 + V5)) # 0.93 with 0.03 SE

# same with scaled yield for yld 6 (very high heritability)
# without GRM:  0.942 with 0.02 SE

```


# Run for all VI's
### STORE BOTH HERITABILITY AND VARIANCE COMPONENTS

 To check variance components as well: store VC in another data frame
 # Days: 131, 138, 145, 152, 160, 165, 173, 179, 187, 192, 201, 206, 215, 223, 228, 237, 242, 257, 263, 270, 278, 283, 293, 299, 307 (CZ REMOVED TIME POINTS BEFORE 1st harvest and 161 imaging not in ET)


### For modeling, check experimental design first, Rep>Range>Row here
```{r For all VI's 08152025}
# Define time points
times <- c(131, 138, 145, 152, 160, 165, 173, 179, 187, 192, 201, 206, 215, 223, 228, 237, 242, 257, 263, 270, 278, 283, 293, 299, 307)

# Initialize heritability results
herit_results_et_23 <- data.frame(
  Time = times,
  Herit_ndvi = NA, Herit_ndvi_spa = NA,
  Herit_gndvi = NA, Herit_gndvi_spa = NA,
  Herit_ndre = NA, Herit_ndre_spa = NA
)

# Initialize variance component storage
varcomp_results <- data.frame(
  Time = numeric(0),
  Index = character(0),
  Model = character(0),
  Component = character(0),
  Estimate = numeric(0),
  stringsAsFactors = FALSE
)

# Function to fit model and return heritability and variance components
fit_heritability <- function(index_prefix, time_point, data) {
  col_name <- paste0(index_prefix, "_", time_point)
  
  if (!col_name %in% colnames(data)) {
    warning(paste("Column", col_name, "not found. Skipping."))
    return(list(h2 = c(NA, NA), vc = NULL))
  }
  
  # --- Simple Model ---
  mod_simple <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE # add grm here
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Simple model failed for", col_name))
    return(NULL)
  })
  
  h2_simple <- NA
  vc_simple <- NULL
  if (!is.null(mod_simple)) {
    h2_simple <- tryCatch({
      vpredict(mod_simple, herit ~ V1 / (V1 + V2))$Estimate
    }, error = function(e) NA)
    
    vc_simple <- tryCatch({
      vc <- summary(mod_simple)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Simple")
    }, error = function(e) NULL)
  }
  
  # --- Spatial Model --- 
  mod_spatial <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE # add grm here
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Spatial model failed for", col_name))
    return(NULL)
  })
  
  h2_spatial <- NA
  vc_spatial <- NULL
  if (!is.null(mod_spatial)) {
    h2_spatial <- tryCatch({
      vpredict(mod_spatial, herit_spatial ~ V1 / (V1 + V5))$Estimate
    }, error = function(e) NA)
    
    vc_spatial <- tryCatch({
      vc <- summary(mod_spatial)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Spatial")
    }, error = function(e) NULL)
  }

  return(list(h2 = c(h2_simple, h2_spatial),
              vc = rbind(vc_simple, vc_spatial)))
}

# Loop through all time points and indices
for (i in seq_along(times)) {
  time <- times[i]
  cat("\nProcessing time point:", time, "\n")
  
  for (index in c("NDVI", "GNDVI", "NDRE")) {
    result <- fit_heritability(index, time, nmsu23_et_pheno)
    
    # Store heritabilities
    herit_results_et_23[i, paste0("Herit_", tolower(index))] <- result$h2[1]
    herit_results_et_23[i, paste0("Herit_", tolower(index), "_spa")] <- result$h2[2]
    
    # Store variance components
    if (!is.null(result$vc)) {
      result$vc$Index <- index
      result$vc$Time <- time
      varcomp_results <- bind_rows(varcomp_results, result$vc)
    }
  }
}

# Round heritability results to 2 decimals
herit_results_et_23 <- herit_results_et_23 %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

# Done!
print(head(herit_results_et_23))
print(head(varcomp_results))


```



Checking some NA's or very high heritabilities or suspiciously differnet estimate among VI's to make sure the realibility of the estimates.
we will also remove a few non-informative time points (h2 ~ 0) during the range of termination. SIT period June 15th through September 11th, 2023 (166 to 254 DOY) so remove time points 215 to 257 (based on results above where the values are mostly 0 or 1(by model failure)). Even after ET, a few time points afterwards shows good values meaning still growing (may be with rainfed water or remaining moisture). 
- remove time points 215, 223, 228, 237, 242, 257 (i.e. after 4th harvest until irrigation started, there was nothing green on the ground)
- also remove 187 and 283 (all 0's)

```{r}
# PART 1: First lets deal with NA's :for those that didn't run
# NDVI: 131: 
# NDRE; 160, 307
# 1. 131 NDVI : set to 0
st_nmsu23_et_ndvi_131 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_131 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_ndvi_131)$varcomp
asreml::vpredict(st_nmsu23_et_ndvi_131, heritA ~ V3 / (V3 + V4)) # Row: 0.04 and bound
  
# NDRE: 160- set to 0.23 (row)
st_nmsu23_et_ndre_160 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_160 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_ndre_160)$varcomp
asreml::vpredict(st_nmsu23_et_ndre_160, heritA ~ V3 / (V3 + V4)) # Row: 0.23, bound with range

# NDRE: 307- set to 0.56 with row
st_nmsu23_et_ndre_307 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_307 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_ndre_307)$varcomp
asreml::vpredict(st_nmsu23_et_ndre_307, heritA ~ V3 / (V3 + V4)) # Row: 0.56 


# PART 2: too high values
 # NDVI: 179
 # GNDVI: 160, 179(much than ndvi), 192(much higher than other two), 278 (much higher than other two),  293 and 299 
 # NDRE: 206 (h2=1)

# NDRE: 206 : set to 0 (not trustworthy)
st_nmsu23_et_ndre_206 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_206 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_ndre_206)$varcomp
asreml::vpredict(st_nmsu23_et_ndre_206, heritA ~ V3 / (V3 + V4)) # still ~ 1, spatial portion pulled from residual
st_nmsu23_et_ndre_206_updated <- update(st_nmsu23_et_ndre_206, start = st_nmsu23_et_ndre_206$gammas)
summary(st_nmsu23_et_ndre_206_updated)$varcomp # still not fixed well. spatial pulled from residual
asreml::vpredict(st_nmsu23_et_ndre_206_updated, heritA ~ V1 / (V1 + V5)) # 0.16

#NDVI
# NDVI 201 :all looks good: still 0.60 (not as high as other two indices) 
st_nmsu23_et_ndvi_201 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_201 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_ndvi_201)$varcomp
asreml::vpredict(st_nmsu23_et_ndvi_201, heritA ~ V1 / (V1 + V5)) #  0.83 


# NDVI 179 :all looks good: still got higher value, 0.83 and everything looks good. 
st_nmsu23_et_ndvi_179 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_179 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_ndvi_179)$varcomp
asreml::vpredict(st_nmsu23_et_ndvi_179, heritA ~ V1 / (V1 + V5)) #  0.83 


# GNDVI's
## GNDVI 160 :all looks good: still got higher value, 0.94 and everything looks good. 
st_nmsu23_et_gndvi_160 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_160 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_gndvi_160)$varcomp
asreml::vpredict(st_nmsu23_et_gndvi_160, heritA ~ V1 / (V1 + V5)) #  0.15 

## GNDVI 179 : change value to 0.91 (update model for convergence)
# full model not converged, with row only 0.27. TRRY TO CONVERGE: did with running model again by starting with values from 1st model (gammas). this acts as starting values for model and sometimes help convergence. HERE IT DID. 
st_nmsu23_et_gndvi_179 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_179 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_gndvi_179)$varcomp # ? in bound for genetic and spatial var components
st_nmsu23_et_gndvi_179_updated <- update.asreml(st_nmsu23_et_gndvi_179, start = st_nmsu23_et_gndvi_179$gammas)
summary(st_nmsu23_et_gndvi_179_updated)$varcomp # converged well, no warnings and no ? in bound
asreml::vpredict(st_nmsu23_et_gndvi_179_updated, heritA ~ V1 / (V1 + V5)) #  0.91


# GNDVI 192: still same 0.97 high value
st_nmsu23_et_gndvi_192 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_192 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_gndvi_192)$varcomp 
asreml::vpredict(st_nmsu23_et_gndvi_192, heritA ~ V1 / (V1 + V5)) #  0.97


## GNDVI 278: its fine as 0.99 but unexpectedly too high ( looks like spatial pulled from residual)
st_nmsu23_et_gndvi_278 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_278 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_gndvi_278)$varcomp
asreml::vpredict(st_nmsu23_et_gndvi_278, heritA ~ V1 / (V1 + V5)) #  0.99 
# update
st_nmsu23_et_gndvi_278_updated <- update.asreml(st_nmsu23_et_gndvi_278, start = st_nmsu23_et_gndvi_278$gammas)
summary(st_nmsu23_et_gndvi_278_updated)$varcomp # same as above...

## GNDVI 293: its fine as 0.94
st_nmsu23_et_gndvi_293 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_293 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_gndvi_293)$varcomp
asreml::vpredict(st_nmsu23_et_gndvi_293, heritA ~ V1 / (V1 + V5)) #  0.94

## GNDVI 299: its fine as 0.96
st_nmsu23_et_gndvi_299 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_299 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_gndvi_299)$varcomp
asreml::vpredict(st_nmsu23_et_gndvi_299, heritA ~ V1 / (V1 + V5)) #  0.96



# now lets fill in this vacant NA in the dataframe with the value we calculated here OR update with new values fixed above
# four values to add: 131 (0) NDVI with spatial, 179(0.91), 206(0.6) GNDVI with spatial, 160(0.23), 206(0), 307(0.56) NDRE with spatial

  #  add values in the dataframe
# NDVI_131 —  (h2 = 0)
herit_results_et_23$Herit_ndvi_spa[herit_results_et_23$Time == 131] <- 0

# GNDVI_179 —  (h2 = 0.91)
herit_results_et_23$Herit_gndvi_spa[herit_results_et_23$Time == 179] <- 0.91
# GNDVI_206 — (h2 = 0.6)
herit_results_et_23$Herit_gndvi_spa[herit_results_et_23$Time == 206] <- 0.88

# NDRE_160 — (h2 = 0.23) row based
herit_results_et_23$Herit_ndre_spa[herit_results_et_23$Time == 160] <- 0.23
# NDRE_206 — (h2 = 0) weird
herit_results_et_23$Herit_ndre_spa[herit_results_et_23$Time == 206] <- 0
# NDRE_307 — (h2 = 0.56) row based
herit_results_et_23$Herit_ndre_spa[herit_results_et_23$Time == 307] <- 0.56


# part 3: remove time points with all ~ 0 heritability ( all <0.1)
# 187, and 283
# also remove time points in SIT period with ~ 0 heritability: 215, 223, 228, 237, 242, 257

# now remove time points
head(herit_results_et_23)
dim(herit_results_et_23) # 25*7
bad_times <- c(187,215, 223,228,237,242,257,283)
herit_results_et_23_clean <- herit_results_et_23[!herit_results_et_23$Time %in% bad_times, ]
dim(herit_results_et_23_clean) # 17*7, thus 8 time points removed


# manually checked and all values are in right place, now lets export this file for future use. 

write.csv(herit_results_et_23_clean, file = "2023/MSI_data/output_files/Heritabilities_MSI_filtered_aa_et_23.csv", row.names = F)
```


Comparison box plots

```{r}

herit_results_filtered_et_23 <- read.csv("2023/MSI_data/output_files/Heritabilities_MSI_filtered_aa_et_23.csv", sep = ",") 


# Reshape all six heritability columns to long format
herit_results_filtered_et_23_long_all <- herit_results_filtered_et_23 %>%
  dplyr::select(Time, Herit_ndvi, Herit_ndvi_spa,
               Herit_gndvi, Herit_gndvi_spa,
               Herit_ndre, Herit_ndre_spa) %>%
  pivot_longer(cols = -Time, names_to = "Type", values_to = "Heritability") %>%
  mutate(
    Index = case_when(
  grepl("gndvi", Type) ~ "GNDVI",
  grepl("ndre", Type) ~ "NDRE",
  grepl("ndvi", Type) ~ "NDVI"
),
    Model = ifelse(grepl("_spa", Type), "Spatial", "Simple")
  )

range(herit_results_filtered_et_23_long_all$Heritability)
# 0.00 0.99

ggplot_herit_results_filtered_et_23_long_all <- ggplot(herit_results_filtered_et_23_long_all, aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") + # Randomly "jitter" (spread out) the data points horizontally by up to ±0.2 units on the x-axis (i.e., within each box)
  # alternate geom_point() where all will be stacked in a vertical line so overlap issue.
  facet_wrap(~ Model) +
  labs(title = "Comparison of Heritability (h²) by VI and Model Type ET 2023 NMSU",
       x = "Vegetation Index",
       y = "Heritability (h²)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")+
  coord_cartesian(ylim = c(0, 1))

ggplot_herit_results_filtered_et_23_long_all
ggsave(filename = "Boxplot_Heritability_Comparison_VI_nmsu_et_2023.png", plot = ggplot_herit_results_filtered_et_23_long_all, path = "2023/output_images", width = 10, height = 8,dpi = 400)

```












 