---
title: "Running multivariate models 2023 NMSU"
author: "Dinesh Ghimire"
date: "2025-09-01"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown




```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr","asreml", "tidyr", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "GGally")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```

#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NMSU"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NMSU 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa_NMSU"
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is misssing please let me know. 
key<-read.csv("2023/meta_data/cultivar_list_key.csv") # contains both CU and NMSU cultivars here. 
# its same for 2022 and 2023 (cultivar list)
```



1. *First treatment NMSU_Normal 2022 *
with spatial in the model, random = ~ vm(Material_Id, ct_geno_inverse) + ar1(Row):ar1(Range),



*Normal irrigation treatment*
Loading and preparing data files
```{r loading files 08262025}
#import merged phenotype file 
nmsu23_nor_pheno <- read.csv("2023/MSI_data/Output_files/nmsu23_nor_combined_filtered_phenotypes_MSI.csv", sep = ",") %>% 
  select(-X) # remove index column
colnames(nmsu23_nor_pheno)
# Before running models, make sure the variables are in correct form. Need to convert into factors and numeric if needed.
  # converting all factorial variables into factor
nmsu23_nor_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu23_nor_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)

#import genotype file
grm_aa <- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1)
grm_aa <- as.matrix(grm_aa)
mean(diag(grm_aa)) # 1
hist(diag(grm_aa))

grm_agh <- grm_aa+diag(.0005,length(grm_aa[,1]),length(grm_aa[,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


Normally if the range of numbers vary a lot between vi's and yield, then we can scale yield to variance level of vi's. but here since yield is in lb and upto 2 digit, should have that issue. If got any, we can always use scaled yield.
```{r scaling yield }
# we only have to scale 6 harvest values based on NDVI's value. Of course the variance between VI's differ but it should be okay to use Yield scaled onto NDVI for analysis with NDRE and GNDVI as well. 

# Define the mapping of yield traits to NDVI columns
yield_traits <- paste0("yld", 2:7)
ndvi_times <- c("NDVI_145", "NDVI_173", "NDVI_206", "NDVI_228", "NDVI_263", "NDVI_299") # selecting one of the time point in each harvest window. selecting based on higher variation time point from heatmap, not actually calculating all. 

# Loop over each yield trait
for (i in 1:length(yield_traits)) {
  
  trait <- yield_traits[i]
  ndvi_col <- ndvi_times[i]
  
  # Calculate variance of NDVI and yield
  ndvi_var <- var(nmsu23_nor_pheno[[ndvi_col]], na.rm = TRUE)
  yld_var <- var(nmsu23_nor_pheno[[trait]], na.rm = TRUE)
  
  # Calculate scaling factor
  sf <- sqrt(ndvi_var) / sqrt(yld_var)
  
  # Create new scaled yield column
  new_col <- paste0("sca_", trait)
  nmsu23_nor_pheno[[new_col]] <- nmsu23_nor_pheno[[trait]] * sf
  
  #  Print scaling factor
  cat("Scaling factor for", trait, "using", ndvi_col, ":", sf, "\n")
}


# thus all scaled yields are added as a new column at last. 
```

one but not confirming now
```{r for checking outliers}
# Fit a linear model
model <- lm(yld2 ~ Rep, data = nmsu23_nor_pheno, na.action = na.exclude)

# Calculate residuals
nmsu23_nor_pheno$residuals <- residuals(model, na.action = na.exclude)
# Standardize the residuals
nmsu23_nor_pheno$residuals_standard <- scale(nmsu23_nor_pheno$residuals, center = T, scale = T) 
# Identify outliers
outliers <- nmsu23_nor_pheno %>% filter(abs(residuals_standard) > 3)
# Print outliers
View(outliers)
# one, not removing for now.

```


## 1.1. For NDVI : not doin for 2023. Based on 2022 results of mt models and 2023 univariate and phenotypic correlation models, NDVI is not a good indicator here compared to GNDVI so moving forward with GNDVI only.




## 1.2. For GNDVI 
- For Harvest window 2:  "138", "145", "152",  "yld2",
- For Harvest window 3: "160", "165", "173", "179",  "yld3",
- For Harvest window 4: "187", "192", 201",  "yld4",
- For Harvest window 5:  "215", "223", "228",, "242",  "yld5", # 237 out due to lower h2
- For Harvest window 6:  "257", "263", "278",  "yld6",
- For Harvest window 7:  "293", "299", "307", "yld7"
= 27

### 1.2.1. Run models first
- Univariate without spatial
- Univariate with spatial
- Bivariate without spatial
- Bivariate with spatial

#### 1.2.1.1. Univariate for yield without spatial


```{r No spatial }
## For yld1: missing in 2023 data so that's out

## For yld2
st_nmsu23_nor_yld2 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld2)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld2, heritA ~ V1 / (V1 + V2)) # 0.134

## For yld3
st_nmsu23_nor_yld3 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld3)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld3, heritA ~ V1 / (V1 + V2)) # 0.25

## For yld4
st_nmsu23_nor_yld4 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld4)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld4, heritA ~ V1 / (V1 + V2)) #0.32

## For yld5
st_nmsu23_nor_yld5 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld5)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld5, heritA ~ V1 / (V1 + V2)) #0.36

## For yld6
st_nmsu23_nor_yld6 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld6 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld6)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld6, heritA ~ V1 / (V1 + V2)) #0.33

## For yld7
st_nmsu23_nor_yld7 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld7 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld7)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld7, heritA ~ V1 / (V1 + V2)) #0.53

```


#### 1.2.1.2. Univariate for yield with spatial

```{r From spatial BLUPS and Univariate models}
asreml.options(extra =200, maxit =200)
## For yld1 :missing harvest 1 in 2023

## For yld2
st_nmsu23_nor_yld2 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld2)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld2, heritA ~ V1 / (V1 + V5)) # 0.58

## For yld3
st_nmsu23_nor_yld3 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld3)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld3, heritA ~ V1 / (V1 + V5)) # 0.71

## For yld4
st_nmsu23_nor_yld4 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld4)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld4, heritA ~ V1 / (V1 + V5)) # 0.69

## For yld5
st_nmsu23_nor_yld5 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld5)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld5, heritA ~ V1 / (V1 + V5)) # 0.69

## For yld6
st_nmsu23_nor_yld6 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld6 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld6)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld6, heritA ~ V1 / (V1 + V5)) # 0.66

## For yld7
st_nmsu23_nor_yld7 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld7 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_nor_pheno)}
summary(st_nmsu23_nor_yld7)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_nor_yld7, heritA ~ V1 / (V1 + V5)) # 0.74

```



#### 1.2.1.3. Bivariate for yield without spatial

```{r Bivariate models without Spatial based on GNDVI}
colnames(nmsu23_nor_pheno)

## lets proceed with calculation for all harvest windows
# 1.   For harvest window 2: "138", "145", "152", "yld2", 
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for GNDVI_138:
mt_nmsu23_nor_gndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_138)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu23_nor_gndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.796384
asreml::vpredict(mt_nmsu23_nor_gndvi_138, res.cor ~ (V6)/sqrt(V5*V7)) # 0.5037916

 ## For GNDVI_145
mt_nmsu23_nor_gndvi_145 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_145) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_145)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_145, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.3467019
asreml::vpredict(mt_nmsu23_nor_gndvi_145, res.cor ~ (V6)/sqrt(V5*V7)) # 0.5062156

## For GNDVI_152
mt_nmsu23_nor_gndvi_152 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_152) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_152)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_152, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9809112
asreml::vpredict(mt_nmsu23_nor_gndvi_152, res.cor ~ (V6)/sqrt(V5*V7)) # 0.7988144


# 2. For harvest window 3:  160, "165", "173", "179", "yld3" 
## For GNDVI_160
mt_nmsu23_nor_gndvi_160 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_160) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_160)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_160, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.9362708
asreml::vpredict(mt_nmsu23_nor_gndvi_160, res.cor ~ (V6)/sqrt(V5*V7)) # - 0.6039955

## For GNDVI_165
mt_nmsu23_nor_gndvi_165 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_165) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_165)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_165, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.3243493
asreml::vpredict(mt_nmsu23_nor_gndvi_165, res.cor ~ (V6)/sqrt(V5*V7)) # 0.3700119

## For GNDVI_173
mt_nmsu23_nor_gndvi_173 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_173) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_173)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_173, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.2138237
asreml::vpredict(mt_nmsu23_nor_gndvi_173, res.cor ~ (V6)/sqrt(V5*V7)) # 0.1076987

## For GNDVI_179
mt_nmsu23_nor_gndvi_179 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_179) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_179)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_179, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8285259
asreml::vpredict(mt_nmsu23_nor_gndvi_179, res.cor ~ (V6)/sqrt(V5*V7)) # 0.6856564


# 3. For harvest window 4: "187", "192", "201", "206", "yld4",
## For GNDVI_187
mt_nmsu23_nor_gndvi_187 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_187) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_187)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_187, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.71454
asreml::vpredict(mt_nmsu23_nor_gndvi_187, res.cor ~ (V6)/sqrt(V5*V7)) # - 0.3947981

## For GNDVI_192
mt_nmsu23_nor_gndvi_192 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_192) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_192)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_192, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.139915
asreml::vpredict(mt_nmsu23_nor_gndvi_192, res.cor ~ (V6)/sqrt(V5*V7)) # 0.5161164

## For GNDVI_201
mt_nmsu23_nor_gndvi_201 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_201) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_201)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_201, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8571291
asreml::vpredict(mt_nmsu23_nor_gndvi_201, res.cor ~ (V6)/sqrt(V5*V7)) # 0.4132969

## For GNDVI_206
mt_nmsu23_nor_gndvi_206 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_206) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_206)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_206, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9921866
asreml::vpredict(mt_nmsu23_nor_gndvi_206, res.cor ~ (V6)/sqrt(V5*V7)) # 0.9256924


# 4. For harvest window 5: "215", "223" , "228", "237", "242", "yld5",
## For GNDVI_215
mt_nmsu23_nor_gndvi_215 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_215) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_215)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_215, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.7372976
asreml::vpredict(mt_nmsu23_nor_gndvi_215, res.cor ~ (V6)/sqrt(V5*V7)) # - 0.08328185

## For GNDVI_223
mt_nmsu23_nor_gndvi_223 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_223) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_223)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_223, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8121997
asreml::vpredict(mt_nmsu23_nor_gndvi_223, res.cor ~ (V6)/sqrt(V5*V7)) # 0.2225667

## For GNDVI_228
mt_nmsu23_nor_gndvi_228 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_228) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_228)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_228, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.830941
asreml::vpredict(mt_nmsu23_nor_gndvi_228, res.cor ~ (V6)/sqrt(V5*V7)) # 0.4828679

## For GNDVI_237: low h2 and didn't work with full model, so removing this time point

## For GNDVI_242
mt_nmsu23_nor_gndvi_242 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_242) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_242)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_242, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8915374
asreml::vpredict(mt_nmsu23_nor_gndvi_242, res.cor ~ (V6)/sqrt(V5*V7)) # 0.7658118


# 5. For harvest window 6: "257", "263", "278"
## For GNDVI_257
mt_nmsu23_nor_gndvi_257 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_257) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_257)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_257, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.6731409
asreml::vpredict(mt_nmsu23_nor_gndvi_257, res.cor ~ (V6)/sqrt(V5*V7)) # 0.05097691

## For GNDVI_263
mt_nmsu23_nor_gndvi_263 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_263) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_263)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_263, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9699347
asreml::vpredict(mt_nmsu23_nor_gndvi_263, res.cor ~ (V6)/sqrt(V5*V7)) # 0.8941899

## For GNDVI_278
mt_nmsu23_nor_gndvi_278 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_278) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_278)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_278, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.419499
asreml::vpredict(mt_nmsu23_nor_gndvi_278, res.cor ~ (V6)/sqrt(V5*V7)) # 0.531547


# 6. For harvest window 7: "293", "299", "307", "yld7"
## For GNDVI_293
mt_nmsu23_nor_gndvi_293 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld7, GNDVI_293) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_293)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_293, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9771094
asreml::vpredict(mt_nmsu23_nor_gndvi_293, res.cor ~ (V6)/sqrt(V5*V7)) # 0.2906468

## For GNDVI_299
mt_nmsu23_nor_gndvi_299 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld7, GNDVI_299) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_299)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_299, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.971444
asreml::vpredict(mt_nmsu23_nor_gndvi_299, res.cor ~ (V6)/sqrt(V5*V7)) # 0.4418684

## For GNDVI_307
mt_nmsu23_nor_gndvi_307 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld7, GNDVI_307) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_307)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_307, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9560128
asreml::vpredict(mt_nmsu23_nor_gndvi_307, res.cor ~ (V6)/sqrt(V5*V7)) # 0.3955942

```



#### 1.2.1.4. Bivariate for yield with spatial

```{r Bivariate models with Spatial based on GNDVI}
colnames(nmsu23_nor_pheno)

## lets proceed with calculation for all harvest windows
# 1.   For harvest window 2: "138", "145", "152", "yld2", 
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for GNDVI_138:
mt_nmsu23_nor_gndvi_138 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_138) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_138)$varcomp # residual variance scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu23_nor_gndvi_138, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8429848
asreml::vpredict(mt_nmsu23_nor_gndvi_138, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7710531
asreml::vpredict(mt_nmsu23_nor_gndvi_138, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8005352

 ## For GNDVI_145
mt_nmsu23_nor_gndvi_145 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_145) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_145)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_145, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9156033
asreml::vpredict(mt_nmsu23_nor_gndvi_145, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7187683
asreml::vpredict(mt_nmsu23_nor_gndvi_145, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7637188

## For GNDVI_152
mt_nmsu23_nor_gndvi_152 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_152) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_152)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_152, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.952244
asreml::vpredict(mt_nmsu23_nor_gndvi_152, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8943059
asreml::vpredict(mt_nmsu23_nor_gndvi_152, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6010083


# 2. For harvest window 3:  160, "165", "173", "179", "yld3" 
## For GNDVI_160: didn't converge with full model, with diag Residual: 
mt_nmsu23_nor_gndvi_160 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_160) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_160)$varcomp # 
# mt_nmsu23_nor_gndvi_160 <- update.asreml(mt_nmsu23_nor_gndvi_160, START= mt_nmsu23_nor_gndvi_160$gammas)
asreml::vpredict(mt_nmsu23_nor_gndvi_160, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6947061
asreml::vpredict(mt_nmsu23_nor_gndvi_160, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4456388
asreml::vpredict(mt_nmsu23_nor_gndvi_160, res.cor ~ (V11)/sqrt(V10*V12)) # DIAG

## For GNDVI_165
mt_nmsu23_nor_gndvi_165 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_165) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_165)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_165, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8373488
asreml::vpredict(mt_nmsu23_nor_gndvi_165, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6561732
asreml::vpredict(mt_nmsu23_nor_gndvi_165, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8148739

## For GNDVI_173
mt_nmsu23_nor_gndvi_173 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_173) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_173)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_173, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9734555
asreml::vpredict(mt_nmsu23_nor_gndvi_173, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8339853
asreml::vpredict(mt_nmsu23_nor_gndvi_173, res.cor ~ (V11)/sqrt(V10*V12)) # diag

## For GNDVI_179
mt_nmsu23_nor_gndvi_179 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_179) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_179)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_179, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9531264
asreml::vpredict(mt_nmsu23_nor_gndvi_179, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8394996
asreml::vpredict(mt_nmsu23_nor_gndvi_179, res.cor ~ (V11)/sqrt(V10*V12)) # 0.700723


# 3. For harvest window 4: "187", "192", "201", "206", "yld4",
## For GNDVI_187
mt_nmsu23_nor_gndvi_187 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_187) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_187)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_187, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8066847
asreml::vpredict(mt_nmsu23_nor_gndvi_187, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.08490591
asreml::vpredict(mt_nmsu23_nor_gndvi_187, res.cor ~ (V11)/sqrt(V10*V12)) # DIAG

## For GNDVI_192
mt_nmsu23_nor_gndvi_192 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_192) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_192)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_192, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.3959558
asreml::vpredict(mt_nmsu23_nor_gndvi_192, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7345305
asreml::vpredict(mt_nmsu23_nor_gndvi_192, res.cor ~ (V11)/sqrt(V10*V12)) # DIAG

## For GNDVI_201
mt_nmsu23_nor_gndvi_201 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_201) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_201)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_201, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9559265
asreml::vpredict(mt_nmsu23_nor_gndvi_201, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6633052
asreml::vpredict(mt_nmsu23_nor_gndvi_201, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8161225

## For GNDVI_206
mt_nmsu23_nor_gndvi_206 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_206) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_206)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_206, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9357265
asreml::vpredict(mt_nmsu23_nor_gndvi_206, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.9760613
asreml::vpredict(mt_nmsu23_nor_gndvi_206, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7412508


# 4. For harvest window 5: "215", "223" , "228", "237", "242", "yld5",
## For GNDVI_215
mt_nmsu23_nor_gndvi_215 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_215) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_215)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_215, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7756483
asreml::vpredict(mt_nmsu23_nor_gndvi_215, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.3114353
asreml::vpredict(mt_nmsu23_nor_gndvi_215, res.cor ~ (V11)/sqrt(V10*V12)) # DIAG

## For GNDVI_223
mt_nmsu23_nor_gndvi_223 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_223) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_223)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_223, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9433547
asreml::vpredict(mt_nmsu23_nor_gndvi_223, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5152588
asreml::vpredict(mt_nmsu23_nor_gndvi_223, res.cor ~ (V11)/sqrt(V10*V12)) # 0.822485

## For GNDVI_228
mt_nmsu23_nor_gndvi_228 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_228) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_228)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_228, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9513843
asreml::vpredict(mt_nmsu23_nor_gndvi_228, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6035596
asreml::vpredict(mt_nmsu23_nor_gndvi_228, res.cor ~ (V11)/sqrt(V10*V12)) # 0.823242

## For GNDVI_237: lower h2 ~ 0.35 so not working, remove. also remove from above

## For GNDVI_242
mt_nmsu23_nor_gndvi_242 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_242) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_242)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_242, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9773571
asreml::vpredict(mt_nmsu23_nor_gndvi_242, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5299082
asreml::vpredict(mt_nmsu23_nor_gndvi_242, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8094152


# 5. For harvest window 6: "257", "263", "278"
## For GNDVI_257
mt_nmsu23_nor_gndvi_257 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_257) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_257)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_257, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9490153
asreml::vpredict(mt_nmsu23_nor_gndvi_257, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1155405
asreml::vpredict(mt_nmsu23_nor_gndvi_257, res.cor ~ (V11)/sqrt(V10*V12)) # 0.9445742

## For GNDVI_263
mt_nmsu23_nor_gndvi_263 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_263) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_263)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_263, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9294808
asreml::vpredict(mt_nmsu23_nor_gndvi_263, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8275514
asreml::vpredict(mt_nmsu23_nor_gndvi_263, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8101848

## For GNDVI_278
mt_nmsu23_nor_gndvi_278 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_278) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_278)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_278, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9399559
asreml::vpredict(mt_nmsu23_nor_gndvi_278, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5823847
asreml::vpredict(mt_nmsu23_nor_gndvi_278, res.cor ~ (V11)/sqrt(V10*V12)) # 0.732109


# 6. For harvest window 7: "293", "299", "307", "yld7"
## For GNDVI_293
mt_nmsu23_nor_gndvi_293 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld7, GNDVI_293) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_293)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_293, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9690862
asreml::vpredict(mt_nmsu23_nor_gndvi_293, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7194917
asreml::vpredict(mt_nmsu23_nor_gndvi_293, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8743123

## For GNDVI_299
mt_nmsu23_nor_gndvi_299 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld7, GNDVI_299) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_299)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_299, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9660682
asreml::vpredict(mt_nmsu23_nor_gndvi_299, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7797229
asreml::vpredict(mt_nmsu23_nor_gndvi_299, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8205428

## For GNDVI_307
mt_nmsu23_nor_gndvi_307 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld7, GNDVI_307) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu23_nor_pheno)}
summary(mt_nmsu23_nor_gndvi_307)$varcomp # 
asreml::vpredict(mt_nmsu23_nor_gndvi_307, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.916531
asreml::vpredict(mt_nmsu23_nor_gndvi_307, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7186621
asreml::vpredict(mt_nmsu23_nor_gndvi_307, res.cor ~ (V11)/sqrt(V10*V12)) # DIAG
```



### 1.2.2. Graphs












################################# Early Termination ################################

*Second treatment NMSU_ET 2023 *
*Early Termination irrigation treatment*

Loading and preparing data files
```{r loading files 09012025}
#import merged phenotype file 
nmsu23_et_pheno <- read.csv("2023/MSI_data/Output_files/nmsu23_et_combined_filtered_phenotypes_MSI.csv", sep = ",") %>% 
  select(-X) # remove index column

# Before running models, make sure the variables are in correct form. Need to convert into factors and numeric if needed.
  # converting all factorial variables into factor
nmsu23_et_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(nmsu23_et_pheno[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
  ## make sure all variables are in correct form (factors or numeric)
ncol(nmsu23_et_pheno) #44
colnames(nmsu23_et_pheno)
#import genotype file (same for 2022 and 2023)
grm_aa <- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1)
grm_aa <- as.matrix(grm_aa)
mean(diag(grm_aa)) # 1
hist(diag(grm_aa))

grm_agh <- grm_aa+diag(.0005,length(grm_aa[,1]),length(grm_aa[,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

#now run the model for now without GRM but need to find GRM.

```


Normally if the range of numbers vary a lot between vi's and yield, then we can scale yield to variance level of vi's. but here since yield is in lb and upto 2 digit, should have that issue. If got any, we can always use scaled yield.
```{r scaling yield }
# we only have to scale 6 harvest values based on NDVI's value. Of course the variance between VI's differ but it should be okay to use Yield scaled onto NDVI for analysis with NDRE and GNDVI as well. 

# Define the mapping of yield traits to NDVI columns
yield_traits <- paste0("yld", 2:6)
ndvi_times <- c("GNDVI_145", "GNDVI_165", "GNDVI_201", "GNDVI_278", "GNDVI_299") # selecting one of the time point in each harvest window. selecting based on higher variation time point from heatmap, not actually calculating all. FOR this check the harvest windows or pheno cor heatmaps to see which time point falls in which window

# Loop over each yield trait
for (i in 1:length(yield_traits)) {
  
  trait <- yield_traits[i]
  ndvi_col <- ndvi_times[i]
  
  # Calculate variance of NDVI and yield
  ndvi_var <- var(nmsu23_et_pheno[[ndvi_col]], na.rm = TRUE)
  yld_var <- var(nmsu23_et_pheno[[trait]], na.rm = TRUE)
  
  # Calculate scaling factor
  sf <- sqrt(ndvi_var) / sqrt(yld_var)
  
  # Create new scaled yield column
  new_col <- paste0("sca_", trait)
  nmsu23_et_pheno[[new_col]] <- nmsu23_et_pheno[[trait]] * sf
  
  #  Print scaling factor
  cat("Scaling factor for", trait, "using", ndvi_col, ":", sf, "\n")
}


# thus all scaled yields are added as a new column at last. 
```

```{r for checking outliers }
# Fit a linear model
model <- lm(yld2 ~ Rep, data = nmsu23_et_pheno, na.action = na.exclude)

# Calculate residuals
nmsu23_et_pheno$residuals <- residuals(model, na.action = na.exclude)
# Standardize the residuals
nmsu23_et_pheno$residuals_standard <- scale(nmsu23_et_pheno$residuals, center = T, scale = T) 
# Identify outliers
outliers <- nmsu23_et_pheno %>% filter(abs(residuals_standard) > 3)
# Print outliers
# View(outliers)
# none

```


Running MT models

## 2.1. For NDVI: not doin for 2023. Based on 2022 results of mt models and 2023 univariate and phenotypic correlation models, NDVI is not a good indicator here compared to GNDVI so moving forward with GNDVI only.

## 2.2. For GNDVI 
- For Harvest window 2:  "131", "145", "152",  "yld2",
- For Harvest window 3: "160", "165", "173", "179",  "yld3",
- For Harvest window 4: "192", "201", "206",  "yld4",
- For Harvest window 5:  "263",  "278",  "yld5", # "270" filtered out bcz of low h2
- For Harvest window 6:  293", "299", "307", "yld6"
= 21

### 2.2.1. Run models first
- Univariate without spatial
- Univariate with spatial
- Bivariate without spatial
- Bivariate with spatial

#### 2.2.1.1. Univariate for yield without spatial

```{r No spatial }
## For yld1: missing in 2023 data so that's out

## For yld2
st_nmsu23_et_yld2 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld2)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld2, heritA ~ V1 / (V1 + V2)) # 0.46

## For yld3
st_nmsu23_et_yld3 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld3)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld3, heritA ~ V1 / (V1 + V2)) # 0.38

## For yld4
st_nmsu23_et_yld4 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld4)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld4, heritA ~ V1 / (V1 + V2)) # 0.31

## For yld5
st_nmsu23_et_yld5 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld5)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld5, heritA ~ V1 / (V1 + V2)) # 0.48

## For yld6
st_nmsu23_et_yld6 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld6 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) ,
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld6)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld6, heritA ~ V1 / (V1 + V2)) # 0.76

```


#### 2.2.1.2. Univariate for yield with spatial

```{r From spatial BLUPS and Univariate models}
asreml.options(extra =200, maxit =200)
## For yld1 :missing harvest 1 in 2023

## For yld2
st_nmsu23_et_yld2 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld2)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld2, heritA ~ V1 / (V1 + V5)) # 0.67

## For yld3
st_nmsu23_et_yld3 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld3)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld3, heritA ~ V1 / (V1 + V5)) # 0.85

## For yld4
st_nmsu23_et_yld4 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld4 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld4)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld4, heritA ~ V1 / (V1 + V5)) # 0.67

## For yld5
st_nmsu23_et_yld5 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld5 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld5)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld5, heritA ~ V1 / (V1 + V5)) # 0.88

## For yld6
st_nmsu23_et_yld6 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld6 ~ 1 + Rep, #rep effect as fixed
         random = ~ vm(Germplasm,  grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = nmsu23_et_pheno)}
summary(st_nmsu23_et_yld6)$varcomp # to check if the model converged well
asreml::vpredict(st_nmsu23_et_yld6, heritA ~ V1 / (V1 + V5)) # 0.93
```



#### 2.2.1.3. Bivariate for yield without spatial

```{r Bivariate models without Spatial based on GNDVI}
colnames(nmsu23_et_pheno)

## lets proceed with calculation for all harvest windows
# 1.   For harvest window 2: "131", "145", "152", "yld2", 
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for GNDVI_131:
mt_nmsu23_et_gndvi_131 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_131) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_131)$varcomp # residual varaince scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu23_et_gndvi_131, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.9830774
asreml::vpredict(mt_nmsu23_et_gndvi_131, res.cor ~ (V6)/sqrt(V5*V7)) # 0.273769

 ## For GNDVI_145
mt_nmsu23_et_gndvi_145 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_145) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_145)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_145, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.5086786
asreml::vpredict(mt_nmsu23_et_gndvi_145, res.cor ~ (V6)/sqrt(V5*V7)) # 0.5276487

## For GNDVI_152
mt_nmsu23_et_gndvi_152 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_152) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_152)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_152, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.5701623
asreml::vpredict(mt_nmsu23_et_gndvi_152, res.cor ~ (V6)/sqrt(V5*V7)) # 0.5568751


# 2. For harvest window 3:  160, "165", "173", "179", "yld3" 
## For GNDVI_160
mt_nmsu23_et_gndvi_160 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_160) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_160)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_160, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.9114225
asreml::vpredict(mt_nmsu23_et_gndvi_160, res.cor ~ (V6)/sqrt(V5*V7)) # - 0.005413357

## For GNDVI_165
mt_nmsu23_et_gndvi_165 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_165) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_165)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_165, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.3136286
asreml::vpredict(mt_nmsu23_et_gndvi_165, res.cor ~ (V6)/sqrt(V5*V7)) # 0.7748986

## For GNDVI_173
mt_nmsu23_et_gndvi_173 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_173) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_173)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_173, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.09916709
asreml::vpredict(mt_nmsu23_et_gndvi_173, res.cor ~ (V6)/sqrt(V5*V7)) # 0.115543

## For GNDVI_179: didnt run
mt_nmsu23_et_gndvi_179 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_179) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):diag(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_179)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_179, gen.cor ~ (V2)/sqrt(V1*V3)) # BOUND
asreml::vpredict(mt_nmsu23_et_gndvi_179, res.cor ~ (V6)/sqrt(V5*V7)) # BOUND


# 3. For harvest window 4: "192", "201", "206", "yld4",
## For GNDVI_192
mt_nmsu23_et_gndvi_192 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_192) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_192)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_192, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.1262285
asreml::vpredict(mt_nmsu23_et_gndvi_192, res.cor ~ (V6)/sqrt(V5*V7)) # 0.6517725

## For GNDVI_201
mt_nmsu23_et_gndvi_201 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_201) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_201)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_201, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.110522
asreml::vpredict(mt_nmsu23_et_gndvi_201, res.cor ~ (V6)/sqrt(V5*V7)) # 0.7491529

## For GNDVI_206
mt_nmsu23_et_gndvi_206 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_206) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_206)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_206, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.3100105
asreml::vpredict(mt_nmsu23_et_gndvi_206, res.cor ~ (V6)/sqrt(V5*V7)) # 0.7478834


# 4. For harvest window 5:  "263", "270", "278" "yld5",
## For GNDVI_263
mt_nmsu23_et_gndvi_263 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_263) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_263)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_263, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9562021
asreml::vpredict(mt_nmsu23_et_gndvi_263, res.cor ~ (V6)/sqrt(V5*V7)) # 0.8732402


## For GNDVI_278
mt_nmsu23_et_gndvi_278 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_278) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_278)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_278, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.951961
asreml::vpredict(mt_nmsu23_et_gndvi_278, res.cor ~ (V6)/sqrt(V5*V7)) # 0.8641695

# 5. For harvest window 6: "293", "299", "307", with yld 6
## For GNDVI_293
mt_nmsu23_et_gndvi_293 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_293) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_293)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_293, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9334501
asreml::vpredict(mt_nmsu23_et_gndvi_293, res.cor ~ (V6)/sqrt(V5*V7)) # 0.5991287

## For GNDVI_299
mt_nmsu23_et_gndvi_299 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_299) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_299)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_299, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9351264
asreml::vpredict(mt_nmsu23_et_gndvi_299, res.cor ~ (V6)/sqrt(V5*V7)) # 0.3622297

## For GNDVI_307: didn't converge
mt_nmsu23_et_gndvi_307 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld6, GNDVI_307) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_307)$varcomp # 
# mt_nmsu23_et_gndvi_307 <- update.asreml(mt_nmsu23_et_gndvi_307, START = mt_nmsu23_et_gndvi_307$Gammas)
asreml::vpredict(mt_nmsu23_et_gndvi_307, gen.cor ~ (V2)/sqrt(V1*V3)) # BOUND
asreml::vpredict(mt_nmsu23_et_gndvi_307, res.cor ~ (V6)/sqrt(V5*V7)) # BOUND


```



#### 2.2.1.4. Bivariate for yield with spatial
remove 270 (too low h2 and thus no gain here)
```{r Bivariate models with Spatial based on GNDVI}
colnames(nmsu23_et_pheno)

## lets proceed with calculation for all harvest windows
# 1.   For harvest window 2: "131", "145", "152", "yld2", 
asreml.options(maxit = 300, extra = 300)
## now run the models multitrait with yld from each window with its imaging events
 ## for GNDVI_131:
mt_nmsu23_et_gndvi_131 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_131) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_131)$varcomp # residual variance scaling for multivariate model (4th row) is set to 1 to make residual correlation structure identifiable.
asreml::vpredict(mt_nmsu23_et_gndvi_131, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.2630543
asreml::vpredict(mt_nmsu23_et_gndvi_131, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.1265622
asreml::vpredict(mt_nmsu23_et_gndvi_131, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6722622

## For GNDVI_145
mt_nmsu23_et_gndvi_145 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_145) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_145)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_145, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7765529
asreml::vpredict(mt_nmsu23_et_gndvi_145, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4286534
asreml::vpredict(mt_nmsu23_et_gndvi_145, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6076202

## For GNDVI_152
mt_nmsu23_et_gndvi_152 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_152) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_152)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_152, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8446707
asreml::vpredict(mt_nmsu23_et_gndvi_152, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7182171
asreml::vpredict(mt_nmsu23_et_gndvi_152, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6715827


# 2. For harvest window 3:  160, "165", "173", "179", "yld3" 
## For GNDVI_160: error with scaled, worked with original yld3.
mt_nmsu23_et_gndvi_160 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_160) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_160)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_160, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.680636
asreml::vpredict(mt_nmsu23_et_gndvi_160, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7128783
asreml::vpredict(mt_nmsu23_et_gndvi_160, res.cor ~ (V11)/sqrt(V10*V12)) # 0.57562

## For GNDVI_165
mt_nmsu23_et_gndvi_165 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_165) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_165)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_165, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7930669
asreml::vpredict(mt_nmsu23_et_gndvi_165, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7816402
asreml::vpredict(mt_nmsu23_et_gndvi_165, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7760096

## For GNDVI_173: didn't run with full, worked with diag for residual...
mt_nmsu23_et_gndvi_173 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_173) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_173)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_173, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8790034
asreml::vpredict(mt_nmsu23_et_gndvi_173, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.861411
asreml::vpredict(mt_nmsu23_et_gndvi_173, res.cor ~ (V11)/sqrt(V10*V12)) # diag

## For GNDVI_179
mt_nmsu23_et_gndvi_179 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_179) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_179)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_179, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9586635
asreml::vpredict(mt_nmsu23_et_gndvi_179, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6801659
asreml::vpredict(mt_nmsu23_et_gndvi_179, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8052792


# 3. For harvest window 4: "192", "201", "206", "yld4",

## For GNDVI_192
mt_nmsu23_et_gndvi_192 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_192) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_192)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_192, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7235934
asreml::vpredict(mt_nmsu23_et_gndvi_192, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7604502
asreml::vpredict(mt_nmsu23_et_gndvi_192, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6325876

## For GNDVI_201
mt_nmsu23_et_gndvi_201 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_201) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_201)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_201, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8093027
asreml::vpredict(mt_nmsu23_et_gndvi_201, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7928979
asreml::vpredict(mt_nmsu23_et_gndvi_201, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7597422

## For GNDVI_206
mt_nmsu23_et_gndvi_206 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld4, GNDVI_206) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_206)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_206, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8498385
asreml::vpredict(mt_nmsu23_et_gndvi_206, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7835405
asreml::vpredict(mt_nmsu23_et_gndvi_206, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6392668


# 4. For harvest window 5: "263",  "278", "yld5",
## For GNDVI_263
mt_nmsu23_et_gndvi_263 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_263) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_263)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_263, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.934741
asreml::vpredict(mt_nmsu23_et_gndvi_263, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.9273239
asreml::vpredict(mt_nmsu23_et_gndvi_263, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6100142


## For GNDVI_278
mt_nmsu23_et_gndvi_278 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld5, GNDVI_278) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_278)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_278, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9665625
asreml::vpredict(mt_nmsu23_et_gndvi_278, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8819769
asreml::vpredict(mt_nmsu23_et_gndvi_278, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6475184


# 5. For harvest window 6: "293", "299", "307", with yld6
## For GNDVI_293
mt_nmsu23_et_gndvi_293 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_293) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_293)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_293, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9492882
asreml::vpredict(mt_nmsu23_et_gndvi_293, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6937897
asreml::vpredict(mt_nmsu23_et_gndvi_293, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7364867

## For GNDVI_299
mt_nmsu23_et_gndvi_299 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_299) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_299)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_299, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.960965
asreml::vpredict(mt_nmsu23_et_gndvi_299, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5564262
asreml::vpredict(mt_nmsu23_et_gndvi_299, res.cor ~ (V11)/sqrt(V10*V12)) # 0.5753187

## For GNDVI_307
mt_nmsu23_et_gndvi_307 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld6, GNDVI_307) ~ 1 + Rep, #rep effect as fixed
         random = ~ us(trait):vm(Germplasm,  grm_inverse) +us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait),
         data = nmsu23_et_pheno)}
summary(mt_nmsu23_et_gndvi_307)$varcomp # 
asreml::vpredict(mt_nmsu23_et_gndvi_307, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.9225547
asreml::vpredict(mt_nmsu23_et_gndvi_307, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7170551
asreml::vpredict(mt_nmsu23_et_gndvi_307, res.cor ~ (V11)/sqrt(V10*V12)) # 0.6832498
```



END of modeling
### 2.2.2. Graphs





 