---
title: "Running univariate models with GRM and spatial to calculate Heritability_ASReml in NY 2023 data"
author: "Dinesh Ghimire"
date: "2026-05-12"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr","asreml", "readxl", "tidyr", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "GGally")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```

#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NY-DG"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NY-DG 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa-NY-DG"
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is misssing please let me know. 
key<-read.csv("2022/meta_data/cultivar_list_key.csv") # contains both CU and NY cultivars here. 
# its same for 2023 and 2023 (cultivar list)
```




#NY 22


################################# Musgrave Trial ################################



1. *NY Musgrave 2023 *

Loading and preparing data files
```{r loading files 05122026}
#import merged phenotype file 
ny_mus_pheno_23 <- read.csv("2023/MSI_data/output_files/Final_NY23_Mus_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_mus_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_mus_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_mus_pheno_23) 
str(ny_mus_pheno_23)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))

grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


- Scaling and outlier check on yield already done in script1, so we can jump into running models


Run the models
VI's
## 125, 136, 144, 150, 152, 156, 164, 166, 172, 177, 179, 186, 192, 195, 198, 213, 223, 228, 244, 248, 251, 255, 263, 274, 278, 300, 305
 # but 3rd harvest was in 255 Julian day meaning after that were flights after all harvests done for that season. Thus, for spatial study to fit RR, we don't need flights after 3rd harvest. So not using 263, 274, 278, 300, 305.
# Final flights upto 3rd harvest:
## 125, 136, 144, 150, 152, 156, 164, 166, 172, 177, 179, 186, 192, 195, 198, 213, 223, 228, 244, 248, 251, 255
 # we have 22 imaging events so 22 heritabilities here. 


```{r run univariate models with spatial and GRM 05122026}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
colnames(ny_mus_pheno_23) 

# first for NDVI
# Final flights upto 3rd harvest:
## 125, 136, 144, 150, 152, 156, 164, 166, 172, 177, 179, 186, 192, 195, 198, 213, 223, 228, 244, 248, 251, 255
 # we have 22 imaging events so 22 heritabilities here.

#  For NDVI_136
st_NY23_mus_ndvi_136 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_136 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndvi_136)$varcomp

asreml::vpredict(st_NY23_mus_ndvi_136, heritA ~ V1 / (V1 + V5)) # 0.3829762 with 0.1366314

```



For yield 

```{r Heritability of yield with GRM 05122026}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
# for yield 1: 0.99 (spatial vc consumes residual)
st_NY23_mus_yld1 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_yld1)$varcomp
asreml::vpredict(st_NY23_mus_yld1, heritA ~ V1 / (V1 + V5)) # 0.99 with 0.0012 SE


# for yield 2
st_NY23_mus_yld2 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_yld2)$varcomp
asreml::vpredict(st_NY23_mus_yld2, heritA ~ V1 / (V1 + V5)) # 0.5853671 with 0.09204607 SE


# for yield 3
st_NY23_mus_yld3 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_yld3)$varcomp
asreml::vpredict(st_NY23_mus_yld3, heritA ~ V1 / (V1 + V5)) # 0.7050828 with 0.09158598 SE

```





# Run for all VI's
### STORE BOTH HERITABILITY AND VARIANCE COMPONENTS

 To check variance components as well: store VC in another data frame
```{r For all VI's 05122026}
asreml.options(extra=300,maxit =300) # to setup more iterations for successful burn-in or convergence
# 125, 136, 144, 150, 152, 156, 164, 166, 172, 177, 179, 186, 192, 195, 198, 213, 223, 228, 244, 248, 251, 255
# Define time points
times_mus_23 <- c(125, 136, 144, 150, 152, 156, 164, 166, 172, 177, 179, 186, 192, 195, 198, 213, 223, 228, 244, 248, 251, 255)

# Initialize heritability results
herit_results <- data.frame(
  Time = times_mus_23,
  Herit_ndvi = NA, Herit_ndvi_spa = NA,
  Herit_gndvi = NA, Herit_gndvi_spa = NA,
  Herit_ndre = NA, Herit_ndre_spa = NA)

# Initialize variance component storage
varcomp_results <- data.frame(
  Time = numeric(0),
  Index = character(0),
  Model = character(0),
  Component = character(0),
  Estimate = numeric(0),
  stringsAsFactors = FALSE)

# Function to fit model and return heritability and variance components
fit_heritability <- function(index_prefix, time_point, data) {
  col_name <- paste0(index_prefix, "_", time_point)
  
  if (!col_name %in% colnames(data)) {
    warning(paste("Column", col_name, "not found. Skipping."))
    return(list(h2 = c(NA, NA), vc = NULL))
  }
  
  # --- Simple Model ---
  mod_simple <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Simple model failed for", col_name))
    return(NULL)
  })
  
  h2_simple <- NA
  vc_simple <- NULL
  if (!is.null(mod_simple)) {
    h2_simple <- tryCatch({
      vpredict(mod_simple, herit ~ V1 / (V1 + V2))$Estimate
    }, error = function(e) NA)
    
    vc_simple <- tryCatch({
      vc <- summary(mod_simple)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Simple")
    }, error = function(e) NULL)
  }
  
  # --- Spatial Model ---
  mod_spatial <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 ")),
           random = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Spatial model failed for", col_name))
    return(NULL)
  })
  
  h2_spatial <- NA
  vc_spatial <- NULL
  if (!is.null(mod_spatial)) {
    h2_spatial <- tryCatch({
      vpredict(mod_spatial, herit_spatial ~ V1 / (V1 + V5))$Estimate
    }, error = function(e) NA)
    
    vc_spatial <- tryCatch({
      vc <- summary(mod_spatial)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Spatial")
    }, error = function(e) NULL)
  }

  return(list(h2 = c(h2_simple, h2_spatial),
              vc = rbind(vc_simple, vc_spatial)))
}

# Loop through all time points and indices
for (i in seq_along(times_mus_23)) {
  time <- times_mus_23[i]
  cat("\nProcessing time point:", time, "\n")
  
  for (index in c("NDVI", "GNDVI", "NDRE")) {
    result <- fit_heritability(index, time, ny_mus_pheno_23)
    
    # Store heritabilities
    herit_results[i, paste0("Herit_", tolower(index))] <- result$h2[1]
    herit_results[i, paste0("Herit_", tolower(index), "_spa")] <- result$h2[2]
    
    # Store variance components
    if (!is.null(result$vc)) {
      result$vc$Index <- index
      result$vc$Time <- time
      varcomp_results <- bind_rows(varcomp_results, result$vc)
    }
  }
}

# Round heritability results to 2 decimals
herit_results_mus_23 <- herit_results %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

# Done!
print(head(herit_results_mus_23))
print(head(varcomp_results))
View(herit_results_mus_23)
```


check for NA's or 1 or 0.99 too
- NDVI with spatial: 152, 195, 244, 248
- GNDVI with spatial: 136, 144, 164, 166, 172, 195, 223, 228, 244, 248, 255
- NDRE with spatial: 152, 156, 177, 213, 248

### NDVI with spatial: 152, 195, 244, 248
```{r NDVI Checking, 05122026}
# NDVI: 152, 0.99 ( spatial borrowed from residual)
st_NY23_mus_ndvi_152 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_152 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndvi_152)$varcomp
asreml::vpredict(st_NY23_mus_ndvi_152, heritA ~ V1 / (V1 + V5)) # 0.99

# NDVI: 195, almost none = 0
st_NY23_mus_ndvi_195 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_195 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndvi_195)$varcomp
asreml::vpredict(st_NY23_mus_ndvi_195, heritA ~ V1 / (V1 + V5)) # 

# NDVI: 244, 0.99 ( spatial borrowed from residual)
st_NY23_mus_ndvi_244 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_244 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndvi_244)$varcomp
asreml::vpredict(st_NY23_mus_ndvi_244, heritA ~ V1 / (V1 + V5)) # 

# NDVI: 248,  0.28(row) 
st_NY23_mus_ndvi_248 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_248 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndvi_248)$varcomp
asreml::vpredict(st_NY23_mus_ndvi_248, heritA ~ V1 / (V1 + V5)) # 
```



### GNDVI with spatial: 136, 144, 164, 166, 172, 195, 223, 228, 244, 248, 255

```{r GNDVI Checking for cleaning, 05072026}
# GNDVI with spatial: 136, 144, 164, 166, 172, 195, 223, 228, 244, 248, 255 
# GNDVI: 136, 0.35. (row only)
st_NY23_mus_gndvi_136 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_136 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_136)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_136, heritA ~ V3 / (V3 + V4)) # 0.3463901

# GNDVI: 144, 0.99 (spatial took all residual)
st_NY23_mus_gndvi_144 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_144 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_144)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_144, heritA ~ V1 / (V1 + V5)) # 

# GNDVI: 164, 0.99 (spatial took all residual)
st_NY23_mus_gndvi_164 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_164 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_164)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_164, heritA ~ V1 / (V1 + V5)) # 

# GNDVI: 166, 0.99 (spatial took all residual)
st_NY23_mus_gndvi_166 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_166 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_166)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_166, heritA ~ V1 / (V1 + V5)) # 

# GNDVI: 172, 0.99 (spatial took all residual)
st_NY23_mus_gndvi_172 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_172 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_172)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_172, heritA ~ V1 / (V1 + V5)) # 

# GNDVI: 195, 0.98 (spatial took all residual)
st_NY23_mus_gndvi_195 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_195 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_195)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_195, heritA ~ V1 / (V1 + V5)) # 

# GNDVI: 223, 0.43 (row)
st_NY23_mus_gndvi_223 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_223 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_223)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_223, heritA ~ V3 / (V3 + V4)) # 

# GNDVI: 228, 0.99 (spatial took all residual)
st_NY23_mus_gndvi_228 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_228 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_228)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_228, heritA ~ V1 / (V1 + V5)) # 

# GNDVI: 244, 0.99 (spatial took all residual)
st_NY23_mus_gndvi_244 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_244 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_244)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_244, heritA ~ V1 / (V1 + V5)) # 

# GNDVI: 248, almost none : 0.1 (row) - check with mt (checked fint, so ran full model here, 0.99 h2)
st_NY23_mus_gndvi_248 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_248 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_248)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_248, heritA ~ V1 / (V1 + V5)) # 

# GNDVI: 255, 0.99 (both) - check with mt model cz only row is less
st_NY23_mus_gndvi_255 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_255 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_gndvi_255)$varcomp
asreml::vpredict(st_NY23_mus_gndvi_255, heritA ~ V1 / (V1 + V5)) #
asreml::vpredict(st_NY23_mus_gndvi_255, heritA ~ V3 / (V3 + V4)) # 

```



### NDRE with spatial: 152, 156, 177, 213, 248
```{r NDRE Checking, 05112026}
## NDRE with spatial: 152, 156, 177, 213, 248
# NDRE: 152, almost none = 0
st_NY23_mus_ndre_152 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_152 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndre_152)$varcomp
asreml::vpredict(st_NY23_mus_ndre_152, heritA ~ V3 / (V3 + V4)) # 

# NDRE: 156, 0.98
st_NY23_mus_ndre_156 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_156 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndre_156)$varcomp
asreml::vpredict(st_NY23_mus_ndre_156, heritA ~ V1 / (V1 + V5)) # 

# NDRE: 177, 0.99 ( spatial took all of residual)
st_NY23_mus_ndre_177 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_177 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndre_177)$varcomp
asreml::vpredict(st_NY23_mus_ndre_177, heritA ~ V1 / (V1 + V5)) # 

# NDRE: 213, almost none = 0 (full and row only both)
st_NY23_mus_ndre_213 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_213 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndre_213)$varcomp
asreml::vpredict(st_NY23_mus_ndre_213, heritA ~ V1 / (V1 + V5)) # 

# NDRE: 248, almost none = 0
st_NY23_mus_ndre_248 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_248 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_mus_pheno_23)}
summary(st_NY23_mus_ndre_248)$varcomp
asreml::vpredict(st_NY23_mus_ndre_248, heritA ~ V1 / (V1 + V5)) # 
```






## make box plot to visulaize and compare 

```{r Box plot to visualize heritability, Mus_23}
herit_results_mus_23 <- read.csv("2023/MSI_data/output_files/Heritabilities_withGRM_NY_Musgrave_23.csv", sep = ",") 
colnames(herit_results_mus_23)

# Reshape all six heritability columns to long format
herit_results_mus_23_long_all <- herit_results_mus_23 %>%
  select(Time, h2_ndvi, h2_ndvi_spa,
               h2_gndvi, h2_gndvi_spa,
               h2_ndre, h2_ndre_spa) %>%
  pivot_longer(cols = -Time, names_to = "Type", values_to = "Heritability") %>%
  mutate(
    Index = case_when(
  grepl("gndvi", Type) ~ "GNDVI",
  grepl("ndvi", Type) ~ "NDVI",
  grepl("ndre", Type) ~ "NDRE"
),
    Model = ifelse(grepl("_spa", Type), "Spatial", "Simple"))


gg_mus_23 <- ggplot(herit_results_mus_23_long_all, aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") + # Randomly "jitter" (spread out) the data points horizontally by up to ±0.2 units on the x-axis (i.e., within each box)
  # alternate geom_point() where all will be stacked in a vertical line so overalp issue.
  facet_wrap(~ Model) +
  labs(title = "h² by VI and Model Type in Musgrave NY 2023",
       x = "Vegetation Index",
       y = "Heritability (h²)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(c(0,1)) +
  scale_fill_brewer(palette = "Set2")

gg_mus_23

ggsave(filename = "Boxplot_Heritability_h2_VI_NY_Musgrave_2023.png", plot = gg_mus_23, path = "2023/output_images", width = 8, height = 7,dpi = 400)


############# Only simple h2 #######

# make boxplot of h2_ndvi, h2_gndvi, h2_ndre ( not spatial ones)

# Reshape only non-spatial heritability columns
herit_results_mus_23_long_simple <- herit_results_mus_23 %>%
  select(Time, h2_ndvi, h2_gndvi, h2_ndre) %>%
  pivot_longer(
    cols = -Time,
    names_to = "Type",
    values_to = "Heritability") %>%
  mutate(
    Index = case_when(
      Type == "h2_ndvi"  ~ "NDVI",
      Type == "h2_gndvi" ~ "GNDVI",
      Type == "h2_ndre"  ~ "NDRE"))

# Boxplot
gg_mus_23_simple <- ggplot(
  herit_results_mus_23_long_simple,
  aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +
  labs(
    title = "h² by Vegetation Index in Musgrave NY 2023",
    x = "Vegetation Index",
    y = "Heritability (h²)" ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(0, 1) +
  scale_fill_brewer(palette = "Set2")

gg_mus_23_simple

ggsave(filename = "Boxplot_Heritability_h2_VI_NY_Musgrave_2023_Simple_only.png",plot = gg_mus_23_simple,path = "2023/output_images",width = 7,height = 6,dpi = 400)

```






################################# Snyder Trial ################################




1. *NY Snyder 2023 *

Loading and preparing data files
```{r loading files 05132026}
#import merged phenotype file 
ny_sny_pheno_23 <- read.csv("2023/MSI_data/output_files/Final_NY23_Sny_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_sny_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_sny_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_sny_pheno_23) 
str(ny_sny_pheno_23)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))

grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


- Scaling and outlier check on yield already done in script1, so we can jump into running models


Run the models
VI's
### 125, 142, 145, 150, 152, 156, 164, 166, 172, 178, 184, 187, 193, 198, 213, 223, 228, 251, 255, 265, 277, 300, 306
 # but 3rd harvest was in 250th Julian day meaning after that were flights after all harvests done for that season. Thus, for spatial study to fit RR, we don't need flights after 3rd harvest. So not using 251, 255, 265, 277, 300, 306
# Final flights upto 3rd harvest:
### 125, 142, 145, 150, 152, 156, 164, 166, 172, 178, 184, 187, 193, 198, 213, 223, 228
 # we have 17 imaging events so 17 heritabilities here. 


```{r run univariate models with spatial and GRM 05122026}
asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
colnames(ny_sny_pheno_23) 

# first for NDVI
# we have 23 imaging events so 23 heritabilities here. 
### 125, 142, 145, 150, 152, 156, 164, 166, 172, 178, 184, 187, 193, 198, 213, 223, 228
#  For NDVI_145
st_ny23_sny_ndvi_145 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_145 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Range):ar1(Row),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_ny23_sny_ndvi_145)$varcomp

asreml::vpredict(st_ny23_sny_ndvi_145, heritA ~ V1 / (V1 + V5)) # 0.72 with 0.11

```



For yield 

```{r Heritability of yield with GRM 05132026}

asreml.options(extra=200,maxit =200) # to setup more iterations for successful burn-in or convergence
# for yield 1
st_ny23_sny_yld1 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld1 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_ny23_sny_yld1)$varcomp
asreml::vpredict(st_ny23_sny_yld1, heritA ~ V1 / (V1 + V5)) # 0.7271587 with 0.11 se


# for yield 2
st_ny23_sny_yld2 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld2 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_ny23_sny_yld2)$varcomp
asreml::vpredict(st_ny23_sny_yld2, heritA ~ V1 / (V1 + V5)) # 0.7529702 with 0.11 SE


# for yield 3
st_ny23_sny_yld3 <- {    
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = yld3 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_ny23_sny_yld3)$varcomp
asreml::vpredict(st_ny23_sny_yld3, heritA ~ V1 / (V1 + V5)) # 0.6835125 with 0.09 SE
```





# Run for all VI's
### STORE BOTH HERITABILITY AND VARIANCE COMPONENTS

 To check variance components as well: store VC in another data frame
```{r For all VI's 05132026}
asreml.options(extra=300,maxit =300) # to setup more iterations for successful burn-in or convergence
# 125, 142, 145, 150, 152, 156, 164, 166, 172, 178, 184, 187, 193, 198, 213, 223, 228
# Define time points
times_sny_23 <- c(125, 142, 145, 150, 152, 156, 164, 166, 172, 178, 184, 187, 193, 198, 213, 223, 228)

# Initialize heritability results
herit_results <- data.frame(
  Time = times_sny_23,
  Herit_ndvi = NA, Herit_ndvi_spa = NA,
  Herit_gndvi = NA, Herit_gndvi_spa = NA,
  Herit_ndre = NA, Herit_ndre_spa = NA)

# Initialize variance component storage
varcomp_results <- data.frame(
  Time = numeric(0),
  Index = character(0),
  Model = character(0),
  Component = character(0),
  Estimate = numeric(0),
  stringsAsFactors = FALSE)

# Function to fit model and return heritability and variance components
fit_heritability <- function(index_prefix, time_point, data) {
  col_name <- paste0(index_prefix, "_", time_point)
  
  if (!col_name %in% colnames(data)) {
    warning(paste("Column", col_name, "not found. Skipping."))
    return(list(h2 = c(NA, NA), vc = NULL))
  }
  
  # --- Simple Model ---
  mod_simple <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 + Rep")),
           random = ~ vm(Germplasm, grm_inverse),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Simple model failed for", col_name))
    return(NULL)
  })
  
  h2_simple <- NA
  vc_simple <- NULL
  if (!is.null(mod_simple)) {
    h2_simple <- tryCatch({
      vpredict(mod_simple, herit ~ V1 / (V1 + V2))$Estimate
    }, error = function(e) NA)
    
    vc_simple <- tryCatch({
      vc <- summary(mod_simple)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Simple")
    }, error = function(e) NULL)
  }
  
  # --- Spatial Model ---
  mod_spatial <- tryCatch({
    attr(grm_inverse, "INVERSE") <- TRUE
    asreml(fixed = as.formula(paste0(col_name, " ~ 1 ")),
           random = ~ vm(Germplasm, grm_inverse) + ar1(Range):ar1(Row),
           residual = ~ id(units),
           data = data)
  }, error = function(e) {
    warning(paste("Spatial model failed for", col_name))
    return(NULL)
  })
  
  h2_spatial <- NA
  vc_spatial <- NULL
  if (!is.null(mod_spatial)) {
    h2_spatial <- tryCatch({
      vpredict(mod_spatial, herit_spatial ~ V1 / (V1 + V5))$Estimate
    }, error = function(e) NA)
    
    vc_spatial <- tryCatch({
      vc <- summary(mod_spatial)$varcomp
      data.frame(Component = rownames(vc), Estimate = vc[, "component"], Model = "Spatial")
    }, error = function(e) NULL)
  }

  return(list(h2 = c(h2_simple, h2_spatial),
              vc = rbind(vc_simple, vc_spatial)))
}

# Loop through all time points and indices
for (i in seq_along(times_sny_23)) {
  time <- times_sny_23[i]
  cat("\nProcessing time point:", time, "\n")
  
  for (index in c("NDVI", "GNDVI", "NDRE")) {
    result <- fit_heritability(index, time, ny_sny_pheno_23)
    
    # Store heritabilities
    herit_results[i, paste0("Herit_", tolower(index))] <- result$h2[1]
    herit_results[i, paste0("Herit_", tolower(index), "_spa")] <- result$h2[2]
    
    # Store variance components
    if (!is.null(result$vc)) {
      result$vc$Index <- index
      result$vc$Time <- time
      varcomp_results <- bind_rows(varcomp_results, result$vc)
    }
  }
}

# Round heritability results to 2 decimals
herit_results_sny_23 <- herit_results %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

# Done!
print(head(herit_results_sny_23))
# print(head(varcomp_results))
View(herit_results_sny_23)

```



check for NA's or 1 or 0.99 too
- NDVI with spatial: 125, 150, 152, 156, 166, 184, 187, 193, 223
- GNDVI with spatial: 125, 142, 145,150, 152, 156, 166, 178, 184, 187, 193, 213, 223, 228
- NDRE with spatial: 164, 184, 187


### NDVI with spatial: 125, 150, 152, 156, 166, 184, 187, 193, 223

```{r Checking NDVI for h2 confirmation, 05132026}
# NDVI: 125, 0.99 ( spatial borrowed from residual), ran without grm to double check if the convergence is real
st_NY23_sny_ndvi_125 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_125 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndvi_125)$varcomp
asreml::vpredict(st_NY23_sny_ndvi_125, heritA ~ V1 / (V1 + V5)) # 0.99


# NDVI: 150, 0.97 ( spatial borrowed from residual), ran without grm
st_NY23_sny_ndvi_150 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_150 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndvi_150)$varcomp
asreml::vpredict(st_NY23_sny_ndvi_150, heritA ~ V1 / (V1 + V5)) # 0.97


# NDVI: 152, 0.99 ( spatial borrowed from residual), but check with mt cz not converged
st_NY23_sny_ndvi_152 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_152 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndvi_152)$varcomp
asreml::vpredict(st_NY23_sny_ndvi_152, heritA ~ V1 / (V1 + V5)) # 0.99


# NDVI: 156, no gen varaince
st_NY23_sny_ndvi_156 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_156 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndvi_156)$varcomp
asreml::vpredict(st_NY23_sny_ndvi_156, heritA ~ V1 / (V1 + V5)) #


# NDVI: 166, 0.99 ( spatial borrowed from residual)
st_NY23_sny_ndvi_166 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_166 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndvi_166)$varcomp
asreml::vpredict(st_NY23_sny_ndvi_166, heritA ~ V1 / (V1 + V5)) # 0.99


# NDVI: 184, 0.99 ( spatial borrowed from residual)
st_NY23_sny_ndvi_184 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_184 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndvi_184)$varcomp
asreml::vpredict(st_NY23_sny_ndvi_184, heritA ~ V1 / (V1 + V5)) # 0.99


# NDVI: 187, 0.99 ( spatial borrowed from residual)
st_NY23_sny_ndvi_187 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_187 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndvi_187)$varcomp
asreml::vpredict(st_NY23_sny_ndvi_187, heritA ~ V1 / (V1 + V5)) # 0.99


# NDVI: 193, 0.96 ( spatial borrowed from residual), similar high without grm
st_NY23_sny_ndvi_193 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_193 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndvi_193)$varcomp
asreml::vpredict(st_NY23_sny_ndvi_193, heritA ~ V1 / (V1 + V5)) # 0.96


# NDVI: 223, 0.99 ( spatial borrowed from residual).. double check with mt
st_NY23_sny_ndvi_223 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDVI_223 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndvi_223)$varcomp
asreml::vpredict(st_NY23_sny_ndvi_223, heritA ~ V1 / (V1 + V5)) # 0.99

```




### GNDVI with spatial: 125, 142, 145,150, 152, 156, 166, 178, 184, 187, 193, 213, 223, 228
```{r Checking GNDVI for h2 confirmation, 05132026}
# GNDVI: 125, 0.99 ( spatial borrowed from residual), double check with mt cz its bound
st_NY23_sny_gndvi_125 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_125 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_125)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_125, heritA ~ V1 / (V1 + V5)) # 0.99

# GNDVI: 142, 0.99 ( spatial borrowed from residual), double check with mt cz its bound
st_NY23_sny_gndvi_142 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_142 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_142)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_142, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 145, 0.99 ( spatial borrowed from residual), , double check with mt cz its bound
st_NY23_sny_gndvi_145 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_145 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_145)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_145, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 150, 0.99 ( spatial borrowed from residual)
st_NY23_sny_gndvi_150 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_150 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_150)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_150, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 152, 0.99 ( spatial borrowed from residual)
st_NY23_sny_gndvi_152 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_152 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_152)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_152, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 156, 0.91 ( spatial borrowed from residual)
st_NY23_sny_gndvi_156 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_156 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_156)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_156, heritA ~ V1 / (V1 + V5)) # 0.91


# GNDVI: 166, 0.99 ( spatial borrowed from residual)
st_NY23_sny_gndvi_166 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_166 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_166)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_166, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 178, 0.99 ( spatial borrowed from residual), double check with mt cz genetic effect is bound
st_NY23_sny_gndvi_178 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_178 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_178)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_178, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 184, 0.99 ( spatial borrowed from residual)
st_NY23_sny_gndvi_184 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_184 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_184)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_184, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 187, 0.99 ( spatial borrowed from residual), double check with mt cz genetic effect is bound
st_NY23_sny_gndvi_187 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_187 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_187)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_187, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 193, 0.99 ( spatial borrowed from residual), double check with mt cz genetic effect is bound
st_NY23_sny_gndvi_193 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_193 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_193)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_193, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 213, 0.99 ( spatial borrowed from residual)
st_NY23_sny_gndvi_213 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_213 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_213)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_213, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 223, 0.99 ( spatial borrowed from residual), not converged so double check with mt models
st_NY23_sny_gndvi_223 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_223 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_223)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_223, heritA ~ V1 / (V1 + V5)) # 0.99


# GNDVI: 228, 0.99 ( spatial borrowed from residual)
st_NY23_sny_gndvi_228 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = GNDVI_228 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_gndvi_228)$varcomp
asreml::vpredict(st_NY23_sny_gndvi_228, heritA ~ V1 / (V1 + V5)) # 0.99

```



###  NDRE with spatial: 164, 184, 187
```{r Checking NDRE for h2 confirmation, 05132026}
# NDRE: 164, 0.99 ( spatial borrowed from residual)
st_NY23_sny_ndre_164 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_164 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndre_164)$varcomp
asreml::vpredict(st_NY23_sny_ndre_164, heritA ~ V1 / (V1 + V5)) # 0.99

# NDRE: 184, 0.99 ( spatial borrowed from residual), double check with mt models cz its bound
st_NY23_sny_ndre_184 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_184 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndre_184)$varcomp
asreml::vpredict(st_NY23_sny_ndre_184, heritA ~ V1 / (V1 + V5)) # 0.99

# NDRE: 187, 0.99 ( spatial borrowed from residual)
st_NY23_sny_ndre_187 <- { 
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = NDRE_187 ~ 1 , 
         random = ~ vm(Germplasm, grm_inverse) +ar1(Row):ar1(Range),
         residual = ~ id(units),
         data = ny_sny_pheno_23)}
summary(st_NY23_sny_ndre_187)$varcomp
asreml::vpredict(st_NY23_sny_ndre_187, heritA ~ V1 / (V1 + V5)) # 0.99

```


## make box plot to visulaize and compare 

```{r Box plot to visualize heritability, Snyder_23}
herit_results_sny_23 <- read.csv("2023/MSI_data/output_files/Heritabilities_withGRM_NY_Snyder_23.csv", sep = ",") 
colnames(herit_results_sny_23)

# Reshape all six heritability columns to long format
herit_results_sny_23_long_all <- herit_results_sny_23 %>%
  select(Time, h2_ndvi, h2_ndvi_spa,
               h2_gndvi, h2_gndvi_spa,
               h2_ndre, h2_ndre_spa) %>%
  pivot_longer(cols = -Time, names_to = "Type", values_to = "Heritability") %>%
  mutate(
    Index = case_when(
  grepl("gndvi", Type) ~ "GNDVI",
  grepl("ndvi", Type) ~ "NDVI",
  grepl("ndre", Type) ~ "NDRE"
),
    Model = ifelse(grepl("_spa", Type), "Spatial", "Simple"))


gg_sny_23 <- ggplot(herit_results_sny_23_long_all, aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") + # Randomly "jitter" (spread out) the data points horizontally by up to ±0.2 units on the x-axis (i.e., within each box)
  # alternate geom_point() where all will be stacked in a vertical line so overalp issue.
  facet_wrap(~ Model) +
  labs(title = "h² by VI and Model Type in Snyder NY 2023",
       x = "Vegetation Index",
       y = "Heritability (h²)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(c(0,1)) +
  scale_fill_brewer(palette = "Set2")

gg_sny_23

ggsave(filename = "Boxplot_Heritability_h2_VI_NY_Snyder_2023.png", plot = gg_sny_23, path = "2023/output_images", width = 8, height = 7,dpi = 400)




############# Only simple h2 #######
# make boxplot of h2_ndvi, h2_gndvi, h2_ndre ( not spatial ones)

# Reshape only non-spatial heritability columns
herit_results_sny_23_long_simple <- herit_results_sny_23 %>%
  select(Time, h2_ndvi, h2_gndvi, h2_ndre) %>%
  pivot_longer(
    cols = -Time,
    names_to = "Type",
    values_to = "Heritability") %>%
  mutate(
    Index = case_when(
      Type == "h2_ndvi"  ~ "NDVI",
      Type == "h2_gndvi" ~ "GNDVI",
      Type == "h2_ndre"  ~ "NDRE"))

# Boxplot
gg_sny_23_simple <- ggplot(
  herit_results_sny_23_long_simple,
  aes(x = Index, y = Heritability, fill = Index)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +
  labs(
    title = "h² by Vegetation Index in Snyder NY 2023",
    x = "Vegetation Index",
    y = "Heritability (h²)" ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(0, 1) +
  scale_fill_brewer(palette = "Set2")

gg_sny_23_simple

ggsave(filename = "Boxplot_Heritability_h2_VI_NY_Snyder_2023_Simple_only.png",plot = gg_sny_23_simple,path = "2023/output_images",width = 7,height = 6,dpi = 400)
```








SNY 2023
For Harvest window 1:  125, 136, 144, 150, and yld1
- For Harvest window 2:  152, 156, 164, 166, 172, 177, 179, 186, 192, 195, 198, and yld2
- For Harvest window 3: 213, 223, 228, 244, 248, 251, 255, and yld3

```{r Variance component for Snyder 2023 for all yield and last gndvi 06052026}
### =====================================================
### Snyder 2023 Variance components
### =====================================================

### =====================================================
### Harvest 1: GNDVI_150 <-> yld1
### =====================================================

# GNDVI_150
st_ny23_sny_gndvi150 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_150 ~ 1 + Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_gndvi150)$varcomp

st_ny23_sny_gndvi150_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_150 ~ 1,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_gndvi150_spa)$varcomp

# yld1
st_ny23_sny_yld1_base <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld1 ~ 1+ Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_yld1_base)$varcomp

st_ny23_sny_yld1 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld1 ~ 1,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_yld1)$varcomp


### =====================================================
### Harvest 2: GNDVI_198 <-> yld2
### =====================================================

# GNDVI_198
st_ny23_sny_gndvi198 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_198 ~ 1+ Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_gndvi198)$varcomp

st_ny23_sny_gndvi198_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_198 ~ 1,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_gndvi198_spa)$varcomp

# yld2
st_ny23_sny_yld2_base <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld2 ~ 1+ Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_yld2_base)$varcomp

st_ny23_sny_yld2 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld2 ~ 1,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_yld2)$varcomp


### =====================================================
### Harvest 3: GNDVI_255 <-> yld3
### =====================================================

# GNDVI_255
st_ny23_sny_gndvi255 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_255 ~ 1+ Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_gndvi255)$varcomp

st_ny23_sny_gndvi255_spa <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = GNDVI_255 ~ 1,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_gndvi255_spa)$varcomp

# yld3
st_ny23_sny_yld3_base <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld3 ~ 1+ Rep,
    random = ~ vm(Germplasm, grm_inverse),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_yld3_base)$varcomp

st_ny23_sny_yld3 <- {
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(
    fixed = yld3 ~ 1,
    random = ~ vm(Germplasm, grm_inverse) + ar1(Row):ar1(Range),
    residual = ~ id(units),
    data = ny_sny_pheno_23)}
summary(st_ny23_sny_yld3)$varcomp



### =====================================================
### Variance component summary table (Base)
### =====================================================

varcomp_sny23_base <- data.frame(
  Trait = c("GNDVI_150","yld1",
            "GNDVI_198","yld2",
            "GNDVI_255","yld3"),
  Genetic = c(
    summary(st_ny23_sny_gndvi150)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_yld1_base)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_gndvi198)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_yld2_base)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_gndvi255)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_yld3_base)$varcomp["vm(Germplasm, grm_inverse)","component"]),
  Residual = c(
    summary(st_ny23_sny_gndvi150)$varcomp["units!R","component"],
    summary(st_ny23_sny_yld1_base)$varcomp["units!R","component"],
    summary(st_ny23_sny_gndvi198)$varcomp["units!R","component"],
    summary(st_ny23_sny_yld2_base)$varcomp["units!R","component"],
    summary(st_ny23_sny_gndvi255)$varcomp["units!R","component"],
    summary(st_ny23_sny_yld3_base)$varcomp["units!R","component"])
)

varcomp_sny23_base




### =====================================================
### Variance component summary table (Spatial)
### =====================================================

varcomp_sny23_spatial <- data.frame(
  Trait = c("GNDVI_150","yld1",
            "GNDVI_198","yld2",
            "GNDVI_255","yld3"),
  Genetic = c(
    summary(st_ny23_sny_gndvi150_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_yld1)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_gndvi198_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_yld2)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_gndvi255_spa)$varcomp["vm(Germplasm, grm_inverse)","component"],
    summary(st_ny23_sny_yld3)$varcomp["vm(Germplasm, grm_inverse)","component"]),
  Spatial = c(
    summary(st_ny23_sny_gndvi150_spa)$varcomp["Row:Range","component"],
    summary(st_ny23_sny_yld1)$varcomp["Row:Range","component"],
    summary(st_ny23_sny_gndvi198_spa)$varcomp["Row:Range","component"],
    summary(st_ny23_sny_yld2)$varcomp["Row:Range","component"],
    summary(st_ny23_sny_gndvi255_spa)$varcomp["Row:Range","component"],
    summary(st_ny23_sny_yld3)$varcomp["Row:Range","component"]),
  Residual = c(
    summary(st_ny23_sny_gndvi150_spa)$varcomp["units!R","component"],
    summary(st_ny23_sny_yld1)$varcomp["units!R","component"],
    summary(st_ny23_sny_gndvi198_spa)$varcomp["units!R","component"],
    summary(st_ny23_sny_yld2)$varcomp["units!R","component"],
    summary(st_ny23_sny_gndvi255_spa)$varcomp["units!R","component"],
    summary(st_ny23_sny_yld3)$varcomp["units!R","component"])
)

varcomp_sny23_spatial
```






 