---
title: "Running bivariate models with GRM and spatial to calculate genetic and spatial correlations"
author: "Dinesh Ghimire"
date: "2026-05-13"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r Libraries}
# packages required for analysis. I rely on the dplyr/tidyr a lot as I find the grouping (group_by,etc..) particularly helpful. I'd be familiar with those before using this script
# List of required packages
required_packages <- c("dplyr","asreml", "readxl", "tidyr", "ggplot2", "reshape2","data.table", "viridis","grid", "gridExtra", "GGally")

# Function to install/load packages
install_and_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Apply to all required packages
invisible(sapply(required_packages, install_and_load))
```

#Git and R project connection
```{r Connecting with GitHub}
#all of this operates within the github project "FFAR-Alfalfa-NY-DG"
#https://github.com/dineshghimire01/FFAR-Alfalfa-NY-DG 
#below is what I get when I call getwd. Just check to make sure that the root for this script is the base of the github project, sometimes it will move to where this file is directly located in "scripts/processing" which you don't want. 
# "/Users/dg663/FFAR-Alfalfa-NY-DG"
getwd()
#below is a kind of master list for all the germplasmName, entry numbers, and genotype ids. If something is misssing please let me know. 
key<-read.csv("2023/meta_data/cultivar_list_key.csv") # contains both CU and NY cultivars here. 
# its same for 2023 and 2023 (cultivar list)
```




#NY 23


################################# Musgrave Trial ################################



1. *NY Musgrave 2023 *

Loading and preparing data files
```{r loading files 05132026}
#import merged phenotype file 
ny_mus_pheno_23 <- read.csv("2023/MSI_data/output_files/Final_NY23_mus_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_mus_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_mus_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_mus_pheno_23) 
str(ny_mus_pheno_23)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))


grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


- Scaling and outlier check on yield already done in script1, so we can jump into running models

RUN for all... 

## 1.2. For GNDVI 
- For Harvest window 1:  125, 136, 144, 150, and yld1
- For Harvest window 2:  152, 156, 164, 166, 172, 177, 179, 186, 192, 195, 198, and yld2
- For Harvest window 3: 213, 223, 228, 244, 248, 251, 255, and yld3
 = 22 flights until 3rd harvest


## only running bivariate for yield with ar spatial to see how MSI and corresponding yield are 
# Bivariate for yield with spatial
- if doesn't run, try sca_yld OR simpler model structure like diag instead of unstructured or one dimensional spatial
- always double check correlation value formula (shortcut used like V10 or V11 being used in the equation)

```{r Multitrait model, GNDVI All flights, 05132026}
### Multitrait model
asreml.options(extra=200, maxit =200) 
unique(ny_mus_pheno_23$Germplasm)
##################### Harvest 1 Window ####################
 # - For Harvest window 1:  125, 136, 144, 150, and yld1

# GNDVI_125 : 
mt_mus_pheno_23_gndvi_125 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_125) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_125)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_125, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6372554
asreml::vpredict(mt_mus_pheno_23_gndvi_125, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4618672
asreml::vpredict(mt_mus_pheno_23_gndvi_125, res.cor ~ (V11)/sqrt(V10*V12)) # 0.7806916

# GNDVI_136 : worked with jst yld or diag for residual trait
mt_mus_pheno_23_gndvi_136 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld1, GNDVI_136) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # diag resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_136)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_136, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6034086
asreml::vpredict(mt_mus_pheno_23_gndvi_136, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8449915
asreml::vpredict(mt_mus_pheno_23_gndvi_136, res.cor ~ (V11)/sqrt(V10*V12)) 

# GNDVI_144 : 
mt_mus_pheno_23_gndvi_144 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_144) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_144)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_144, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.4025349
asreml::vpredict(mt_mus_pheno_23_gndvi_144, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4440798
asreml::vpredict(mt_mus_pheno_23_gndvi_144, res.cor ~ (V11)/sqrt(V10*V12)) 

# GNDVI_150 : 
mt_mus_pheno_23_gndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_150) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), 
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_150)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.3643384
asreml::vpredict(mt_mus_pheno_23_gndvi_150, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4862491
asreml::vpredict(mt_mus_pheno_23_gndvi_150, res.cor ~ (V11)/sqrt(V10*V12)) # 0.5703683






##################### Harvest 2 Window ####################
# For Harvest window 2:  152, 156, 164, 166, 172, 177, 179, 186, 192, 195, 198, and yld2

# GNDVI_152 
mt_mus_pheno_23_gndvi_152 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_152) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_152)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_152, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.6453829
asreml::vpredict(mt_mus_pheno_23_gndvi_152, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.072413
asreml::vpredict(mt_mus_pheno_23_gndvi_152, res.cor ~ (V11)/sqrt(V10*V12)) # 0.05061395

# GNDVI_156 
mt_mus_pheno_23_gndvi_156 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_156) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_156)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_156, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.2359866
asreml::vpredict(mt_mus_pheno_23_gndvi_156, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.003086009
asreml::vpredict(mt_mus_pheno_23_gndvi_156, res.cor ~ (V11)/sqrt(V10*V12)) # diag...

# GNDVI_164 
mt_mus_pheno_23_gndvi_164 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_164) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_164)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_164, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6415622
asreml::vpredict(mt_mus_pheno_23_gndvi_164, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1226018
asreml::vpredict(mt_mus_pheno_23_gndvi_164, res.cor ~ (V11)/sqrt(V10*V12))

# GNDVI_166 
mt_mus_pheno_23_gndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_166) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_166)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6935922
asreml::vpredict(mt_mus_pheno_23_gndvi_166, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.06649855
asreml::vpredict(mt_mus_pheno_23_gndvi_166, res.cor ~ (V11)/sqrt(V10*V12)) # 0.8792663

# GNDVI_172 
mt_mus_pheno_23_gndvi_172 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_172) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_172)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_172, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6758091
asreml::vpredict(mt_mus_pheno_23_gndvi_172, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.2293952
asreml::vpredict(mt_mus_pheno_23_gndvi_172, res.cor ~ (V11)/sqrt(V10*V12)) # 0.88

# GNDVI_177 
mt_mus_pheno_23_gndvi_177 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_177) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_177)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_177, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6307984
asreml::vpredict(mt_mus_pheno_23_gndvi_177, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.009508416
asreml::vpredict(mt_mus_pheno_23_gndvi_177, res.cor ~ (V11)/sqrt(V10*V12)) # 0.5739082

# GNDVI_179 
mt_mus_pheno_23_gndvi_179 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_179) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_179)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_179, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.5738695
asreml::vpredict(mt_mus_pheno_23_gndvi_179, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6094249
asreml::vpredict(mt_mus_pheno_23_gndvi_179, res.cor ~ (V11)/sqrt(V10*V12)) # 0.5853588

# GNDVI_186 
mt_mus_pheno_23_gndvi_186 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_186) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_186)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_186, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.4049283
asreml::vpredict(mt_mus_pheno_23_gndvi_186, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.01458153
asreml::vpredict(mt_mus_pheno_23_gndvi_186, res.cor ~ (V11)/sqrt(V10*V12)) # 0.1468794

# GNDVI_192 (worked with jst yld)
mt_mus_pheno_23_gndvi_192 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld2, GNDVI_192) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_192)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_192, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.1453776
asreml::vpredict(mt_mus_pheno_23_gndvi_192, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.2091636
asreml::vpredict(mt_mus_pheno_23_gndvi_192, res.cor ~ (V11)/sqrt(V10*V12))

# GNDVI_195 
mt_mus_pheno_23_gndvi_195 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld2, GNDVI_195) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_195)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_195, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.1579136
asreml::vpredict(mt_mus_pheno_23_gndvi_195, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1818738
asreml::vpredict(mt_mus_pheno_23_gndvi_195, res.cor ~ (V11)/sqrt(V10*V12)) # - 0.7783282

# GNDVI_198 
mt_mus_pheno_23_gndvi_198 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_198) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_198)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_198, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.2191799
asreml::vpredict(mt_mus_pheno_23_gndvi_198, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1437625
asreml::vpredict(mt_mus_pheno_23_gndvi_198, res.cor ~ (V11)/sqrt(V10*V12)) # - 0.1511012


##################### Harvest 3 Window ####################
#For Harvest window 3: 213, 223, 228, 244, 248, 251, 255, yld3

# GNDVI_213 : 
mt_mus_pheno_23_gndvi_213 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_213) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_213)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_213, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.2927097
asreml::vpredict(mt_mus_pheno_23_gndvi_213, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.6289137
asreml::vpredict(mt_mus_pheno_23_gndvi_213, res.cor ~ (V11)/sqrt(V10*V12)) # -0.11

# GNDVI_223 : (worked with jst yld3, not sca)
mt_mus_pheno_23_gndvi_223 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld3, GNDVI_223) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_223)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_223, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.4821267
asreml::vpredict(mt_mus_pheno_23_gndvi_223, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.5849669
asreml::vpredict(mt_mus_pheno_23_gndvi_223, res.cor ~ (V11)/sqrt(V10*V12)) # 


# GNDVI_228 : 
mt_mus_pheno_23_gndvi_228 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_228) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_228)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_228, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.5055674
asreml::vpredict(mt_mus_pheno_23_gndvi_228, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.7598246
asreml::vpredict(mt_mus_pheno_23_gndvi_228, res.cor ~ (V11)/sqrt(V10*V12)) # 0.3936578

# GNDVI_244 : 
mt_mus_pheno_23_gndvi_244 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_244) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_244)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_244, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8033733
asreml::vpredict(mt_mus_pheno_23_gndvi_244, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8859017
asreml::vpredict(mt_mus_pheno_23_gndvi_244, res.cor ~ (V11)/sqrt(V10*V12)) # 0.1983045


# GNDVI_248 :
mt_mus_pheno_23_gndvi_248 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_248) ~ trait ,
         random = ~ us(trait):vm(Germplasm, grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_248)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_248, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8964026 (without grm, converged)
asreml::vpredict(mt_mus_pheno_23_gndvi_248, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.786806
asreml::vpredict(mt_mus_pheno_23_gndvi_248, res.cor ~ (V11)/sqrt(V10*V12)) # 


# GNDVI_251 : 
mt_mus_pheno_23_gndvi_251 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_251) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_251)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_251, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7294716
asreml::vpredict(mt_mus_pheno_23_gndvi_251, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.8434454
asreml::vpredict(mt_mus_pheno_23_gndvi_251, res.cor ~ (V11)/sqrt(V10*V12)) # 0.3186641


# GNDVI_255 : 
mt_mus_pheno_23_gndvi_255 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_255) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_mus_pheno_23)}
summary(mt_mus_pheno_23_gndvi_255)$varcomp
asreml::vpredict(mt_mus_pheno_23_gndvi_255, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8460815
asreml::vpredict(mt_mus_pheno_23_gndvi_255, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.4331947
asreml::vpredict(mt_mus_pheno_23_gndvi_255, res.cor ~ (V11)/sqrt(V10*V12)) # diag
```






## Graphs: Genetic and Spatial trend throughout the season
```{r Mt models graphs GNDVI Musgrave 2023 05122026}
mt_models_gndvi_mus_23 <- read.csv("2023/MSI_data/output_files/MT_models_GC_SC_values_GNDVI_Musgrave_2023.csv")
colnames(mt_models_gndvi_mus_23)
# Use your dataframe
df_gndvi_mus_23 <- mt_models_gndvi_mus_23
# Preserve order of Trait
df_gndvi_mus_23$Trait <- factor(df_gndvi_mus_23$Trait, levels = df_gndvi_mus_23$Trait)

# ---- Graph 1: Genetic1 only ----
df_gndvi_mus_23_no_spa <- df_gndvi_mus_23 %>%
  select(Trait, Genetic) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value")

gndvi_mus_23_p1 <- ggplot(df_gndvi_mus_23_no_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI without Spatial Musgrave 2023",
       y = "Correlation")
gndvi_mus_23_p1
# ---- Graph 2: Genetic, & Spatial  ----
df_gndvi_mus_23_spa <- df_gndvi_mus_23 %>%
  select(Trait, Genetic, Spatial) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value") %>%
  filter(!is.na(Value))

gndvi_mus_23_p2 <- ggplot(df_gndvi_mus_23_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Genetic" = "forestgreen", 
                                "Spatial" = "darkorange")) +
  scale_linetype_manual(values = c("Genetic" = "solid", 
                                   "Spatial" = "solid")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "GNDVI Trend over Musgrave 2023",
       y = "Correlation")

# Show both plots
gndvi_mus_23_p1 # only genetic
gndvi_mus_23_p2 # genetic and spatial

# ggsave
ggsave("2023/output_images/Genetic_Spatial_Correlations_NY_Musgrave_2023.png", gndvi_mus_23_p2, width = 8, height = 6, dpi = 400)
```






################################# Snyder Trial ################################

1. *NY Snyder 2023 *

Loading and preparing data files
```{r loading files 05132026}
#import merged phenotype file 
ny_sny_pheno_23 <- read.csv("2023/MSI_data/output_files/Final_NY23_sny_combined_phenotypes_All_MSI.csv")  %>%
  select(-X) # remove index column

# need to convert into factors
ny_sny_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")] <- lapply(ny_sny_pheno_23[, c("Range", "Rep","Germplasm","Plot", "Row")], as.factor)
colnames(ny_sny_pheno_23) 
str(ny_sny_pheno_23)


#import genotype file
grm_aa_ny<- read.csv("2022/genotype_data/grm_final.csv", sep = ",", row.names = 1, check.names = FALSE)
grm_aa_ny  <- as.matrix(grm_aa_ny)
mean(diag(grm_aa_ny )) # 1
hist(diag(grm_aa_ny ))


grm_agh <- grm_aa_ny +diag(.0005,length(grm_aa_ny [,1]),length(grm_aa_ny [,1])) # adding 0.05 nugget/ridge to ensure matrix is strictly positive definite
grm_inverse <- solve(grm_agh)

```


- Scaling and outlier check on yield already done in script1, so we can jump into running models

RUN for all... 

## 1.2. For GNDVI 
- For Harvest window 1:  125, 142, 145, 150, 152, 156, and yld1
- For Harvest window 2:  164, 166, 172, 178, 184, 187, 193, 198, and yld2
- For Harvest window 3: 213, 223, 228, yld3
= total 17 flights


## only running bivariate for yield with ar spatial to see how MSI and corresponding yield are 
# Bivariate for yield with spatial
- if doesn't run, try sca_yld OR simpler model structure like diag instead of unstructured or one dimensional spatial



```{r Multitrait model, GNDVI All flights, Snyder 05122026}
### Multitrait model
asreml.options(extra=200, maxit =200) 
unique(ny_sny_pheno_23$Germplasm)
##################### Harvest 1 Window ####################
 # - For Harvest window 1:  125, 142, 145, 150, 152, 156, and yld1

# GNDVI_125 : bad
mt_sny_pheno_23_gndvi_125 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_125) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_125)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_125, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.09838533
asreml::vpredict(mt_sny_pheno_23_gndvi_125, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.04567734
asreml::vpredict(mt_sny_pheno_23_gndvi_125, res.cor ~ (V11)/sqrt(V10*V12)) # diag

# GNDVI_142 : 
mt_sny_pheno_23_gndvi_142 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_142) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_142)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_142, gen.cor ~ (V2)/sqrt(V1*V3)) # 0
asreml::vpredict(mt_sny_pheno_23_gndvi_142, spa.cor ~ (V5)/sqrt(V4*V6)) # 0
asreml::vpredict(mt_sny_pheno_23_gndvi_142, res.cor ~ (V11)/sqrt(V10*V12)) 

# GNDVI_145 : worked with just yld
mt_sny_pheno_23_gndvi_145 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld1, GNDVI_145) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_145)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_145, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.16
asreml::vpredict(mt_sny_pheno_23_gndvi_145, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.17
asreml::vpredict(mt_sny_pheno_23_gndvi_145, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_150 : 
mt_sny_pheno_23_gndvi_150 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_150) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_150)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_150, gen.cor ~ (V2)/sqrt(V1*V3)) # 0
asreml::vpredict(mt_sny_pheno_23_gndvi_150, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.27
asreml::vpredict(mt_sny_pheno_23_gndvi_150, res.cor ~ (V11)/sqrt(V10*V12)) # diag

# GNDVI_152 : 
mt_sny_pheno_23_gndvi_152 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_152) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_152)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_152, gen.cor ~ (V2)/sqrt(V1*V3)) # 0
asreml::vpredict(mt_sny_pheno_23_gndvi_152, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.23
asreml::vpredict(mt_sny_pheno_23_gndvi_152, res.cor ~ (V11)/sqrt(V10*V12)) # diag

# GNDVI_156 : 
mt_sny_pheno_23_gndvi_156 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld1, GNDVI_156) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_156)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_156, gen.cor ~ (V2)/sqrt(V1*V3)) # - 0.2533751
asreml::vpredict(mt_sny_pheno_23_gndvi_156, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.252678
asreml::vpredict(mt_sny_pheno_23_gndvi_156, res.cor ~ (V11)/sqrt(V10*V12)) # 0.2574392



##################### Harvest 2 Window ####################
# For Harvest window 2:  164, 166, 172, 178, 184, 187, 193, 198, and yld2

# GNDVI_164 :  
mt_sny_pheno_23_gndvi_164 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_164) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_164)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_164, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.72 
asreml::vpredict(mt_sny_pheno_23_gndvi_164, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.24
asreml::vpredict(mt_sny_pheno_23_gndvi_164, res.cor ~ (V11)/sqrt(V10*V12)) # 


# GNDVI_166 :  
mt_sny_pheno_23_gndvi_166 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_166) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_166)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_166, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.739496 
asreml::vpredict(mt_sny_pheno_23_gndvi_166, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.3247779
asreml::vpredict(mt_sny_pheno_23_gndvi_166, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_172 :  
mt_sny_pheno_23_gndvi_172 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_172) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_172)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_172, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8186812 
asreml::vpredict(mt_sny_pheno_23_gndvi_172, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.4602847
asreml::vpredict(mt_sny_pheno_23_gndvi_172, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_178 :  
mt_sny_pheno_23_gndvi_178 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_178) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_178)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_178, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.8687644 
asreml::vpredict(mt_sny_pheno_23_gndvi_178, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.2119893
asreml::vpredict(mt_sny_pheno_23_gndvi_178, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_184 :  
mt_sny_pheno_23_gndvi_184 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_184) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_184)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_184, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7941721 
asreml::vpredict(mt_sny_pheno_23_gndvi_184, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.2451191
asreml::vpredict(mt_sny_pheno_23_gndvi_184, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_187 :  
mt_sny_pheno_23_gndvi_187 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_187) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # diag resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_187)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_187, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7834961 
asreml::vpredict(mt_sny_pheno_23_gndvi_187, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.1475376
asreml::vpredict(mt_sny_pheno_23_gndvi_187, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_193 :  
mt_sny_pheno_23_gndvi_193 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(yld2, GNDVI_193) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_193)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_193, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.7544318 
asreml::vpredict(mt_sny_pheno_23_gndvi_193, spa.cor ~ (V5)/sqrt(V4*V6)) # 0
asreml::vpredict(mt_sny_pheno_23_gndvi_193, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_198 :  
mt_sny_pheno_23_gndvi_198 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld2, GNDVI_198) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_198)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_198, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.2354209 
asreml::vpredict(mt_sny_pheno_23_gndvi_198, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1107793
asreml::vpredict(mt_sny_pheno_23_gndvi_198, res.cor ~ (V11)/sqrt(V10*V12)) # 



##################### Harvest 3 Window ####################
#For Harvest window 3: 213, 223, 228, yld3

# GNDVI_213 
mt_sny_pheno_23_gndvi_213 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_213) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_213)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_213, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.6122953
asreml::vpredict(mt_sny_pheno_23_gndvi_213, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.2846863
asreml::vpredict(mt_sny_pheno_23_gndvi_213, res.cor ~ (V11)/sqrt(V10*V12)) # 0.90561

# GNDVI_223 
mt_sny_pheno_23_gndvi_223 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_223) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):diag(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_223)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_223, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.674427
asreml::vpredict(mt_sny_pheno_23_gndvi_223, spa.cor ~ (V5)/sqrt(V4*V6)) # 0.1895156
asreml::vpredict(mt_sny_pheno_23_gndvi_223, res.cor ~ (V11)/sqrt(V10*V12)) # 

# GNDVI_228 
mt_sny_pheno_23_gndvi_228 <- {  
  attr(grm_inverse, "INVERSE") <- TRUE
  asreml(fixed = cbind(sca_yld3, GNDVI_228) ~ trait ,
         random = ~ us(trait):vm(Germplasm,  grm_inverse) + us(trait):ar1(Row):ar1(Range),
         residual = ~ id(units):us(trait), # us resid
         data = ny_sny_pheno_23)}
summary(mt_sny_pheno_23_gndvi_228)$varcomp
asreml::vpredict(mt_sny_pheno_23_gndvi_228, gen.cor ~ (V2)/sqrt(V1*V3)) # 0.622699
asreml::vpredict(mt_sny_pheno_23_gndvi_228, spa.cor ~ (V5)/sqrt(V4*V6)) # - 0.04027907
asreml::vpredict(mt_sny_pheno_23_gndvi_228, res.cor ~ (V11)/sqrt(V10*V12)) # 

```




## Graphs: Genetic and Spatial trend throughout the season
```{r Mt models graphs GNDVI Snyder 2023 05132026}
mt_models_gndvi_sny_23 <- read.csv("2023/MSI_data/output_files/MT_models_GC_SC_values_GNDVI_Snyder_2023.csv")
colnames(mt_models_gndvi_sny_23)
# Use your dataframe
df_gndvi_sny_23 <- mt_models_gndvi_sny_23
# Preserve order of Trait
df_gndvi_sny_23$Trait <- factor(df_gndvi_sny_23$Trait, levels = df_gndvi_sny_23$Trait)

# ---- Graph 1: Genetic1 only ----
df_gndvi_sny_23_no_spa <- df_gndvi_sny_23 %>%
  select(Trait, Genetic) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value")

gndvi_sny_23_p1 <- ggplot(df_gndvi_sny_23_no_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Bivariate GNDVI without Spatial  Snyder 2023",
       y = "Correlation")
gndvi_sny_23_p1
# ---- Graph 2: Genetic, & Spatial  ----
df_gndvi_sny_23_spa <- df_gndvi_sny_23 %>%
  select(Trait, Genetic, Spatial) %>%
  pivot_longer(-Trait, names_to = "Component", values_to = "Value") %>%
  filter(!is.na(Value))

gndvi_sny_23_p2 <- ggplot(df_gndvi_sny_23_spa, aes(x = Trait, y = Value, color = Component, group = Component)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Genetic" = "forestgreen", 
                                "Spatial" = "darkorange")) +
  scale_linetype_manual(values = c("Genetic" = "solid", 
                                   "Spatial" = "solid")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "GNDVI Trend over Snyder 2023",
       y = "Correlation")

# Show both plots
gndvi_sny_23_p1 # only genetic
gndvi_sny_23_p2 # genetic and spatial

# ggsave
ggsave("2023/output_images/Genetic_Spatial_Correlations_NY_Snyder_2023.png", gndvi_sny_23_p2, width = 8, height = 6, dpi = 400)
```
```




 
 